---
title: "Lab2"
output:
  pdf_document: default
  html_document: default
date: "2024-09-06"
---

[Stat 5000]{.smallcaps}[Lab #2]{.smallcaps}\
[Fall 2024]{.smallcaps} [Due Tue Sep 10th]{.smallcaps}
[Name: Sam Olson]{.smallcaps}

**Directions:** Complete the exercises below. When you are finished,
turn in any required files online in Canvas, then check-in with the Lab
TA for dismissal.\

**[Introduction to t-Tests in R]{.underline}**

Refer to the `fuel_economy.csv` data file posted in Canvas. This data
set has information about an observational study of automobiles driven
in Canada, including the following two columns:

`Cylinders`:

:   category variable with two levels - 4 or 6

`Consumption`:

:   numeric response variable with the fuel consumption in miles per
    gallon (mpg)

Researchers are interested in exploring whether there is a difference in
the average fuel consumption of vehicles with engines built using
differing numbers of cylinders. The code to conduct a two-sample t-test
in R is explained below. The full R program is provided in the file
`fuel_economy_Lab2.R` posted on Canvas.

-   First, load in the data using the *Import Dataset* tool in R Studio.
    Be sure to change the variable type on the Cylinders column to
    "factor\" and enter "4, 6\" as the levels.

```{r}
library(readr)
fuel <- read_csv("fuel_economy.csv", 
                 col_types=cols(Cylinders=col_factor(levels=c("4", "6"))))
View(fuel)
```

-   Next, compute the corresponding summary statistics within in group.

```{r}
library(tidyverse)
sum_stats = fuel |> 
    group_by(Cylinders) |>
    summarize(
        Y_n = n(),
        Y_mean = mean(Consumption.mpg),
        Y_sd = sd(Consumption.mpg)
    )
sum_stats
```                

-   Then, use the `t.test()` function to conduct a test for the
    difference in mean fuel consumption between 4 and 6 cylinder
    vehicles. Indicate the response variable name before the $\sim$ and
    the category variable name after, use the `data` option to provide
    the name of the dataset, and use the `var.equal` option set to
    "TRUE\" to indicate the population variances are assumed equal.

```{r}
HT = t.test(Consumption.mpg~Cylinders, data=fuel, var.equal=TRUE)
HT
```

You can see what pieces of information are stored in the `HT`
variable using the `names()` function. You can access these pieces
of information using the `$` operator, e.g.

```{r}
names(HT)
HT$null.value
```

\newpage

**[Assignment]{.underline}**

> 1.  State the hypotheses for the two-sided test.

$H_0: \mu_{4 \space \text{Cylinder}} = \mu_{6 \space \text{Cylinder}}$, with $H_A: \mu_{4 \space \text{Cylinder}} \neq \mu_{6 \space \text{Cylinder}}$, descriptively stated as: Null Hypothesis is the mean consumption (mpg) of 4 Cylinder cars is equal to the mean consumption of 6 Cylinder cars with alternate hypothesis that the mean consumption of 4 Cylinder cars is **not equal** to the mean consumption of 6 Cylinder cars. 

\newpage 

> 2.  From the output, find/compute the difference in the two sample means.

```{r}
HT = t.test(Consumption.mpg~Cylinders, data=fuel, var.equal=TRUE)
HT
```

```{r}
sampleMean1 <- 34.08100
sampleMean2 <- 24.51297 
difference <- sampleMean1 - sampleMean2 
difference
```
The difference in the two sample means is $\approx 9.568$ mpg (4 Cylinder - 6 Cylinder).


\newpage 

> 3.  From the output, find/compute the estimate of the pooled standard deviation.

Formula: 

$s_p = \sqrt{ \frac{(n_1 - 1)s_1^2 + (n_2 - 1)s_2^2}{(n_1+n_2-2)}}$

```{r}
library(dplyr)
data1 <- fuel %>% 
  filter(fuel$Cylinders == 4)

data2 <- fuel %>% 
  filter(fuel$Cylinders == 6)

#Step 2: Finding standard deviation 
s1 <- sd(data1$Consumption.mpg)
s2 <- sd(data2$Consumption.mpg)

#Step 3: Finding sample size 
n1 <- length(data1$Consumption.mpg)
n2 <- length(data2$Consumption.mpg)

#Step 4: Calculate pooled standard deviation
pooled <- sqrt(((n1-1)*s1^2 + (n2-1)*s2^2) / (n1+n1-2))
pooled 

sd(fuel$Consumption.mpg)
```

From the above this yields an estimated pooled standard deviation of $\approx 7.228$ mpg.

\newpage 

> 4.  From the output, find/compute the test statistic for the hypothesis test.

```{r}
library(dplyr)
m1 <- mean(data1$Consumption.mpg)

m2 <- mean(data2$Consumption.mpg)

manualTStatistic <- (m1 - m2) / (pooled * sqrt((1/n1) + (1/n2)))
manualTStatistic
```

```{r}
HT$statistic
```

\newpage 

> 5.  From the output, find/compute the degrees of freedom for the test.

```{r}
HT$parameter
```
df = $n_1 + n_2 - 2 = 2111 + 1041 - 2 = 3150$

\newpage 

> 6.  From the output, find/compute the $p$-value for the two-sided hypothesis test.

```{r}
HT
HT$p.value
```

$p-value < 2.2e-16$ or $\approx 0$ (so small R rounds it to 0).


\newpage 

> 7.  Interpret the results of the two-sided test in the context of the research question.

The low p-value means we estimate with very low probability the observed results given the null hypothesis is true (no difference in average fuel consumption between the Cylinder types). This provides evidence to reject the null hypothesis and consider the alternative hypothesis, that is we have support that there is a difference in average fuel consumption (mpg) between 4 and 6 cylinder vehicles. 

\newpage 

> 8.  By default, R conducts the two-sided hypothesis test. You can change this by adding the parameter "`alternative=greater`" or "`alternative=less`" inside the `t.test()` function. Provide a research question corresponding to either the "greater\" or "less\" one-sided test.

```{r}
altHT <- t.test(Consumption.mpg~Cylinders, data=fuel, var.equal=TRUE, alternative = "greater")
altHT
```

For input $\text{"alternative = "greater""}$: 

We are specifying an alternate hypothesis that the average mpg of 4 cylinder vehicles is higher (greater than) the average mpg of 6 cylinder vehicles. (Though not explicitly called for, it is worth noting that this hypothesis provides a similar p-value, evidence, and conclusions to the above, namely in support of 4 cylinder vehicles having greater mpg, or on average more efficient fuel consumption.)

**Total:** 25 points **\# correct:** **%:**
