---
title: "Take-Home Part I"
output: pdf_document
author: "Sam Olson"
header-includes:
  - \usepackage{float}
  - \usepackage{placeins}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r loading, echo = F, message = F, warning = F}
library(here)
library(knitr)
library(kableExtra)
dat <- read.table(here("STAT_6010/Take-Home 1/exam1dat.txt"), header = T)
source(here("STAT_6010/Take-Home 1/nonlin.txt"))
source(here("STAT_6010/Take-Home 1/basicglm.txt"))

x   <- dat$x
y   <- dat$y
n   <- length(y)

# Expectation function: mu_i = alpha * x_i^(beta * x_i^(-gamma))
efctn <- function(xv, ps) {
  alpha <- ps[1]; beta <- ps[2]; gamma <- ps[3]
  alpha * xv^(beta * xv^(-gamma))
}

# Derivatives: d(mu)/d(alpha), d(mu)/d(beta), d(mu)/d(gamma)
# d(mu)/d(alpha) = x^(beta * x^(-gamma))
# d(mu)/d(beta)  = mu * x^(-gamma) * log(x)
# d(mu)/d(gamma) = -mu * beta * log(x)^2 * x^(-gamma)
eders <- function(xv, ps) {
  alpha <- ps[1]; beta <- ps[2]; gamma <- ps[3]
  mu    <- efctn(xv, ps)
  da    <- xv^(beta * xv^(-gamma))
  db    <- mu * xv^(-gamma) * log(xv)
  dg    <- -mu * beta * log(xv)^2 * xv^(-gamma)
  cbind(da, db, dg)
}

# Constant variance: weights = I
ewts <- function(xv, ps) diag(length(xv))

# Starting values (rough visual guess from scatterplot)
ps0 <- c(alpha = 0.5, beta = 1.2, gamma = 0.16)

fit1 <- nonlin(xmat = x, ys = y, ps = ps0,
               fctn = efctn, ders = eders, wts = ewts)
```

# The Model and Data

Let $\{Y_i : i = 1, \ldots, n\}$ denote a set of response variables and let $\{x_i : i = 1, \ldots, n\}$ represent a corresponding set of non-decreasing covariate values. The model for the response variables is

$$Y_i = \mu_i + w_i \tag{1}$$

where

$$\mu_i = \alpha x_i^{\beta x_i^{-\gamma}}. \tag{2}$$

In (2), $\alpha \geq 0$, $\beta \geq 0$, and $\gamma \geq 0$ are unknown parameters, and for $-1 < \theta < 1$,

$$w_i = \theta w_{i-1} + \epsilon_i, \tag{3}$$

where the $\epsilon_i$ are independent and identically distributed with $\epsilon_i \sim N(0, \sigma^2)$.

# Q1

Consider a simplified model in which the $w_i$ in (1) are iid $N(0, \sigma^2)$, i.e., a traditional additive error nonlinear regression model with constant variance. Fit this model to the project data using generalized least squares to estimate the parameters of the expectation function.

## **(a)** and **(b)**: Report GLS estimates of $\alpha$, $\beta$, $\gamma$ and the moment-based estimate of $\sigma^2$. Approximate 95% confidence intervals for $\alpha$, $\beta$, $\gamma$.

Starting values are chosen by reasoning through $\mu_i = \alpha x_i^{\beta x_i^{-\gamma}}$, starting with $\gamma$, followed by $\beta$, and then $\alpha$. The scatterplot of response by covariate shows the response rising gradually near $x = 400$, suggesting slow exponent decay and thus a small $\gamma$, certainly less than 1; we settle on $\gamma_0 \approx 0.16$ after graphical exploration. Next, for $x = 100$, the effective power at $x$ is $\beta \cdot 100^{-0.16} \approx 0.48\beta$; further graphical exploration indicated an effective exponent $\approx 0.55$ near this point, giving $\beta_0 \approx 0.55/0.48 \approx 1.2$. Finally, using these values of $\gamma_0$ and $\beta_0$, we back-calculate from $y \approx 6.5$, $x \approx 100$, giving $\alpha_0 \approx 0.5$. However, other starting values were also considered to ensure appropriate convergence, which was met.
  
```{r q1-b, echo = F, message = F, warning = F}
bhat  <- fit1$bs
s2hat <- fit1$sshat

se    <- sqrt(diag(fit1$covb))
z95   <- qnorm(0.975)
lower <- bhat - z95 * se
upper <- bhat + z95 * se

ci_tab <- data.frame(
  Parameter = c("$\\alpha$", "$\\beta$", "$\\gamma$"),
  Estimate  = unname(round(bhat,  3)),
  Lower     = unname(round(lower, 3)),
  Upper     = unname(round(upper, 3))
)

kable(ci_tab,
      format    = "latex",
      booktabs  = TRUE,
      escape    = FALSE,
      align     = c("l", "r", "r", "r"),
      col.names = c("Parameter", "Estimate", "Lower", "Upper"),
      caption   = "GLS estimates with approximate 95\\% confidence intervals.") |>
  add_header_above(c(" " = 2, "95% CI" = 2)) |>
  kable_styling(latex_options = c("HOLD_position"),
                font_size = 10)
```

As noted in the solution to Question 1: The lower endpoint of the interval for $\hat\alpha$ is negative, which falls outside the scientifically meaningful ("desired") parameter space $\alpha \geq 0$, though it does not prevent $\mu_i$ from being defined. Also, the moment-based estimate of $\sigma^2$ is $\hat\sigma^2 = `r round(s2hat, 3)`$.

## **(c)** and **(d)**: A plot with the estimated expectation function overlaid on the scatterplot along with an approximate 95% pointwise confidence band. A plot of standardized residuals against the covariate.

**(c)**: As noted in the solution to Question 1: An approximate pointwise confidence band is constructed via the delta method, whose derivatives of $\mu_i$ with respect to $(\alpha, \beta, \gamma)$ are

$$
\frac{\partial \mu_i}{\partial \alpha} = x_i^{\beta x_i^{-\gamma}}, \qquad
  \frac{\partial \mu_i}{\partial \beta}  = \mu_i x_i^{-\gamma} \log x_i, \qquad
  \frac{\partial \mu_i}{\partial \gamma} = -\mu_i \beta (\log x_i)^2 x_i^{-\gamma}.
$$

The band is pointwise in that at each fixed $x_i$, the interval $\hat{\mu}(x_i) \pm 1.96 \cdot \widehat{\text{SE}}[\hat{\mu}(x_i)]$ is constructed to have approximate 95% coverage for each $\mu(x_i)$. The band is approximate because the delta method uses an asymptotic result.

**(d)**: Standardized residuals were computed using the hat-matrix correction $r_i / [\hat{\sigma}^2(1 - h_{i,i})]^{1/2}$, expression (9.40) of the Stat 520 notes, returned directly by `nonlin` via `fit1$stdres`. The solution sheet cites (9.37), the general plug-in form; (9.40), its constant-variance simplification, is used here and obtained by substituting $\hat{\sigma}_i^2 = \hat{\sigma}^2$. The residual plot shows a good deal of remaining structure as a function of the covariate, consistent with Figure 3 of Sibbesen (1981). This pattern motivates the use of an autoregressive structure in the “regression equation error terms”. This is to say: The remaining pattern in the residual plot motivates the inclusion of an AR(1) small-scale component in the full model, treating the serial dependence across ordered covariate values analogously to serial correlation over time.

```{r q1-cd, echo = F, message = F, warning = F, fig.height=8, fig.width=5, fig.cap="(Top) Fitted extended Freundlich curve with approximate 95\\% pointwise confidence band. (Bottom) Standardized residuals versus covariate."}
xgrid  <- seq(min(x[x > 0]), max(x), length.out = 300)
mugrid <- efctn(xgrid, bhat)
Dgrid  <- eders(xgrid, bhat)
pvar   <- apply(Dgrid, 1, function(d) t(d) %*% fit1$covb %*% d)
hw     <- z95 * sqrt(pvar)

par(mfrow = c(2, 1))

plot(x, y, xlab = "Covariate Value", ylab = "Response Value", main = "")
lines(xgrid, mugrid, lwd = 2)
lines(xgrid, mugrid + hw, lty = 2)
lines(xgrid, mugrid - hw, lty = 2)

plot(x, fit1$stdres,
     xlab = "Covariate Value", ylab = "Standardized Residual", main = "")
abline(h = 0, lty = 3)
```

\newpage

# Q2

Consider now just the portion of the model for $w_i$ in (3). Suppose we were able to observe the $w_i$. (Removed: Take $W_0 = 0$, and) Use without derivation that under second-order stationarity, $E(W_i) = 0$ and $E(W_i^2) = \sigma^2 / (1 - \theta^2)$ for all $i$.

## **(a)** Show that the estimating equation for $\theta$ based on the **naive** estimating function

$$v_i(\mathbf{W}, \theta) = W_i - \theta W_{i-1}$$

is an unbiased estimating equation.

As given, the naive estimating function is $v_i(\mathbf{W}, \theta) = W_i - \theta W_{i-1}$. 

Substituting $W_i = \theta W_{i-1} + \epsilon_i$ and taking expectation gives

$$
E[v_i(\mathbf{W}, \theta)] = E[W_i - \theta W_{i-1}] = E[\theta W_{i-1} + \epsilon_i - \theta W_{i-1}] = E[\epsilon_i] = 0,
$$

since $\epsilon_i \sim N(0, \sigma^2)$ is independent of $W_{i-1}$. 

Because this holds for every $i$ and all $|\theta| < 1$, the estimating equation (EE) $\sum_{i=2}^n v_i(\mathbf{W}, \theta) = 0$ is unbiased. Setting the EE to zero and solving gives

$$
\tilde{\theta}_{\text{naive}} = \frac{\sum_{i=2}^n W_i}{\sum_{i=2}^n W_{i-1}}.
$$

Note: While this form is algebraically valid, the denominator $\sum W_{i-1}$ can be near zero, contributing to the numerical instability ($\tilde\theta_{\text{naive}} > 1$) observed when this estimating equation is applied to the Q1 residuals in Q2(d), *spoilers*.

## **(b)** Show that the estimating equation for $\theta$ based on the **optimal** estimating function

$$v_i(\mathbf{W}, \theta) = W_{i-1}(W_i - \theta W_{i-1})$$

is an unbiased estimating equation.

The given optimal estimating function is $v_i(\mathbf{W},\theta) = W_{i-1}(W_i - \theta W_{i-1})$.

Taking the expectation and substituting $W_i = \theta W_{i-1} + \epsilon_i$,

$$
E[v_i(\mathbf{W},\theta)]
  = E[W_{i-1}(W_i - \theta W_{i-1})]
  = E[W_{i-1}\epsilon_i].
$$

Since $W_{i-1}$ is a function of $\epsilon_1, \ldots, \epsilon_{i-1}$ alone and $\epsilon_i \sim N(0,\sigma^2)$ is independent of those error terms, 

$E[W_{i-1}\epsilon_i] = E[W_{i-1}]E[\epsilon_i] = 0$.

The EE $\sum_{i=2}^n W_{i-1}(W_i - \theta W_{i-1}) = 0$ is therefore unbiased, and its solution yields the estimator

$$
\tilde\theta_{\text{opt}}
  = \frac{\sum_{i=2}^n W_i W_{i-1}}{\sum_{i=2}^n W_{i-1}^2}.
$$

Note: This EE is "optimal" in the sense of Godambe's criterion: Among linear estimating functions in $W_{i-1}$, multiplying the naive function by $W_{i-1}$ maximizes the ratio $[E\{\partial v/\partial\theta\}]^2 / \operatorname{var}\{v\}$ (see Ch. 8.3 of the textbook). Also, while both estimating equations yield similar closed-form estimators, in general the two estimators are in numerically distinct.

## **(c)** Develop a moment-based estimator for $\sigma^2$ that could be used with either estimating equation, and show that it is unbiased.

Because $\epsilon_i = W_i - \theta W_{i-1}$ and $\epsilon_i \sim N(0,\sigma^2)$, we have $E[\epsilon_i^2] = \sigma^2$ $\forall i$. Plugging in $\tilde\theta$ from either estimating equation would motivate a possible estimator of the form

$$
\hat\sigma^2 = \frac{1}{n-2}\sum_{i=2}^{n}(W_i - \tilde\theta\, W_{i-1})^2.
$$

(Dropping the assumption $W_0 = 0$ restricts the sum to $i \geq 2$.)

The denominator $n - 2$ is chosen by analogy with the residual mean square: One degree of freedom is notionally attributed to the estimation of $\theta$ from the $n - 1$ available pairs $(W_{i-1}, W_i)$, paralleling the use of $n - p$ in the denominator for linear regression.

To see approximate unbiasedness, note

$$
E\!\left[\frac{1}{n-2}\sum_{i=2}^n (W_i - \theta W_{i-1})^2\right] = \sigma^2.
$$

Once $\theta$ is replaced by $\tilde\theta$, however, strict unbiasedness no longer holds, i.e., $E[(W_i - \tilde\theta W_{i-1})^2] \neq \sigma^2$ exactly, because $\tilde\theta$ is itself a function of the data.

## **(d)** Apply both estimating equations using the raw residuals from Q1 as surrogate observations of $w_i$. Compute $\hat\sigma^2$ for each and compare all three estimates of $\sigma^2$ (naive EE, optimal EE, Q1 GLS). Are these estimators all estimating the same quantity?

```{r q2-d, echo = F, message = F, warning = F}
# Raw residuals from Q1 as surrogate w_i observations
ehat  <- y - fit1$yhat
n_e   <- length(ehat)

e_curr <- ehat[2:n_e]
e_prev <- ehat[1:(n_e - 1)]

# --- Naive estimating equation: sum(e_i - theta * e_{i-1}) = 0
theta_naive <- sum(e_curr) / sum(e_prev)

# --- Optimal estimating equation: sum(e_{i-1} * (e_i - theta * e_{i-1})) = 0
theta_opt <- sum(e_curr * e_prev) / sum(e_prev^2)

# --- sigma^2 estimates (include i = 1 with W_0 = 0)
# e_all_curr <- c(ehat[1],  e_curr)   # W_1, W_2, ..., W_n
# e_all_prev <- c(0,        e_prev)   # W_0=0, W_1, ..., W_{n-1}
e_all_curr <- e_curr   # W_2, ..., W_n
e_all_prev <- e_prev   # W_1, ..., W_{n-1}

innov_naive <- e_all_curr - theta_naive * e_all_prev
innov_opt   <- e_all_curr - theta_opt   * e_all_prev

# s2_naive <- sum(innov_naive^2) / (n_e - 1)
# s2_opt   <- sum(innov_opt^2)   / (n_e - 1)
s2_naive <- sum(innov_naive^2) / (n_e - 2)
s2_opt   <- sum(innov_opt^2)   / (n_e - 2)
s2_q1    <- fit1$sshat

# cat(sprintf("theta (naive EE):    %.4f\n", theta_naive))
# cat(sprintf("theta (optimal EE):  %.4f\n", theta_opt))
# cat(sprintf("sigma^2 (naive EE):  %.4f\n", s2_naive))
# cat(sprintf("sigma^2 (optimal EE):%.4f\n", s2_opt))
# cat(sprintf("sigma^2 (Q1 GLS):    %.4f\n", s2_q1))
results_tab <- data.frame(
  Method = c("Naive EE", "Optimal EE", "Q1 GLS"),
  Theta  = c(round(theta_naive, 4), round(theta_opt, 4), "---"),
  Sigma2 = round(c(s2_naive, s2_opt, s2_q1), 4)
)

kable(results_tab,
      format    = "latex",
      booktabs  = TRUE,
      align     = c("l", "r", "r"),
      col.names = c("Method", "$\\hat{\\theta}$", "$\\hat{\\sigma}^2$"),
      escape    = FALSE,
      caption   = "Comparison of $\\hat{\\theta}$ and $\\hat{\\sigma}^2$ estimates. Note: $\\tilde\\theta_{\\text{naive}} > 1$ falls outside the desired parameter space $|\\theta| < 1$.") |>
  kable_styling(latex_options = "hold_position",
                font_size = 10)
```

Because the Q1 residuals $\hat{e}_i$ contain mean-estimation error, i.e., they are $y_i - \mu_i(\hat\alpha,\hat\beta,\hat\gamma)$ rather than $w_i = y_i - \mu_i(\alpha,\beta,\gamma)$, the estimating equations applied here are only approximations to those derived in Q2(a) and (b). The two estimating equations yield different estimates of $\theta$, though both target the same population parameter, that parameter being the AR(1) autocorrelation coefficient $\theta$ in model (3). The numerical difference between $\tilde{\theta}_{\text{naive}} =$ `r round(theta_naive, 3)` and $\tilde{\theta}_{\text{opt}} =$ `r round(theta_opt, 3)` reflects estimation error rather than any difference in what the equations are targetting, with the optimal estimate being the more reliable of the two (again, "better" by Godambe's criterion.)

More importantly, the three $\hat\sigma^2$ values are **not** all estimating the same quantity. The Q1 GLS estimate was obtained under the assumption that the $w_i$ are iid, so it targets $\text{Var}(w_i) = \sigma^2/(1-\theta^2)$; the estimator itself does not account for the AR(1) structure that connects that marginal variance to $\sigma^2$. The two EE-based estimates, by contrast, are obtained by squaring residuals of the form $W_i - \tilde\theta W_{i-1} \approx \epsilon_i$, so they target the innovation variance $\sigma^2$ directly. Because $\theta > 0$ is typical for sorption data, as noted, the marginal variance exceeds the innovation variance, and we expect $\hat\sigma^2_{\text{Q1}} > \hat\sigma^2_{\text{naive}}$ and $\hat\sigma^2_{\text{Q1}} > \hat\sigma^2_{\text{opt}}$. Also notably the two innovation-variance estimates share that common target but can differ substantially, since $\hat\sigma^2_{\text{naive}}$ is based on distorted squared innovations from the inadmissible $\tilde\theta_{\text{naive}} = 1.005$.

# Q3

Assume $x_0 = 0$ and $w_0 = 0$. Derive the full log likelihood for the model of (1), (2), and (3), and maximize it to find simultaneous estimates of $\alpha$, $\beta$, $\gamma$, $\theta$, and $\sigma^2$.

## **(a)** An expression for the log likelihood.

Given $w_0 = 0$, the residuals $w_i = Y_i - \mu_i$ satisfy $w_i = \theta w_{i-1} + \epsilon_i$ with $\epsilon_i \overset{iid}{\sim} N(0,\sigma^2)$. 

The likelihood derived here is a *conditional* likelihood, conditioning on $w_0 = 0$ rather than treating $W_1$ as drawn from the stationary marginal distribution, $N(0, \sigma^2/(1-\theta^2))$. 

Given this conditioning, each innovation $\epsilon_i = w_i - \theta w_{i-1}$ is independent, so the joint density of $(Y_1,\ldots,Y_n)$ may be written as a product of $n$ normal densities. 

Noting the above, the log likelihood may be written

$$
\ell(\alpha,\beta,\gamma,\theta,\sigma^2)
= \sum_{i=1}^n \left[ -\frac{1}{2}\log(2\pi) - \frac{1}{2}\log(\sigma^2)
  - \frac{(w_i - \theta w_{i-1})^2}{2\sigma^2} \right]
= -\frac{n}{2}\log(2\pi)
  -\frac{n}{2}\log(\sigma^2)
  -\frac{1}{2\sigma^2}
    \sum_{i=1}^n \bigl(w_i - \theta w_{i-1}\bigr)^2,
$$

where $w_i = Y_i - \mu_i = Y_i - \alpha x_i^{\beta x_i^{-\gamma}}$ and $w_0 = 0$.

```{r q3-a, echo = F, message = F, warning = F}
loglik_q3 <- function(par, x, y) {
  alpha  <- par[1]
  beta   <- par[2]
  gamma  <- par[3]
  theta  <- par[4]
  sigma2 <- par[5]

  if (alpha  <= 0) return(-Inf)
  if (beta   <= 0) return(-Inf)
  if (gamma  <= 0) return(-Inf)
  if (sigma2 <= 0) return(-Inf)
  if (abs(theta) >= 1) return(-Inf)

  mu     <- alpha * x^(beta * x^(-gamma))
  w      <- y - mu
  w_prev <- c(0, w[-length(w)])
  eps    <- w - theta * w_prev

  n  <- length(y)
  ll <- -(n/2) * log(2 * pi) - (n/2) * log(sigma2) - sum(eps^2) / (2 * sigma2)
  return(ll)
}
```

## **(b)** A description of how you maximized the log likelihood. Note that Q1 estimates provide natural starting values and $\theta_0 = 0$; `optim` with `method = "BFGS"` is recommended over Nelder-Mead for five parameters.

### Wald 

The log likelihood was maximized using `optim` with `method = "BFGS"`. Starting values for $(\alpha, \beta, \gamma)$ were taken from the Q1 GLS fit; $\theta_0 = 0$ and $\sigma^2_0 = \hat\sigma^2_{\text{Q1}}$ completed the five-parameter starting vector. The Q1 fit is the $\theta = 0$ special case of the full model, placing the initial iterate near the (perhaps "an") expected solution. Multiple starting values were also considered to confirm convergence. Because `optim` minimizes by default, we pass the negative log likelihood. The Hessian was then extracted by specifying `hessian = TRUE` for Wald inference in part (c).

```{r q3-b, echo = F, message = F, warning = F}
# Negative log likelihood for optim
negll <- function(par, x, y) -loglik_q3(par, x, y)

ps0_q3 <- c(bhat[1], bhat[2], bhat[3], 0, s2hat)

fit3 <- optim(
  par     = ps0_q3,
  fn      = negll,
  x       = x,
  y       = y,
  method  = "BFGS",
  hessian = TRUE
)

if (fit3$convergence != 0)
  warning("optim did not converge: code ", fit3$convergence)

# q3-b chunk — add these lines after fit3
mle        <- fit3$par
names(mle) <- c("alpha", "beta", "gamma", "theta", "sigma2")
obs_info   <- fit3$hessian
cov_mle    <- solve(obs_info)
se_mle     <- sqrt(diag(cov_mle))
z95        <- qnorm(0.975)
```

### Profile Method 

Profile likelihood confidence intervals were constructed to complement the Wald intervals, as Wald intervals can perform poorly when the log-likelihood is asymmetric or the MLEs are near parameter boundaries---both plausible concerns in this application. For each parameter $\psi \in \{\alpha, \beta, \gamma, \theta, \sigma^2\}$, the profile log-likelihood is

$$
\ell_p(\psi) = \max_{\boldsymbol\lambda} \; \ell(\psi, \boldsymbol\lambda),
$$

where $\boldsymbol\lambda$ denotes the remaining four parameters. At each point of a grid spanning $\hat{\psi} \pm 12$ Wald standard errors for $(\alpha, \beta, \gamma)$ and $\hat{\psi} \pm 4$ Wald standard errors for $(\theta, \sigma^2)$---the wider range for the regression parameters reflecting the asymmetric likelihood surface and the need to capture the $\alpha > 0$ boundary region---where the Wald SE is taken from the observed information matrix at the MLE, truncated to the parameter's natural support. The inner maximization over $\boldsymbol\lambda$ was performed using `optim` with `method = "BFGS"`, initialized at the full MLE. Results were confirmed with Nelder-Mead, though BFGS was retained for later use. 

The 95\% confidence interval for $\psi$ is the set of grid values satisfying

$$
\ell_p(\psi) \geq \hat\ell - \tfrac{1}{2}\chi^2_{0.95,\,1}.
$$

Note: The choice to profile over single parameters is one of convenience; profiling over pairs or higher-dimensional subsets is equally valid but produced negligible differences in my preliminary analysis.

```{r q3-b-profile, echo = F, message = F, warning = F}
cutoff      <- qchisq(0.95, df = 1) / 2
ll_max      <- -fit3$value
multipliers <- c(12, 12, 12, 4, 4)

profile_ci_one <- function(par_idx, n_grid = 100) {
  mle_val <- mle[par_idx]
  se_val  <- se_mle[par_idx]
  hard_lo <- c(1e-6, 1e-6, 1e-6, -1 + 1e-4, 1e-6)[par_idx]
  hard_hi <- c(Inf,  Inf,  Inf,   1 - 1e-4,  Inf )[par_idx]
  lo_raw  <- mle_val - multipliers[par_idx] * se_val
  hi_raw  <- mle_val + multipliers[par_idx] * se_val
  grid    <- seq(max(lo_raw, hard_lo), min(hi_raw, hard_hi), length.out = n_grid)
  pl <- vapply(grid, function(psi) {
    obj <- function(p_other) {
      par_full           <- numeric(5)
      par_full[-par_idx] <- p_other
      par_full[ par_idx] <- psi
      negll(par_full, x, y)
    }
    res <- tryCatch(
      optim(mle[-par_idx], obj, method = "BFGS",
            control = list(maxit = 2000, reltol = 1e-10)),
      error = function(e) list(value = Inf)
    )
    if (!is.finite(res$value)) {
      res <- tryCatch(
        optim(mle[-par_idx], obj, method = "Nelder-Mead",
              control = list(maxit = 5000, reltol = 1e-10)),
        error = function(e) list(value = Inf)
      )
    }
    -res$value
  }, numeric(1))
  target <- ll_max - cutoff
  interp_cross <- function(side) {
    idx       <- if (side == "lower") which(grid < mle_val) else which(grid > mle_val)
    vals_side <- pl[idx]
    grid_side <- grid[idx]
    finite    <- is.finite(vals_side)
    vals_side <- vals_side[finite]
    grid_side <- grid_side[finite]
    below     <- which(vals_side < target)
    if (length(below) == 0) return(NA_real_)
    b    <- if (side == "lower") max(below) else min(below)
    b_in <- if (side == "lower") b + 1      else b - 1
    if (b_in < 1 || b_in > length(grid_side)) return(grid_side[b])
    approx(x = vals_side[c(b, b_in)],
           y = grid_side[c(b, b_in)],
           xout = target)$y
  }
  list(
    ci   = c(lower = interp_cross("lower"), upper = interp_cross("upper")),
    grid = grid,
    pl   = pl
  )
}

prof_bounds <- lapply(1:5, profile_ci_one)
prof_lower  <- sapply(prof_bounds, function(b) b$ci["lower"])
prof_upper  <- sapply(prof_bounds, function(b) b$ci["upper"])
```

## **(c)** The MLEs, 95% confidence intervals (stating whether Wald or profile), and the maximized log likelihood value.

### Wald

Wald confidence intervals were constructed from the observed information matrix (the negative Hessian of $\ell$ evaluated at the MLE, returned directly by `optim` via `hessian = TRUE`) used in place of the expected Fisher information. These are asymptotically equivalent under standard regularity conditions, though the distinction is worth noting here. With $n = 23$ and $p = 5$, these intervals rely heavily on asymptotic normality and should be interpreted with caution, as results will show.

### Profile Method

Profile 95\% confidence intervals are reported alongside the Wald intervals in Table 3. The most consequential difference is for $\alpha$: the Wald lower bound of $-0.4834$ is negative and therefore inadmissible under the model, which requires $\alpha \geq 0$. The profile lower bound of $0.0801$ correctly respects this constraint, reflecting the asymmetry of the likelihood surface near the boundary. Right-skew is also evident for $\sigma^2$. Where the two interval types agree closely (for $\gamma$ and $\theta$) the likelihood surface is approximately quadratic and the Wald approximation is adequate. Given the inadmissibility of the Wald interval for $\alpha$, profile confidence intervals are used as the basis for later analysis, when noted.

The profile upper bound for $\theta$ of $0.9151$ also warrants some note: Its estimated value lies close to the boundary $|\theta| = 1$, meaning the data are consistent with near-unit-root serial dependence. Under near-unit-root conditions, the marginal variance $\sigma^2 / (1 - \theta^2)$ becomes large and poorly estimated, and the conditional likelihood approximation based on $w_0 = 0$ may be less accurate than asymptotic theory suggests. This is a limitation that cannot be resolved here for $n = 23$.

Note: Profile log-likelihood maxima matched $\hat\ell = -29.598$ to within $0.001$ across all parameters, confirming that the inner BFGS optimization successfully recovered the MLE at each fixed value.

```{r q3-c, echo = F, message = F, warning = F}
prof_tab <- data.frame(
  Parameter = c("$\\alpha$", "$\\beta$", "$\\gamma$", "$\\theta$", "$\\sigma^2$"),
  MLE       = unname(round(mle,                4)),
  W_Lower   = unname(round(mle - z95 * se_mle, 4)),
  W_Upper   = unname(round(mle + z95 * se_mle, 4)),
  P_Lower   = unname(round(prof_lower,         4)),
  P_Upper   = unname(round(prof_upper,         4)),
  row.names = NULL
)

kable(prof_tab,
      format    = "latex",
      booktabs  = TRUE,
      escape    = FALSE,
      align     = c("l", "r", "r", "r", "r", "r"),
      col.names = c("Parameter", "MLE", "Lower", "Upper", "Lower", "Upper"),
      caption   = paste0("Wald and profile 95\\% confidence intervals at the MLE. ",
                         "Maximized log likelihood: ", round(ll_max, 4), ".")) |>
  add_header_above(c(" " = 2, "Wald 95% CI" = 2, "Profile 95% CI" = 2)) |>
  kable_styling(latex_options = "HOLD_position", font_size = 10)
```

## **(d)** A plot of the data with the estimated expectation function overlaid along with an approximate 95% pointwise confidence band.

The confidence band for $\mu_i = \alpha x_i^{\beta x_i^{-\gamma}}$ uses the delta method with the $3 \times 3$ submatrix of $\widehat{\text{Cov}}$ corresponding to $(\alpha, \beta, \gamma)$. The derivatives are the same as in Q1, in addition to its method of construction.

```{r q3-d, echo = F, message = F, warning = F, fig.height=4.5, fig.width=6, fig.cap="Fitted extended Freundlich curve with approximate 95\\% pointwise confidence band via the delta method (not profile-based)."}

alpha3 <- mle["alpha"]; beta3 <- mle["beta"]; gamma3 <- mle["gamma"]

mugrid3 <- efctn(xgrid, mle[1:3])
Dgrid3  <- eders(xgrid, mle[1:3])

# 3x3 sub-covariance for (alpha, beta, gamma)
cov_abg <- cov_mle[1:3, 1:3]

pvar3 <- apply(Dgrid3, 1,
               function(d) as.numeric(t(d) %*% cov_abg %*% d))
hw3   <- z95 * sqrt(pvar3)

plot(x, y, xlab = "Covariate Value", ylab = "Response Value", main = "")
lines(xgrid, mugrid3, lwd = 2)
lines(xgrid, mugrid3 + hw3, lty = 2)
lines(xgrid, mugrid3 - hw3, lty = 2)
```

## **(e)** A comparison of the large-scale component ($\alpha$, $\beta$, $\gamma$) to Q1: point estimates, interval estimates, estimated correlations among parameter estimators, and confidence band widths. Discuss any differences.

```{r q3-e, echo = F, message = F, warning = F}
# width_q1 <- 2 * z95 * sqrt(diag(fit1$covb))
# width_q3 <- 2 * z95 * se_mle[1:3]
# --- Point estimates and CI widths
width_q1 <- 2 * z95 * sqrt(diag(fit1$covb))
width_q3 <- prof_upper[1:3] - prof_lower[1:3]

comp_tab <- data.frame(
  Parameter  = c("$\\alpha$", "$\\beta$", "$\\gamma$"),
  Q1_Est     = unname(round(bhat,      4)),
  Q3_Est     = unname(round(mle[1:3], 4)),
  Q1_Width   = unname(round(width_q1, 4)),
  Q3_Width   = unname(round(width_q3, 4))
)

kable(comp_tab,
      format    = "latex",
      booktabs  = TRUE,
      escape    = FALSE,
      align     = c("l", "r", "r", "r", "r"),
      col.names = c("Parameter", "Q1 GLS", "Q3 MLE", "Q1 GLS", "Q3 Profile"),
      caption   = "Point estimates and 95\\% CI widths by method (Q1 Wald, Q3 profile).") |>
  add_header_above(c(" " = 1, "Point Estimate" = 2, "CI Width" = 2)) |>
  kable_styling(latex_options = "HOLD_position", font_size = 10)

# --- Correlation matrices
cor_q1 <- round(cov2cor(fit1$covb),          3)
cor_q3 <- round(cov2cor(cov_mle[1:3, 1:3]), 3)
pnames <- c("$\\alpha$", "$\\beta$", "$\\gamma$")
rownames(cor_q1) <- colnames(cor_q1) <- pnames
rownames(cor_q3) <- colnames(cor_q3) <- pnames

kable(cor_q1,
      format   = "latex",
      booktabs = TRUE,
      escape   = FALSE,
      caption  = "Estimated correlation matrix for $(\\alpha, \\beta, \\gamma)$: Q1 GLS.") |>
  kable_styling(latex_options = "HOLD_position", font_size = 10)

kable(cor_q3,
      format   = "latex",
      booktabs = TRUE,
      escape   = FALSE,
      caption  = "Estimated correlation matrix for $(\\alpha, \\beta, \\gamma)$: Q3 MLE.") |>
  kable_styling(latex_options = "HOLD_position", font_size = 10)
```

The point estimates of $(\alpha, \beta, \gamma)$ shift noticeably between Q1 and Q3, most substantially $\hat\alpha$ moving from $0.5063$ to $0.7395$. This may be attributed to the fact that jointly estimating $\theta$ reweights the observations according to the AR(1) correlation structure rather than treating them as independent. When correlation parameters are known, GLS with $\Sigma^{-1}(\theta)$ as a fixed weight matrix improves efficiency over the iid estimator. When they must be estimated jointly, however, the additional uncertainty in $\theta$ and $\sigma^2$ propagates into the regression parameters and can widen confidence intervals, as evidenced here.

The profile CI widths under Q3 are wider than Q1 Wald for $\alpha$ and $\gamma$, but narrower for $\beta$. This is likely because the shift in $\hat\beta$ on the likelihood surface places it in a region of higher curvature. Additionally, the mean confidence band width increases from $0.6225$ to $0.9666$. The inadmissible Wald lower bound of $-0.4834$ for $\alpha$ is replaced by a profile lower bound of $0.0801$, respecting the $\alpha \geq 0$ constraint. This reflects the additional uncertainty introduced by estimating $\theta$ and $\sigma^2$ jointly with the regression parameters: the likelihood surface is more diffuse in the $(\alpha, \beta, \gamma)$ directions once the correlated error structure is built in, and the Q1 covariance matrix, which ignored this, was understating that uncertainty. It should also be noted that Table 4 compares Q1 Wald intervals to Q3 profile intervals --- which are distinct constructions. Profile intervals are generally wider than Wald when the likelihood surface is asymmetric, so part of the observed width difference reflects the interval type rather than solely the model specification change.

Part of the observed width difference is also attributable to $\hat\sigma^2_{\text{Q1}} =$ `r round(s2_q1, 3)` $>$ $\hat\sigma^2_{\text{MLE}} =$ `r round(mle["sigma2"], 3)`. Because Q1 uses a larger variance estimate, its standard errors are inflated relative to what they would be under the "correct" innovation variance. The structural difference between the models (ignoring vs.\ jointly estimating the AR(1) structure) accounts for more of the width increase than the raw comparison in Table 4 suggests. The currently observed Q3 $>$ Q1 width comparison therefore understates the true inferential cost of the iid misspecification. The estimated correlations among the regression parameter estimators also shift under Q3, with $|\hat\rho_{\alpha,\beta}| \approx 0.97$ under both methods indicating that the likelihood surface is extremely flat along the $(\alpha, \beta)$ surface, indicating that individual parameter estimates are highly sensitive to fluctuations regardless of the fitting method, and the wide confidence intervals are intrinsic to the model structure rather than an artefact of the estimation procedure.

## **(f)** A comparison of the small-scale component ($\theta$, $\sigma^2$) to the optimal estimating equation results from Q2. Two estimates of $\theta$ and three estimates of $\sigma^2$ are available (Q3 MLE, Q1 GLS, Q2 moment-based). Discuss any differences.

```{r q3-f, echo = F, message = F, warning = F}
comp_small <- data.frame(
  Quantity = c("$\\theta$", "$\\sigma^2$"),
  Q2_OptEE = unname(round(c(theta_opt, s2_opt),          4)),
  Q3_MLE   = unname(round(mle[c("theta", "sigma2")],     4))
)

kable(comp_small,
      format    = "latex",
      booktabs  = TRUE,
      escape    = FALSE,
      align     = c("l", "r", "r"),
      col.names = c("Quantity", "Q2 Optimal EE", "Q3 MLE"),
      caption   = "Comparison of small-scale parameter estimates.") |>
  kable_styling(latex_options = "HOLD_position", font_size = 10)

all_s2 <- data.frame(
  Estimator = c("Q1 GLS (marginal variance)", "Q2 moment-based", "Q3 MLE"),
  sigma2    = round(c(s2_q1, s2_opt, mle["sigma2"]), 4)
)

kable(all_s2,
      format    = "latex",
      booktabs  = TRUE,
      escape    = FALSE,
      align     = c("l", "r"),
      col.names = c("Estimator", "$\\hat{\\sigma}^2$"),
      caption   = "All estimates of $\\sigma^2$ across methods.") |>
  kable_styling(latex_options = "HOLD_position", font_size = 10)
```

The two estimates of $\theta$ largely agree: the Q2 optimal estimating equation yields $\hat\theta = 0.5422$ and the Q3 MLE yields $\hat\theta = 0.5598$. Both target the same parameter but differ in how they handle the regression function. The Q2 estimate used raw residuals from the misspecified iid model of Q1 as surrogates for $w_i$, introducing noise from the first-stage estimation; the Q3 MLE estimates $\theta$ simultaneously with $(\alpha, \beta, \gamma)$, which should in theory reduce that bias. The small discrepancy is consistent with this explanation.

For $\sigma^2$, the three estimators target different quantities, as established in Q2(d). As such, the Q1 GLS estimate of `r round(s2_q1, 4)` targets the marginal variance $\sigma^2 / (1 - \theta^2)$ and is therefore expected to be the largest when $\theta > 0$, as evidenced in the above. Also, both the Q2 moment-based estimate (`r round(s2_opt, 4)`) and the Q3 MLE (`r round(mle["sigma2"], 4)`) target the innovation variance $\sigma^2$, and are indeed substantially smaller. The Q2 version "inherits" approximation error from using Q1 residuals instead of true $w_i$ values, while the Q3 MLE is obtained from the correctly specified joint model; the remaining discrepancy of `r round(abs(s2_opt - mle["sigma2"]), 4)` between them is then attributable to the first-stage approximation.

\newpage 

# Q4

Assume $x_0 = 0$ and $w_0 = 0$. Using the events $A_i = (Y_i = y_i \mid Y_{i-1} = y_{i-1})$ for $i = 2, \ldots, n$, derive a pairwise conditional log composite likelihood and fit the model. Report the same components as in Q3.

Additionally, discuss which asymptotic context for composite likelihood is most appropriate given $n = 23$. Address potential issues with estimating the $H$ and $J$ matrices in the Godambe information; compute them if possible and use them to construct interval estimates for the regression function parameters.

## **(a)** Expression for the Log Composite Likelihood

We are given the model $w_i = Y_i - \mu_i$ and the AR(1) structure $w_i = \theta w_{i-1} + \epsilon_i$ with $\epsilon_i \sim N(0, \sigma^2)$. 

Since $w_i \mid w_{i-1} \sim N(\theta w_{i-1}, \sigma^2)$, the conditional distribution of $Y_i$ given $Y_{i-1} = y_{i-1}$ is

$$
Y_i \mid Y_{i-1} = y_{i-1} \sim N\!\bigl(\mu_i + \theta(y_{i-1} - \mu_{i-1}),\; \sigma^2\bigr).
$$

The corresponding conditional density is

$$
f(y_i \mid y_{i-1}) = \frac{1}{\sqrt{2\pi\sigma^2}} \exp\!\left(-\frac{\bigl[(y_i - \mu_i) - \theta(y_{i-1} - \mu_{i-1})\bigr]^2}{2\sigma^2}\right).
$$

Letting $\epsilon_i = (y_i - \mu_i) - \theta(y_{i-1} - \mu_{i-1})$ and applying log simplification rules, each conditional log-likelihood piece is

$$
\ell_i = \log f(y_i \mid y_{i-1})
       = \log\!\left[(2\pi\sigma^2)^{-1/2}\right] + \log\!\left[\exp\!\left(-\frac{\epsilon_i^2}{2\sigma^2}\right)\right]
       = -\frac{1}{2}\log(2\pi) - \frac{1}{2}\log\sigma^2 - \frac{\epsilon_i^2}{2\sigma^2}.
$$

Summing over $i = 2, \ldots, n$, the constants $-\tfrac{1}{2}\log(2\pi)$ and $-\tfrac{1}{2}\log\sigma^2$ are summed $n-1$ times, with a factor of $\sigma^2$ able to be pulled out of the sum in the latter, giving the pairwise conditional log composite likelihood

$$
\ell_C = \sum_{i=2}^n \ell_i
       = \sum_{i=2}^n \left[-\frac{1}{2}\log(2\pi) - \frac{1}{2}\log\sigma^2 - \frac{\epsilon_i^2}{2\sigma^2}\right]
       = -\frac{n-1}{2}\log(2\pi) - \frac{n-1}{2}\log\sigma^2 - \frac{1}{2\sigma^2}\sum_{i=2}^n \epsilon_i^2.
$$

Note: The composite likelihood conditions on $Y_1$ (i.e., sums from $i = 2$).

```{r q4-loglik, echo = F, message = F, warning = F}
# Log composite likelihood: sum of conditional pieces i = 2, ..., n
logcl_q4 <- function(par, x, y) {
  alpha  <- par[1]
  beta   <- par[2]
  gamma  <- par[3]
  theta  <- par[4]
  sigma2 <- par[5]

  if (alpha  <= 0) return(-Inf)
  if (beta   <= 0) return(-Inf)
  if (gamma  <= 0) return(-Inf)
  if (sigma2 <= 0) return(-Inf)
  if (abs(theta) >= 1) return(-Inf)

  mu    <- alpha * x^(beta * x^(-gamma))
  w     <- y - mu
  # epsilon_i = w_i - theta * w_{i-1}, summed from i = 2
  eps   <- w[-1] - theta * w[-length(w)]

  m  <- length(eps)   # n - 1
  lc <- -(m/2) * log(2 * pi) - (m/2) * log(sigma2) -
        sum(eps^2) / (2 * sigma2)
  return(lc)
}

negcl <- function(par, x, y) -logcl_q4(par, x, y)
```

```{r q4-fit, echo = F, message = F, warning = F}
# Starting values from Q1 GLS (same as Q3)
ps0_q4 <- c(bhat[1], bhat[2], bhat[3], 0, s2hat)

fit4 <- optim(
  par     = ps0_q4,
  fn      = negcl,
  x       = x,
  y       = y,
  method  = "BFGS",
  hessian = TRUE
)

if (fit4$convergence != 0)
  warning("optim did not converge: code ", fit4$convergence)

mcle        <- fit4$par
names(mcle) <- c("alpha", "beta", "gamma", "theta", "sigma2")
```

## **(b)** Maximization

The log composite likelihood was maximized using `optim` with `method = "BFGS"`, using Q1 GLS estimates as starting values for $(\alpha, \beta, \gamma)$ and $\theta_0 = 0$, $\sigma^2_0 = \hat\sigma^2_{\text{Q1}}$, though other starting values were also considered to ensure appropriate convergence.

## **(c)** MCLEs, Confidence Intervals, and Maximized Log Composite 

Two sets of 95\% confidence intervals are reported in Table~9: Godambe Wald (sandwich covariance $\hat{G}^{-1}$) and Godambe-adjusted profile. The adjusted profile threshold requires the $\lambda_\psi$ scaling correction detailed and derived below. Similar to Q3, the adjusted profile intervals are preferred as they correctly respect the $\alpha \geq 0$ (and $\beta \geq 0$) boundary and account for likelihood asymmetry.

## Asymptotic Context and Godambe Information

This problem falls under Asymptotic Context II (Ch. 9.3.2): a single realization of a length-$n = 23$ dependent sequence. Neither formal asymptotic result from Ch. 9 applies directly. Context I requires iid replicates, which are unavailable. The Comets--Janzura result for Context II requires translation-invariant conditional specification, which fails here because $\mu_i$ varies with the covariate. The question is then whether computing empirical analogs of $H$ and $J$ has any theoretical justification. What follows is a good-faith attempt at a justification, though it is not a fully rigorous treatment of the question.

A justification lies in the structure of the AR(1) model itself. Each piece of the conditional composite likelihood conditions on $Y_{i-1}$, so the residual entering $\ell_i$ is exactly $\epsilon_i = e_i \sim \mathrm{iid}\; N(0, \sigma^2)$. The piece score $\nabla\ell_i$ is a function of $e_i$ and $Y_{i-1}$, and because $e_i \perp Y_{i-1}$ by the AR(1) structure,

$$
E[\nabla\ell_i \mid Y_{i-1}] = 0.
$$

The sequence $\{\nabla\ell_i(\boldsymbol\psi_0)\}$ evaluated at the true parameter $\boldsymbol\psi_0$ is a martingale difference sequence (MDS). Notably, this concept is not covered in this course; that point notwithstanding, MDS admits a CLT and LLN under conditions that do not require iid replicates or second-order stationarity of the $Y_i$, though those conditions cannot be verified with the information and assumptions given. At the estimated $\hat{\boldsymbol\psi}$, the MDS property holds only approximately. In the language of Geyer (2020), the composite score is an unbiased estimating function whose asymptotic variance takes the sandwich form $U^{-1}V(U^{-1})^T$, which is inverse Godambe information. The MDS property is what makes the natural estimators $\hat{H}$ and $\hat{J}$ consistent for those matrices (conditions (8) and (9) of Geyer (2020)).

With $n = 23$ observations and five parameters, both matrices should be interpreted cautiously regardless: because $\hat{H}$ and $\hat{J}$ must be estimated empirically from sums of $n - 1 = 22$ score contributions, small samples can produce unstable estimates of the sandwich variance, and their finite-sample performance is uncertain. The MDS argument provides the theoretical grounding that makes computing Godambe intervals meaningful rather than purely heuristic, and guarantees a limiting distribution for any bootstrap procedure to approximate. As noted in Ch. 9.3.3, the parametric bootstrap remains the more reliable alternative in practice, particularly fpr $n = 23$.

Note: The Geyer references are noted at the end of this exam under "Outside Resources". 

## $\hat{H}$, $\hat{J}$, and Godambe-Based Intervals

With this justification, imperfect as it may be, the empirical analogs are computed as follows. The Godambe information matrix is $G = H^{-1} J H^{-1}$, with $H = -E[\nabla^2 \ell_C]$ and $J = \text{Var}[\nabla \ell_C]$, such that $\hat{G}^{-1} = \hat{H}^{-1} \hat{J} \hat{H}^{-1}$ serves as the sandwich covariance for the MCLE. $\hat{H}$ is taken as the negative Hessian of $\ell_C$ evaluated at the MCLE, returned directly by \texttt{optim}. $\hat{J}$ is the sum of outer products of individual piece score vectors $\nabla\ell_i$, approximated by numerical central differences at the MCLE. This outer-product form parallels the Context I estimator in Ch. 9.3.1.

Examining Tables 10 and 11, the $(3,3)$ elements of both $\hat{H}$ (38593.78) and $\hat{J}$ (40035.62) are very large relative to all others, reflecting high curvature in the $\gamma$ direction and indicating that $\gamma$ is the most precisely determined parameter locally. The large negative off-diagonal entries in rows/columns 1--3 and 2--3 are consistent with the high $(\alpha, \gamma)$ and $(\beta, \gamma)$ correlations seen in Table 15, and confirm that the identifiability challenges in this model are concentrated in $(\alpha, \beta, \gamma)$

## Profile Composite Likelihoods

Unfortunately though perhaps unsurprisingly, the Godambe Wald intervals in Table~9 do not respect parameter boundaries and inherit the symmetric normal approximation. Profile intervals address both limitations, as in Q3, but the composite likelihood ratio statistic requires a Godambe-based correction before comparison to a $\chi^2_1$ threshold.

Profile composite likelihood intervals are constructed analogously to Q3, 
with one modification. The composite likelihood ratio statistic

$$
W_C(\psi_0) = 2\bigl[\ell_C(\hat\psi) - \ell_C(\psi_0,\,\hat{\boldsymbol\lambda}_{\psi_0})\bigr]
$$

is not asymptotically $\chi^2_1$; instead, it converges to $\lambda_\psi \chi^2_1$, where the scalar correction

$$
\lambda_\psi
  = \frac{(\hat{G}^{-1})_{\psi, \psi}}{(\hat{H}^{-1})_{\psi, \psi}}
  = \frac{\text{Godambe variance for } \psi}{\text{Wald variance for } \psi}
$$

is the ratio of the Godambe to Wald marginal variance for parameter $\psi$, evaluated at the MCLE. The adjusted profile threshold is therefore

$$
\ell_C(\psi) \;\geq\; \hat\ell_C - \tfrac{1}{2}\,\lambda_\psi\,\chi^2_{0.95,\,1}.
$$

Because $\hat{J} \approx \hat{H}$ (see Tables 10 and 11), $\lambda_\psi \approx 1$ for all parameters, so the correction has modest practical effect here. The adjusted profile intervals are reported alongside the Godambe Wald intervals in Table~9. Note that $\hat{H}$ and $\hat{J}$ are not available until after the Godambe computation; the profile grid is evaluated first and the threshold is applied retrospectively once $\lambda_\psi$ is known.

Importantly, the adjusted profile lower bound for $\gamma$ does not exist at the 95% level: the composite likelihood surface to the left of $\hat\gamma = 0.145$ remains above the adjusted threshold all the way to the parameter boundary, in contrast to the full likelihood of Q3, which yields a lower bound of $0.068$. This is a direct consequence of the composite likelihood using only $n - 1 = 22$ pairs rather than all $n = 23$ observations. This reduction in information is just enough (in this instance) to render $\gamma$'s lower bound undefined under the composite likelihood. Figure~3 confirms this is a structural issue. The normed profile composite likelihood (y-axis of Figure 3) enters above the adjusted cutoff of $\approx 0.1465$ at the left boundary of the admissible parameter space ($\gamma > 0$) and never descends through it, rather than exhibiting the smooth symmetric crossing seen for the remaining parameters.

```{r q4-intervals, echo = F, message = F, warning = F}
cov_wald4 <- solve(fit4$hessian)
se_wald4  <- sqrt(diag(cov_wald4))
# Numerical score for a single piece i (i indexes into 2:n, so piece i
# corresponds to the pair (y[i], y[i-1]) with i in 2:n)
score_piece <- function(par, x, y, i) {
  h   <- 1e-5 * abs(par)
  h[h < 1e-8] <- 1e-8
  p   <- length(par)
  scr <- numeric(p)
  for (k in seq_len(p)) {
    par_hi       <- par; par_hi[k] <- par[k] + h[k]
    par_lo       <- par; par_lo[k] <- par[k] - h[k]
    # single-piece log CL for pair (y[i-1], y[i])
    piece <- function(pv) {
      a <- pv[1]; b <- pv[2]; g <- pv[3]
      th <- pv[4]; s2 <- pv[5]
      mu   <- a * x^(b * x^(-g))
      w    <- y - mu
      ei   <- w[i] - th * w[i - 1]
      -0.5 * log(2 * pi) - 0.5 * log(s2) - ei^2 / (2 * s2)
    }
    scr[k] <- (piece(par_hi) - piece(par_lo)) / (2 * h[k])
  }
  scr
}

n_obs <- length(y)
p     <- 5

# J: sum of outer products of piece scores
J_hat <- matrix(0, p, p)
for (i in 2:n_obs) {
  si    <- score_piece(mcle, x, y, i)
  J_hat <- J_hat + outer(si, si)
}

# H: negative Hessian of composite LL at MCLE (already computed by optim)
H_hat <- fit4$hessian   # Hessian of *negative* CL, so this is already -d^2 lC

# Godambe information and sandwich covariance
# G = H J^{-1} H  =>  G^{-1} = H^{-1} J H^{-1}
H_inv    <- solve(H_hat)
cov_god  <- H_inv %*% J_hat %*% H_inv
se_god   <- sqrt(diag(cov_god))
names(se_god) <- names(mcle)

# Godambe adjustment factors (requires cov_god and H_inv from Godambe chunk)
cutoff_cl   <- qchisq(0.95, df = 1) / 2
ll_max_cl   <- -fit4$value
lambda_psi  <- diag(cov_god) / diag(H_inv)
cutoff_adj  <- lambda_psi * cutoff_cl          # length-5 vector

multipliers_cl <- c(12, 12, 12, 4, 4)

profile_godambe_one <- function(par_idx, n_grid = 100) {
  mle_val <- mcle[par_idx]
  se_val  <- se_god[par_idx]
  hard_lo <- c(1e-6, 1e-6, 1e-6, -1 + 1e-4, 1e-6)[par_idx]
  hard_hi <- c(Inf,  Inf,  Inf,   1 - 1e-4,  Inf )[par_idx]
  lo_raw  <- mle_val - multipliers_cl[par_idx] * se_val
  hi_raw  <- mle_val + multipliers_cl[par_idx] * se_val
  # Allow grid to extend past hard_lo so -Inf wall produces a natural crossing
  grid    <- seq(lo_raw, min(hi_raw, hard_hi), length.out = n_grid)
  pl <- vapply(grid, function(psi) {
    obj <- function(p_other) {
      par_full           <- numeric(5)
      par_full[-par_idx] <- p_other
      par_full[ par_idx] <- psi
      negcl(par_full, x, y)
    }
    res <- tryCatch(
      optim(mcle[-par_idx], obj, method = "BFGS",
            control = list(maxit = 2000, reltol = 1e-10)),
      error = function(e) list(value = Inf)
    )
    if (!is.finite(res$value)) {
      res <- tryCatch(
        optim(mcle[-par_idx], obj, method = "Nelder-Mead",
              control = list(maxit = 5000, reltol = 1e-10)),
        error = function(e) list(value = Inf)
      )
    }
    -res$value
  }, numeric(1))
  target <- ll_max_cl - cutoff_adj[par_idx]
  interp_cross <- function(side) {
    idx       <- if (side == "lower") which(grid < mle_val) else which(grid > mle_val)
    vals_side <- pl[idx]
    grid_side <- grid[idx]
    if (side == "lower") {
      vals_side[!is.finite(vals_side)] <- -Inf
    } else {
      finite    <- is.finite(vals_side)
      vals_side <- vals_side[finite]
      grid_side <- grid_side[finite]
    }
    below <- which(vals_side < target)
    if (length(below) == 0) return(NA_real_)
    b    <- if (side == "lower") max(below) else min(below)
    b_in <- if (side == "lower") b + 1      else b - 1
    if (b_in < 1 || b_in > length(grid_side)) return(grid_side[b])
    # If the wall point itself is -Inf, interpolation is undefined — return boundary
    if (!is.finite(vals_side[b])) return(hard_lo)
    result <- approx(x = vals_side[c(b, b_in)],
                     y = grid_side[c(b, b_in)],
                     xout = target)$y
    if (side == "lower") result <- max(result, hard_lo)
    result
  }
  list(
    ci     = c(lower = interp_cross("lower"), upper = interp_cross("upper")),
    grid   = grid,
    pl     = pl,
    lambda = lambda_psi[par_idx]
  )
}

prof_god_bounds <- lapply(1:5, profile_godambe_one)
prof_god_lower  <- sapply(prof_god_bounds, function(b) b$ci["lower"])
prof_god_upper  <- sapply(prof_god_bounds, function(b) b$ci["upper"])

hard_lo_vec <- c(1e-6, 1e-6, 1e-6, -1 + 1e-4, 1e-6)
gp_lower_clean <- ifelse(
  !is.na(prof_god_lower) & abs(prof_god_lower - hard_lo_vec) < 1e-5,
  NA_real_,
  prof_god_lower
)

prof_god_tab <- data.frame(
  Parameter = c("$\\alpha$", "$\\beta$", "$\\gamma$", "$\\theta$", "$\\sigma^2$"),
  MCLE      = unname(round(mcle,                  4)),
  GW_Lower  = unname(round(mcle - z95 * se_god,   4)),
  GW_Upper  = unname(round(mcle + z95 * se_god,   4)),
  GP_Lower  = unname(round(gp_lower_clean,         4)),
  GP_Upper  = unname(round(prof_god_upper,         4)),
  row.names = NULL
)

kable(prof_god_tab,
      format    = "latex",
      booktabs  = TRUE,
      escape    = FALSE,
      align     = c("l", "r", "r", "r", "r", "r"),
      col.names = c("Parameter", "MCLE", "Lower", "Upper", "Lower", "Upper"),
      caption   = paste0("Godambe Wald and adjusted profile 95\\% confidence intervals. ",
                         "Maximized composite log likelihood: ", round(ll_max_cl, 4), ".")) |>
  add_header_above(c(" " = 2, "Godambe Wald 95% CI" = 2,
                               "Godambe Adj. Profile 95% CI" = 2)) |>
  kable_styling(latex_options = "HOLD_position", font_size = 10)
```

```{r q4-godambe-info, echo = F, message = F, warning = F}
pnames <- c("$\\alpha$", "$\\beta$", "$\\gamma$", "$\\theta$", "$\\sigma^2$")
H_tab <- as.data.frame(unname(round(H_hat, 4)))
rownames(H_tab) <- pnames
colnames(H_tab) <- pnames

kable(H_tab,
      format   = "latex",
      booktabs = TRUE,
      escape   = FALSE,
      caption  = "Estimated $H$ matrix (negative Hessian of composite log likelihood at MCLE).") |>
  kable_styling(latex_options = "HOLD_position", font_size = 9)

J_tab <- as.data.frame(unname(round(J_hat, 4)))
rownames(J_tab) <- pnames
colnames(J_tab) <- pnames

kable(J_tab,
      format   = "latex",
      booktabs = TRUE,
      escape   = FALSE,
      caption  = "Estimated $J$ matrix (sum of outer score products at MCLE).") |>
  kable_styling(latex_options = "HOLD_position", font_size = 9)
```

```{r q4-godambe-attempt-issues, echo=F, message=F, warning=F, fig.height=4.5, fig.width=6, fig.pos="H", fig.cap="Normed adjusted profile composite likelihood for $\\gamma$."}
cutoff_normed <- exp(-qchisq(0.95, df = 1) / 2)

pg       <- prof_god_bounds[[3]]
R_norm_g <- exp((pg$pl - ll_max_cl) / pg$lambda)

plot(pg$grid, R_norm_g,
     type = "l", lwd = 2, col = "black",
     xlab = expression(gamma),
     ylab = "Normed Profile Composite Likelihood",
     # main = expression("Normed Profile CL: " * gamma),
     xlim = c(-0.1, max(pg$grid)),
     ylim = c(0, 1.05))
abline(h = cutoff_normed,     lty = 2, col = "red")
abline(v = mcle[3],           lty = 1, col = "blue")
abline(v = prof_god_upper[3], lty = 2, col = "blue")
legend("topright",
       legend = c("Normed profile CL",
                  "95% cutoff (0.1465)",
                  expression(hat(gamma)),
                  "Upper CI bound"),
       lty = c(1, 2, 1, 2),
       col = c("black", "red", "blue", "blue"),
       lwd = c(2, 1, 1, 1),
       bty = "n", cex = 0.85)
```

\FloatBarrier

## **(d)** Fitted Curve and 95% Pointwise Confidence Band

Constructing a pointwise confidence band via the delta method requires a joint covariance matrix over $(\alpha, \beta, \gamma)$, not a collection of marginal profile intervals. The Godambe Wald covariance $\hat{G}^{-1}$ is therefore used for the band as the natural sandwich covariance for the composite likelihood. Note that this choice is independent of the $\gamma$ lower-bound issue: even if all five adjusted profile CIs had well-defined bounds, the delta-method band would still require $\hat{G}^{-1}$.

```{r q4-band, echo=F, message=F, warning=F, fig.height=4.5, fig.width=6, fig.pos="H", fig.cap="Fitted extended Freundlich curve with approximate 95\\% pointwise confidence band based on the Godambe covariance."}
mugrid4 <- efctn(xgrid, mcle[1:3])
Dgrid4  <- eders(xgrid, mcle[1:3])
cov_abg4 <- cov_god[1:3, 1:3]
pvar4    <- apply(Dgrid4, 1,
                  function(d) as.numeric(t(d) %*% cov_abg4 %*% d))
hw4      <- z95 * sqrt(pvar4)
plot(x, y, xlab = "Covariate Value", ylab = "Response Value", main = "")
lines(xgrid, mugrid4, lwd = 2)
lines(xgrid, mugrid4 + hw4, lty = 2)
lines(xgrid, mugrid4 - hw4, lty = 2)
```

\FloatBarrier

## **(e)** Comparison of Large-Scale Component to Q1 and Q3

```{r q4-comp-large, echo = F, message = F, warning = F}
pnames3 <- c("$\\alpha$", "$\\beta$", "$\\gamma$")

# --- Point estimates
comp_pt4 <- data.frame(
  Parameter = pnames3,
  Q1_GLS    = unname(round(bhat,      4)),
  Q3_MLE    = unname(round(mle[1:3],  4)),
  Q4_MCLE   = unname(round(mcle[1:3], 4))
)

kable(comp_pt4,
      format    = "latex",
      booktabs  = TRUE,
      escape    = FALSE,
      align     = c("l", "r", "r", "r"),
      col.names = c("Parameter", "Q1 GLS", "Q3 MLE", "Q4 MCLE"),
      caption   = "Point estimates of large-scale parameters across methods.") |>
  kable_styling(latex_options = "HOLD_position", font_size = 10)

# --- CI widths
# comp_width4 <- data.frame(
#   Parameter      = pnames3,
#   Width_Q1       = unname(round(2 * z95 * sqrt(diag(fit1$covb)), 4)),
#   # Width_Q3       = unname(round(2 * z95 * se_mle[1:3],           4)),
#   Width_Q3 = unname(round(prof_upper[1:3] - prof_lower[1:3], 4)),
#   Width_Q4_Wald  = unname(round(2 * z95 * se_wald4[1:3],         4)),
#   Width_Q4_God   = unname(round(2 * z95 * se_god[1:3],           4))
# )
# 
# kable(comp_width4,
#       format    = "latex",
#       booktabs  = TRUE,
#       escape    = FALSE,
#       align     = c("l", "r", "r", "r", "r"),
#       col.names = c("Parameter", "Q1 GLS", "Q3 MLE", "Q4 Wald", "Q4 Godambe"),
#       caption   = paste0("95\\% CI widths for large-scale parameters across methods ",
#                          "(Q1 Wald, Q3 profile, Q4 Wald and Godambe). ",
#                          "Mean confidence band half-width (Q4): ",
#                          round(mean(hw4), 4), ".")) |>
#   add_header_above(c(" " = 1, " " = 1, " " = 1, "Q4" = 2)) |>
#   kable_styling(latex_options = "HOLD_position", font_size = 10)
prof_god_width <- prof_god_upper[1:3] - gp_lower_clean[1:3]
# gamma lower is NA so its width is NA (one-sided interval)

comp_width4 <- data.frame(
  Parameter      = pnames3,
  Width_Q1       = unname(round(2 * z95 * sqrt(diag(fit1$covb)), 4)),
  Width_Q3       = unname(round(prof_upper[1:3] - prof_lower[1:3], 4)),
  Width_Q4_Prof  = unname(round(prof_god_width, 4)),
  Width_Q4_God   = unname(round(2 * z95 * se_god[1:3],           4))
)

kable(comp_width4,
      format    = "latex",
      booktabs  = TRUE,
      escape    = FALSE,
      align     = c("l", "r", "r", "r", "r"),
      col.names = c("Parameter", "Q1 GLS", "Q3 MLE", "Q4 Profile", "Q4 Godambe"),
      caption   = paste0("95\\% CI widths for large-scale parameters across methods ",
                         "(Q1 Wald, Q3 profile, Q4 adjusted profile and Godambe Wald). ",
                         "Q4 profile width for $\\gamma$ is one-sided (lower bound undefined).")) |>
  add_header_above(c(" " = 1, " " = 1, " " = 1, "Q4" = 2)) |>
  kable_styling(latex_options = "HOLD_position", font_size = 10)

# --- Correlation matrix
cor_q4 <- round(cov2cor(cov_god[1:3, 1:3]), 3)
rownames(cor_q4) <- colnames(cor_q4) <- pnames3

kable(cor_q4,
      format   = "latex",
      booktabs = TRUE,
      escape   = FALSE,
      caption  = "Estimated correlation matrix for $(\\alpha, \\beta, \\gamma)$: Q4 Godambe.") |>
  kable_styling(latex_options = "HOLD_position", font_size = 10)
```

```{r resid-comparison, echo = F, message = F, warning = F, fig.height=12, fig.width=9, fig.cap="Standardized residuals versus covariate under Q1 GLS (left), Q3 MLE (center), and Q4 MCLE (right)."}
# Q3 innovation residuals
w_q3      <- y - efctn(x, mle[1:3])
innov_q3  <- w_q3 - mle["theta"] * c(0, w_q3[-n])
stdres_q3 <- innov_q3 / sqrt(mle["sigma2"])

# Q4 innovation residuals
w_q4      <- y - efctn(x, mcle[1:3])
innov_q4  <- w_q4 - mcle["theta"] * c(0, w_q4[-n])
stdres_q4 <- innov_q4 / sqrt(mcle["sigma2"])

ylim_r <- range(c(fit1$stdres, stdres_q3, stdres_q4))

par(mfrow = c(3, 1))

plot(x, fit1$stdres, xlab = "Covariate Value", ylab = "Standardized Residual",
     main = "Q1 GLS", ylim = ylim_r)
abline(h = 0, lty = 3)

plot(x, stdres_q3, xlab = "Covariate Value", ylab = "",
     main = "Q3 MLE", ylim = ylim_r)
abline(h = 0, lty = 3)

plot(x, stdres_q4, xlab = "Covariate Value", ylab = "",
     main = "Q4 MCLE", ylim = ylim_r)
abline(h = 0, lty = 3)
```

```{r resid-comparison-sqrt, echo = F, message = F, warning = F, eval = F, fig.height=12, fig.width=9, fig.cap="Standardized residuals versus square root of covariate under Q1 GLS (left), Q3 MLE (center), and Q4 MCLE (right)."}
# Q3 innovation residuals
w_q3      <- y - efctn(x, mle[1:3])
innov_q3  <- w_q3 - mle["theta"] * c(0, w_q3[-n])
stdres_q3 <- innov_q3 / sqrt(mle["sigma2"])
# Q4 innovation residuals
w_q4      <- y - efctn(x, mcle[1:3])
innov_q4  <- w_q4 - mcle["theta"] * c(0, w_q4[-n])
stdres_q4 <- innov_q4 / sqrt(mcle["sigma2"])
ylim_r <- range(c(fit1$stdres, stdres_q3, stdres_q4))
par(mfrow = c(3, 1))
plot(sqrt(x), fit1$stdres, xlab = expression(sqrt("Covariate Value")), ylab = "Standardized Residual",
     main = "Q1 GLS", ylim = ylim_r)
abline(h = 0, lty = 3)
plot(sqrt(x), stdres_q3, xlab = expression(sqrt("Covariate Value")), ylab = "",
     main = "Q3 MLE", ylim = ylim_r)
abline(h = 0, lty = 3)
plot(sqrt(x), stdres_q4, xlab = expression(sqrt("Covariate Value")), ylab = "",
     main = "Q4 MCLE", ylim = ylim_r)
abline(h = 0, lty = 3)
```

The point estimates show a monotonically increasing progression across methods: $\hat\alpha$ increases from 0.506 (Q1) to 0.740 (Q3) to 1.063 (Q4), while $\hat\beta$ and $\hat\gamma$ decrease correspondingly. This reflects movement along the likelihood surface defined by the near-collinearity  among the large-scale parameters (Table 14: $\hat{\rho}_{\alpha,\beta} = -0.980$, $\hat{\rho}_{\alpha,\gamma} = -0.845$). It's worth emphasizing that Q1 ignores $\theta$ entirely, Q3 conditions on $w_0 = 0$ using all $n$ terms, and Q4 conditions on $Y_1$ using only $n - 1$ terms.

The CI widths in Table 13 compare Q1 Wald, Q3 profile, and Q4 adjusted profile alongside Q4 Godambe Wald. For scalar inference the adjusted profile is preferred for the same reasons as in Q3: Profiling respects parameter boundaries and accounts for likelihood asymmetry. However, the $\gamma$ lower bound remains undefined under the composite likelihood, and becomes a one-sided interval. The Godambe Wald widths are included for reference and because the delta-method confidence band in (d) must use $\hat{G}^{-1}$.

The Q4 adjusted profile widths for $\alpha$ and $\beta$ are wider than Q3 profile, and substantially so for $\alpha$ (3.311 vs 2.170). The Godambe-based intervals are approximately equal to the inverse-Hessian Wald intervals because $\hat{J} \approx \hat{H}$ numerically (Tables 10 and 11), so the sandwich correction $\hat{H}^{-1}\hat{J}\hat{H}^{-1}$ is close to $\hat{H}^{-1}$ alone rather than strictly narrower. With only $n - 1 = 22$ outer products contributing to $\hat{J}$, the agreement may partly reflect insufficient sample size (to reliably distinguish the two matrices), though under correct model specification, $J = H$ is expected asymptotically, so the agreement is consistent with an adequate composite likelihood specification.

The standardized residual plots of Figure 5 look visually similar across all three methods, but the comparison merits greater specificity. There is the mild cyclical pattern present in all three panels --- positive residuals at low and high covariate values, with a slight dip in the mid-range --- which points to a small remaining misspecification in $\mu_i$ itself rather than the error structure. This would be consistent with the extended Freundlich curve being an empirical approximation that may not perfectly capture the sorption mechanism across the full covariate range (alluded to in Sibbesen, 1981), though for $n = 23$ this cannot be formally tested.

## **(f)** Comparison of Small-Scale Component

```{r q4-comp-small, echo = F, message = F, warning = F}
# comp_small4 <- data.frame(
#   Quantity = c("$\\theta$", "$\\sigma^2$"),
#   Q2_OptEE = unname(round(c(theta_opt, s2_opt),       4)),
#   Q3_MLE   = unname(round(mle[c("theta", "sigma2")],  4)),
#   Q4_MCLE  = unname(round(mcle[c("theta", "sigma2")], 4))
# )
# 
# kable(comp_small4,
#       format    = "latex",
#       booktabs  = TRUE,
#       escape    = FALSE,
#       align     = c("l", "r", "r", "r"),
#       col.names = c("Quantity", "Q2 Optimal EE", "Q3 MLE", "Q4 MCLE"),
#       caption   = "Comparison of small-scale parameter estimates across methods.") |>
#   kable_styling(latex_options = "HOLD_position", font_size = 10)
# 
comp_small4 <- data.frame(
  Quantity   = c("$\\theta$", "$\\sigma^2$"),
  Q2_OptEE   = unname(round(c(theta_opt, s2_opt),       4)),
  Q3_MLE     = unname(round(mle[c("theta", "sigma2")],  4)),
  Q3_CI      = c(paste0("(", round(prof_lower[4], 3), ", ", round(prof_upper[4], 3), ")"),
                 paste0("(", round(prof_lower[5], 3), ", ", round(prof_upper[5], 3), ")")),
  Q4_MCLE    = unname(round(mcle[c("theta", "sigma2")], 4)),
  Q4_CI      = c(paste0("(", round(prof_god_lower[4], 3), ", ", round(prof_god_upper[4], 3), ")"),
                 paste0("(", round(prof_god_lower[5], 3), ", ", round(prof_god_upper[5], 3), ")"))
)

kable(comp_small4,
      format    = "latex",
      booktabs  = TRUE,
      escape    = FALSE,
      align     = c("l", "r", "r", "l", "r", "l"),
      col.names = c("Quantity", "Q2 Optimal EE", "Q3 MLE", "Q3 95\\% Profile CI",
                    "Q4 MCLE", "Q4 95\\% Adj. Profile CI"),
      caption   = "Comparison of small-scale parameter estimates with profile confidence intervals. Q3 uses standard profile; Q4 uses Godambe-adjusted profile.") |>
  kable_styling(latex_options = "HOLD_position", font_size = 10)

all_s2_4 <- data.frame(
  Estimator = c("Q1 GLS (marginal variance)", "Q2 moment-based",
                "Q3 MLE", "Q4 MCLE"),
  sigma2    = round(c(s2_q1, s2_opt, mle["sigma2"], mcle["sigma2"]), 4)
)

kable(all_s2_4,
      format    = "latex",
      booktabs  = TRUE,
      escape    = FALSE,
      align     = c("l", "r"),
      col.names = c("Estimator", "$\\hat{\\sigma}^2$"),
      caption   = "All estimates of $\\sigma^2$ across methods.") |>
  kable_styling(latex_options = "HOLD_position", font_size = 10)
```

The Q4 estimates of $\theta$ and $\sigma^2$ are close to Q3, as expected given that the two likelihoods differ by exactly one term. The small (but monotonically increasing) progression in $\hat\theta$ (Q2: 0.542, Q3: 0.560, Q4: 0.575) is consistent with successive reduction of first-stage approximation error: Q2 uses residuals from misspecified large-scale parameters, Q3 jointly estimates all parameters from the full conditional likelihood, and Q4 does the same from the pairwise conditional likelihood. The $\sigma^2$ estimates follow the same hierarchy noted in Q2(d) and Q3(f): Q1 GLS is substantially larger (1.015) because it targets the marginal variance, while Q2, Q3, and Q4 all target the innovation variance directly and are in close agreement with one another (`r round(s2_opt, 3)`, `r round(mle["sigma2"], 3)`, `r round(mcle["sigma2"], 3)`).

\newpage 

# Q5

A two-stage estimation strategy would consist of estimating $\alpha$, $\beta$, $\gamma$ from the nonlinear constant variance model of Q1, and then using the residuals from that fit to estimate $\theta$ and $\sigma^2$ via the optimal estimating equation of Q2. Based on what you have learned in completing this exam, what would you conclude about the possibility of using such a two-stage procedure to fit the overall extended Freundlich model to data from this problem?

## Concluding Thoughts in Sequence

The first-stage GLS fit assumes $w_i \overset{iid}{\sim} N(0,\sigma^2)$. When the true errors follow an AR(1) process with $\theta \approx 0.56$, this misspecification has two compounding consequences.

First, the GLS estimating equations use weight matrix $W = I$ in place of the efficient $\Sigma^{-1}(\theta)$, producing a consistent but inefficient estimator of $(\alpha, \beta, \gamma)$. The Q1 covariance matrix, which assumes $W = I$, is therefore incorrect even as a large-sample approximation, and the resulting standard errors are unreliable for inference regardless of what happens in Stage 2.

Second, and more critically, the Q1 raw residuals $\hat{e}_i = y_i - \mu_i(\hat\alpha,\hat\beta,\hat\gamma)$ are noise-filled surrogates for the true $w_i$. 

Consequently, the difference

$$
\hat{e}_i - w_i = \mu_i(\alpha,\beta,\gamma) - \mu_i(\hat\alpha,\hat\beta,\hat\gamma)
$$

is a smooth nonlinear function of the Stage 1 parameter estimation error. Because $\mu_i$ varies smoothly with the ordered covariate, runs of positive or negative estimation error across adjacent $x_i$'s compound the lag-1 autocorrelation of $\{\hat{e}_i\}$ relative to that of $\{w_i\}$. With $n = 23$ the effect of this distortion is not negligible.

The numerical results confirm this distortion. The naive estimating equation applied to Q1 residuals returned $\tilde\theta_{\text{naive}} = 1.005$, which is outside the desired parameter space $|\theta| < 1$. This is a symptom of the distortion described above, and not just a byproduct of $n-23$. The systematic residual pattern visible in the Q1 panel of Figure 5, i.e., positive residuals at low and high covariate values, with a slight dip in the mid-range, is exactly the signature of a misspecified error structure, and the naive estimating function has no mechanism to separate genuine autocorrelation from it.

The optimal estimating equation fares better ($\tilde\theta_{\text{opt}} = 0.542$) because multiplying by $W_{i-1}$ down-weights large residuals and focuses on local lag-1 relationships, making it more robust to the global trend. The direction of the remaining discrepancy between $\tilde\theta_{\text{opt}}$ and the Q3 MLE is not straightforward to pin down though. Writing $\hat{e}_i = w_i + d_i$ where $d_i = \mu_i(\alpha,\beta,\gamma) - \mu_i(\hat\alpha,\hat\beta,\hat\gamma)$ is the Stage 1 estimation error, the numerator of $\tilde\theta_{\text{opt}}$ acquires cross terms involving $d_i$ and the denominator acquires $\sum d_{i-1}^2$, so the ratio can shift in either direction. The observed difference between $\tilde\theta_{\text{opt}}$ and $\hat\theta_{\text{MLE}}$ is $0.018$, within expected sampling variability at $n = 23$; by itself, this is not sufficient to establish the direction or source of bias. The substantive point stands: $\tilde\theta_{\text{opt}}$ uses noise-contaminated surrogates for $w_i$, and the Q3 MLE attempts to circumvent this by estimating all parameters jointly.

```{r q5-bands, echo = F, message = F, warning = F, fig.height=4.5, fig.width=9, fig.cap="Fitted extended Freundlich curves with approximate 95\\% pointwise confidence bands."}
ylim_shared <- range(c(mugrid  - hw,  mugrid  + hw,
                       mugrid3 - hw3, mugrid3 + hw3,
                       mugrid4 - hw4, mugrid4 + hw4, y))

par(mfrow = c(1, 3))

plot(x, y, xlab = "Covariate Value", ylab = "Response Value",
     main = "Q1 GLS", ylim = ylim_shared)
lines(xgrid, mugrid, lwd = 2)
lines(xgrid, mugrid + hw,  lty = 2)
lines(xgrid, mugrid - hw,  lty = 2)

plot(x, y, xlab = "Covariate Value", ylab = "",
     main = "Q3 MLE", ylim = ylim_shared)
lines(xgrid, mugrid3, lwd = 2)
lines(xgrid, mugrid3 + hw3, lty = 2)
lines(xgrid, mugrid3 - hw3, lty = 2)

plot(x, y, xlab = "Covariate Value", ylab = "",
     main = "Q4 MCLE", ylim = ylim_shared)
lines(xgrid, mugrid4, lwd = 2)
lines(xgrid, mugrid4 + hw4, lty = 2)
lines(xgrid, mugrid4 - hw4, lty = 2)
```

```{r q5-summary, echo = F, message = F, warning = F}
summary_tab <- data.frame(
  Method  = c("Q1 GLS", "Q2 Optimal EE", "Q3 MLE", "Q4 MCLE"),
  Theta   = c("---",
              round(theta_opt,       3),
              round(mle["theta"],    3),
              round(mcle["theta"],   3)),
  Sigma2  = round(c(s2_q1, s2_opt,
                    mle["sigma2"], mcle["sigma2"]), 3),
  BandHW  = c(round(2 * mean(hw),  3), "---",
              round(2 * mean(hw3), 3),
              round(2 * mean(hw4), 3))
)

kable(summary_tab,
      format    = "latex",
      booktabs  = TRUE,
      escape    = FALSE,
      align     = c("l", "r", "r", "r"),
      col.names = c("Method", "$\\hat{\\theta}$",
                    "$\\hat{\\sigma}^2$", "Mean Confidence Band Width"),
      caption = "Small-scale estimates and mean confidence band widths across all methods.")|>   
  kable_styling(latex_options = "HOLD_position", font_size = 10)
```

Another key idea is efficiency loss relative to the joint MLE. As established in Q2(d), Q1 GLS targets the marginal variance $\sigma^2/(1-\theta^2)$ while all remaining methods target the innovation variance $\sigma^2$ directly.

Beyond the finite-sample distortion already described, efficiency is lost relative to the joint MLE even asymptotically. The Q3 MLE shifts $\hat\alpha$ from $0.506$ to $0.739$ and $\hat\beta$ from $1.222$ to $1.028$ — reflecting the reweighting of observations once the AR(1) structure is incorporated (or, perhaps, imposed). This loss is a meaningful finite-sample concern. Because the likelihood couples the regression and AR(1) parameters, the two-stage estimator with a fixed $W = I$ weight matrix is inefficient. In large samples, an appropriate GLS procedure that first estimates $\theta$ consistently and then uses $\hat\Sigma^{-1}(\hat\theta)$ as the weight matrix would recover asymptotic efficiency. The two-stage procedure as described here cannot, because Stage 1 *fixes* $W = I$ rather than iterating.

Figure 6 and Table 17 make the practical consequence visible across all methods simultaneously. The Q1 confidence band is the narrowest of the three, but this narrowness is illusory. It reflects the iid assumption forcing underestimation of uncertainty rather than genuine precision. The Q3 and Q4 bands are wider throughout, and correctly so, because they acknowledge the curvature of the (five-parameter) likelihood surface and the correlation structure of the errors jointly. The mean band widths in Table 17 (`r round(2 * mean(hw), 3)` for Q1, `r round(2 * mean(hw3), 3)` for Q3, `r round(2 * mean(hw4), 3)` for Q4) quantify this directly. The Q4 band is the widest because the Godambe sandwich covariance, even approximately, propagates more uncertainty than the inverse Hessian alone. Taken together, Figure 6 and Table 17 provide the clearest single summary of why the two-stage procedure understates inferential uncertainty: Its narrow band is a consequence of ignoring what Q3 and Q4 explicitly model.

Lastly, the Q4 composite likelihood provides corroborating evidence. The Q4 MCLE is close to Q3 ($\hat\theta = 0.575$, $\hat\sigma^2 = 0.644$), confirming that adjacent pairs carry most of the information about $\theta$ and that the two approaches are in close numerical agreement. That agreement is reassuring, but it comes at a cost: constructing even approximately valid interval estimates for Q4 required invoking martingale difference sequence theory that I will be the first to admit sits at the edge of my own understanding (though self-imposed by my curiosity-desire to determine whether a justification existed in this setting). What I can say with confidence is that such theory carries its own conditions, none of which can be meaningfully verified in this setting. The intervals are not without a theoretical basis, but the finite-sample warrant for that basis is genuinely weak. The inferential difficulty that Q4 makes explicit is the same difficulty the two-stage procedure sidesteps by ignoring the correlation structure entirely.

## Conclusion

To summarize, the two-stage procedure as implemented here fails for compounding reasons: a misspecified Stage 1 weight matrix ($W = I$ rather than $\hat{\Sigma}^{-1}(\hat\theta)$), systematically distorted residuals passed to Stage 2, and an efficiency loss relative to the joint MLE reflecting the specific choice to fix the weight matrix rather than iterate. The numerical evidence is consistent: $\tilde\theta_{\text{naive}} > 1$ signals trend distortion, $\tilde\theta_{\text{opt}}$ is displaced from the true value, and the Q1 confidence band is artificially narrow, evidenced by Figure 6 and Table 17.

Among the estimators considered, the joint MLE of Q3 is the most defensible. More broadly, the Godambe-based inference explored in Q4, while interesting, is, in practice, ill-advised in this setting. A single realization of length $n = 23$ provides no replication, and neither formal asymptotic context from Ch. 9 applies cleanly. The MDS argument offers a theoretical framework, but it cannot substitute for the asymptotics the problem requires. For formal inference at this sample size, a parametric bootstrap calibrated to the Q3 model would be the natural and preferred alternative. Such an approach would require no appeal to asymptotic normality and would respect parameter boundaries naturally, two conditions that make both Wald and profile intervals imprecise at $n = 23$ (perhaps hinting at what's to come).

# Outside Resources

Geyer, C. J. (2013). Asymptotics of maximum likelihood without the LLN or CLT or sample size going to infinity. In *Advances in Modern Statistical Theory and Applications: A Festschrift in honor of Morris L. Eaton*, G. L. Jones and X. Shen eds. IMS Collections, Vol. 10, pp. 1--24. Institute of Mathematical Statistics: Hayward, CA. \url{https://www.stat.umn.edu/geyer/lecam/simple.pdf}

Geyer, C. J. (2020). Unbiased estimating equations. Stat 8112 Lecture Notes, University of Minnesota. September 26, 2020. \url{https://www.stat.umn.edu/geyer/8112/notes/equations.pdf}