---
title: "Take-Home Part I"
output: pdf_document
author: "Sam Olson"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r loading, echo = F, message = F, warning = F}
library(here)
library(knitr)
library(kableExtra)
dat <- read.table(here("STAT_6010/Take-Home 1/exam1dat.txt"), header = T)
source(here("STAT_6010/Take-Home 1/nonlin.txt"))
source(here("STAT_6010/Take-Home 1/basicglm.txt"))

x   <- dat$x
y   <- dat$y
n   <- length(y)

# Expectation function: mu_i = alpha * x_i^(beta * x_i^(-gamma))
efctn <- function(xv, ps) {
  alpha <- ps[1]; beta <- ps[2]; gamma <- ps[3]
  alpha * xv^(beta * xv^(-gamma))
}

# Derivatives: d(mu)/d(alpha), d(mu)/d(beta), d(mu)/d(gamma)
# d(mu)/d(alpha) = x^(beta * x^(-gamma))
# d(mu)/d(beta)  = mu * x^(-gamma) * log(x)
# d(mu)/d(gamma) = -mu * beta * log(x)^2 * x^(-gamma)
eders <- function(xv, ps) {
  alpha <- ps[1]; beta <- ps[2]; gamma <- ps[3]
  mu    <- efctn(xv, ps)
  da    <- xv^(beta * xv^(-gamma))
  db    <- mu * xv^(-gamma) * log(xv)
  dg    <- -mu * beta * log(xv)^2 * xv^(-gamma)
  cbind(da, db, dg)
}

# Constant variance: weights = I
ewts <- function(xv, ps) diag(length(xv))

# Starting values (rough visual guess from scatterplot)
ps0 <- c(alpha = 0.5, beta = 1.2, gamma = 0.16)

fit1 <- nonlin(xmat = x, ys = y, ps = ps0,
               fctn = efctn, ders = eders, wts = ewts)
```

# The Model and Data

Let $\{Y_i : i = 1, \ldots, n\}$ denote a set of response variables and let $\{x_i : i = 1, \ldots, n\}$ represent a corresponding set of non-decreasing covariate values. The model for the response variables is

$$Y_i = \mu_i + w_i \tag{1}$$

where

$$\mu_i = \alpha x_i^{\beta x_i^{-\gamma}}. \tag{2}$$

In (2), $\alpha \geq 0$, $\beta \geq 0$, and $\gamma \geq 0$ are unknown parameters, and for $-1 < \theta < 1$,

$$w_i = \theta w_{i-1} + \epsilon_i, \tag{3}$$

where the $\epsilon_i$ are independent and identically distributed with $\epsilon_i \sim N(0, \sigma^2)$.

A few comments about the model and data are in order.

- The extended Freundlich curve is rather extravagant, having an exponent raised to an additional exponent with the covariate appearing several times. It appears to have been developed partly as a purely empirical exercise and partly through reasoning connected with the theoretical underpinnings of sorption (Sibbesen 1981, Barrow 2023).

- These data present a difficult situation for autoregressive errors, having only $n = 23$ observations. Typically in a time series problem we would like a sample size on the order of 100 or more, because estimating autocorrelation is difficult with limited data.

- For the same reason, we might question how well asymptotic approximations will function in producing inferential quantities such as standard errors and confidence intervals. But we may have no choice but to use such methods.

\newpage 

# Q1

Consider a simplified model in which the $w_i$ in (1) are iid $N(0, \sigma^2)$, i.e., a traditional additive error nonlinear regression model with constant variance. Fit this model to the project data using generalized least squares.

## **(a)** and **(b)**: Report GLS estimates of $\alpha$, $\beta$, $\gamma$ and the moment-based estimate of $\sigma^2$. Approximate 95% confidence intervals for $\alpha$, $\beta$, $\gamma$.

```{r q1-b, echo = F, message = F, warning = F}
bhat  <- fit1$bs
s2hat <- fit1$sshat

se    <- sqrt(diag(fit1$covb))
z95   <- qnorm(0.975)
lower <- bhat - z95 * se
upper <- bhat + z95 * se

ci_tab <- data.frame(
  Parameter = c("$\\alpha$", "$\\beta$", "$\\gamma$"),
  Estimate  = unname(round(bhat,  3)),
  Lower     = unname(round(lower, 3)),
  Upper     = unname(round(upper, 3))
)

kable(ci_tab,
      format    = "latex",
      booktabs  = TRUE,
      escape    = FALSE,
      align     = c("l", "r", "r", "r"),
      col.names = c("Parameter", "Estimate", "Lower", "Upper"),
      caption   = "GLS estimates with approximate 95\\% confidence intervals.") |>
  add_header_above(c(" " = 2, "95% CI" = 2)) |>
  kable_styling(latex_options = c("HOLD_position"),
                font_size = 10)
```

Note that the lower endpoint of the interval for $\hat\alpha$ is negative, which falls outside the scientifically meaningful parameter space $\alpha \geq 0$, though it does not prevent $\mu_i$ from being defined. The moment-based estimate of $\sigma^2$ is $\hat\sigma^2 = `r round(s2hat, 3)`$.

## **(c)** and **(d)**: A plot with the estimated expectation function overlaid on the scatterplot along with an approximate 95% pointwise confidence band. A plot of standardized residuals against the covariate.

**(c)**: An approximate pointwise confidence band is constructed via the delta method. The necessary derivatives of $\mu_i$ with respect to $(\alpha, \beta, \gamma)$ are

$$
\frac{\partial \mu_i}{\partial \alpha} = x_i^{\beta x_i^{-\gamma}}, \qquad
  \frac{\partial \mu_i}{\partial \beta}  = \mu_i x_i^{-\gamma} \log x_i, \qquad
  \frac{\partial \mu_i}{\partial \gamma} = -\mu_i \beta (\log x_i)^2 x_i^{-\gamma}.
$$

**(d)**: Standardized residuals were computed using the hat-matrix correction $r_i / \{\hat\sigma^2(1 - h_{ii})\}^{1/2}$ as in expression (9.37) of the Stat 520 notes; these are returned directly by `nonlin` in `fit1$stdres`. The residual plot shows a good deal of remaining structure as a function of the covariate, consistent with Figure 3 of Sibbesen (1981). This pattern motivates the inclusion of an AR(1) small-scale component in the full model, treating the serial dependence across ordered covariate values analogously to serial correlation over time.

```{r q1-cd, echo = F, message = F, warning = F, fig.height=8, fig.width=5, fig.cap="(Top) Fitted extended Freundlich curve with approximate 95\\% pointwise confidence band. (Bottom) Standardized residuals versus covariate."}
xgrid  <- seq(min(x[x > 0]), max(x), length.out = 300)
mugrid <- efctn(xgrid, bhat)
Dgrid  <- eders(xgrid, bhat)
pvar   <- apply(Dgrid, 1, function(d) t(d) %*% fit1$covb %*% d)
hw     <- z95 * sqrt(pvar)

par(mfrow = c(2, 1))

plot(x, y, xlab = "Covariate Value", ylab = "Response Value", main = "")
lines(xgrid, mugrid, lwd = 2)
lines(xgrid, mugrid + hw, lty = 2)
lines(xgrid, mugrid - hw, lty = 2)

plot(x, fit1$stdres,
     xlab = "Covariate Value", ylab = "Standardized Residual", main = "")
abline(h = 0, lty = 3)
```

\newpage

# Q2

Consider now just the portion of the model for $w_i$ in (3). Suppose we were able to observe the $w_i$. Take $W_0 = 0$, and use without derivation that under second-order stationarity, $E(W_i) = 0$ and $E(W_i^2) = \sigma^2 / (1 - \theta^2)$ for all $i$.

## **(a)** Show that the estimating equation for $\theta$ based on the **naive** estimating function

$$v_i(W, \theta) = W_i - \theta W_{i-1}$$

is an unbiased estimating equation.

The naive estimating function is $v_i(W, \theta) = W_i - \theta W_{i-1}$. Substituting $W_i = \theta W_{i-1} + \epsilon_i$ directly gives

$$
E[v_i(W, \theta)] = E[\theta W_{i-1} + \epsilon_i - \theta W_{i-1}] = E[\epsilon_i] = 0,
$$

since $\epsilon_i \sim N(0, \sigma^2)$ is independent of $W_{i-1}$. Because this holds for every $i$ and all $\theta \in (-1, 1)$, the estimating equation $\sum_{i=2}^n v_i(W, \theta) = 0$ is unbiased. Setting it to zero and solving gives

$$
\tilde{\theta}_{\text{naive}} = \frac{\sum_{i=2}^n W_i}{\sum_{i=2}^n W_{i-1}}.
$$

While this form is algebraically valid, the denominator $\sum W_{i-1}$ can be near zero in practice, which contributes to the poor numerical behavior seen in Q2(d), **spoilers**.

## **(b)** Show that the estimating equation for $\theta$ based on the **optimal** estimating function

$$v_i(W, \theta) = W_{i-1}(W_i - \theta W_{i-1})$$

is an unbiased estimating equation.

The optimal estimating function is $v_i(W,\theta) = W_{i-1}(W_i - \theta W_{i-1})$. Expanding the expectation and using the model $W_i = \theta W_{i-1} + \epsilon_i$,

$$
E[W_{i-1} W_i]
  = E[W_{i-1}(\theta W_{i-1} + \epsilon_i)]
  = \theta E[W_{i-1}^2] + E[W_{i-1}\,\epsilon_i].
$$

Because $W_{i-1}$ is a function of $\epsilon_1, \ldots, \epsilon_{i-1}$ alone and $\epsilon_i$ is independent of those, $E[W_{i-1}\epsilon_i] = 0$.

Under the assumption of second-order (weak) stationarity of $\{W_i\}$ --- meaning $E(W_i) = 0$, $E(W_i^2) = \sigma^2/(1 - \theta^2)$ for all $i$, and $\text{Cov}(W_i, W_j)$ depending only on $|i - j|$, which holds when $|\theta| < 1$ --- we have $E[W_{i-1}^2] = \sigma^2/(1 - \theta^2)$,

$$
E[v_i(W,\theta)]
  = \theta \frac{\sigma^2}{1-\theta^2} - \theta \frac{\sigma^2}{1-\theta^2} = 0.
$$

The estimating equation $\sum_{i=2}^n W_{i-1}(W_i - \theta W_{i-1}) = 0$ is therefore unbiased, and its solution is the lag-1 autocorrelation estimator

$$
\tilde\theta_{\text{opt}}
  = \frac{\sum_{i=2}^n W_i W_{i-1}}{\sum_{i=2}^n W_{i-1}^2}.
$$

This is called "optimal" in the sense of Godambe's criterion: among linear estimating functions in $W_{i-1}$, multiplying the naive function by $W_{i-1}$ maximizes the ratio $[E\{\partial v/\partial\theta\}]^2 / \operatorname{var}\{v\}$ (see Ch. 8.3 of the course notes). Note that both estimating equations yield closed-form estimators, and the two estimators are in general numerically distinct.

## **(c)** Develop a moment-based estimator for $\sigma^2$ that could be used with either estimating equation, and show that it is unbiased.

Because $\epsilon_i = W_i - \theta W_{i-1}$ and $\epsilon_i \sim N(0,\sigma^2)$, we have $E[\epsilon_i^2] = \sigma^2$ for all $i$. This immediately suggests the estimator (plugging in $\tilde\theta$ from either estimating equation and using $W_0 = 0$),

$$
\hat\sigma^2 = \frac{1}{n-1}\sum_{i=1}^{n}(W_i - \tilde\theta\, W_{i-1})^2,
$$

where $W_0 = 0$. Note that the $i = 1$ term contributes $W_1 - \tilde{\theta} W_0 = W_1$, which equals $\epsilon_1$ only by virtue of the boundary condition $W_0 = 0$; this is a boundary artifact rather than a general property of the innovation structure. To see that this estimator is approximately unbiased, note that at the true $\theta$ each term satisfies $E[(W_i - \theta W_{i-1})^2] = E[\epsilon_i^2] = \sigma^2$, so

$$
E\!\left[\frac{1}{n}\sum_{i=1}^n (W_i - \theta W_{i-1})^2\right] = \sigma^2.
$$

Once $\theta$ is replaced by $\tilde\theta$, however, strict unbiasedness no longer holds: $E[(W_i - \tilde\theta W_{i-1})^2] \neq \sigma^2$ exactly, because $\tilde\theta$ is itself a function of the data. The estimator is therefore only approximately unbiased, in the same spirit as $s^2$ in ordinary regression, with the $n-1$ denominator providing a degrees-of-freedom correction by analogy with the linear case — though unlike $s^2$ in ordinary regression, this correction does not achieve exact unbiasedness here because the innovation residuals are not linear projections in the standard sense.

## **(d)** Apply both estimating equations using the raw residuals from Q1 as surrogate observations of $w_i$. Compute $\hat\sigma^2$ for each and compare all three estimates of $\sigma^2$ (naive EE, optimal EE, Q1 GLS). Are these estimators all estimating the same quantity?

```{r q2-d, echo = F, message = F, warning = F}
# Raw residuals from Q1 as surrogate w_i observations
ehat  <- y - fit1$yhat
n_e   <- length(ehat)

e_curr <- ehat[2:n_e]
e_prev <- ehat[1:(n_e - 1)]

# --- Naive estimating equation: sum(e_i - theta * e_{i-1}) = 0
theta_naive <- sum(e_curr) / sum(e_prev)

# --- Optimal estimating equation: sum(e_{i-1} * (e_i - theta * e_{i-1})) = 0
theta_opt <- sum(e_curr * e_prev) / sum(e_prev^2)

# --- sigma^2 estimates (include i = 1 with W_0 = 0)
e_all_curr <- c(ehat[1],  e_curr)   # W_1, W_2, ..., W_n
e_all_prev <- c(0,        e_prev)   # W_0=0, W_1, ..., W_{n-1}

innov_naive <- e_all_curr - theta_naive * e_all_prev
innov_opt   <- e_all_curr - theta_opt   * e_all_prev

s2_naive <- sum(innov_naive^2) / (n_e - 1)
s2_opt   <- sum(innov_opt^2)   / (n_e - 1)
s2_q1    <- fit1$sshat

# cat(sprintf("theta (naive EE):    %.4f\n", theta_naive))
# cat(sprintf("theta (optimal EE):  %.4f\n", theta_opt))
# cat(sprintf("sigma^2 (naive EE):  %.4f\n", s2_naive))
# cat(sprintf("sigma^2 (optimal EE):%.4f\n", s2_opt))
# cat(sprintf("sigma^2 (Q1 GLS):    %.4f\n", s2_q1))
results_tab <- data.frame(
  Method = c("Naive EE", "Optimal EE", "Q1 GLS"),
  Theta  = c(round(theta_naive, 4), round(theta_opt, 4), "---"),
  Sigma2 = round(c(s2_naive, s2_opt, s2_q1), 4)
)

kable(results_tab,
      format    = "latex",
      booktabs  = TRUE,
      align     = c("l", "r", "r"),
      col.names = c("Method", "$\\hat{\\theta}$", "$\\hat{\\sigma}^2$"),
      escape    = FALSE,
      caption   = "Comparison of $\\hat{\\theta}$ and $\\hat{\\sigma}^2$ estimates. Note that $\\tilde\\theta_{\\text{naive}} > 1$ falls outside the stationary parameter space $|\\theta| < 1$.") |>
  kable_styling(latex_options = "hold_position",
                font_size = 10)
```

Because the Q1 residuals $\hat{e}_i$ contain mean-estimation error — they are $y_i - \mu_i(\hat\alpha,\hat\beta,\hat\gamma)$ rather than the true $w_i = y_i - \mu_i(\alpha,\beta,\gamma)$ — the estimating equations applied here are only approximations to those derived in Q2(a)–(b). The two estimating equations yield different estimates of $\theta$, but both target the same population parameter --- the AR(1) autocorrelation coefficient $\theta$ in model (3). The numerical difference between $\tilde{\theta}_{\text{naive}} = 1.005$ and $\tilde{\theta}_{\text{opt}} = 0.542$ reflects estimation error rather than any difference in target, with the optimal estimate being the more reliable of the two (lower asymptotic variance by Godambe's criterion).

More importantly, the three $\hat\sigma^2$ values are **not** all estimating the same quantity. The Q1 GLS estimate was obtained under the assumption that the $w_i$ are iid, so it targets $\text{Var}(w_i)$; because the true model is AR(1), $\text{Var}(w_i) = \sigma^2/(1-\theta^2)$, but the estimator itself has no knowledge of that structure. In contrast, the two EE-based estimates are obtained by squaring residuals of the form $W_i - \tilde\theta W_{i-1} \approx \epsilon_i$, so they target the innovation variance $\sigma^2$. Because the AR(1) process has positive autocorrelation ($\theta > 0$ is typical for sorption data), the marginal variance exceeds the innovation variance, and we expect $\hat\sigma^2_{\text{Q1}} > \hat\sigma^2_{\text{naive}} \approx \hat\sigma^2_{\text{opt}}$.

\newpage

# Q3

Assume $x_0 = 0$ and $w_0 = 0$. Derive the full log likelihood for the model of (1), (2), and (3), and maximize it to find simultaneous estimates of $\alpha$, $\beta$, $\gamma$, $\theta$, and $\sigma^2$.

## **(a)** An expression for the log likelihood.

Given $w_0 = 0$, the residuals $w_i = Y_i - \mu_i$ satisfy $w_i = \theta w_{i-1} + \epsilon_i$ with $\epsilon_i \overset{iid}{\sim} N(0,\sigma^2)$. The likelihood derived here is a *conditional* likelihood, conditioning on $w_0 = 0$ rather than treating $W_1$ as drawn from the stationary marginal distribution $N(0, \sigma^2/(1-\theta^2))$. Given this conditioning, each innovation $\epsilon_i = w_i - \theta w_{i-1}$ is independent, so the joint density of $(Y_1,\ldots,Y_n)$ factors as a product of $n$ normal densities in the $\epsilon_i$. The log likelihood is

$$
\ell(\alpha,\beta,\gamma,\theta,\sigma^2)
= -\frac{n}{2}\log(2\pi)
  -\frac{n}{2}\log(\sigma^2)
  -\frac{1}{2\sigma^2}
    \sum_{i=1}^n \bigl(w_i - \theta w_{i-1}\bigr)^2,
$$

where $w_i = Y_i - \mu_i = Y_i - \alpha x_i^{\beta x_i^{-\gamma}}$ and $w_0 = 0$.

```{r q3-a, echo = F, message = F, warning = F}
loglik_q3 <- function(par, x, y) {
  alpha  <- par[1]
  beta   <- par[2]
  gamma  <- par[3]
  theta  <- par[4]
  sigma2 <- par[5]

  if (alpha  <= 0) return(-Inf)
  if (beta   <= 0) return(-Inf)
  if (gamma  <= 0) return(-Inf)
  if (sigma2 <= 0) return(-Inf)
  if (abs(theta) >= 1) return(-Inf)

  mu     <- alpha * x^(beta * x^(-gamma))
  w      <- y - mu
  w_prev <- c(0, w[-length(w)])
  eps    <- w - theta * w_prev

  n  <- length(y)
  ll <- -(n/2) * log(2 * pi) - (n/2) * log(sigma2) - sum(eps^2) / (2 * sigma2)
  return(ll)
}
```

## **(b)** A description of how you maximized the log likelihood. Note that Q1 estimates provide natural starting values and $\theta_0 = 0$; `optim` with `method = "BFGS"` is recommended over Nelder-Mead for five parameters.

### Wald 

The log likelihood was maximized using `optim` with `method = "BFGS"`. Starting values for $(\alpha, \beta, \gamma)$ were taken from the Q1 GLS fit; $\theta_0 = 0$ and $\sigma^2_0 = \hat\sigma^2_{\text{Q1}}$ completed the five-parameter starting vector. These are natural starting values given that the Q1 fit is the $\theta = 0$ special case of the full model, placing the initial iterate close to the region of the likelihood surface where the solution is expected to lie and reducing sensitivity to the starting value concern noted for five-parameter problems of this type. Because `optim` minimizes by default, we pass the negative log likelihood. The Hessian was requested via `hessian = TRUE` to support Wald inference.

```{r q3-b, echo = F, message = F, warning = F}
# Negative log likelihood for optim
negll <- function(par, x, y) -loglik_q3(par, x, y)

ps0_q3 <- c(bhat[1], bhat[2], bhat[3], 0, s2hat)

fit3 <- optim(
  par     = ps0_q3,
  fn      = negll,
  x       = x,
  y       = y,
  method  = "BFGS",
  hessian = TRUE
)

if (fit3$convergence != 0)
  warning("optim did not converge: code ", fit3$convergence)

# q3-b chunk — add these lines after fit3
mle        <- fit3$par
names(mle) <- c("alpha", "beta", "gamma", "theta", "sigma2")
obs_info   <- fit3$hessian
cov_mle    <- solve(obs_info)
se_mle     <- sqrt(diag(cov_mle))
z95        <- qnorm(0.975)
```

### Profile Method 

Profile likelihood confidence intervals were constructed to complement the Wald intervals and avoid reliance on asymptotic normality of the MLEs. For each parameter $\psi \in \{\alpha, \beta, \gamma, \theta, \sigma^2\}$, the profile log likelihood is

$$
\ell_p(\psi) = \max_{\boldsymbol\lambda} \; \ell(\psi, \boldsymbol\lambda),
$$

where $\boldsymbol\lambda$ denotes the remaining four parameters. At each point of a 60-point grid spanning $\hat{\psi} \pm 4$ Wald standard errors---where the Wald SE is taken from the observed information matrix at the MLE---truncated to the parameter's natural support, the inner maximization over $\boldsymbol\lambda$ was performed using `optim` with `method = "Nelder-Mead"`, initialized at the full MLE. Nelder-Mead was preferred over BFGS for the inner step because constraint boundaries (e.g., $\sigma^2 > 0$, $|\theta| < 1$) can cause finite-difference gradient evaluations to return non-finite values. The 95% confidence interval for $\psi$ is the set of grid values satisfying

$$
\ell_p(\psi) \geq \hat\ell - \tfrac{1}{2}\chi^2_{0.95,\,1},
$$

with endpoints obtained by linear interpolation at the threshold crossing.

```{r q3-b-profile, echo = F, message = F, warning = F}
# --- Profile likelihood CIs ---
# Threshold: profile log-lik must stay within qchisq(0.95, 1)/2 of the max
cutoff <- qchisq(0.95, df = 1) / 2          # ≈ 1.9208
ll_max <- -fit3$value

profile_ci_one <- function(par_idx, n_grid = 60) {
  mle_val <- mle[par_idx]
  se_val  <- se_mle[par_idx]

  hard_lo <- c(1e-6, 1e-6, 1e-6, -1 + 1e-4, 1e-6)[par_idx]
  hard_hi <- c(Inf,  Inf,  Inf,   1 - 1e-4,  Inf )[par_idx]

  lo_raw <- mle_val - 4 * se_val
  hi_raw <- mle_val + 4 * se_val
  grid   <- seq(max(lo_raw, hard_lo), min(hi_raw, hard_hi), length.out = n_grid)

  pl <- vapply(grid, function(psi) {
    obj <- function(p_other) {
      par_full <- numeric(5)
      par_full[-par_idx] <- p_other
      par_full[ par_idx] <- psi
      negll(par_full, x, y)
    }
    res <- tryCatch(
      optim(mle[-par_idx], obj, method = "Nelder-Mead",
            control = list(maxit = 2000, reltol = 1e-10)),
      error = function(e) list(value = Inf)
    )
    -res$value
  }, numeric(1))

  target <- ll_max - cutoff

  interp_cross <- function(side) {
    idx       <- if (side == "lower") which(grid < mle_val) else which(grid > mle_val)
    vals_side <- pl[idx]
    grid_side <- grid[idx]
    below     <- which(vals_side < target)
    if (length(below) == 0) return(NA_real_)
    b    <- if (side == "lower") max(below) else min(below)
    b_in <- if (side == "lower") b + 1 else b - 1
    if (b_in < 1 || b_in > length(grid_side)) return(grid_side[b])
    approx(x = vals_side[c(b, b_in)],
           y = grid_side[c(b, b_in)],
           xout = target)$y
  }

  list(
    ci   = c(lower = interp_cross("lower"), upper = interp_cross("upper")),
    grid = grid,
    pl   = pl
  )
}

# Re-run with updated function
prof_bounds <- lapply(1:5, profile_ci_one)

# Update downstream extractions to use $ci
prof_lower <- sapply(prof_bounds, function(b) b$ci["lower"])
prof_upper <- sapply(prof_bounds, function(b) b$ci["upper"])
```

## **(c)** The MLEs, 95% confidence intervals (stating whether Wald or profile), and the maximized log likelihood value.

### Wald 

Wald confidence intervals were constructed from the observed information matrix --- the negative Hessian of $\ell$ evaluated at the MLE, returned directly by `optim` via `hessian = TRUE` --- used in place of the expected Fisher information $I(\theta) = -E(\nabla^2 \ell)$. These are asymptotically equivalent under standard regularity conditions, though the distinction is worth noting given $n = 23$. With $n = 23$ and $p = 5$, these intervals rely heavily on asymptotic normality and should be interpreted with corresponding caution.

### Profile Method 

Profile 95\% confidence intervals are reported alongside the Wald intervals in Table 3. The most consequential difference is for $\alpha$: the Wald lower bound of $-0.4834$ is negative and therefore inadmissible under the model, which requires $\alpha > 0$. The profile lower bound of $0.2060$ correctly respects this constraint, reflecting the asymmetry of the likelihood surface near the boundary. Right-skew is also evident for $\sigma^2$, consistent with its known small-sample behavior. Where the two interval types agree closely --- as for $\gamma$ and $\theta$ --- the likelihood surface is approximately quadratic and the Wald approximation is adequate. Given the inadmissibility of the Wald interval for $\alpha$, profile confidence intervals are used as the basis for all subsequent inference.

```{r q3-c, echo = F, message = F, warning = F}
prof_tab <- data.frame(
  Parameter = c("$\\alpha$", "$\\beta$", "$\\gamma$", "$\\theta$", "$\\sigma^2$"),
  MLE       = unname(round(mle,                4)),
  W_Lower   = unname(round(mle - z95 * se_mle, 4)),
  W_Upper   = unname(round(mle + z95 * se_mle, 4)),
  P_Lower   = unname(round(prof_lower,         4)),
  P_Upper   = unname(round(prof_upper,         4)),
  row.names = NULL
)

kable(prof_tab,
      format    = "latex",
      booktabs  = TRUE,
      escape    = FALSE,
      align     = c("l", "r", "r", "r", "r", "r"),
      col.names = c("Parameter", "MLE", "Lower", "Upper", "Lower", "Upper"),
      caption   = paste0("Wald and profile 95\\% confidence intervals at the MLE. ",
                         "Maximized log likelihood: ", round(ll_max, 4), ".")) |>
  add_header_above(c(" " = 2, "Wald 95% CI" = 2, "Profile 95% CI" = 2)) |>
  kable_styling(latex_options = "HOLD_position", font_size = 10)
```

```{r q3-profile-check, echo = F, message = F, warning = F}
pnames <- c("$\\alpha$", "$\\beta$", "$\\gamma$", "$\\theta$", "$\\sigma^2$")

prof_check <- data.frame(
  Parameter   = pnames,
  Profile_Max = unname(round(sapply(prof_bounds, function(b) max(b$pl)), 4)),
  LL_Max      = round(ll_max, 4),
  Difference  = unname(round(ll_max - sapply(prof_bounds, function(b) max(b$pl)), 4)),
  row.names   = NULL
)

kable(prof_check,
      format    = "latex",
      booktabs  = TRUE,
      escape    = FALSE,
      align     = c("l", "r", "r", "r"),
      col.names = c("Parameter", "Profile Max $\\ell_p$", "MLE $\\hat{\\ell}$", "Difference"),
      caption   = "Profile log-likelihood maxima versus the maximized log likelihood. Differences near zero indicate successful inner Nelder-Mead convergence for each parameter.") |>
  kable_styling(latex_options = "HOLD_position", font_size = 10)
```

Table 4 reports the maximum achieved value of each profile log likelihood $\ell_p(\psi)$ against $\hat\ell = -29.598$. All differences are at most $0.0012$ in absolute value, confirming that the inner Nelder-Mead optimization successfully recovered the MLE when each parameter was held fixed at its estimated value --- the profile surface is tracing the correct likelihood contour in every case. The one negative difference for $\theta$ ($-0.0001$) is numerical noise from the inner optimizer slightly overshooting, not a genuine improvement over the MLE. Given these negligible discrepancies, the profile surfaces are tracing the correct likelihood contours and the profile intervals are reliable. Combined with the inadmissibility of the Wald interval for $\alpha$, this confirms that profile confidence intervals are the appropriate choice for subsequent inference (when applicable).

The profile upper bound for $\theta$ of $0.9151$ also warrants note: it lies close to the stationarity boundary $|\theta| = 1$, meaning the data are consistent with near-unit-root serial dependence. Under near-unit-root conditions, the moment $E(W_i^2) = \sigma^2/(1 - \theta^2)$ used throughout Q2 becomes large and poorly estimated, and the conditional likelihood approximation based on $w_0 = 0$ may be less accurate than the asymptotic theory suggests. This is a limitation that cannot be resolved at $n = 23$.

## **(d)** A plot of the data with the estimated expectation function overlaid along with an approximate 95% pointwise confidence band.

The confidence band for $\mu_i = \alpha x_i^{\beta x_i^{-\gamma}}$ uses the delta method with the $3 \times 3$ submatrix of $\widehat{\text{Cov}}$ corresponding to $(\alpha, \beta, \gamma)$. The derivatives are the same as in Q1.

```{r q3-d, echo = F, message = F, warning = F, fig.height=4.5, fig.width=6, fig.cap="Fitted extended Freundlich curve with approximate 95\\% pointwise confidence band via the delta method. Scalar parameter CIs in (c) use profile intervals; the band uses the Wald delta method as profile bands for functions of multiple parameters are not standard."}

alpha3 <- mle["alpha"]; beta3 <- mle["beta"]; gamma3 <- mle["gamma"]

mugrid3 <- efctn(xgrid, mle[1:3])
Dgrid3  <- eders(xgrid, mle[1:3])

# 3x3 sub-covariance for (alpha, beta, gamma)
cov_abg <- cov_mle[1:3, 1:3]

pvar3 <- apply(Dgrid3, 1,
               function(d) as.numeric(t(d) %*% cov_abg %*% d))
hw3   <- z95 * sqrt(pvar3)

plot(x, y, xlab = "Covariate Value", ylab = "Response Value", main = "")
lines(xgrid, mugrid3, lwd = 2)
lines(xgrid, mugrid3 + hw3, lty = 2)
lines(xgrid, mugrid3 - hw3, lty = 2)
```

## **(e)** A comparison of the large-scale component ($\alpha$, $\beta$, $\gamma$) to Q1: point estimates, interval estimates, estimated correlations among parameter estimators, and confidence band widths. Discuss any differences.

```{r q3-e, echo = F, message = F, warning = F}
# width_q1 <- 2 * z95 * sqrt(diag(fit1$covb))
# width_q3 <- 2 * z95 * se_mle[1:3]
# --- Point estimates and CI widths
width_q1 <- 2 * z95 * sqrt(diag(fit1$covb))
width_q3 <- prof_upper[1:3] - prof_lower[1:3]

comp_tab <- data.frame(
  Parameter  = c("$\\alpha$", "$\\beta$", "$\\gamma$"),
  Q1_Est     = unname(round(bhat,      4)),
  Q3_Est     = unname(round(mle[1:3], 4)),
  Q1_Width   = unname(round(width_q1, 4)),
  Q3_Width   = unname(round(width_q3, 4))
)

kable(comp_tab,
      format    = "latex",
      booktabs  = TRUE,
      escape    = FALSE,
      align     = c("l", "r", "r", "r", "r"),
      col.names = c("Parameter", "Q1 GLS", "Q3 MLE", "Q1 GLS", "Q3 MLE"),
      caption   = paste0("Point estimates and 95\\% CI widths by method (Q1 Wald, Q3 profile). ",
                         "Mean confidence band half-width: Q1 = ",
                         round(mean(hw), 4), ", Q3 = ", round(mean(hw3), 4), ".")) |>
  add_header_above(c(" " = 1, "Point Estimate" = 2, "CI Width" = 2)) |>
  kable_styling(latex_options = "HOLD_position", font_size = 10)

# --- Correlation matrices
cor_q1 <- round(cov2cor(fit1$covb),          3)
cor_q3 <- round(cov2cor(cov_mle[1:3, 1:3]), 3)
pnames <- c("$\\alpha$", "$\\beta$", "$\\gamma$")
rownames(cor_q1) <- colnames(cor_q1) <- pnames
rownames(cor_q3) <- colnames(cor_q3) <- pnames

kable(cor_q1,
      format   = "latex",
      booktabs = TRUE,
      escape   = FALSE,
      caption  = "Estimated correlation matrix for $(\\alpha, \\beta, \\gamma)$: Q1 GLS.") |>
  kable_styling(latex_options = "HOLD_position", font_size = 10)

kable(cor_q3,
      format   = "latex",
      booktabs = TRUE,
      escape   = FALSE,
      caption  = "Estimated correlation matrix for $(\\alpha, \\beta, \\gamma)$: Q3 MLE.") |>
  kable_styling(latex_options = "HOLD_position", font_size = 10)
```

The point estimates of $(\alpha, \beta, \gamma)$ shift noticeably between Q1 and Q3 — most visibly $\hat\alpha$ moving from 0.506 to 0.740 — because jointly estimating $\theta$ reweights the observations according to the AR(1) correlation structure rather than treating them as independent. When correlation parameters are known, GLS with $\Sigma^{-1}(\theta)$ as a fixed weight matrix improves efficiency over the iid estimator. When they must be estimated jointly, however, the additional uncertainty in $\theta$ and $\sigma^2$ propagates into the regression parameters and can widen confidence intervals---which is exactly what Table 5 reflects. The profile CI widths under Q3 are wider than Q1 Wald for $\gamma$, essentially unchanged for $\beta$, and shifted substantially upward for $\alpha$---the inadmissible Wald lower bound of $-0.4834$ is replaced by a profile lower bound of $0.2060$, correctly respecting the $\alpha > 0$ constraint, and the mean confidence band half-width increases from $0.627$ to $0.967$. This reflects the additional uncertainty introduced by estimating $\theta$ and $\sigma^2$ jointly with the regression parameters: the likelihood surface is more diffuse in the $(\alpha, \beta, \gamma)$ directions once the correlated error structure is acknowledged, and the Q1 covariance matrix, which ignored this coupling, was understating that uncertainty. It should also be noted that Table 5 compares Q1 Wald intervals to Q3 profile intervals---methodologically distinct constructions. Profile intervals are generally wider than Wald when the likelihood surface is asymmetric, so part of the observed width difference reflects the interval type rather than purely the model specification change.

This effect is partially offset by the fact that $\hat\sigma^2_{\text{Q1}} = 1.015 > \hat\sigma^2_{\text{MLE}} = 0.625$, which inflates the Q1 standard errors; the Q3 intervals would be even wider if $\sigma^2$ were held equal across methods. The estimated correlations among the regression parameter estimators also shift under Q3, with $|\hat\rho_{\alpha,\beta}| \approx 0.97$ under both methods indicating that the likelihood surface is extremely flat along the $(\alpha, \beta)$ ridge — meaning individual parameter estimates are highly sensitive to small perturbations regardless of the fitting method, and the wide confidence intervals are intrinsic to the model structure rather than an artifact of the estimation procedure.

## **(f)** A comparison of the small-scale component ($\theta$, $\sigma^2$) to the optimal estimating equation results from Q2. Two estimates of $\theta$ and three estimates of $\sigma^2$ are available (Q3 MLE, Q1 GLS, Q2 moment-based). Discuss any differences.

```{r q3-f, echo = F, message = F, warning = F}
comp_small <- data.frame(
  Quantity = c("$\\theta$", "$\\sigma^2$"),
  Q2_OptEE = unname(round(c(theta_opt, s2_opt),          4)),
  Q3_MLE   = unname(round(mle[c("theta", "sigma2")],     4))
)

kable(comp_small,
      format    = "latex",
      booktabs  = TRUE,
      escape    = FALSE,
      align     = c("l", "r", "r"),
      col.names = c("Quantity", "Q2 Optimal EE", "Q3 MLE"),
      caption   = "Comparison of small-scale parameter estimates.") |>
  kable_styling(latex_options = "HOLD_position", font_size = 10)

all_s2 <- data.frame(
  Estimator = c("Q1 GLS (marginal variance)", "Q2 moment-based", "Q3 MLE"),
  sigma2    = round(c(s2_q1, s2_opt, mle["sigma2"]), 4)
)

kable(all_s2,
      format    = "latex",
      booktabs  = TRUE,
      escape    = FALSE,
      align     = c("l", "r"),
      col.names = c("Estimator", "$\\hat{\\sigma}^2$"),
      caption   = "All estimates of $\\sigma^2$ across methods.") |>
  kable_styling(latex_options = "HOLD_position", font_size = 10)
```

The two estimates of $\theta$ (Q2 optimal EE and Q3 MLE) both target the same parameter but differ in how they handle the regression function. The Q2 estimate used raw residuals from the misspecified iid model of Q1 as surrogates for $w_i$, introducing noise from the first-stage estimation; the Q3 MLE estimates $\theta$ simultaneously with $(\alpha,\beta,\gamma)$, which should reduce that bias.

For $\sigma^2$, the three estimators target different quantities, as established in Q2(d): the Q1 GLS estimate targets the marginal variance $\sigma^2/(1 - \theta^2)$ and is therefore expected to be the largest when $\theta > 0$. Both the Q2 moment-based and Q3 MLE estimates target the innovation variance $\sigma^2$, but the Q2 version inherits the approximation error of using Q1 residuals instead of true $w_i$ values, while the Q3 MLE is obtained from the correctly specified joint model. Any remaining discrepancy between Q2 and Q3 for $\sigma^2$ can be attributed to that first-stage approximation.

\newpage 

# Q4

Assume $x_0 = 0$ and $w_0 = 0$. Using the events $A_i = (Y_i = y_i \mid Y_{i-1} = y_{i-1})$ for $i = 2, \ldots, n$, derive a pairwise conditional log composite likelihood and fit the model. Report the same components as in Q3.

Additionally, discuss which asymptotic context for composite likelihood is most appropriate given $n = 23$. Address potential issues with estimating the $H$ and $J$ matrices in the Godambe information; compute them if possible and use them to construct interval estimates for the regression function parameters.

## Pairwise Conditional Log Composite Likelihood

Given the model, $w_i = Y_i - \mu_i$, and the AR(1) structure $w_i = \theta w_{i-1} + \epsilon_i$ with $\epsilon_i \sim N(0,\sigma^2)$, the conditional distribution of $Y_i$ given $Y_{i-1} = y_{i-1}$ is

$$
Y_i \mid Y_{i-1} = y_{i-1}
  \sim N\!\bigl(\mu_i + \theta(y_{i-1} - \mu_{i-1}),\; \sigma^2\bigr),
$$

since $w_i \mid w_{i-1} \sim N(\theta w_{i-1}, \sigma^2)$. Each conditional log likelihood piece is therefore

$$
\ell_i = -\frac{1}{2}\log(2\pi)
         - \frac{1}{2}\log\sigma^2
         - \frac{\epsilon_i^2}{2\sigma^2}, \qquad
  \epsilon_i = (y_i - \mu_i) - \theta(y_{i-1} - \mu_{i-1}),
$$

and the pairwise conditional log composite likelihood is

$$
\ell_C = \sum_{i=2}^n \ell_i
         = -\frac{n-1}{2}\log(2\pi)
           -\frac{n-1}{2}\log\sigma^2
           -\frac{1}{2\sigma^2}\sum_{i=2}^n \epsilon_i^2.
$$

The composite likelihood thus conditions on $Y_1$ (i.e., sums from $i=2$), whereas the full likelihood in Q3 used $w_0 = 0$ to include $i = 1$, so the two differ by exactly one term.

## Asymptotic Context

This problem falls under Asymptotic Context II (Ch.\ 9.3.2): a single realization of a dependent sequence of length $n = 23$, not independent replicates of a multivariate vector. Under Context II, the Comets--Janzura result would require translation-invariant conditional specification, which fails here because $\mu_i$ varies with the covariate. The Context I estimators of $H$ and $J$---averaging over iid replicates---are therefore unavailable. We instead compute empirical analogs: $\hat{H}$ from the negative Hessian of $\ell_C$ at the MCLE, and $\hat{J}$ from the sum of outer products of piece scores $\nabla \ell_i$. With $n = 23$ these should be interpreted cautiously; parametric bootstrap would be the principled alternative. Godambe-based intervals are reported as the analytic option, with this caveat noted.

```{r q4-loglik, echo = F, message = F, warning = F}
# Log composite likelihood: sum of conditional pieces i = 2, ..., n
logcl_q4 <- function(par, x, y) {
  alpha  <- par[1]
  beta   <- par[2]
  gamma  <- par[3]
  theta  <- par[4]
  sigma2 <- par[5]

  if (alpha  <= 0) return(-Inf)
  if (beta   <= 0) return(-Inf)
  if (gamma  <= 0) return(-Inf)
  if (sigma2 <= 0) return(-Inf)
  if (abs(theta) >= 1) return(-Inf)

  mu    <- alpha * x^(beta * x^(-gamma))
  w     <- y - mu
  # epsilon_i = w_i - theta * w_{i-1}, summed from i = 2
  eps   <- w[-1] - theta * w[-length(w)]

  m  <- length(eps)   # n - 1
  lc <- -(m/2) * log(2 * pi) - (m/2) * log(sigma2) -
        sum(eps^2) / (2 * sigma2)
  return(lc)
}

negcl <- function(par, x, y) -logcl_q4(par, x, y)
```

```{r q4-fit, echo = F, message = F, warning = F}
# Starting values from Q1 GLS (same as Q3)
ps0_q4 <- c(bhat[1], bhat[2], bhat[3], 0, s2hat)

fit4 <- optim(
  par     = ps0_q4,
  fn      = negcl,
  x       = x,
  y       = y,
  method  = "BFGS",
  hessian = TRUE
)

if (fit4$convergence != 0)
  warning("optim did not converge: code ", fit4$convergence)

mcle        <- fit4$par
names(mcle) <- c("alpha", "beta", "gamma", "theta", "sigma2")
```

## **(a)** Expression for the Log Composite Likelihood

See derivation above; implemented in `logcl_q4`.

## **(b)** Maximization

The log composite likelihood was maximized using `optim` with `method = "BFGS"`, using Q1 GLS estimates as starting values for $(\alpha, \beta, \gamma)$ and $\theta_0 = 0$, $\sigma^2_0 = \hat\sigma^2_{\text{Q1}}$.

## **(c)** MCLEs, Confidence Intervals, and Maximized Log Composite 

Likelihood Wald intervals use the standard inverse-Hessian covariance, computed from the Hessian of the negative composite log likelihood returned by `optim`. These are reported alongside the Godambe-based intervals further below.

```{r q4-wald, echo = F, message = F, warning = F}
cov_wald4 <- solve(fit4$hessian)
se_wald4  <- sqrt(diag(cov_wald4))

ci_wald4 <- data.frame(
  Parameter = c("$\\alpha$", "$\\beta$", "$\\gamma$", "$\\theta$", "$\\sigma^2$"),
  MCLE      = unname(round(mcle,                    4)),
  SE        = unname(round(se_wald4,                4)),
  Lower     = unname(round(mcle - z95 * se_wald4,   4)),
  Upper     = unname(round(mcle + z95 * se_wald4,   4))
)

kable(ci_wald4,
      format    = "latex",
      booktabs  = TRUE,
      escape    = FALSE,
      align     = c("l", "r", "r", "r", "r"),
      col.names = c("Parameter", "MCLE", "SE", "Lower", "Upper"),
      caption   = paste0("MCLEs with standard errors and Wald 95\\% confidence intervals. ",
                         "Maximized log composite likelihood: ",
                         round(-fit4$value, 4), ".")) |>
  add_header_above(c(" " = 3, "95% CI" = 2)) |>
  kable_styling(latex_options = "HOLD_position", font_size = 10)
```

## H and J Matrices and Godambe-Based Intervals

For $\hat{J}$ we compute the sum of outer products of individual piece score vectors $\nabla\ell_i$, approximated by numerical central differences. This uses the Context I outer-product form (Ch. 9.3.1); under the Comets-Janzura result for Context II, $J_m$ would instead involve neighborhood-based sums (Ch. 9, eq. 9.34), but since that result does not apply here, the Context I formula is used as a practical approximation with no direct theoretical justification under Context II. The resulting Godambe intervals should therefore be interpreted as heuristic rather than formally valid. Examining Tables 11 and 12, the $(3,3)$ elements of both $\hat{H}$ (38593.78) and $\hat{J}$ (40035.62) are very large relative to all others, reflecting high curvature in the $\gamma$ direction and suggesting $\gamma$ is the most precisely determined parameter locally. The large negative off-diagonal entries in rows/columns 1 and 3, and 2 and 3, are consistent with the high $(\alpha,\gamma)$ and $(\beta,\gamma)$ correlations seen in Table 15, and confirm that the identifiability challenges in this model are concentrated in the $(\alpha,\beta,\gamma)$ block.

```{r q4-godambe, echo = F, message = F, warning = F}
# Numerical score for a single piece i (i indexes into 2:n, so piece i
# corresponds to the pair (y[i], y[i-1]) with i in 2:n)
score_piece <- function(par, x, y, i) {
  h   <- 1e-5 * abs(par)
  h[h < 1e-8] <- 1e-8
  p   <- length(par)
  scr <- numeric(p)
  for (k in seq_len(p)) {
    par_hi       <- par; par_hi[k] <- par[k] + h[k]
    par_lo       <- par; par_lo[k] <- par[k] - h[k]
    # single-piece log CL for pair (y[i-1], y[i])
    piece <- function(pv) {
      a <- pv[1]; b <- pv[2]; g <- pv[3]
      th <- pv[4]; s2 <- pv[5]
      mu   <- a * x^(b * x^(-g))
      w    <- y - mu
      ei   <- w[i] - th * w[i - 1]
      -0.5 * log(2 * pi) - 0.5 * log(s2) - ei^2 / (2 * s2)
    }
    scr[k] <- (piece(par_hi) - piece(par_lo)) / (2 * h[k])
  }
  scr
}

n_obs <- length(y)
p     <- 5

# J: sum of outer products of piece scores
J_hat <- matrix(0, p, p)
for (i in 2:n_obs) {
  si    <- score_piece(mcle, x, y, i)
  J_hat <- J_hat + outer(si, si)
}

# H: negative Hessian of composite LL at MCLE (already computed by optim)
H_hat <- fit4$hessian   # Hessian of *negative* CL, so this is already -d^2 lC

# Godambe information and sandwich covariance
# G = H J^{-1} H  =>  G^{-1} = H^{-1} J H^{-1}
H_inv    <- solve(H_hat)
cov_god  <- H_inv %*% J_hat %*% H_inv
se_god   <- sqrt(diag(cov_god))
names(se_god) <- names(mcle)

ci_god <- data.frame(
  Parameter = c("$\\alpha$", "$\\beta$", "$\\gamma$", "$\\theta$", "$\\sigma^2$"),
  MCLE      = unname(round(mcle,                 4)),
  SE        = unname(round(se_god,               4)),
  Lower     = unname(round(mcle - z95 * se_god,  4)),
  Upper     = unname(round(mcle + z95 * se_god,  4))
)

kable(ci_god,
      format    = "latex",
      booktabs  = TRUE,
      escape    = FALSE,
      align     = c("l", "r", "r", "r", "r"),
      col.names = c("Parameter", "MCLE", "SE", "Lower", "Upper"),
      caption   = "MCLEs with Godambe-based 95\\% confidence intervals.") |>
  add_header_above(c(" " = 3, "95\\% CI" = 2)) |>
  kable_styling(latex_options = "HOLD_position", font_size = 10)

pnames <- c("$\\alpha$", "$\\beta$", "$\\gamma$", "$\\theta$", "$\\sigma^2$")

H_tab <- as.data.frame(round(H_hat, 4))
rownames(H_tab) <- pnames
colnames(H_tab) <- pnames

kable(H_tab,
      format   = "latex",
      booktabs = TRUE,
      escape   = FALSE,
      caption  = "Estimated $H$ matrix (negative Hessian of composite log likelihood at MCLE).") |>
  kable_styling(latex_options = "HOLD_position", font_size = 9)

J_tab <- as.data.frame(round(J_hat, 4))
rownames(J_tab) <- pnames
colnames(J_tab) <- pnames

kable(J_tab,
      format   = "latex",
      booktabs = TRUE,
      escape   = FALSE,
      caption  = "Estimated $J$ matrix (sum of outer score products at MCLE).") |>
  kable_styling(latex_options = "HOLD_position", font_size = 9)
```

## **(d)** Fitted Curve and 95% Pointwise Confidence Band

The band uses the delta method with the $(\alpha,\beta,\gamma)$ block of the Godambe covariance $\hat{G}^{-1}$.

```{r q4-band, echo=F, message=F, warning=F, fig.height=4.5, fig.width=6}
mugrid4 <- efctn(xgrid, mcle[1:3])
Dgrid4  <- eders(xgrid, mcle[1:3])
cov_abg4 <- cov_god[1:3, 1:3]
pvar4    <- apply(Dgrid4, 1,
                  function(d) as.numeric(t(d) %*% cov_abg4 %*% d))
hw4      <- z95 * sqrt(pvar4)
plot(x, y, xlab = "Covariate Value", ylab = "Response Value", main = "")
lines(xgrid, mugrid4, lwd = 2)
lines(xgrid, mugrid4 + hw4, lty = 2)
lines(xgrid, mugrid4 - hw4, lty = 2)
```

*Figure: Fitted extended Freundlich curve with approximate 95\% pointwise confidence band based on the Godambe covariance. Band uses the delta method with the $(\alpha,\beta,\gamma)$ block of $\hat{G}^{-1}$; see Q3(d) for the analogous Wald construction.*

## **(e)** Comparison of Large-Scale Component to Q1 and Q3

```{r q4-comp-large, echo = F, message = F, warning = F}
pnames3 <- c("$\\alpha$", "$\\beta$", "$\\gamma$")

# --- Point estimates
comp_pt4 <- data.frame(
  Parameter = pnames3,
  Q1_GLS    = unname(round(bhat,      4)),
  Q3_MLE    = unname(round(mle[1:3],  4)),
  Q4_MCLE   = unname(round(mcle[1:3], 4))
)

kable(comp_pt4,
      format    = "latex",
      booktabs  = TRUE,
      escape    = FALSE,
      align     = c("l", "r", "r", "r"),
      col.names = c("Parameter", "Q1 GLS", "Q3 MLE", "Q4 MCLE"),
      caption   = "Point estimates of large-scale parameters across methods.") |>
  kable_styling(latex_options = "HOLD_position", font_size = 10)

# --- CI widths
comp_width4 <- data.frame(
  Parameter      = pnames3,
  Width_Q1       = unname(round(2 * z95 * sqrt(diag(fit1$covb)), 4)),
  # Width_Q3       = unname(round(2 * z95 * se_mle[1:3],           4)),
  Width_Q3 = unname(round(prof_upper[1:3] - prof_lower[1:3], 4)),
  Width_Q4_Wald  = unname(round(2 * z95 * se_wald4[1:3],         4)),
  Width_Q4_God   = unname(round(2 * z95 * se_god[1:3],           4))
)

kable(comp_width4,
      format    = "latex",
      booktabs  = TRUE,
      escape    = FALSE,
      align     = c("l", "r", "r", "r", "r"),
      col.names = c("Parameter", "Q1 GLS", "Q3 MLE", "Q4 Wald", "Q4 Godambe"),
      caption   = paste0("95\\% CI widths for large-scale parameters across methods ",
                         "(Q1 Wald, Q3 profile, Q4 Wald and Godambe). ",
                         "Mean confidence band half-width (Q4): ",
                         round(mean(hw4), 4), ".")) |>
  add_header_above(c(" " = 1, " " = 1, " " = 1, "Q4" = 2)) |>
  kable_styling(latex_options = "HOLD_position", font_size = 10)

# --- Correlation matrix
cor_q4 <- round(cov2cor(cov_god[1:3, 1:3]), 3)
rownames(cor_q4) <- colnames(cor_q4) <- pnames3

kable(cor_q4,
      format   = "latex",
      booktabs = TRUE,
      escape   = FALSE,
      caption  = "Estimated correlation matrix for $(\\alpha, \\beta, \\gamma)$: Q4 Godambe.") |>
  kable_styling(latex_options = "HOLD_position", font_size = 10)
```

```{r resid-comparison, echo = F, message = F, warning = F, fig.height=12, fig.width=9, fig.cap="Standardized residuals versus covariate under Q1 GLS (left), Q3 MLE (center), and Q4 MCLE (right)."}
# Q3 innovation residuals
w_q3      <- y - efctn(x, mle[1:3])
innov_q3  <- w_q3 - mle["theta"] * c(0, w_q3[-n])
stdres_q3 <- innov_q3 / sqrt(mle["sigma2"])

# Q4 innovation residuals
w_q4      <- y - efctn(x, mcle[1:3])
innov_q4  <- w_q4 - mcle["theta"] * c(0, w_q4[-n])
stdres_q4 <- innov_q4 / sqrt(mcle["sigma2"])

ylim_r <- range(c(fit1$stdres, stdres_q3, stdres_q4))

par(mfrow = c(3, 1))

plot(x, fit1$stdres, xlab = "Covariate Value", ylab = "Standardized Residual",
     main = "Q1 GLS", ylim = ylim_r)
abline(h = 0, lty = 3)

plot(x, stdres_q3, xlab = "Covariate Value", ylab = "",
     main = "Q3 MLE", ylim = ylim_r)
abline(h = 0, lty = 3)

plot(x, stdres_q4, xlab = "Covariate Value", ylab = "",
     main = "Q4 MCLE", ylim = ylim_r)
abline(h = 0, lty = 3)
```

```{r resid-comparison-sqrt, echo = F, message = F, warning = F, eval = F, fig.height=12, fig.width=9, fig.cap="Standardized residuals versus square root of covariate under Q1 GLS (left), Q3 MLE (center), and Q4 MCLE (right)."}
# Q3 innovation residuals
w_q3      <- y - efctn(x, mle[1:3])
innov_q3  <- w_q3 - mle["theta"] * c(0, w_q3[-n])
stdres_q3 <- innov_q3 / sqrt(mle["sigma2"])
# Q4 innovation residuals
w_q4      <- y - efctn(x, mcle[1:3])
innov_q4  <- w_q4 - mcle["theta"] * c(0, w_q4[-n])
stdres_q4 <- innov_q4 / sqrt(mcle["sigma2"])
ylim_r <- range(c(fit1$stdres, stdres_q3, stdres_q4))
par(mfrow = c(3, 1))
plot(sqrt(x), fit1$stdres, xlab = expression(sqrt("Covariate Value")), ylab = "Standardized Residual",
     main = "Q1 GLS", ylim = ylim_r)
abline(h = 0, lty = 3)
plot(sqrt(x), stdres_q3, xlab = expression(sqrt("Covariate Value")), ylab = "",
     main = "Q3 MLE", ylim = ylim_r)
abline(h = 0, lty = 3)
plot(sqrt(x), stdres_q4, xlab = expression(sqrt("Covariate Value")), ylab = "",
     main = "Q4 MCLE", ylim = ylim_r)
abline(h = 0, lty = 3)
```

The point estimates show a clear monotone progression across methods: $\hat\alpha$ increases from 0.506 (Q1) to 0.740 (Q3) to 1.063 (Q4), while $\hat\beta$ and $\hat\gamma$ decrease correspondingly. This pattern is consistent with movement along the likelihood ridge defined by the near-collinearity among the large-scale parameters — Table 15 shows $\hat\rho_{\alpha,\beta} = -0.980$ and $\hat\rho_{\alpha,\gamma} = -0.845$ under Q4, so small shifts in the likelihood surface translate into substantial opposing movements in individual parameter estimates. The progression is not noise; it reflects the successive reduction in the number of terms contributing to the likelihood (Q1 ignores $\theta$ entirely, Q3 conditions on $w_0 = 0$ using all $n$ terms, Q4 conditions on $Y_1$ using only $n-1$ terms).

The CI widths (Table 14) tell a nuanced story. Q4 Wald intervals are the widest of all four methods for every parameter — substantially so for $\alpha$ (5.467 under Q4 Wald vs 2.044 under Q3 profile) — because the inverse-Hessian covariance of the composite likelihood inherits the near-singular curvature of the five-parameter surface without the stabilizing contribution of the first observation. The Godambe-based intervals are narrower than Q4 Wald for all three parameters, because $\hat{J} \approx \hat{H}$ numerically (Tables 11 and 12 are in close agreement), so the sandwich correction $\hat{H}^{-1}\hat{J}\hat{H}^{-1}$ deflates slightly relative to $\hat{H}^{-1}$ alone rather than inflating as it would when score variability substantially exceeds curvature. It is worth noting, however, that this agreement may partly reflect the small number of outer products ($n - 1 = 22$) used to estimate $\hat{J}$, which may not be sufficient to reliably distinguish $J$ from $H$ at this sample size. Under correct specification of the composite likelihood, $J = H$ is expected to hold asymptotically---so the approximate equality $\hat{J} \approx \hat{H}$ provides informal evidence that the pairwise conditional model is not grossly misspecified, though the small sample makes this interpretation tentative.

Turning to the standardized residual plots, all three methods produce visually similar patterns — the residuals are confined roughly to $(-2, 1)$ across the covariate range, with no obvious trend or fan. This is somewhat surprising given how different the Q1 and Q3/Q4 model specifications are, and warrants caution in interpretation. Two observations: first, the innovation residuals used for Q3 and Q4 ($\hat\epsilon_i = \hat{w}_i - \hat\theta \hat{w}_{i-1}$) are conceptually different objects from the Q1 standardized residuals — they have had the AR(1) dependence explicitly removed, so the absence of trend in the Q3/Q4 panels is the expected behavior of correctly whitened residuals, not evidence that Q1 was adequate. Second, there does appear to be mild cyclical structure across all three panels — positive residuals cluster at low and high covariate values, with a slight dip in the mid-range — which may reflect a small remaining misspecification in the mean function $\mu_i$ itself rather than the error structure. This would be consistent with Sibbesen's (1981) observation that the extended Freundlich curve is an empirical approximation and may not perfectly capture the sorption mechanism at all covariate values. With $n = 23$ it is not possible to formally test for this, but it is a limitation worth noting. Plotting residuals against $\sqrt{x}$ or $\log(x)$ in place of $x$ compresses the dense cluster of small covariate values and can make structure in that region easier to see, but does not alter the residuals themselves or change any of the above comparisons — the same patterns are present under any monotone transformation of the horizontal axis, and the substantive conclusions about whitening, cyclical structure, and mean misspecification are unaffected by the choice of scale.

## **(f)** Comparison of Small-Scale Component

```{r q4-comp-small, echo = F, message = F, warning = F}
comp_small4 <- data.frame(
  Quantity = c("$\\theta$", "$\\sigma^2$"),
  Q2_OptEE = unname(round(c(theta_opt, s2_opt),       4)),
  Q3_MLE   = unname(round(mle[c("theta", "sigma2")],  4)),
  Q4_MCLE  = unname(round(mcle[c("theta", "sigma2")], 4))
)

kable(comp_small4,
      format    = "latex",
      booktabs  = TRUE,
      escape    = FALSE,
      align     = c("l", "r", "r", "r"),
      col.names = c("Quantity", "Q2 Optimal EE", "Q3 MLE", "Q4 MCLE"),
      caption   = "Comparison of small-scale parameter estimates across methods.") |>
  kable_styling(latex_options = "HOLD_position", font_size = 10)

all_s2_4 <- data.frame(
  Estimator = c("Q1 GLS (marginal variance)", "Q2 moment-based",
                "Q3 MLE", "Q4 MCLE"),
  sigma2    = round(c(s2_q1, s2_opt, mle["sigma2"], mcle["sigma2"]), 4)
)

kable(all_s2_4,
      format    = "latex",
      booktabs  = TRUE,
      escape    = FALSE,
      align     = c("l", "r"),
      col.names = c("Estimator", "$\\hat{\\sigma}^2$"),
      caption   = "All estimates of $\\sigma^2$ across methods.") |>
  kable_styling(latex_options = "HOLD_position", font_size = 10)
```

The Q4 estimates of $\theta$ and $\sigma^2$ are close to Q3, as expected given that conditioning on $Y_1$ rather than fixing $w_0 = 0$ differs by a single term. The small monotone progression $\tilde\theta_{\text{opt}} = 0.542 \to \hat\theta_{\text{Q3}} = 0.560 \to \hat\theta_{\text{Q4}} = 0.575$ is consistent with successive reduction of first-stage approximation error: Q2 uses misspecified residuals, Q3 jointly estimates all parameters from the full conditional likelihood, and Q4 does the same from the pairwise conditional likelihood which down-weights the first observation. Any divergence again reflects the single omitted term. The same $\sigma^2$ hierarchy established in Q2(d) and Q3(f) applies: Q1 GLS targets $\sigma^2/(1 - \theta^2)$, while Q2, Q3, and Q4 all target the innovation variance $\sigma^2$.

\newpage 

# Q5

A two-stage estimation strategy would consist of estimating $\alpha$, $\beta$, $\gamma$ from the nonlinear constant variance model of Q1, and then using the residuals from that fit to estimate $\theta$ and $\sigma^2$ via the optimal estimating equation of Q2. Based on what you have learned in completing this exam, what would you conclude about the possibility of using such a two-stage procedure to fit the overall extended Freundlich model to data from this problem?

## Overall 

The two-stage procedure is superficially appealing because it decouples a hard five-parameter optimization into two simpler steps. The numerical results from Q1–Q4 allow a fairly detailed assessment of whether this decoupling actually works, and the picture that emerges is not encouraging — for reasons that go well beyond the sample size.

## Stage 1 is misspecified in a way that contaminates Stage 2

The first-stage GLS fit assumes $w_i \overset{iid}{\sim} N(0,\sigma^2)$. When the true errors follow an AR(1) process with $\theta \approx 0.56$, this misspecification has two consequences that compound each other.

First, the GLS estimating equations used in Stage 1 take the form $(D^\top W D)^{-1} D^\top W y$ where $D$ is the $n \times 3$ matrix of derivatives of $\mu_i$ and $W$ is the weight matrix; under the iid assumption of Stage 1, $W = I$, reducing this to $(D^\top D)^{-1} D^\top y$. The efficient estimator under AR(1) errors requires replacing the identity weight matrix with $\Sigma^{-1}(\theta)$, the inverse of the AR(1) covariance matrix. Ignoring this produces a consistent but *inefficient* estimator of $(\alpha,\beta,\gamma)$ in large samples — though consistency itself requires regularity conditions and a unique solution to the estimating equations that cannot be verified at $n = 23$ with a highly nonlinear mean function; the Q1 covariance matrix, which assumes $W = I$, is therefore incorrect even as a large-sample
approximation. The confidence intervals from Q1 are too narrow or too wide depending on the local curvature of $\mu_i$ relative to the correlation structure, and cannot be trusted for inference regardless of what happens in Stage 2.

Second, and more critically for Stage 2, the Q1 raw residuals $\hat{e}_i = y_i - \mu_i(\hat{\alpha}, \hat{\beta}, \hat{\gamma})$ are noisy surrogates for the true $w_i = y_i - \mu_i(\alpha, \beta, \gamma)$, subject to systematic finite-sample distortion. In large samples, as $(\hat{\alpha}, \hat{\beta}, \hat{\gamma}) \to (\alpha, \beta, \gamma)$, the residuals do converge to the true $w_i$; the problem is that $n = 23$ is far too small for this asymptotic regime to be operative. The difference is

$$\hat{e}_i - w_i = \mu_i(\alpha,\beta,\gamma) - \mu_i(\hat\alpha,\hat\beta,\hat\gamma),$$

which is a smooth nonlinear function of the Stage 1 parameter estimation error $(\hat\alpha - \alpha, \hat\beta - \beta, \hat\gamma - \gamma)$. This error does not vanish term-by-term; rather, it induces a systematic distortion in the expected autocorrelation structure of $\{\hat{e}_i\}$ relative to $\{w_i\}$, because the regression function $\mu_i$ varies smoothly with the ordered covariate. In particular, runs of positive or negative estimation error in $\mu_i$ across adjacent $x_i$ values will inflate or deflate the lag-1 autocorrelation of $\hat{e}_i$ relative to that of $w_i$.

## The numerical results confirm this distortion

The naive estimating equation applied to Q1 residuals returned $\tilde\theta_{\text{naive}} = 1.005$, which lies outside the stationary parameter space $|\theta| < 1$. This is not simply a small-sample anomaly; it is a symptom of the deterministic trend contamination described above. The sum $\sum_{i=2}^n \hat{e}_i$ in the numerator of $\tilde\theta_{\text{naive}}$ is strongly influenced by the systematic pattern visible in the Q1 residual plot (consistent positive residuals in the mid-range of $x$, negative at the extremes) --- the same residual structure Sibbesen (1981) observed and left unmodeled, and which motivated the AR(1) component in the first place --- which is precisely the signature of the misspecified error structure from Stage 1. The naive estimating function, which has no mechanism to separate genuine autocorrelation from trend contamination, is essentially useless here.

The optimal estimating equation fares better ($\tilde\theta_{\text{opt}} = 0.542$) because multiplying by $W_{i-1}$ down-weights the contribution of large residuals and focuses on local lag-1 relationships, making it more robust to the smooth global trend in $\hat{e}_i$. Nevertheless, $\tilde\theta_{\text{opt}}$ still deviates from the Q3 MLE of $0.560$, and the direction of the difference is consistent with attenuation toward zero caused by the residual contamination.

The $\sigma^2$ estimates reinforce this: the Q1 GLS moment estimate ($1.015$) targets the marginal variance $\sigma^2/(1-\theta^2)$ under the iid assumption, not the innovation variance $\sigma^2$. Feeding this into a two-stage procedure would produce inconsistent starting conditions for any subsequent inference on the full model, since the two quantities are related by a factor of $1/(1-\theta^2) \approx 1.45$ at $\hat\theta \approx 0.56$.

## Efficiency loss relative to the joint MLE

```{r q5-bands, echo = F, message = F, warning = F, fig.height=4.5, fig.width=9, fig.cap="Fitted extended Freundlich curves with approximate 95\\% pointwise confidence bands: Q1 GLS (left), Q3 MLE (center), Q4 MCLE/Godambe (right)."}
ylim_shared <- range(c(mugrid  - hw,  mugrid  + hw,
                       mugrid3 - hw3, mugrid3 + hw3,
                       mugrid4 - hw4, mugrid4 + hw4, y))

par(mfrow = c(1, 3))

plot(x, y, xlab = "Covariate Value", ylab = "Response Value",
     main = "Q1 GLS", ylim = ylim_shared)
lines(xgrid, mugrid, lwd = 2)
lines(xgrid, mugrid + hw,  lty = 2)
lines(xgrid, mugrid - hw,  lty = 2)

plot(x, y, xlab = "Covariate Value", ylab = "",
     main = "Q3 MLE", ylim = ylim_shared)
lines(xgrid, mugrid3, lwd = 2)
lines(xgrid, mugrid3 + hw3, lty = 2)
lines(xgrid, mugrid3 - hw3, lty = 2)

plot(x, y, xlab = "Covariate Value", ylab = "",
     main = "Q4 MCLE", ylim = ylim_shared)
lines(xgrid, mugrid4, lwd = 2)
lines(xgrid, mugrid4 + hw4, lty = 2)
lines(xgrid, mugrid4 - hw4, lty = 2)
```

Even if consistency of the two-stage estimator could be established under regularity conditions (it can, in principle, when $n \to \infty$ and Stage 1 gives consistent first-stage estimates), efficiency is lost relative to the joint MLE. The joint Q3 MLE accounts simultaneously for the curvature of the likelihood in all five parameters. The Stage 1 GLS, being based on a misspecified weight matrix, solves a different set of estimating equations entirely, and the resulting parameter estimates sit at a different point on the likelihood surface. The Q3 results illustrate this concretely: the MLE shifts $\hat\alpha$ from $0.506$ to $0.739$ and $\hat\beta$ from $1.222$ to $1.028$, non-trivial movements that reflect the reweighting of observations that occurs when the AR(1) structure is correctly incorporated. The profile CIs under Q3 correctly exclude inadmissible values for $\alpha$ and are wider for $\gamma$ than under Q1 Wald, and the confidence band is wider under Q3, reflecting the increased uncertainty in $(\alpha,\beta,\gamma)$ once the correlated error structure is properly acknowledged. A two-stage procedure cannot reproduce this because Stage 1 is oblivious to $\theta$.

## The Q4 composite likelihood provides corroborating evidence

The Q4 pairwise conditional composite likelihood, which uses only the $n-1$ adjacent conditional pairs rather than the full joint distribution, produced estimates close to Q3 ($\hat\theta = 0.575$, $\hat\sigma^2 = 0.644$), further demonstrating that the information about $\theta$ is primarily carried by adjacent pairs and that both Q3 and Q4 are substantially self-consistent. The two-stage approach, by contrast, implicitly assumes that the information in the regression function and the information in the autocorrelation structure are separable, which they are not: the AR(1) structure of the errors means that the likelihood contributions from adjacent $Y_i$ pairs jointly depend on all five parameters simultaneously.

## Conclusion

```{r q5-summary, echo = F, message = F, warning = F}
summary_tab <- data.frame(
  Method  = c("Q1 GLS", "Q2 Optimal EE", "Q3 MLE", "Q4 MCLE"),
  Theta   = c("---",
              round(theta_opt,       3),
              round(mle["theta"],    3),
              round(mcle["theta"],   3)),
  Sigma2  = round(c(s2_q1, s2_opt,
                    mle["sigma2"], mcle["sigma2"]), 3),
  BandHW  = c(round(mean(hw),  3), "---",
              round(mean(hw3), 3),
              round(mean(hw4), 3))
)

kable(summary_tab,
      format    = "latex",
      booktabs  = TRUE,
      escape    = FALSE,
      align     = c("l", "r", "r", "r"),
      col.names = c("Method", "$\\hat{\\theta}$",
                    "$\\hat{\\sigma}^2$", "Mean Band HW"),
      caption = "Small-scale estimates and mean confidence band half-widths across all methods.") |>
  kable_styling(latex_options = "HOLD_position", font_size = 10)
```

Q1 GLS targets the marginal variance $\sigma^2/(1-\theta^2)$; all remaining methods target the innovation variance $\sigma^2$. Band half-widths are delta-method throughout; scalar parameter CIs for Q3 use profile intervals (Table 3).

The two-stage procedure should not be used for this model. The Stage 1 GLS fit is based on an incorrect error specification, producing residuals that are contaminated by a systematic, covariate-dependent estimation error that distorts the autocorrelation structure used in Stage 2. Consistency of the full procedure requires that this contamination vanish asymptotically, but $n = 23$ is far too small for that approximation to be credible, and the numerical evidence — particularly $\tilde\theta_{\text{naive}} > 1$ — suggests the distortion is materially large at this sample size. More fundamentally, the efficiency loss relative to the joint MLE is not merely a finite-sample concern: because the likelihood surface couples the regression parameters and the AR(1) parameters, decoupled estimation cannot recover the efficiency of the joint estimator even in large samples. Among the analytic estimators considered here, the joint MLE of Q3 is the most appropriate; a parametric bootstrap procedure, as suggested by the course notes for Context II applications, would be the natural next step if formal interval estimates were required.