---
title: "HW8"
output: pdf_document
date: "2024-11-06"
---

# Q1 

Write a function which takes 2 arguments n and k which are positive integers. It should return the nXn matrix:

$$
\begin{pmatrix}
k & 1 & 0 & 0 & \dots & 0 & 0 \\
1 & k & 1 & 0 & \dots & 0 & 0 \\
0 & 1 & k & 1 & \dots & 0 & 0 \\
0 & 0 & 1 & k & \dots & 0 & 0 \\
\vdots & \vdots & \vdots & \vdots & \ddots & \vdots & \vdots \\
0 & 0 & 0 & 0 & \dots & k & 1 \\
0 & 0 & 0 & 0 & \dots & 1 & k
\end{pmatrix}
$$

Call the function defined above for n = 6 and k = 5, and provide the matrix you obtain.

```{r}
create_matrix <- function(n, k) {
  mat <- matrix(0, n, n)
  
  diag(mat) <- k
  
  for (i in 1:(n-1)) {
    mat[i, i+1] <- 1
    mat[i+1, i] <- 1
  }
  
  return(mat)
}

matrix_result <- create_matrix(6, 5)
matrix_result
```

```{r}
create_matrix <- function(n, k) {
  # Create an n x n matrix filled with zeros
  mat <- matrix(0, n, n)
  
  # Fill the diagonal with k
  diag(mat) <- k
  
  # Fill the diagonals above and below the main diagonal with 1
  mat[row(mat) == col(mat) + 1] <- 1
  mat[row(mat) == col(mat) - 1] <- 1
  
  return(mat)
}

# Call the function for n = 6 and k = 5
matrix_result <- create_matrix(6, 5)
matrix_result
```

\newpage

# Q2 

Consider the continuous function

$$
f(x) = 
\begin{cases} 
x^2 + 2x + 3 & \text{if } x < 0 \\
x + 3 & \text{if } 0 \leq x < 2 \\
x^2 + 4x - 7 & \text{if } 2 \leq x 
\end{cases}
$$

Write a function tmpFn which takes a single argument xVec. The function should return the vector of values of the function f(x) evaluated at the values xVec. Plot the function f(x) for -3 < x < 3.

```{r}
tmpFn <- function(xVec) {
  result <- numeric(length(xVec))
  
  result[xVec < 0] <- xVec[xVec < 0]^2 + 2 * xVec[xVec < 0] + 3
  result[xVec >= 0 & xVec < 2] <- xVec[xVec >= 0 & xVec < 2] + 3
  result[xVec >= 2] <- xVec[xVec >= 2]^2 + 4 * xVec[xVec >= 2] - 7
  
  return(result)
}


xValues <- seq(-3, 3, length.out = 100)
yValues <- tmpFn(xValues)

plot(x = xValues, y = yValues, 
     type = "l",
     col = "black", 
     lwd = 1.5,
     xlab = "x", 
     ylab = "f(x)", 
     main = "Plot of f(x) for -3 < x < 3")
```

\newpage

# Q3 

Greatest common divisor of two integers The greatest common divisor (gcd) of two integers m and n can be calculated using Euclid’s Algorithm: Divide m by n. If the remainder is zero, the gcd is n. If not, divide n by the remainder. If the remainder is zero, then the previous remainder is the gcd. If not, continue dividing the remainder into previous remainder until a remainder of zero is obtained. The gcd is the value of the last nonzero remainder. Write a function gcd(m,n) using a while loop to find the gcd of two integers m and n. 

```{r}
gcd <- function(m, n) {
  m <- abs(m)
  n <- abs(n)
  
  while (n != 0) {
    remainder <- m %% n
    m <- n
    n <- remainder
  }
  
  return(m)
}
```

```{r}
gcd(4*7, 4*4*4*7)
gcd(47, 93)
```

\newpage

# Q4

eQTL mapping. (The following problem was suggested by Professor Dan Nettleton.) Write a function order.matrix which takes in a matrix x and returns a matrix containing the row and column indices of the sorted values of x. Test this function on a 4X3 matrix of independent $\chi_{1}^2$ pseudo-random deviates. 

```{r}
order.matrix <- function(x) {
  sorted_indices <- order(x)
  
  row_indices <- (sorted_indices - 1) %% nrow(x) + 1
  col_indices <- (sorted_indices - 1) %/% nrow(x) + 1
  
  result <- cbind(row_indices, col_indices)
  return(result)
}

set.seed(37)  
test_matrix <- matrix(rchisq(4 * 3, df = 1), nrow = 4, ncol = 3)
dim(test_matrix)
test_matrix

indices <- order.matrix(test_matrix)
indices
dim(indices)
```

\newpage

# Q5 

Polar representation of a number. Let $x \in \mathbb{R}^p$. The polar respresentation of $\textbf{x} = (x_1, x_2, ..., x_p)$ is given by ($R, \theta_1, \theta_2, ..., \theta_{p-1}$), where: 

$$
\begin{aligned}
x_1 &= R \cos \theta_1 \\
x_2 &= R \sin \theta_1 \cos \theta_2 \\
x_3 &= R \sin \theta_1 \sin \theta_2 \cos \theta_3 \\
&\;\;\dots \quad = \quad \dots \\
x_{p-1} &= R \prod_{i=1}^{p-2} \sin \theta_i \cos \theta_{p-1} \\
x_p &= R \prod_{i=1}^{p-1} \sin \theta_i,
\end{aligned}
$$

where $0 \leq R < \infty, 0 \leq \theta_1 < 2\pi \text{ and } 0 \leq \theta_i < \pi$ i = 2, 3, . . . , p - 1.

## (a)

Write a function polaroid which takes in an arbitrary p-dimensional vector $\mathbf{x}$ and provides its polar representation as a vector, with the first element as $R$ and the remainder being $\theta_1, \theta_2, \dots, \theta_{p-1}$. 

```{r}
polaroid <- function(x) {
  R <- sqrt(sum(x^2))
  
  p <- length(x)
  theta <- numeric(p - 1)
  
  if (R > 0) {
    theta[1] <- atan2(x[2], x[1])
    if (theta[1] < 0) {
      theta[1] <- theta[1] + 2 * pi  
    }
    
    for (i in 2:(p - 1)) {
      numerator <- sqrt(sum(x[(i + 1):p]^2))
      denominator <- sqrt(sum(x[i:p]^2))
      theta[i] <- acos(numerator / denominator)
    }
  }
  
  return(c(R, theta))
}

x <- c(1, 2, 3)
polar_representation <- polaroid(x)
polar_representation
```

```{r}
polaroid <- function(x) {
  R <- sqrt(sum(x^2))
  
  if (R == 0) return(c(R))
  
  theta1 <- atan2(x[2], x[1])
  if (theta1 < 0) {
    theta1 <- theta1 + 2 * pi  
  }
  
  p <- length(x)
  if (p > 2) {
    numerators <- sqrt(rev(cumsum(rev(x[-c(1, 2)]^2))))
    denominators <- sqrt(cumsum(rev(x[-1]^2)))
    theta_rest <- acos(numerators / denominators)
  } else {
    theta_rest <- numeric(0)  
  }
  
  return(c(R, theta1, theta_rest))
}

x <- c(1, 2, 3)
polar_representation <- polaroid(x)
print(polar_representation)
```

## (b)

Write a function normalize which takes in a matrix and returns its normalized form: i.e., the matrix with rows scaled such that the sum of squares of each row is equal to 1. 

```{r}
normalize <- function(x) {
  row_norms <- sqrt(rowSums(x^2))
  
  normalized_matrix <- x / row_norms
  
  normalized_matrix[is.nan(normalized_matrix)] <- 0
  
  return(normalized_matrix)
}
```

```{r}
set.seed(37)  
test_matrix <- matrix(rnorm(12), nrow = 4, ncol = 3)
test_matrix
rowSums(test_matrix^2)

normalized_matrix <- normalize(test_matrix)
normalized_matrix
rowSums(normalized_matrix^2)
```

## (c)

Obtain a 1000X5 matrix $\mathbf{y}$ of $N(0, 1)$ pseudo-random deviates. Use apply and normalize to obtain the normalized values. Call this matrix $\mathbf{z}$. We test whether the columns of $\mathbf{z}$ are uniform on $U(-1, 1)$. One may test whether a sample $x \sim U(-1, 1)$ using ks.test(x, "punif", min=-1, max=1) where punif represents the cumulative distribution function of the uniform over range $(-1, 1)$. Summarize your results. 

```{r}
set.seed(37)  

y <- matrix(rnorm(1000 * 5), nrow = 1000, ncol = 5)

normalize <- function(x) {
  row_norms <- sqrt(rowSums(x^2))
  normalized_matrix <- sweep(x, 1, row_norms, FUN = "/")
  normalized_matrix[is.nan(normalized_matrix)] <- 0
  return(normalized_matrix)
}

z <- apply(X = y, 
           MARGIN = 2,
           FUN = function(col) normalize(matrix(col, ncol = 1))
           )

ks_tests <- apply(X = z, 
                  MARGIN = 2, 
                  FUN = function(column) ks.test(column, "punif", min = -1, max = 1)
                  )

ks_tests
```

The KS test results strongly suggest that the columns of the normalized matrix z are not uniformly distributed on (-1,1). The deviations are significant, implying that the normalization did not transform the data into a uniform distribution over this range. Further investigation or a different transformation might be needed if the goal is to achieve a uniform distribution.

## (d)

Obtain polar representations of $\mathbf{y}$ using your function polaroid and test whether $R^2 \sim \chi^2_5$ distribution. Provide a page of histograms or boxplots of $\theta_1, \theta_2, \theta_3, \theta_4$. Test whether these are from the uniform distributions on their respective ranges, i.e., $[0, 2\pi)$ for $\theta_1 \text{ and } [0, \pi)$ for $\theta_2, \theta_3, \theta_4$. 

```{r, warning = F}
set.seed(37) 

polaroid <- function(x) {
  # Calculate R
  R <- sqrt(sum(x^2))
  
  # If R is zero, return R with no angles (undefined case)
  if (R == 0) return(c(R))
  
  # Calculate the first angle theta1 using atan2
  theta1 <- atan2(x[2], x[1])
  if (theta1 < 0) theta1 <- theta1 + 2 * pi
  
  p <- length(x)
  
  # Calculate the remaining angles, only if p > 2
  if (p > 2) {
    numerators <- sqrt(rev(cumsum(rev(x[-c(1, 2)]^2))))
    denominators <- sqrt(cumsum(rev(x[-1]^2))[-1])
    
    # Ensure lengths match for division
    if (length(numerators) != length(denominators)) {
      stop("Numerators and denominators lengths do not match.")
    }
    
    # Calculate theta_rest and clamp values to be in [-1, 1]
    cos_values <- numerators / denominators
    cos_values <- pmin(pmax(cos_values, -1), 1)  # Clamp values to [-1, 1]
    theta_rest <- acos(cos_values)
  } else {
    theta_rest <- numeric(0)  # No additional angles if p <= 2
  }
  
  # Return the polar representation
  return(c(R, theta1, theta_rest))
}

# Test the function on a matrix y
set.seed(123)
y <- matrix(rnorm(1000 * 5), nrow = 1000, ncol = 5)
polar_representation <- t(apply(y, 1, polaroid))

# Extract R values and theta values
R_values <- polar_representation[, 1]
theta_values <- if (ncol(polar_representation) > 1) polar_representation[, -1] else NULL
```

```{r}
R_squared <- R_values^2

chisq_test <- ks.test(R_squared, "pchisq", df = 5)
chisq_test
```

```{r}
hist(theta_values[, 1], breaks = 30, main = expression(theta[1]), xlab = expression(theta[1]), xlim = c(0, 2 * pi))

hist(theta_values[, 2], breaks = 30, main = expression(theta[2]), xlab = expression(theta[2]), xlim = c(0, pi))
hist(theta_values[, 3], breaks = 30, main = expression(theta[3]), xlab = expression(theta[3]), xlim = c(0, pi))
hist(theta_values[, 4], breaks = 30, main = expression(theta[4]), xlab = expression(theta[4]), xlim = c(0, pi))
```

```{r}
theta1_test <- ks.test(theta_values[, 1] / (2 * pi), "punif", min = 0, max = 1)

theta2_test <- ks.test(theta_values[, 2] / pi, "punif", min = 0, max = 1)
theta3_test <- ks.test(theta_values[, 3] / pi, "punif", min = 0, max = 1)
theta4_test <- ks.test(theta_values[, 4] / pi, "punif", min = 0, max = 1)

list(theta1_test = theta1_test, theta2_test = theta2_test,
     theta3_test = theta3_test, theta4_test = theta4_test)
```