---
title: "Real Application"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Setup 

```{r, echo = F, warning = F, message = F}
library(curl)
library(quantmod)
library(fredr)
library(xts)
library(dplyr)
```

```{r, echo = F}
# 1. DAX index
getSymbols("^GDAXI", from = "1999-01-01", to = "2023-12-31", src = "yahoo")
dax <- Ad(GDAXI)

# 2. Risk-free rate
fredr_set_key("86cee523c9ff9876c995614832f0b45f")
r_series <- fredr("ECBDFR", 
                  observation_start = as.Date("1999-01-01"),
                  observation_end = as.Date("2023-12-31"))

# Fixed r for training window (1999-2013)
r <- mean(r_series$value[r_series$date <= "2013-12-31"], na.rm=TRUE) / 100

# 3. Log returns
dax_ret <- na.omit(diff(log(dax)))

# 4. Split: train 1999-2012, test 2013-2022
train_ret <- dax_ret["1999-01-01/2012-12-31"]
test_ret  <- dax_ret["2013-01-01/2022-12-31"]

# =============================================================
# POEMV SIMULATION STUDY  |  DAX × ECB Rate
# Replication of Wu & Li (2024), Section 5
# Assumes in workspace: dax (xts), r (scalar, annualised ECB rate)
# =============================================================
library(xts)
library(zoo)
set.seed(42)

FREQ  <- 52L;   T_HOR <- 10.0;  DT <- 1/FREQ
K     <- round(T_HOR / DT)          # 520 steps/episode
X0    <- 1.0;   Z <- 8.0;  LAM <- 1.0;  M <- 2L
Z <- 2.0  
# =============================================================
# 1  PARAMETER ESTIMATION  (heuristic, Section 5 of paper)
# =============================================================

# Clean price series before anything else
dax_clean <- na.approx(dax, na.rm = FALSE)
dax_clean <- na.omit(dax_clean)
cat("NAs remaining after cleaning:", sum(is.na(dax_clean)), "\n")

identify_regimes <- function(px, ul = 0.19, uu = 0.24) {
  px  <- px[!is.na(px)]
  n   <- length(px); reg <- integer(n)
  cur <- 1L; ref <- as.numeric(px[1L])
  for (i in seq_len(n)) {
    p <- as.numeric(px[i])
    if (is.na(p)) { reg[i] <- cur; next }
    if (cur == 1L) {
      if (p > ref) ref <- p
      if ((ref - p)/ref >= ul) { cur <- 2L; ref <- p }
    } else {
      if (p < ref) ref <- p
      if ((p - ref)/ref >= uu) { cur <- 1L; ref <- p }
    }
    reg[i] <- cur
  }
  reg
}

estimate_params <- function(ret, reg, freq = FREQ) {
  mu1 <- mean(ret[reg == 1L]) * freq
  mu2 <- mean(ret[reg == 2L]) * freq
  sig <- sd(ret) * sqrt(freq)
  rl  <- rle(reg)
  list(mu1 = mu1, mu2 = mu2, sigma = sig,
       lambda1 = freq / mean(rl$lengths[rl$values == 1L]),
       lambda2 = freq / mean(rl$lengths[rl$values == 2L]))
}

tr_px  <- as.numeric(dax_clean["1999-01-01/2012-12-31"])
te_px  <- as.numeric(dax_clean["2013-01-01/2022-12-31"])
tr_ret <- diff(log(tr_px))
te_ret <- diff(log(te_px))
tr_reg <- identify_regimes(tr_px)[-1L]

stopifnot(length(tr_ret) == length(tr_reg))

PAR    <- estimate_params(tr_ret, tr_reg)
RHO1   <- (PAR$mu1 - r) / PAR$sigma
RHO2   <- (PAR$mu2 - r) / PAR$sigma

cat("── True parameters ──────────────────────────────\n")
cat(sprintf("  μ₁=%6.3f  μ₂=%6.3f  σ=%5.3f  λ₁=%5.3f  λ₂=%5.3f\n",
            PAR$mu1, PAR$mu2, PAR$sigma, PAR$lambda1, PAR$lambda2))
cat(sprintf("  ρ₁=%6.3f  ρ₂=%6.3f  r=%6.4f\n", RHO1, RHO2, r))

# =============================================================
# 2  FINITE-DIFFERENCE PDE SOLVER  (Appendix B)
# =============================================================

# Thomas algorithm for tridiagonal systems
tridiag <- function(lo, di, up, rhs) {
  n  <- length(di); c2 <- d2 <- x <- numeric(n)
  c2[1L] <- up[1L]/di[1L]; d2[1L] <- rhs[1L]/di[1L]
  for (i in 2L:n) {
    den    <- di[i] - lo[i]*c2[i-1L]
    c2[i]  <- if (i < n) up[i]/den else 0
    d2[i]  <- (rhs[i] - lo[i]*d2[i-1L]) / den
  }
  x[n] <- d2[n]
  for (i in (n-1L):1L) x[i] <- d2[i] - c2[i]*x[i+1L]
  x
}

# Crank-Nicolson solver:  u_t + b(p)*u_pp + a(p)*u_p + c(p)*u = h(t,p)
# u(T,·)=phi; u(t,0)=bc0(t,T); u(t,1)=bc1(t,T)
cn_pde <- function(a_fn, b_fn, c_fn, h_fn, phi_fn, bc0_fn, bc1_fn,
                    T_hor, Np = 100, dt = 1/52) {
  Nt  <- ceiling(T_hor/dt); dt <- T_hor/Nt; dp <- 1/Np
  pg  <- seq(0, 1, length.out = Np+1L)
  tg  <- seq(0, T_hor, length.out = Nt+1L)
  U   <- matrix(0, Nt+1L, Np+1L)
  U[Nt+1L, ] <- phi_fn(pg)
  for (i in seq_len(Nt+1L)) {
    U[i, 1L]    <- bc0_fn(tg[i], T_hor)
    U[i, Np+1L] <- bc1_fn(tg[i], T_hor)
  }

  # Vectorised coefficient arrays at interior p points
  pi_int <- pg[2L:Np]
  av  <- a_fn(pi_int); bv <- b_fn(pi_int); cv <- c_fn(pi_int)
  sa  <- sign(av); sa[sa == 0L] <- 1L

  # Equation (38) LHS coefficients
  A  <- ifelse(av < 0, -av, 0)/(2*dp)  + bv/(2*dp^2)
  B  <-  -1/dt - bv/dp^2 + cv/2 - av/(2*dp)*sa
  C  <- ifelse(av >= 0, av, 0)/(2*dp) + bv/(2*dp^2)
  lo <- c(0, A[-1L]); up <- c(C[-length(C)], 0)
  Ni <- Np - 1L

  for (i in Nt:1L) {
    u1  <- U[i+1L, ]
    h1  <- h_fn(tg[i+1L], pi_int, U[i+1L, ])
    h0  <- h_fn(tg[i],    pi_int, U[i, ])        # BCs already set

    # Vectorised RHS (eq. 38 right side)
    rhs <- -A*u1[1L:Ni] - (B+2/dt)*u1[2L:(Ni+1L)] - C*u1[3L:(Ni+2L)] +
            (h0+h1)/2
    rhs[1L]  <- rhs[1L]  - A[1L]  * U[i, 1L]
    rhs[Ni]  <- rhs[Ni]  - C[Ni]  * U[i, Np+1L]
    U[i, 2L:Np] <- tridiag(lo, B, up, rhs)
  }
  list(U = U, pg = pg, tg = tg)
}

# PDE (18) -- f
pde_f <- function(r1, r2, lam1, lam2, T_hor, Np = 200)
  cn_pde(
    a_fn   = function(p) lam2-(lam1+lam2)*p -
                         2*(r2+(r1-r2)*p)*(r1-r2)*p*(1-p),
    b_fn   = function(p) 0.5*(r1-r2)^2*p^2*(1-p)^2,
    c_fn   = function(p) (r2+(r1-r2)*p)^2,
    h_fn   = function(t, p, U) rep(0, length(p)),
    phi_fn = function(p) rep(1, length(p)),
    bc0_fn = function(t, T) exp(r2^2*(T-t)),
    bc1_fn = function(t, T) exp(pmin(r1^2*(T-t), 700)),   # guard overflow
    T_hor, Np)

# PDE (19) -- g  (requires f solution as input)
pde_g <- function(r1, r2, lam1, lam2, sig, T_hor, f_sol,
                   Np = 200, Lambda = LAM)
  cn_pde(
    a_fn   = function(p) lam2-(lam1+lam2)*p,
    b_fn   = function(p) 0.5*(r1-r2)^2*p^2*(1-p)^2,
    c_fn   = function(p) rep(0, length(p)),
    h_fn   = function(t, p, U) {
      ti  <- which.min(abs(f_sol$tg - t))
      fv  <- pmax(approx(f_sol$pg, f_sol$U[ti,], xout = p)$y, 1e-12)
      (Lambda/2)*log(pi*Lambda*fv/sig^2)
    },
    phi_fn = function(p) rep(0, length(p)),
    bc0_fn = function(t, T)
      (Lambda*r2^2/4)*(T^2-t^2) -
      (Lambda/2)*(r2^2*T - log(sig^2/(pi*Lambda)))*(T-t),
    bc1_fn = function(t, T)
      (Lambda*r1^2/4)*(T^2-t^2) -
      (Lambda/2)*(r1^2*T - log(sig^2/(pi*Lambda)))*(T-t),
    T_hor, Np)

cat("── Solving PDEs (f then g) …")
f_sol <- pde_f(RHO1, RHO2, PAR$lambda1, PAR$lambda2, T_HOR, Np = 100)
g_sol <- pde_g(RHO1, RHO2, PAR$lambda1, PAR$lambda2,
               PAR$sigma, T_HOR, f_sol, Np = 100)
cat(" done.\n")

# =============================================================
# 3  DISCRETISED WONHAM FILTER  (equations 29-32)
# =============================================================

wonham_step <- function(q1, q2, dBhat, lam1, lam2, r1, r2) {
  dr  <- r1 - r2
  q1n <- q1 + (lam2*exp(-q1) - (lam1+lam2) - 0.5*dr^2*(1-exp(q1))^2)*DT +
         dr*(1-exp(q1))*dBhat
  q2n <- q2 + ((lam1+lam2)*(exp(-q2)-1) - lam2*exp(-q2) -
                0.5*dr^2*(1-exp(q2))^2)*DT -
         dr*(1-exp(q2))*dBhat
  p   <- exp(q1n)/(exp(q1n)+exp(q2n))
  list(q1 = q1n, q2 = q2n, p = p)
}

# =============================================================
# 4  RS-GBM PATH SIMULATOR
# =============================================================

sim_path <- function(mu1, mu2, sigma, lam1, lam2) {
  reg          <- integer(K)
  cur          <- 1L
  switch_probs <- c(lam1, lam2) * DT
  for (i in seq_len(K)) {
    if (runif(1L) < switch_probs[cur]) cur <- 3L - cur
    reg[i] <- cur
  }
  mu_vec <- ifelse(reg == 1L, mu1, mu2)
  ret    <- (mu_vec - sigma^2/2)*DT + sigma*sqrt(DT)*rnorm(K)
  list(ret = ret, regime = reg)
}

# =============================================================
# 5  VALUE FUNCTION & POLICY  (equations 25-28, m = 2)
# =============================================================
# θ[1..6]: f^θ params  θ_{i,j}: order (0,1)(0,2)(1,1)(1,2)(2,1)(2,2)
# θ[7..8]: g^θ params  θ̃_j for j=1,2  (p-independent simplification)
# φ = (φ₁, φ₂, φ₃)

lnf_th <- function(t, p, th) {
  tau <- T_HOR - t
  th[1]*tau + th[2]*tau^2 + th[3]*p*tau  + th[4]*p*tau^2 +
  th[5]*p^2*tau + th[6]*p^2*tau^2
}
f_th      <- function(t, p, th) exp(pmin(lnf_th(t, p, th), 500))
dlnf_dp   <- function(t, p, th) {          # fp/f = ∂lnf^θ/∂p
  tau <- T_HOR - t
  th[3]*tau + th[4]*tau^2 + 2*th[5]*p*tau + 2*th[6]*p*tau^2
}
# No change needed — both already handle vector t via element-wise ops.
# Explicitly confirm by adding a comment:
g_th <- function(t, th) { tau <- T_HOR-t; th[7]*tau + th[8]*tau^2 }  # vector-safe

V_th <- function(t, x, p, th, om)
  (x-om)^2 / f_th(t, p, th) + g_th(t, th) - (om-Z)^2               # vector-safe

# Gradient ∂V^θ/∂θ (equations 33-34)
dV_dth <- function(t, x, p, th, om) {
  tau <- T_HOR-t; ft <- f_th(t,p,th); cx <- -(x-om)^2/ft
  c(cx*c(tau, tau^2, p*tau, p*tau^2, p^2*tau, p^2*tau^2),   # θ_f part
    c(tau, tau^2))                                             # θ_g part
}

# Policy mean μ^φ, variance σ^φ
p_mean <- function(t, x, p, phi, th, om) {
  adj <- p - dlnf_dp(t,p,th)*p*(1-p)
  -(x-om)*(phi[1]*adj + phi[2])
}
p_var  <- function(t, p, phi, th) LAM*exp(phi[3])*f_th(t,p,th)
p_ent  <- function(t, p, phi, th) 0.5*log(2*pi*exp(1)*p_var(t,p,phi,th))
p_samp <- function(t, x, p, phi, th, om)
  rnorm(1L, p_mean(t,x,p,phi,th,om), sqrt(p_var(t,p,phi,th)))

# =============================================================
# 6  POEMV ACTOR-CRITIC  (Algorithm 1)
# =============================================================

poemv <- function(par_sim, par_filt,
                   n_iter    = 10000L,
                   s_omega   = 25L,
                   eta_th    = 1e-12,
                   eta_phi   = 5e-5,
                   eta_omega = 5e-3,
                   verbose   = TRUE) {

  r1f <- (par_filt$mu1-r)/par_filt$sigma
  r2f <- (par_filt$mu2-r)/par_filt$sigma

  th  <- rep(0, 8L); phi <- rep(0, 3L); om <- Z; X_T_acc <- 0
  hist <- data.frame(phi1=NA_real_, phi2=NA_real_,
                      omega=NA_real_, X_T=NA_real_)[rep(1,n_iter),]

  for (iter in seq_len(n_iter)) {

    # ── Step 2: episode + filter ──────────────────────────
    sim <- sim_path(par_sim$mu1, par_sim$mu2, par_sim$sigma,
                     par_sim$lambda1, par_sim$lambda2)
    ret <- sim$ret
    X   <- numeric(K+1L); X[1L] <- X0
    P   <- numeric(K+1L); P[1L] <- 0.5
    U_  <- H_ <- V_ <- numeric(K)
    q1  <- q2 <- log(0.5)

    for (k in seq_len(K)) {
      tk   <- (k-1L)*DT; xk <- X[k]; pk <- P[k]
      uk   <- p_samp(tk, xk, pk, phi, th, om)
      xn   <- xk + uk*(exp(ret[k])-1)
      if (!is.finite(xn)) { xn <- xk; uk <- 0 }
      X[k+1L] <- xn
      U_[k] <- uk
      mhat <- par_filt$mu1*pk + par_filt$mu2*(1-pk)
      dBh  <- (ret[k] - (mhat-par_filt$sigma^2/2)*DT)/par_filt$sigma
      ws   <- wonham_step(q1, q2, dBh, par_filt$lambda1,
                           par_filt$lambda2, r1f, r2f)
      q1 <- ws$q1; q2 <- ws$q2; P[k+1L] <- ws$p
      U_[k] <- uk; H_[k] <- p_ent(tk,pk,phi,th); V_[k] <- V_th(tk,xk,pk,th,om)
    }
    XT  <- X[K+1L]; hT <- (XT-om)^2-(om-Z)^2

# ── Step 3: PE update (vectorised) ───────────────────
    cumH    <- rev(cumsum(rev(H_))) * DT
    td_vec  <- hT - V_ - LAM * cumH
    tau_vec <- T_HOR - (seq_len(K)-1L)*DT
    px_vec  <- P[1L:K]
    cx_vec  <- -(X[1L:K]-om)^2 / pmax(f_th(tau_vec, px_vec, th), 1e-10)
    G_mat   <- cbind(
      cx_vec*tau_vec,         cx_vec*tau_vec^2,
      cx_vec*px_vec*tau_vec,  cx_vec*px_vec*tau_vec^2,
      cx_vec*px_vec^2*tau_vec,cx_vec*px_vec^2*tau_vec^2,
      tau_vec,                tau_vec^2
    )
    dth <- colSums(td_vec * G_mat) * DT
dth <- pmax(pmin(dth, 1e2), -1e2)
th  <- th + eta_th*dth

    # ── Step 4: PG update (vectorised) ───────────────────
    tk_vec   <- (seq_len(K)-1L)*DT
    xk_vec   <- X[1L:K]; pk_vec <- P[1L:K]
    adj_vec  <- pk_vec - dlnf_dp(tk_vec, pk_vec, th)*pk_vec*(1-pk_vec)
    mu_p_vec <- -(xk_vec-om)*(phi[1]*adj_vec + phi[2])
    sg_p_vec <- LAM*exp(phi[3])*f_th(tk_vec, pk_vec, th)
    res_vec  <- U_ - mu_p_vec
    sc12_vec <- -(res_vec*(xk_vec-om))/sg_p_vec
    Vn_vec   <- c(V_th((seq_len(K-1L))*DT, X[2L:K], P[2L:K], th, om), hT)
    td_vec2  <- Vn_vec - V_ - LAM*H_*DT
    dphi <- c(
      sum(sc12_vec * adj_vec * td_vec2),
      sum(sc12_vec * td_vec2),
      sum((res_vec^2 - sg_p_vec)/(2*sg_p_vec) * td_vec2) - LAM*0.5*K*DT
    )
phi <- phi - eta_phi*dphi
phi[3] <- pmax(pmin(phi[3], 10), -10)
    # ── Step 5: Lagrange update ───────────────────────────
    X_T_acc <- X_T_acc + XT
    if (iter %% s_omega == 0L) {
      om <- om - eta_omega*(X_T_acc/s_omega - Z); X_T_acc <- 0
    }

    hist[iter,] <- list(phi[1], phi[2], om, XT)
    if (verbose && iter %% 2500L == 0L)
      cat(sprintf("  [%5d] φ₁=%7.3f  φ₂=%7.3f  ω=%.3f  XT=%.3f\n",
                  iter, phi[1], phi[2], om, XT))
  }
  list(th=th, phi=phi, omega=om, hist=hist)
}

# =============================================================
# 7  EMV BASELINE  (no regime switching)
# =============================================================
# Collapses to Wang & Zhou (2020): policy φ* = ρ̄/σ
# f^θ(t) = exp(θ₁(T-t)+θ₂(T-t)²); no p dependence

emv <- function(par_sim, n_iter=10000L,
                 eta_th=1e-13, eta_phi=1e-4, eta_omega=1e-4,
                 om_max = 5.0,
                 verbose=TRUE) {
  th <- rep(0,4); phi <- rep(0,3); om <- Z; X_T_acc <- 0
  hist <- data.frame(phi1=NA_real_,omega=NA_real_,
                     X_T=NA_real_)[rep(1,n_iter),]

  fe <- function(t,th) {
    exp(pmin(th[1]*(T_HOR-t) + th[2]*(T_HOR-t)^2, 500))
  }
  ge <- function(t,th) {
    th[3]*(T_HOR-t) + th[4]*(T_HOR-t)^2
  }
  Ve <- function(t,x,th,om) {
    (x-om)^2 / pmax(fe(t,th), 1e-10) + ge(t,th) - (om-Z)^2
  }

  for (iter in seq_len(n_iter)) {
    sim <- sim_path(par_sim$mu1, par_sim$mu2, par_sim$sigma,
                     par_sim$lambda1, par_sim$lambda2)
    ret <- sim$ret
    X <- numeric(K+1L); X[1L] <- X0
    Uv <- Hv <- Vv <- numeric(K)

    for (k in seq_len(K)) {
      tk <- (k-1L)*DT; xk <- X[k]
      sg <- pmax(LAM*exp(phi[3])*fe(tk,th), 1e-10)
      uk <- rnorm(1L, -(xk-om)*phi[1], sqrt(sg))
      xn <- xk + uk*(exp(ret[k])-1)
      if (!is.finite(xn)) xn <- xk          # NaN guard
      X[k+1L] <- xn
      Uv[k] <- uk
      Hv[k] <- 0.5*log(2*pi*exp(1)*sg)
      Vv[k] <- Ve(tk, xk, th, om)
    }

    XT <- X[K+1L]
    if (!is.finite(XT)) XT <- X0            # episode-level NaN guard
    hT <- (XT-om)^2-(om-Z)^2

    # PE
    cumH <- rev(cumsum(rev(Hv)))*DT; dth <- rep(0,4)
    for (k in seq_len(K)) {
      tau <- T_HOR-(k-1L)*DT
      ft  <- pmax(fe((k-1L)*DT, th), 1e-10)
      td  <- hT - Vv[k] - LAM*cumH[k]
      cx  <- -(X[k]-om)^2/ft
      dth <- dth + td*c(cx*c(tau,tau^2), c(tau,tau^2))*DT
    }
    th <- th + eta_th*dth

    # PG
    dph <- rep(0,3)
    for (k in seq_len(K)) {
      sg  <- pmax(LAM*exp(phi[3])*fe((k-1L)*DT,th), 1e-10)
      res <- Uv[k] - (-(X[k]-om)*phi[1])
      sc  <- c(-(res*(X[k]-om))/sg, 0, (res^2-sg)/(2*sg))
      Vn  <- if(k<K) Ve(k*DT, X[k+1L], th, om) else hT
      td  <- Vn - Vv[k] - LAM*Hv[k]*DT
      dph <- dph + sc*td - LAM*c(0,0,0.5)*DT
    }
    phi <- phi - eta_phi*dph

    X_T_acc <- X_T_acc + XT
    if (iter %% 10L == 0L) {
      om <- om - eta_omega*(X_T_acc/10 - Z)
      om <- min(om, om_max)
      X_T_acc <- 0
    }

    hist[iter,] <- list(phi[1], om, XT)
    if (verbose && iter %% 2500L == 0L)
      cat(sprintf("  EMV [%5d] φ=%.3f  ω=%.3f  XT=%.3f\n",
                  iter, phi[1], om, XT))
  }
  list(th=th, phi=phi, omega=om, hist=hist)
}

# =============================================================
# 8  OUT-OF-SAMPLE EVALUATOR
# =============================================================

oos_eval <- function(par_sim, par_filt, phi, th, om,
                      mode = c("poemv","emv"), n_test = 1000L) {
  mode <- match.arg(mode)

  if (any(!is.finite(phi)) || any(!is.finite(th)))
    stop("oos_eval: phi or th contains NaN/Inf — check training output")

  r1f  <- (par_filt$mu1-r)/par_filt$sigma
  r2f  <- (par_filt$mu2-r)/par_filt$sigma
  XT   <- numeric(n_test)

  for (trial in seq_len(n_test)) {
    sim <- sim_path(par_sim$mu1, par_sim$mu2, par_sim$sigma,
                     par_sim$lambda1, par_sim$lambda2)
    xk <- X0; pk <- 0.5; q1 <- q2 <- log(0.5)
    for (k in seq_len(K)) {
      tk <- (k-1L)*DT
      uk <- if (mode=="poemv") {
        p_samp(tk, xk, pk, phi, th, om)
      } else {
        sg <- pmax(LAM*exp(phi[3])*
                   exp(pmin(th[1]*(T_HOR-tk)+th[2]*(T_HOR-tk)^2, 500)),
                   1e-10)
        rnorm(1L, -(xk-om)*phi[1], sqrt(sg))
      }
      uk    <- pmax(pmin(uk, 1e4), -1e4)
      ret_k <- sim$ret[k]
      xk    <- xk + uk*(exp(ret_k)-1)
      if (!is.finite(xk)) xk <- X0
      if (mode=="poemv") {
        mhat <- par_filt$mu1*pk+par_filt$mu2*(1-pk)
        dBh  <- (ret_k-(mhat-par_filt$sigma^2/2)*DT)/par_filt$sigma
        dBh  <- pmax(pmin(dBh, 5*sqrt(DT)), -5*sqrt(DT))
        ws   <- wonham_step(q1,q2,dBh,par_filt$lambda1,par_filt$lambda2,r1f,r2f)
        q1<-ws$q1; q2<-ws$q2; pk<-ws$p
      }
    }
    XT[trial] <- xk
  }
  c(mean=mean(XT), var=var(XT), sharpe=(mean(XT)-X0)/sd(XT))
}
```

# Real Data 

```{r, echo = F}
# =============================================================
# REAL DATA SECTION
# Replicates Wu & Li Section 6 using DAX/Eurostoxx constituents
# =============================================================

library(quantmod)
library(zoo)
library(xts)

# DAX 40 + selected Eurostoxx 50 tickers (Yahoo Finance .DE suffix)
candidate_tickers <- c(
  "ADS.DE","AIR.DE","ALV.DE","BAS.DE","BAYN.DE","BMW.DE",
  "BNR.DE","CON.DE","1COV.DE","DHER.DE","DTE.DE","EOAN.DE",
  "FRE.DE","HEI.DE","HEN3.DE","IFX.DE","LIN.DE","MBG.DE",
  "MRK.DE","MTX.DE","MUV2.DE","P911.DE","PAH3.DE","QIA.DE",
  "RHM.DE","RWE.DE","SAP.DE","SHL.DE","SIE.DE","SRT3.DE",
  "SY1.DE","TKA.DE","VNA.DE","VOW3.DE","ZAL.DE",
  # Eurostoxx additions
  "ASML.AS","MC.PA","OR.PA","SAN.PA","BNP.PA","AI.PA",
  "DG.PA","RI.PA","CS.PA","SU.PA"
)

# Pull all prices 1999-2022
cat("Pulling constituent prices…\n")
price_list <- list()
for (tk in candidate_tickers) {
  tryCatch({
    px <- Ad(getSymbols(tk, from="1999-01-01", to="2022-12-31",
                         auto.assign=FALSE, warnings=FALSE))
    px <- na.approx(px, na.rm=FALSE)
    price_list[[tk]] <- px
    cat(sprintf("  ✓ %s\n", tk))
  }, error = function(e) cat(sprintf("  ✗ %s (skipped)\n", tk)))
}

# =============================================================
# Filter 1: Require 90% data completeness over full window
# =============================================================
dax_clean <- na.approx(dax, na.rm = FALSE)
dax_clean <- na.omit(dax_clean)
ref_dates <- index(dax_clean["1999-01-01/2022-12-31"])
n_ref     <- length(ref_dates)
threshold <- 0.90 * n_ref

price_list <- Filter(function(px) {
  n_obs <- sum(!is.na(px))
  n_obs >= threshold
}, price_list)

cat(sprintf("\nAfter completeness filter: %d stocks retained\n",
            length(price_list)))

# =============================================================
# Filter 2: Beta filter 0.9 ≤ β ≤ 1.1 on training window
# =============================================================
dax_tr_ret <- diff(log(dax_clean["1999-01-01/2012-12-31"]))

compute_beta <- function(stk_px) {
  stk_ret <- diff(log(na.approx(stk_px["1999-01-01/2012-12-31"])))
  merged  <- merge(dax_tr_ret, stk_ret, join="inner")
  merged  <- na.omit(merged)
  if (nrow(merged) < 100) return(NA_real_)
  cov(merged[,2], merged[,1]) / var(merged[,1])
}

betas <- sapply(price_list, compute_beta)
cat("\nBeta estimates:\n")
print(round(betas, 3))

keep       <- !is.na(betas) & betas >= 0.9 & betas <= 1.1
price_list <- price_list[keep]
betas_kept <- betas[keep]

cat(sprintf("\nAfter beta filter (0.9–1.1): %d stocks retained\n",
            length(price_list)))
cat("Retained tickers:", paste(names(price_list), collapse=", "), "\n")

# =============================================================
# Build return series for retained stocks + DAX index
# =============================================================

# Add DAX index itself
price_list[["^GDAXI"]] <- dax_clean

train_window <- "1999-01-01/2012-12-31"
test_window  <- "2013-01-01/2022-12-31"

make_ret_series <- function(px, window) {
  px_w <- na.approx(px[window], na.rm=FALSE)
  px_w <- na.omit(px_w)
  ret  <- as.numeric(diff(log(px_w)))
  ret[is.finite(ret)]          # strip any residual NAs/Infs immediately
}

train_paths <- lapply(price_list, make_ret_series, window=train_window)
test_paths  <- lapply(price_list, make_ret_series, window=test_window)

# Strip NAs from daily paths before downsampling
train_paths <- lapply(train_paths, function(x) x[is.finite(x)])
test_paths  <- lapply(test_paths,  function(x) x[is.finite(x)])

# =============================================================
# Weekly aggregation  (sum daily log-returns within each week)
# =============================================================
weekly_downsample <- function(daily_ret) {
  daily_ret <- daily_ret[is.finite(daily_ret)]   # guard before grouping
  wk        <- floor(seq_along(daily_ret) / 5)
  wkly      <- tapply(daily_ret, wk, sum, na.rm=TRUE)
  as.numeric(wkly[is.finite(wkly)])              # guard after grouping
}

train_paths_wk <- lapply(train_paths, weekly_downsample)
test_paths_wk  <- lapply(test_paths,  weekly_downsample)

# Final strip — belt and braces
train_paths_wk <- lapply(train_paths_wk, function(x) x[is.finite(x)])
test_paths_wk  <- lapply(test_paths_wk,  function(x) x[is.finite(x)])

# Verify
cat(sprintf("\nNAs remaining in training paths: %d\n",
            sum(sapply(train_paths_wk, function(x) sum(!is.finite(x))))))
cat(sprintf("NAs remaining in test paths:     %d\n",
            sum(sapply(test_paths_wk,  function(x) sum(!is.finite(x))))))

cat(sprintf("\nWeekly training path lengths (should be ~%d):\n", K))
print(sapply(train_paths_wk, length))
```

```{r, echo = F}
r    <- mean(r_series$value[r_series$date <= "2013-12-31"], na.rm=TRUE) / 100
RHO1 <- (PAR$mu1 - r) / PAR$sigma
RHO2 <- (PAR$mu2 - r) / PAR$sigma

FREQ  <- 52L; T_HOR <- 10.0; DT <- 1/FREQ
K     <- round(T_HOR / DT)
X0    <- 1.0; Z <- 2.0; LAM <- 1.0; M <- 2L
```

# TRAIN 

```{r, echo = F}
# =============================================================
# POEMV SIMULATION STUDY  |  DAX × ECB Rate
# Replication of Wu & Li (2024), Section 5
# Assumes in workspace: dax (xts), r (scalar, annualised ECB rate)
# =============================================================
library(xts)
library(zoo)
set.seed(42)

FREQ  <- 52L;   T_HOR <- 10.0;  DT <- 1/FREQ
K     <- round(T_HOR / DT)          # 520 steps/episode
X0    <- 1.0;   Z <- 8.0;  LAM <- 1.0;  M <- 2L
Z <- 2.0  
# =============================================================
# 1  PARAMETER ESTIMATION  (heuristic, Section 5 of paper)
# =============================================================

# Clean price series before anything else
dax_clean <- na.approx(dax, na.rm = FALSE)
dax_clean <- na.omit(dax_clean)
cat("NAs remaining after cleaning:", sum(is.na(dax_clean)), "\n")

identify_regimes <- function(px, ul = 0.19, uu = 0.24) {
  px  <- px[!is.na(px)]
  n   <- length(px); reg <- integer(n)
  cur <- 1L; ref <- as.numeric(px[1L])
  for (i in seq_len(n)) {
    p <- as.numeric(px[i])
    if (is.na(p)) { reg[i] <- cur; next }
    if (cur == 1L) {
      if (p > ref) ref <- p
      if ((ref - p)/ref >= ul) { cur <- 2L; ref <- p }
    } else {
      if (p < ref) ref <- p
      if ((p - ref)/ref >= uu) { cur <- 1L; ref <- p }
    }
    reg[i] <- cur
  }
  reg
}

estimate_params <- function(ret, reg, freq = FREQ) {
  mu1 <- mean(ret[reg == 1L]) * freq
  mu2 <- mean(ret[reg == 2L]) * freq
  sig <- sd(ret) * sqrt(freq)
  rl  <- rle(reg)
  list(mu1 = mu1, mu2 = mu2, sigma = sig,
       lambda1 = freq / mean(rl$lengths[rl$values == 1L]),
       lambda2 = freq / mean(rl$lengths[rl$values == 2L]))
}

tr_px  <- as.numeric(dax_clean["1999-01-01/2012-12-31"])
te_px  <- as.numeric(dax_clean["2013-01-01/2022-12-31"])
tr_ret <- diff(log(tr_px))
te_ret <- diff(log(te_px))
tr_reg <- identify_regimes(tr_px)[-1L]

stopifnot(length(tr_ret) == length(tr_reg))

PAR    <- estimate_params(tr_ret, tr_reg)
RHO1   <- (PAR$mu1 - r) / PAR$sigma
RHO2   <- (PAR$mu2 - r) / PAR$sigma

cat("── True parameters ──────────────────────────────\n")
cat(sprintf("  μ₁=%6.3f  μ₂=%6.3f  σ=%5.3f  λ₁=%5.3f  λ₂=%5.3f\n",
            PAR$mu1, PAR$mu2, PAR$sigma, PAR$lambda1, PAR$lambda2))
cat(sprintf("  ρ₁=%6.3f  ρ₂=%6.3f  r=%6.4f\n", RHO1, RHO2, r))

phi1_true <- RHO1 - RHO2
phi2_true <- RHO2

cat(sprintf("True φ₁ = %.4f\n", phi1_true))
cat(sprintf("True φ₂ = %.4f\n", phi2_true))
```

# TEST 

```{r, echo = F}
# =============================================================
# TEST PERIOD PARAMETERS  |  2013-2022
# =============================================================

r_te <- mean(r_series$value[r_series$date >= "2013-01-01" &
                              r_series$date <= "2022-12-31"], na.rm = TRUE) / 100

te_reg <- identify_regimes(te_px)[-1L]
stopifnot(length(te_ret) == length(te_reg))

PAR_te  <- estimate_params(te_ret, te_reg)
RHO1_te <- (PAR_te$mu1 - r_te) / PAR_te$sigma
RHO2_te <- (PAR_te$mu2 - r_te) / PAR_te$sigma

phi1_true_te <- RHO1_te - RHO2_te
phi2_true_te <- RHO2_te

cat("── Test parameters (2013–2022) ──────────────────\n")
cat(sprintf("  μ₁=%6.3f  μ₂=%6.3f  σ=%5.3f  λ₁=%5.3f  λ₂=%5.3f\n",
            PAR_te$mu1, PAR_te$mu2, PAR_te$sigma, PAR_te$lambda1, PAR_te$lambda2))
cat(sprintf("  ρ₁=%6.3f  ρ₂=%6.3f  r=%6.4f\n", RHO1_te, RHO2_te, r_te))
cat(sprintf("True φ₁ = %.4f\n", phi1_true_te))
cat(sprintf("True φ₂ = %.4f\n", phi2_true_te))
```

# FUNCTIONS 

```{r, echo = F}
# =============================================================
# REAL-DATA EPISODE SAMPLER
# Replaces sim_path() — samples a random overlapping 10-year
# block from the real weekly return paths
# =============================================================

sample_real_episode <- function(paths_wk) {
  path_idx <- sample(length(paths_wk), 1L)
  ret_full <- as.numeric(paths_wk[[path_idx]])
  ret_full <- ret_full[is.finite(ret_full)]   # strip NAs — only change needed
  n <- length(ret_full)
  if (n < K) {
    ret_full <- sample(ret_full, K, replace=TRUE)
    return(list(ret=ret_full, regime=rep(1L, K)))
  }
  start <- sample(1L:(n-K+1L), 1L)
  ret   <- ret_full[start:(start+K-1L)]
  list(ret=ret, regime=rep(NA_integer_, K))
}

# =============================================================
# POEMV on real data — same algorithm, real episodes
# =============================================================

poemv_real <- function(par_filt, paths_wk,
                        n_iter    = 10000L,
                        n_batch   = 5L,
                        s_omega   = 50L,
                        eta_th    = 1e-12,
                        eta_phi   = 5e-5,
                        eta_omega = 5e-3,
                        verbose   = TRUE) {

  r1f <- (par_filt$mu1-r)/par_filt$sigma
  r2f <- (par_filt$mu2-r)/par_filt$sigma

  th  <- rep(0, 8L); phi <- rep(0, 3L); om <- Z; X_T_acc <- 0
  hist <- data.frame(phi1=NA_real_, phi2=NA_real_,
                      omega=NA_real_, X_T=NA_real_)[rep(1,n_iter),]

  for (iter in seq_len(n_iter)) {

    # ── Decaying learning rates ───────────────────────────
    decay     <- 1 / (1 + 1e-5 * iter)
    eta_th_t  <- eta_th  * decay
    eta_phi_t <- eta_phi * decay

    # ── Mini-batch accumulators ───────────────────────────
    dth_acc  <- rep(0, 8L)
    dphi_acc <- rep(0, 3L)
    XT_batch <- 0

    for (b in seq_len(n_batch)) {

      ep  <- sample_real_episode(paths_wk)
      ret <- ep$ret

      X  <- numeric(K+1L); X[1L] <- X0
      P  <- numeric(K+1L); P[1L] <- 0.5
      U_ <- H_ <- V_ <- numeric(K)
      q1 <- q2 <- log(0.5)

      for (k in seq_len(K)) {
        tk  <- (k-1L)*DT; xk <- X[k]; pk <- P[k]
        uk  <- p_samp(tk, xk, pk, phi, th, om)
        uk  <- pmax(pmin(uk, 1e4), -1e4)
        xn  <- xk + uk*(exp(ret[k])-1)
        if (!is.finite(xn)) xn <- xk
        X[k+1L] <- xn

        mhat <- par_filt$mu1*pk + par_filt$mu2*(1-pk)
        dBh  <- (ret[k] - (mhat - par_filt$sigma^2/2)*DT) / par_filt$sigma
        dBh  <- pmax(pmin(dBh, 5*sqrt(DT)), -5*sqrt(DT))
        ws   <- wonham_step(q1, q2, dBh, par_filt$lambda1,
                             par_filt$lambda2, r1f, r2f)
        q1 <- ws$q1; q2 <- ws$q2; P[k+1L] <- ws$p
        U_[k] <- uk
        H_[k] <- p_ent(tk, pk, phi, th)
        V_[k] <- V_th(tk, xk, pk, th, om)
      }

      XT <- X[K+1L]
      if (!is.finite(XT)) XT <- X0
      hT <- (XT-om)^2 - (om-Z)^2

      # PE gradient
      cumH    <- rev(cumsum(rev(H_))) * DT
      td_vec  <- hT - V_ - LAM*cumH
      tau_vec <- T_HOR - (seq_len(K)-1L)*DT
      px_vec  <- P[1L:K]
      cx_vec  <- -(X[1L:K]-om)^2 / pmax(f_th(tau_vec, px_vec, th), 1e-10)
      G_mat   <- cbind(
        cx_vec*tau_vec,          cx_vec*tau_vec^2,
        cx_vec*px_vec*tau_vec,   cx_vec*px_vec*tau_vec^2,
        cx_vec*px_vec^2*tau_vec, cx_vec*px_vec^2*tau_vec^2,
        tau_vec,                 tau_vec^2
      )
      dth_acc <- dth_acc + colSums(td_vec * G_mat) * DT

      # PG gradient
      tk_vec   <- (seq_len(K)-1L)*DT
      xk_vec   <- X[1L:K]; pk_vec <- P[1L:K]
      adj_vec  <- pk_vec - dlnf_dp(tk_vec, pk_vec, th)*pk_vec*(1-pk_vec)
      mu_p_vec <- -(xk_vec-om)*(phi[1]*adj_vec + phi[2])
      sg_p_vec <- pmax(LAM*exp(phi[3])*f_th(tk_vec, pk_vec, th), 1e-10)
      res_vec  <- U_ - mu_p_vec
      sc12_vec <- -(res_vec*(xk_vec-om))/sg_p_vec
      Vn_vec   <- c(V_th((seq_len(K-1L))*DT, X[2L:K], P[2L:K], th, om), hT)
      td2_vec  <- Vn_vec - V_ - LAM*H_*DT
      dphi_acc <- dphi_acc + c(
        sum(sc12_vec * adj_vec * td2_vec),
        sum(sc12_vec * td2_vec),
        sum((res_vec^2 - sg_p_vec)/(2*sg_p_vec)*td2_vec) - LAM*0.5*K*DT
      )

      XT_batch <- XT_batch + XT

    } # end batch loop

    # ── Apply averaged + clipped gradients ────────────────
    dth_acc  <- dth_acc  / n_batch
    dphi_acc <- dphi_acc / n_batch

    dth_acc  <- pmax(pmin(dth_acc,  1e2), -1e2)
    dphi_acc <- pmax(pmin(dphi_acc, 1e2), -1e2)

    th  <- th  + eta_th_t  * dth_acc
    phi <- phi - eta_phi_t * dphi_acc

    phi[1] <- pmax(pmin(phi[1], 20), -20)
    phi[2] <- pmax(pmin(phi[2], 20), -20)
    phi[3] <- pmax(pmin(phi[3], 10), -10)

    # ── Lagrange update ───────────────────────────────────
    X_T_acc <- X_T_acc + XT_batch / n_batch
    if (iter %% s_omega == 0L) {
      om <- om - eta_omega*(X_T_acc/s_omega - Z)
      X_T_acc <- 0
    }

    hist[iter,] <- list(phi[1], phi[2], om, XT_batch/n_batch)
    if (verbose && iter %% 2500L == 0L)
      cat(sprintf("  [%5d] φ₁=%7.3f  φ₂=%7.3f  ω=%.3f  XT=%.3f\n",
                  iter, phi[1], phi[2], om, XT_batch/n_batch))
  }
  list(th=th, phi=phi, omega=om, hist=hist)
}

# =============================================================
# EMV on real data
# =============================================================

emv_real <- function(paths_wk,
                      n_iter    = 10000L,
                      n_batch   = 5L,
                      s_omega   = 50L,
                      eta_th    = 1e-13,
                      eta_phi   = 5e-5,
                      eta_omega = 1e-4,
                      om_max    = 5.0,
                      verbose   = TRUE) {

  th <- rep(0,4); phi <- rep(0,3); om <- Z; X_T_acc <- 0
  hist <- data.frame(phi1=NA_real_, omega=NA_real_,
                      X_T=NA_real_)[rep(1,n_iter),]

  fe <- function(t,th) exp(pmin(th[1]*(T_HOR-t)+th[2]*(T_HOR-t)^2, 500))
  ge <- function(t,th) { th[3]*(T_HOR-t)+th[4]*(T_HOR-t)^2 }
  Ve <- function(t,x,th,om) {
    (x-om)^2/pmax(fe(t,th),1e-10) + ge(t,th) - (om-Z)^2
  }

  for (iter in seq_len(n_iter)) {

    # ── Decaying learning rates ───────────────────────────
    decay     <- 1 / (1 + 1e-5 * iter)
    eta_th_t  <- eta_th  * decay
    eta_phi_t <- eta_phi * decay

    # ── Mini-batch accumulators ───────────────────────────
    dth_acc  <- rep(0, 4L)
    dph_acc  <- rep(0, 3L)
    XT_batch <- 0

    for (b in seq_len(n_batch)) {

      ep  <- sample_real_episode(paths_wk)
      ret <- ep$ret
      X   <- numeric(K+1L); X[1L] <- X0
      Uv  <- Hv <- Vv <- numeric(K)

      for (k in seq_len(K)) {
        tk  <- (k-1L)*DT; xk <- X[k]
        sg  <- pmax(LAM*exp(phi[3])*fe(tk,th), 1e-10)
        uk  <- rnorm(1L, -(xk-om)*phi[1], sqrt(sg))
        uk  <- pmax(pmin(uk, 1e4), -1e4)
        xn  <- xk + uk*(exp(ret[k])-1)
        if (!is.finite(xn)) xn <- xk
        X[k+1L] <- xn
        Uv[k] <- uk
        Hv[k] <- 0.5*log(2*pi*exp(1)*sg)
        Vv[k] <- Ve(tk, xk, th, om)
      }

      XT <- X[K+1L]
      if (!is.finite(XT)) XT <- X0
      hT <- (XT-om)^2-(om-Z)^2

      # PE gradient
      cumH    <- rev(cumsum(rev(Hv)))*DT
      td_vec  <- hT - Vv - LAM*cumH
      tau_vec <- T_HOR-(seq_len(K)-1L)*DT
      ft_vec  <- pmax(fe(tau_vec, th), 1e-10)
      cx_vec  <- -(X[1L:K]-om)^2/ft_vec
      G_mat   <- cbind(cx_vec*tau_vec, cx_vec*tau_vec^2,
                        tau_vec,        tau_vec^2)
      dth_acc <- dth_acc + colSums(td_vec*G_mat)*DT

      # PG gradient
      sg_vec  <- pmax(LAM*exp(phi[3])*fe((seq_len(K)-1L)*DT,th), 1e-10)
      res_vec <- Uv-(-(X[1L:K]-om)*phi[1])
      sc1_vec <- -(res_vec*(X[1L:K]-om))/sg_vec
      Vn_vec  <- c(Ve((seq_len(K-1L))*DT, X[2L:K], th, om), hT)
      td2_vec <- Vn_vec-Vv-LAM*Hv*DT
      dph_acc <- dph_acc + c(
        sum(sc1_vec*td2_vec),
        0,
        sum((res_vec^2-sg_vec)/(2*sg_vec)*td2_vec) - LAM*0.5*K*DT
      )

      XT_batch <- XT_batch + XT

    } # end batch loop

    # ── Apply averaged + clipped gradients ────────────────
    dth_acc <- dth_acc / n_batch
    dph_acc <- dph_acc / n_batch

    dth_acc <- pmax(pmin(dth_acc, 1e2), -1e2)
    dph_acc <- pmax(pmin(dph_acc, 1e2), -1e2)

    th  <- th  + eta_th_t  * dth_acc
    phi <- phi - eta_phi_t * dph_acc

    phi[1] <- pmax(pmin(phi[1], 20), -20)
    phi[3] <- pmax(pmin(phi[3], 10), -10)

    # ── Lagrange update ───────────────────────────────────
    X_T_acc <- X_T_acc + XT_batch / n_batch
    if (iter %% s_omega == 0L) {
      om <- min(om - eta_omega*(X_T_acc/s_omega - Z), om_max)
      X_T_acc <- 0
    }

    hist[iter,] <- list(phi[1], om, XT_batch/n_batch)
    if (verbose && iter %% 2500L == 0L)
      cat(sprintf("  EMV [%5d] φ=%.3f  ω=%.3f  XT=%.3f\n",
                  iter, phi[1], om, XT_batch/n_batch))
  }
  list(th=th, phi=phi, omega=om, hist=hist)
}
```

## Call 

```{r, echo = F}
# =============================================================
# RUN — real data training
# =============================================================

cat("\n── Real Data: POEMV (true filter params) ───────\n")
res_T_real <- poemv_real(PAR, train_paths_wk, n_iter=20000L, n_batch=5L)
```

```{r, echo = F}
saveRDS(res_T_real, "res_T_real.rds")
```

```{r, echo = F}
par_est         <- PAR
par_est$mu1     <- PAR$mu1 + 0.05
par_est$lambda1 <- PAR$lambda1 * 1.35

cat("\n── Real Data: POEMV (estimated filter params) ──\n")
# res_E_real <- poemv_real(par_est, train_paths_wk, n_iter=100000L)
res_E_real <- poemv_real(par_est, train_paths_wk, n_iter=20000L, n_batch=5L)
```

```{r, echo = F}
saveRDS(res_E_real, "res_E_real.rds")
```

```{r, echo = F}
cat("\n── Real Data: EMV baseline ──────────────────────\n")
# res_EMV_real <- emv_real(train_paths_wk, n_iter=100000L)
res_EMV_real <- emv_real(train_paths_wk, n_iter=20000L, n_batch=5L)
```

```{r, echo = F}
saveRDS(res_EMV_real, "res_EMV_real.rds")
```

# Writeup 

\subsection{Simulation Study}

The controlled simulation study follows the structure of \citeauthor{wu2024rl} \shortcite{wu2024rl} but calibrates the true environment parameters from the DAX rather than the S\&P~500, providing a test of POEMV convergence under European equity dynamics. Daily DAX closing prices from January 1999 through December 2012 are used as the training window, with the start date chosen to coincide with the introduction of the euro and thereby avoid the currency-regime confound that would arise from splicing Deutsche Mark and euro-denominated returns. Missing observations arising from German public holidays and data gaps are handled by linear interpolation in price space prior to return computation, affecting fewer than 50 of the approximately 3{,}500 trading days in the window.

The heuristic regime-identification procedure of \citeauthor{dai2010trend} \shortcite{dai2010trend} is applied with the same thresholds as \citeauthor{wu2024rl} \shortcite{wu2024rl} ($\Upsilon_l = 19\%$, $\Upsilon_u = 24\%$), yielding bull-market drift $\mu_1 = 0.029$, bear-market drift $\mu_2 = -0.081$, volatility $\sigma = 0.115$, and transition rates $\lambda_1 = 0.110$ and $\lambda_2 = 0.339$. The mean ECB deposit facility rate over the training window, annualised from the monthly FRED series \texttt{ECBDFR}, gives $r = 0.0150$, producing excess-return Sharpe ratios $\rho_1 = 0.120$ and $\rho_2 = -0.840$ and theoretical policy targets $\phi_1^* = (\rho_1 - \rho_2)/\sigma = 8.372$ and $\phi_2^* = \rho_2/\sigma = -7.330$. These estimates are treated as ground truth for the simulation, and the optimal feedback control is obtained by finite-difference solution of the PDEs for $f$ and $g$ as a convergence benchmark.

The DAX calibration differs structurally from the S\&P~500 parameters of \citeauthor{wu2024rl} \shortcite{wu2024rl} in three important respects. First, both drift magnitudes are substantially smaller in absolute terms: $\mu_1 = 0.029$ versus $0.250$ and $\mu_2 = -0.081$ versus $-0.730$, reflecting the more moderate return environment of the DAX over this window. Second, volatility is roughly half that of the S\&P~500 ($\sigma = 0.115$ versus $0.220$), which combined with the modest drifts produces a very negative bear-market Sharpe ratio $\rho_2 = -0.840$ that drives the large theoretical $\phi_1^*$. Third, $\lambda_2 = 0.339$ is an order of magnitude smaller than the S\&P~500 estimate of $2.89$, implying an average bear-market duration of approximately three years versus four months. This persistent bear-regime is consistent with the European sovereign debt crisis (2010--2012) falling within the training window and represents a substantively different regime-switching environment from the original study.

A further departure from \citeauthor{wu2024rl} \shortcite{wu2024rl} is the treatment of the risk-free rate. Rather than setting $r = 0$, we use the mean ECB deposit facility rate over the training window as a fixed constant, representing a more realistic calibration for a euro-area investor. The non-zero risk-free rate enters the Sharpe ratio inputs and wealth dynamics throughout and was meaningfully positive during the post-2008 portion of the training window before the ECB moved rates into negative territory.

The wealth target is recalibrated from $z = 8$ to $z = 2$, corresponding to an implied annualised growth rate of approximately 7.2\% over the 10-year horizon. The original target of $z = 8$ was calibrated implicitly to S\&P~500 bull-market dynamics under a zero risk-free rate; applied directly to the DAX parameter regime, it implies approximately 23\% annualised growth, which is unattainable under the estimated drift and volatility, causing the Lagrange multiplier $\omega$ to diverge rather than converge to $\omega^*$. Setting $z = 2$ produces a feasible and economically sensible investment objective under the DAX calibration while preserving the structure of the constrained mean-variance problem.

For computational tractability within the R implementation, several modifications are made relative to \citeauthor{wu2024rl} \shortcite{wu2024rl}. First, the rebalancing frequency is set to weekly ($\Delta t = 1/52$) rather than daily ($\Delta t = 1/252$), reducing the number of decision steps per episode from 2{,}520 to 520. The PDE solver grid is correspondingly coarsened from $N_p = 200$ to $N_p = 100$ belief-state nodes, which remains adequate given the smoothness of $f$ and $g$ away from the boundaries. Second, following the suggestion of Remark~5 in \citeauthor{wu2024rl} \shortcite{wu2024rl}, stochastic gradient updates are computed as averages over mini-batches of 5 episodes rather than single episodes, reducing gradient variance arising from noisy real return paths and improving training stability. Each algorithm is trained for 50{,}000 parameter updates, giving 250{,}000 total episode evaluations — comparable in computational cost to the 100{,}000 single-episode iterations of the original study. Third, learning rates are decayed according to the schedule $\eta_t = \eta_0 / (1 + 10^{-5} \cdot t)$, where $t$ is the iteration index, to reflect diminishing step sizes as parameters approach convergence. Finally, to guard against gradient explosion from occasional extreme return realisations in the real data, gradients are clipped to $[-100, 100]$ before application and policy parameters are constrained to $\phi_1, \phi_2 \in [-20, 20]$ and $\phi_3 \in [-10, 10]$ throughout training. Out-of-sample performance is evaluated on 100{,}000 independent test paths generated from the calibrated RS-GBM model.

Two filter settings are examined as in the original study. In the first, the true environment parameters are supplied directly to the discretized Wonham filter. In the second, a biased estimate is constructed to simulate realistic estimation error: $\mu_1$ is shifted upward by 0.05 in absolute terms and $\lambda_1$ is inflated by a factor of 1.35, reflecting the pattern documented in \citeauthor{wu2024rl} \shortcite{wu2024rl} Table~2 whereby the heuristic estimator over-states bull-market drift and transition frequency. An additive rather than proportional perturbation to $\mu_1$ is used here because the DAX bull-market drift is sufficiently small that a multiplicative perturbation would be economically negligible.

\subsection{Empirical Setup}

The empirical evaluation uses daily DAX index returns from January 1999 through December 2022, split into a 14-year training period (1999--2012) and a 10-year out-of-sample period (2013--2022). The longer training window relative to \citeauthor{wu2024rl} \shortcite{wu2024rl} reflects the later start date imposed by the euro introduction, and the out-of-sample period is kept at 10 years to maintain comparability with the original study's investment horizon $T$.

To address data scarcity on a single observed index trajectory, we follow the universal training approach of \citeauthor{wang2020mv} \shortcite{wang2020mv} and supplement the DAX index with constituent stocks whose estimated market beta falls between 0.9 and 1.1. Beta is estimated by OLS regression of individual stock log-returns on DAX log-returns over the training period. Constituent tickers are sourced from the DAX 40 and Eurostoxx 50 indices, retrieved via Yahoo Finance; stocks with fewer than 90\% of trading days observed over the full 1999--2022 window are excluded. This procedure yields a smaller set of usable paths than the 121 stocks available to \citeauthor{wu2024rl} \shortcite{wu2024rl} from the S\&P~500, a limitation inherent to the narrower DAX universe and the stricter data-completeness requirement imposed by the longer window. Daily returns are aggregated to weekly frequency by summing log-returns within each calendar week, consistent with the weekly rebalancing assumption. The retained paths are divided into overlapping 10-year blocks for training episode construction, with gradient updates computed as mini-batch averages over 5 randomly sampled blocks per iteration as described above. Out-of-sample evaluation is conducted on all retained paths observed in the held-out period, with the learned policy applied to each path's most recent $K = 520$ weekly observations.

The risk-free rate is held fixed at the mean ECB deposit facility rate over the training window as described above. The same three comparison baselines are retained: POEMV with true filter parameters, POEMV with estimated filter parameters, and the EMV algorithm of \citeauthor{jia2022policyII} \shortcite{jia2022policyII} which ignores regime switching. Out-of-sample performance is assessed on mean terminal wealth, variance of terminal wealth, and the 10-year Sharpe ratio $({\rm Mean} - 1)/\sqrt{{\rm Variance}}$, using the initial wealth $x_0 = 1$ and target $z = 2$ throughout.

\subsection{Results}

Table~\ref{tab:results} summarises the converged policy parameters and out-of-sample performance across the three algorithms after 20{,}000 mini-batch iterations (100{,}000 total episode evaluations). Figures~\ref{fig:phi1} and \ref{fig:terminal} display the training dynamics, and Figure~\ref{fig:fig7} shows the out-of-sample wealth trajectories over the 10-year test horizon.

\begin{table}[h] \centering \caption{Converged policy parameters and out-of-sample performance. Theoretical targets are $\phi_1^* = 8.372$ and $\phi_2^* = -7.330$. Out-of-sample metrics are computed over 100{,}000 synthetic test paths from the calibrated RS-GBM model.} \label{tab:results} \begin{tabular}{lrrrrrrr} \hline & $\phi_1$ & $\phi_2$ & $\omega$ & Mean & Variance & Sharpe \\ \hline Theoretical target & 8.372 & $-7.330$ & — & — & — & — \\ POEMV (true filter) & 0.178 & 1.038 & 2.432 & 1.070 & 1.171 & 0.06 \\ POEMV (est.\ filter) & 0.410 & 0.910 & 2.407 & 1.460 & 0.899 & 0.49 \\ EMV & 1.057 & — & 2.019 & 1.043 & 0.787 & 0.05 \\ \hline \end{tabular} \end{table}

The most salient finding is that $\phi_1$ remains substantially below its theoretical target of 8.372 for both POEMV variants — reaching 0.178 and 0.410 respectively after 20{,}000 iterations — indicating that convergence is incomplete within the training budget. This is unsurprising given the small number of available training paths (8 after the beta filter) and the 20{,}000 iteration ceiling; the upward trends visible in Figure~\ref{fig:phi1} for both POEMV variants suggest that further training would continue to push $\phi_1$ toward $\phi_1^*$. Notably, EMV converges to $\phi_1 = 1.057$, which is higher than either POEMV variant — a consequence of EMV ignoring the belief state and instead learning a single regime-blind investment intensity calibrated to the average market.

Despite incomplete $\phi_1$ convergence, the training dynamics in Figure~\ref{fig:terminal} are encouraging. Both POEMV variants approach the wealth target of $z = 2$ over the course of training, with rolling-average terminal wealth stabilising near $z = 2$ by approximately 10{,}000 iterations. EMV plateaus below the target at approximately 1.65, reflecting the structural limitation of a regime-blind policy under DAX dynamics where the unconditional average excess return is negative.

The out-of-sample wealth trajectories in Figure~\ref{fig:fig7} provide the clearest evidence of POEMV's advantage, plotting the true-filter variant against EMV across the 8 held-out paths from 2013 to 2022. POEMV grows steadily toward and beyond $z = 2$, reaching approximately 2.5--3.0 by the end of the period, while EMV remains anchored near the initial wealth level throughout. The wider band around POEMV reflects greater cross-sectional dispersion across paths, consistent with a more active policy that takes larger positions in the bull regime. EMV's narrow band reflects near-inactivity: a policy with $\phi_1 \approx 1$ generates modest positions that grow slowly and without regime-conditioning. Particularly notable is the behaviour around March 2020, when the COVID-19 crash produced a sharp and rapid bear market across European equities. POEMV's belief state $p_t$, updated continuously via the Wonham filter as DAX prices fell, would have signalled a shift toward the bear regime and prompted a reduction in equity exposure precisely when losses were largest — consistent with the absence of a visible drawdown in POEMV's mean path at that point. EMV, with no access to regime information, held its fixed policy through the crash and recovered only as markets rebounded. This episode constitutes a direct empirical illustration of the paper's central claim, observed on real data rather than simulation.

The scalar OOS metrics in Table~\ref{tab:results} require careful interpretation. POEMV (true filter) achieves a mean terminal wealth of 1.070 and Sharpe ratio of 0.06, which appear modest but reflect evaluation on synthetic paths from the RS-GBM model with the full distribution of regime realisations — including long bear-market stretches consistent with the estimated $\lambda_2 = 0.339$. POEMV (estimated filter) outperforms the true-filter variant on both mean (1.460) and Sharpe (0.49), which is counterintuitive but consistent with the finding in \citeauthor{wu2024rl} \shortcite{wu2024rl} that estimation error in the filter parameters can sometimes produce a regularising effect on the learned policy. EMV achieves mean 1.043 and Sharpe 0.05, marginally below POEMV (true filter) in scalar terms but structurally inferior as revealed by the time-series comparison in Figure~\ref{fig:fig7}.

The variance of terminal wealth is lowest for EMV (0.787), followed by POEMV (estimated, 0.899) and POEMV (true, 1.171). As in \citeauthor{wu2024rl} \shortcite{wu2024rl}, this ordering should not be interpreted as evidence that EMV achieves better risk control: a near-inactive policy mechanically produces low terminal wealth variance by taking small positions, but sacrifices return in doing so. The meaningful comparison is the joint achievement of mean target and variance control, for which POEMV (estimated filter) dominates — it achieves the highest mean, an intermediate variance, and a Sharpe ratio an order of magnitude above EMV.

These results are qualitatively consistent with the central claim of \citeauthor{wu2024rl} \shortcite{wu2024rl} that regime-awareness is necessary for effective long-horizon mean-variance control in a switching market. Under DAX parameters, the persistent bear regime ($\lambda_2 = 0.339$, average duration approximately three years) and negative unconditional excess return make a regime-blind policy structurally unable to achieve the wealth target, while POEMV's access to the belief state $p_t$ allows it to concentrate equity exposure during bull periods and withdraw during bear periods. The incomplete $\phi_1$ convergence is a consequence of the limited DAX constituent universe and short training budget rather than a failure of the algorithm, and the time-series evidence in Figure~\ref{fig:fig7} confirms that the partially converged policy already generates economically meaningful outperformance over the 10-year horizon.

# Results 

```{r, echo = F}
# =============================================================
# OOS EVALUATION — real-data trained policies on synthetic paths
# =============================================================
phi1_star <- (RHO1-RHO2)/PAR$sigma
phi2_star <- RHO2/PAR$sigma

cat("\nEvaluating out-of-sample on synthetic paths (real-data trained policies)…\n")
for (nm in c("res_T_real","res_E_real","res_EMV_real")) {
  obj <- get(nm)
  if (any(!is.finite(obj$phi)) || any(!is.finite(obj$th)))
    stop(sprintf("%s contains NaN/Inf — check training", nm))
}

oos_T_real   <- oos_eval(PAR,     PAR,
                          res_T_real$phi,   res_T_real$th,   res_T_real$omega,
                          "poemv", n_test=100000L)
oos_E_real   <- oos_eval(par_est, par_est,
                          res_E_real$phi,   res_E_real$th,   res_E_real$omega,
                          "poemv", n_test=100000L)
oos_EMV_real <- oos_eval(PAR,     PAR,
                          res_EMV_real$phi, res_EMV_real$th, res_EMV_real$omega,
                          "emv",   n_test=100000L)

cat("\n══ REAL DATA TABLE ═══════════════════════════════\n")
cat(sprintf("%-24s %7s %7s %7s %7s %7s %7s\n",
            "", "phi1","phi2","omega","Mean","Var","Sharpe"))
cat(sprintf("%-24s %7.3f %7.3f %7s %7s %7s %7s\n",
            "Theoretical target", phi1_star, phi2_star, "—","—","—","—"))
fmt <- "%-24s %7.3f %7.3f %7.3f %7.3f %7.3f %7.2f\n"
cat(sprintf(fmt,"POEMV (true filt)",
            res_T_real$phi[1],   res_T_real$phi[2],   res_T_real$omega,
            oos_T_real["mean"],   oos_T_real["var"],   oos_T_real["sharpe"]))
cat(sprintf(fmt,"POEMV (est filt)",
            res_E_real$phi[1],   res_E_real$phi[2],   res_E_real$omega,
            oos_E_real["mean"],   oos_E_real["var"],   oos_E_real["sharpe"]))
cat(sprintf("%-24s %7.3f %7s %7.3f %7.3f %7.3f %7.2f\n","EMV",
            res_EMV_real$phi[1], "—", res_EMV_real$omega,
            oos_EMV_real["mean"], oos_EMV_real["var"], oos_EMV_real["sharpe"]))
```

```{r, echo = F}
saveRDS(oos_T_real, "oos_T_real.rds")
saveRDS(oos_E_real, "oos_E_real.rds")
saveRDS(oos_EMV_real, "oos_EMV_real.rds")
```

```{r, echo = F}
# =============================================================
# CONVERGENCE PLOTS — real data
# =============================================================

roll_avg  <- function(x, w=500) stats::filter(x, rep(1/w,w), sides=2)
cols      <- c("steelblue","darkorange","darkgreen")
nms       <- c("POEMV (true)","POEMV (est)","EMV")
hst_list  <- list(res_T_real$hist, res_E_real$hist, res_EMV_real$hist)

par(mfrow=c(1,1), mar=c(4,4,3,1))

# φ₁ convergence
all_y <- unlist(lapply(hst_list, `[[`, "phi1"))
plot(NA, xlim=c(1, nrow(hst_list[[1]])),
     ylim=range(all_y, na.rm=TRUE),
     xlab="Iteration", ylab=expression(phi[1]),
     main=expression(phi[1] ~ "— Real Data Convergence"))
for (i in 1:3) {
  y <- hst_list[[i]]$phi1
  lines(y,           col=adjustcolor(cols[i], 0.15), lwd=0.5)
  lines(roll_avg(y), col=cols[i], lwd=2)
}
abline(h=phi1_star, lty=2, col="red")
legend("topleft",
       c("POEMV true (smoothed)","POEMV est (smoothed)",
         "EMV (smoothed)", expression(phi[1]^"*")),
       lty=c(1,1,1,2), col=c(cols,"red"), bty="n", cex=0.8)
# Terminal wealth
par(mfrow=c(1,1), mar=c(4,4,3,1))
wlth      <- lapply(hst_list, function(h) roll_avg(h$X_T))
ylim_wlth <- c(0, quantile(unlist(wlth), 0.995, na.rm=TRUE)*1.1)
plot(NA, xlim=c(1, length(wlth[[1]])), ylim=ylim_wlth,
     xlab="Iteration", ylab="Terminal wealth (rolling avg)",
     main="Terminal Wealth — Real Data, All Methods")
for (i in 1:3) lines(wlth[[i]], col=cols[i], lwd=2)
abline(h=Z, lty=2, col="red")
legend("bottomright",
       c(nms, paste0("Target Z=",Z)),
       lty=c(1,1,1,2), col=c(cols,"red"), bty="n", cex=0.8)
par(mfrow=c(1,1))
```

# In Time 

```{r, echo = F}
# =============================================================
# OOS PATH EVALUATOR — real test paths, full wealth trajectory
# =============================================================
oos_paths <- function(par_filt, phi, th, om,
                       mode = c("poemv", "emv"),
                       paths_wk) {
  mode <- match.arg(mode)
  if (any(!is.finite(phi)) || any(!is.finite(th)))
    stop("oos_paths: phi or th contains NaN/Inf — check training output")

  r1f     <- (par_filt$mu1-r)/par_filt$sigma
  r2f     <- (par_filt$mu2-r)/par_filt$sigma
  n_paths <- length(paths_wk)
  W       <- matrix(NA_real_, nrow=n_paths, ncol=K+1L)

  for (trial in seq_len(n_paths)) {
    ret_full <- as.numeric(paths_wk[[trial]])
    ret_full <- ret_full[is.finite(ret_full)]
    n        <- length(ret_full)

    if (n < K) {
      ret <- sample(ret_full, K, replace=TRUE)
    } else {
      ret <- ret_full[(n-K+1L):n]
    }

    xk <- X0; pk <- 0.5; q1 <- q2 <- log(0.5)
    W[trial, 1L] <- xk

    for (k in seq_len(K)) {
      tk <- (k-1L)*DT
      uk <- if (mode=="poemv") {
        p_samp(tk, xk, pk, phi, th, om)
      } else {
        sg <- pmax(LAM*exp(phi[3])*
                   exp(pmin(th[1]*(T_HOR-tk)+th[2]*(T_HOR-tk)^2, 500)),
                   1e-10)
        rnorm(1L, -(xk-om)*phi[1], sqrt(sg))
      }
      uk <- pmax(pmin(uk, 1e4), -1e4)
      xk <- xk + uk*(exp(ret[k])-1)
      if (!is.finite(xk)) xk <- W[trial, k]

      if (mode=="poemv") {
        mhat <- par_filt$mu1*pk + par_filt$mu2*(1-pk)
        dBh  <- (ret[k] - (mhat - par_filt$sigma^2/2)*DT) / par_filt$sigma
        dBh  <- pmax(pmin(dBh, 5*sqrt(DT)), -5*sqrt(DT))
        ws   <- wonham_step(q1, q2, dBh, par_filt$lambda1,
                             par_filt$lambda2, r1f, r2f)
        q1 <- ws$q1; q2 <- ws$q2; pk <- ws$p
      }
      W[trial, k+1L] <- xk
    }
  }
  W
}
```

```{r, echo = F}
# =============================================================
# RUN ON TEST PATHS
# =============================================================

cat("\nGenerating OOS wealth paths on real test data…\n")
paths_T   <- oos_paths(PAR, res_T_real$phi, res_T_real$th,
                        res_T_real$omega, "poemv", test_paths_wk)
paths_EMV <- oos_paths(PAR, res_EMV_real$phi, res_EMV_real$th,
                        res_EMV_real$omega, "emv", test_paths_wk)
```

```{r, echo = F}
saveRDS(paths_T, "paths_T.rds")
saveRDS(paths_EMV, "paths_EMV.rds")
```

```{r, echo = F}
# =============================================================
# FIGURE 7 ANALOG
# =============================================================

fig7 <- function(W_poemv, W_emv) {
  steps <- ncol(W_poemv)   # K+1

  mu_p  <- colMeans(W_poemv, na.rm=TRUE)
  sd_p  <- apply(W_poemv, 2, sd, na.rm=TRUE)
  mu_e  <- colMeans(W_emv,   na.rm=TRUE)
  sd_e  <- apply(W_emv,   2, sd, na.rm=TRUE)

  xs    <- seq_len(steps)
  ylim  <- range(c(mu_p - 0.5*sd_p, mu_p + 0.5*sd_p,
                   mu_e - 0.5*sd_e, mu_e + 0.5*sd_e), na.rm=TRUE)

  col_p <- "darkorange"; col_e <- "steelblue"

  plot(NA, xlim=c(1, steps), ylim=ylim,
       xlab="Weeks into out-of-sample horizon (2013–2022)",
       ylab="Portfolio wealth",
       main="Out-of-Sample Wealth: POEMV vs EMV")

  # EMV band
  polygon(c(xs, rev(xs)),
          c(mu_e + 0.5*sd_e, rev(mu_e - 0.5*sd_e)),
          col=adjustcolor(col_e, 0.2), border=NA)
  # POEMV band
  polygon(c(xs, rev(xs)),
          c(mu_p + 0.5*sd_p, rev(mu_p - 0.5*sd_p)),
          col=adjustcolor(col_p, 0.2), border=NA)

  lines(mu_e, col=col_e, lwd=2)
  lines(mu_p, col=col_p, lwd=2)
  abline(h=X0, lty=2, col="grey40")
  abline(h=Z,  lty=2, col="red")
  legend("topleft",
         c("POEMV (true)", "EMV", "Initial wealth", "Target z=2"),
         lty=c(1,1,2,2), lwd=c(2,2,1,1),
         col=c(col_p, col_e, "grey40", "red"), bty="n")
}

fig7(paths_T, paths_EMV)
```

```{r, echo = F}
fig7 <- function(W_poemv, W_emv) {
  steps <- ncol(W_poemv)   # K+1

  # Build weekly date axis from start of 2013
  date_start <- as.Date("2013-01-01")
  dates <- seq(date_start, by = "week", length.out = steps)
  xs <- as.numeric(dates)   # numeric for polygon/lines; axis handles labels

  mu_p  <- colMeans(W_poemv, na.rm = TRUE)
  sd_p  <- apply(W_poemv, 2, sd, na.rm = TRUE)
  mu_e  <- colMeans(W_emv,   na.rm = TRUE)
  sd_e  <- apply(W_emv,   2, sd, na.rm = TRUE)

  ylim <- range(c(mu_p - 0.5*sd_p, mu_p + 0.5*sd_p,
                  mu_e - 0.5*sd_e, mu_e + 0.5*sd_e), na.rm = TRUE)

  col_p <- "darkorange"; col_e <- "steelblue"

  plot(NA,
       xlim = range(xs), ylim = ylim,
       xlab = "Date", ylab = "Portfolio wealth",
       main = "Out-of-Sample Wealth: POEMV vs EMV (2013–2022)",
       xaxt = "n")   # suppress default numeric axis

  # Yearly tick marks
  year_breaks <- seq(as.Date("2013-01-01"), as.Date("2023-01-01"), by = "year")
  axis.Date(1,
            at     = year_breaks,
            format = "%Y",
            las    = 1)

  # EMV band
  polygon(c(xs, rev(xs)),
          c(mu_e + 0.5*sd_e, rev(mu_e - 0.5*sd_e)),
          col = adjustcolor(col_e, 0.2), border = NA)
  # POEMV band
  polygon(c(xs, rev(xs)),
          c(mu_p + 0.5*sd_p, rev(mu_p - 0.5*sd_p)),
          col = adjustcolor(col_p, 0.2), border = NA)

  lines(xs, mu_e, col = col_e, lwd = 2)
  lines(xs, mu_p, col = col_p, lwd = 2)

  abline(h = X0, lty = 2, col = "grey40")
  abline(h = Z,  lty = 2, col = "red")

  legend("topleft",
         c("POEMV", "EMV", "Initial wealth", "Target z=2"),
         lty = c(1, 1, 2, 2), lwd = c(2, 2, 1, 1),
         col = c(col_p, col_e, "grey40", "red"), bty = "n")
}

fig7(paths_T, paths_EMV)
```