---
title: "Attempt1"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r, echo = F, warning = F, message = F}
library(curl)
library(quantmod)
library(fredr)
library(xts)
library(dplyr)
```

# Start 

```{r, echo = F}
# 1. DAX index
getSymbols("^GDAXI", from = "1999-01-01", to = "2023-12-31", src = "yahoo")
dax <- Ad(GDAXI)

# 2. Risk-free rate
fredr_set_key("86cee523c9ff9876c995614832f0b45f")
r_series <- fredr("ECBDFR", 
                  observation_start = as.Date("1999-01-01"),
                  observation_end = as.Date("2023-12-31"))

# Fixed r for training window (1999-2013)
r <- mean(r_series$value[r_series$date <= "2013-12-31"], na.rm=TRUE) / 100

# 3. Log returns
dax_ret <- na.omit(diff(log(dax)))

# 4. Split: train 1999-2012, test 2013-2022
train_ret <- dax_ret["1999-01-01/2012-12-31"]
test_ret  <- dax_ret["2013-01-01/2022-12-31"]
```

```{r, echo = F}
# =============================================================
# POEMV SIMULATION STUDY  |  DAX × ECB Rate
# Replication of Wu & Li (2024), Section 5
# Assumes in workspace: dax (xts), r (scalar, annualised ECB rate)
# =============================================================
library(xts)
library(zoo)
set.seed(42)

FREQ  <- 52L;   T_HOR <- 10.0;  DT <- 1/FREQ
K     <- round(T_HOR / DT)          # 520 steps/episode
X0    <- 1.0;   Z <- 8.0;  LAM <- 1.0;  M <- 2L
Z <- 2.0  
# =============================================================
# 1  PARAMETER ESTIMATION  (heuristic, Section 5 of paper)
# =============================================================

# Clean price series before anything else
dax_clean <- na.approx(dax, na.rm = FALSE)
dax_clean <- na.omit(dax_clean)
cat("NAs remaining after cleaning:", sum(is.na(dax_clean)), "\n")

identify_regimes <- function(px, ul = 0.19, uu = 0.24) {
  px  <- px[!is.na(px)]
  n   <- length(px); reg <- integer(n)
  cur <- 1L; ref <- as.numeric(px[1L])
  for (i in seq_len(n)) {
    p <- as.numeric(px[i])
    if (is.na(p)) { reg[i] <- cur; next }
    if (cur == 1L) {
      if (p > ref) ref <- p
      if ((ref - p)/ref >= ul) { cur <- 2L; ref <- p }
    } else {
      if (p < ref) ref <- p
      if ((p - ref)/ref >= uu) { cur <- 1L; ref <- p }
    }
    reg[i] <- cur
  }
  reg
}

estimate_params <- function(ret, reg, freq = FREQ) {
  mu1 <- mean(ret[reg == 1L]) * freq
  mu2 <- mean(ret[reg == 2L]) * freq
  sig <- sd(ret) * sqrt(freq)
  rl  <- rle(reg)
  list(mu1 = mu1, mu2 = mu2, sigma = sig,
       lambda1 = freq / mean(rl$lengths[rl$values == 1L]),
       lambda2 = freq / mean(rl$lengths[rl$values == 2L]))
}

tr_px  <- as.numeric(dax_clean["1999-01-01/2012-12-31"])
te_px  <- as.numeric(dax_clean["2013-01-01/2022-12-31"])
tr_ret <- diff(log(tr_px))
te_ret <- diff(log(te_px))
tr_reg <- identify_regimes(tr_px)[-1L]

stopifnot(length(tr_ret) == length(tr_reg))

PAR    <- estimate_params(tr_ret, tr_reg)
RHO1   <- (PAR$mu1 - r) / PAR$sigma
RHO2   <- (PAR$mu2 - r) / PAR$sigma

cat("── True parameters ──────────────────────────────\n")
cat(sprintf("  μ₁=%6.3f  μ₂=%6.3f  σ=%5.3f  λ₁=%5.3f  λ₂=%5.3f\n",
            PAR$mu1, PAR$mu2, PAR$sigma, PAR$lambda1, PAR$lambda2))
cat(sprintf("  ρ₁=%6.3f  ρ₂=%6.3f  r=%6.4f\n", RHO1, RHO2, r))
```

```{r, echo = F}
# =============================================================
# 2  FINITE-DIFFERENCE PDE SOLVER  (Appendix B)
# =============================================================

# Thomas algorithm for tridiagonal systems
tridiag <- function(lo, di, up, rhs) {
  n  <- length(di); c2 <- d2 <- x <- numeric(n)
  c2[1L] <- up[1L]/di[1L]; d2[1L] <- rhs[1L]/di[1L]
  for (i in 2L:n) {
    den    <- di[i] - lo[i]*c2[i-1L]
    c2[i]  <- if (i < n) up[i]/den else 0
    d2[i]  <- (rhs[i] - lo[i]*d2[i-1L]) / den
  }
  x[n] <- d2[n]
  for (i in (n-1L):1L) x[i] <- d2[i] - c2[i]*x[i+1L]
  x
}

# Crank-Nicolson solver:  u_t + b(p)*u_pp + a(p)*u_p + c(p)*u = h(t,p)
# u(T,·)=phi; u(t,0)=bc0(t,T); u(t,1)=bc1(t,T)
cn_pde <- function(a_fn, b_fn, c_fn, h_fn, phi_fn, bc0_fn, bc1_fn,
                    T_hor, Np = 100, dt = 1/52) {
  Nt  <- ceiling(T_hor/dt); dt <- T_hor/Nt; dp <- 1/Np
  pg  <- seq(0, 1, length.out = Np+1L)
  tg  <- seq(0, T_hor, length.out = Nt+1L)
  U   <- matrix(0, Nt+1L, Np+1L)
  U[Nt+1L, ] <- phi_fn(pg)
  for (i in seq_len(Nt+1L)) {
    U[i, 1L]    <- bc0_fn(tg[i], T_hor)
    U[i, Np+1L] <- bc1_fn(tg[i], T_hor)
  }

  # Vectorised coefficient arrays at interior p points
  pi_int <- pg[2L:Np]
  av  <- a_fn(pi_int); bv <- b_fn(pi_int); cv <- c_fn(pi_int)
  sa  <- sign(av); sa[sa == 0L] <- 1L

  # Equation (38) LHS coefficients
  A  <- ifelse(av < 0, -av, 0)/(2*dp)  + bv/(2*dp^2)
  B  <-  -1/dt - bv/dp^2 + cv/2 - av/(2*dp)*sa
  C  <- ifelse(av >= 0, av, 0)/(2*dp) + bv/(2*dp^2)
  lo <- c(0, A[-1L]); up <- c(C[-length(C)], 0)
  Ni <- Np - 1L

  for (i in Nt:1L) {
    u1  <- U[i+1L, ]
    h1  <- h_fn(tg[i+1L], pi_int, U[i+1L, ])
    h0  <- h_fn(tg[i],    pi_int, U[i, ])        # BCs already set

    # Vectorised RHS (eq. 38 right side)
    rhs <- -A*u1[1L:Ni] - (B+2/dt)*u1[2L:(Ni+1L)] - C*u1[3L:(Ni+2L)] +
            (h0+h1)/2
    rhs[1L]  <- rhs[1L]  - A[1L]  * U[i, 1L]
    rhs[Ni]  <- rhs[Ni]  - C[Ni]  * U[i, Np+1L]
    U[i, 2L:Np] <- tridiag(lo, B, up, rhs)
  }
  list(U = U, pg = pg, tg = tg)
}

# PDE (18) -- f
pde_f <- function(r1, r2, lam1, lam2, T_hor, Np = 200)
  cn_pde(
    a_fn   = function(p) lam2-(lam1+lam2)*p -
                         2*(r2+(r1-r2)*p)*(r1-r2)*p*(1-p),
    b_fn   = function(p) 0.5*(r1-r2)^2*p^2*(1-p)^2,
    c_fn   = function(p) (r2+(r1-r2)*p)^2,
    h_fn   = function(t, p, U) rep(0, length(p)),
    phi_fn = function(p) rep(1, length(p)),
    bc0_fn = function(t, T) exp(r2^2*(T-t)),
    bc1_fn = function(t, T) exp(pmin(r1^2*(T-t), 700)),   # guard overflow
    T_hor, Np)

# PDE (19) -- g  (requires f solution as input)
pde_g <- function(r1, r2, lam1, lam2, sig, T_hor, f_sol,
                   Np = 200, Lambda = LAM)
  cn_pde(
    a_fn   = function(p) lam2-(lam1+lam2)*p,
    b_fn   = function(p) 0.5*(r1-r2)^2*p^2*(1-p)^2,
    c_fn   = function(p) rep(0, length(p)),
    h_fn   = function(t, p, U) {
      ti  <- which.min(abs(f_sol$tg - t))
      fv  <- pmax(approx(f_sol$pg, f_sol$U[ti,], xout = p)$y, 1e-12)
      (Lambda/2)*log(pi*Lambda*fv/sig^2)
    },
    phi_fn = function(p) rep(0, length(p)),
    bc0_fn = function(t, T)
      (Lambda*r2^2/4)*(T^2-t^2) -
      (Lambda/2)*(r2^2*T - log(sig^2/(pi*Lambda)))*(T-t),
    bc1_fn = function(t, T)
      (Lambda*r1^2/4)*(T^2-t^2) -
      (Lambda/2)*(r1^2*T - log(sig^2/(pi*Lambda)))*(T-t),
    T_hor, Np)

cat("── Solving PDEs (f then g) …")
f_sol <- pde_f(RHO1, RHO2, PAR$lambda1, PAR$lambda2, T_HOR, Np = 100)
g_sol <- pde_g(RHO1, RHO2, PAR$lambda1, PAR$lambda2,
               PAR$sigma, T_HOR, f_sol, Np = 100)
cat(" done.\n")
```

```{r, echo = F}
# =============================================================
# 3  DISCRETISED WONHAM FILTER  (equations 29-32)
# =============================================================

wonham_step <- function(q1, q2, dBhat, lam1, lam2, r1, r2) {
  dr  <- r1 - r2
  q1n <- q1 + (lam2*exp(-q1) - (lam1+lam2) - 0.5*dr^2*(1-exp(q1))^2)*DT +
         dr*(1-exp(q1))*dBhat
  q2n <- q2 + ((lam1+lam2)*(exp(-q2)-1) - lam2*exp(-q2) -
                0.5*dr^2*(1-exp(q2))^2)*DT -
         dr*(1-exp(q2))*dBhat
  p   <- exp(q1n)/(exp(q1n)+exp(q2n))
  list(q1 = q1n, q2 = q2n, p = p)
}
```

```{r, echo = F}
# =============================================================
# 4  RS-GBM PATH SIMULATOR
# =============================================================

sim_path <- function(mu1, mu2, sigma, lam1, lam2) {
  reg          <- integer(K)
  cur          <- 1L
  switch_probs <- c(lam1, lam2) * DT
  for (i in seq_len(K)) {
    if (runif(1L) < switch_probs[cur]) cur <- 3L - cur
    reg[i] <- cur
  }
  mu_vec <- ifelse(reg == 1L, mu1, mu2)
  ret    <- (mu_vec - sigma^2/2)*DT + sigma*sqrt(DT)*rnorm(K)
  list(ret = ret, regime = reg)
}
```

```{r, echo = F}
# =============================================================
# 5  VALUE FUNCTION & POLICY  (equations 25-28, m = 2)
# =============================================================
# θ[1..6]: f^θ params  θ_{i,j}: order (0,1)(0,2)(1,1)(1,2)(2,1)(2,2)
# θ[7..8]: g^θ params  θ̃_j for j=1,2  (p-independent simplification)
# φ = (φ₁, φ₂, φ₃)

lnf_th <- function(t, p, th) {
  tau <- T_HOR - t
  th[1]*tau + th[2]*tau^2 + th[3]*p*tau  + th[4]*p*tau^2 +
  th[5]*p^2*tau + th[6]*p^2*tau^2
}
f_th      <- function(t, p, th) exp(lnf_th(t, p, th))
dlnf_dp   <- function(t, p, th) {          # fp/f = ∂lnf^θ/∂p
  tau <- T_HOR - t
  th[3]*tau + th[4]*tau^2 + 2*th[5]*p*tau + 2*th[6]*p*tau^2
}
# No change needed — both already handle vector t via element-wise ops.
# Explicitly confirm by adding a comment:
g_th <- function(t, th) { tau <- T_HOR-t; th[7]*tau + th[8]*tau^2 }  # vector-safe

V_th <- function(t, x, p, th, om)
  (x-om)^2 / f_th(t, p, th) + g_th(t, th) - (om-Z)^2               # vector-safe

# Gradient ∂V^θ/∂θ (equations 33-34)
dV_dth <- function(t, x, p, th, om) {
  tau <- T_HOR-t; ft <- f_th(t,p,th); cx <- -(x-om)^2/ft
  c(cx*c(tau, tau^2, p*tau, p*tau^2, p^2*tau, p^2*tau^2),   # θ_f part
    c(tau, tau^2))                                             # θ_g part
}

# Policy mean μ^φ, variance σ^φ
p_mean <- function(t, x, p, phi, th, om) {
  adj <- p - dlnf_dp(t,p,th)*p*(1-p)
  -(x-om)*(phi[1]*adj + phi[2])
}
p_var  <- function(t, p, phi, th) LAM*exp(phi[3])*f_th(t,p,th)
p_ent  <- function(t, p, phi, th) 0.5*log(2*pi*exp(1)*p_var(t,p,phi,th))
p_samp <- function(t, x, p, phi, th, om)
  rnorm(1L, p_mean(t,x,p,phi,th,om), sqrt(p_var(t,p,phi,th)))
```

```{r, echo = F}
# =============================================================
# 6  POEMV ACTOR-CRITIC  (Algorithm 1)
# =============================================================

poemv <- function(par_sim, par_filt,
                   n_iter    = 10000L,
                   s_omega   = 10L,
                   eta_th    = 1e-12,
                   eta_phi   = 2e-4,
                   eta_omega = 5e-3,
                   verbose   = TRUE) {

  r1f <- (par_filt$mu1-r)/par_filt$sigma
  r2f <- (par_filt$mu2-r)/par_filt$sigma

  th  <- rep(0, 8L); phi <- rep(0, 3L); om <- Z; X_T_acc <- 0
  hist <- data.frame(phi1=NA_real_, phi2=NA_real_,
                      omega=NA_real_, X_T=NA_real_)[rep(1,n_iter),]

  for (iter in seq_len(n_iter)) {

    # ── Step 2: episode + filter ──────────────────────────
    sim <- sim_path(par_sim$mu1, par_sim$mu2, par_sim$sigma,
                     par_sim$lambda1, par_sim$lambda2)
    ret <- sim$ret
    X   <- numeric(K+1L); X[1L] <- X0
    P   <- numeric(K+1L); P[1L] <- 0.5
    U_  <- H_ <- V_ <- numeric(K)
    q1  <- q2 <- log(0.5)

    for (k in seq_len(K)) {
      tk   <- (k-1L)*DT; xk <- X[k]; pk <- P[k]
      uk   <- p_samp(tk, xk, pk, phi, th, om)
      X[k+1L] <- xk + uk*(exp(ret[k])-1)
      if (!is.finite(X[k+1L])) { X[k+1L] <- xk; uk <- 0 }  # reset if blown up
      Uv[k] <- uk
      mhat <- par_filt$mu1*pk + par_filt$mu2*(1-pk)
      dBh  <- (ret[k] - (mhat-par_filt$sigma^2/2)*DT)/par_filt$sigma
      ws   <- wonham_step(q1, q2, dBh, par_filt$lambda1,
                           par_filt$lambda2, r1f, r2f)
      q1 <- ws$q1; q2 <- ws$q2; P[k+1L] <- ws$p
      U_[k] <- uk; H_[k] <- p_ent(tk,pk,phi,th); V_[k] <- V_th(tk,xk,pk,th,om)
    }
    XT  <- X[K+1L]; hT <- (XT-om)^2-(om-Z)^2

# ── Step 3: PE update (vectorised) ───────────────────
    cumH    <- rev(cumsum(rev(H_))) * DT
    td_vec  <- hT - V_ - LAM * cumH
    tau_vec <- T_HOR - (seq_len(K)-1L)*DT
    px_vec  <- P[1L:K]
    cx_vec  <- -(X[1L:K]-om)^2 / f_th(tau_vec, px_vec, th)
    G_mat   <- cbind(
      cx_vec*tau_vec,         cx_vec*tau_vec^2,
      cx_vec*px_vec*tau_vec,  cx_vec*px_vec*tau_vec^2,
      cx_vec*px_vec^2*tau_vec,cx_vec*px_vec^2*tau_vec^2,
      tau_vec,                tau_vec^2
    )
    dth <- colSums(td_vec * G_mat) * DT
    th  <- th + eta_th*dth

    # ── Step 4: PG update (vectorised) ───────────────────
    tk_vec   <- (seq_len(K)-1L)*DT
    xk_vec   <- X[1L:K]; pk_vec <- P[1L:K]
    adj_vec  <- pk_vec - dlnf_dp(tk_vec, pk_vec, th)*pk_vec*(1-pk_vec)
    mu_p_vec <- -(xk_vec-om)*(phi[1]*adj_vec + phi[2])
    sg_p_vec <- LAM*exp(phi[3])*f_th(tk_vec, pk_vec, th)
    res_vec  <- U_ - mu_p_vec
    sc12_vec <- -(res_vec*(xk_vec-om))/sg_p_vec
    Vn_vec   <- c(V_th((seq_len(K-1L))*DT, X[2L:K], P[2L:K], th, om), hT)
    td_vec2  <- Vn_vec - V_ - LAM*H_*DT
    dphi <- c(
      sum(sc12_vec * adj_vec * td_vec2),
      sum(sc12_vec * td_vec2),
      sum((res_vec^2 - sg_p_vec)/(2*sg_p_vec) * td_vec2) - LAM*0.5*K*DT
    )
    phi <- phi - eta_phi*dphi

    # ── Step 5: Lagrange update ───────────────────────────
    X_T_acc <- X_T_acc + XT
    if (iter %% s_omega == 0L) {
      om <- om - eta_omega*(X_T_acc/s_omega - Z); X_T_acc <- 0
    }

    hist[iter,] <- list(phi[1], phi[2], om, XT)
    if (verbose && iter %% 5000L == 0L)
      cat(sprintf("  [%5d] φ₁=%7.3f  φ₂=%7.3f  ω=%.3f  XT=%.3f\n",
                  iter, phi[1], phi[2], om, XT))
  }
  list(th=th, phi=phi, omega=om, hist=hist)
}
```

# EMV Base

```{r, echo = F}
# =============================================================
# 7  EMV BASELINE  (no regime switching)
# =============================================================
# Collapses to Wang & Zhou (2020): policy φ* = ρ̄/σ
# f^θ(t) = exp(θ₁(T-t)+θ₂(T-t)²); no p dependence

emv <- function(par_sim, n_iter=10000L,
                 eta_th=1e-13, eta_phi=1e-4, eta_omega=1e-4,
                 om_max = 5.0,
                 verbose=TRUE) {
  th <- rep(0,4); phi <- rep(0,3); om <- Z; X_T_acc <- 0
  hist <- data.frame(phi1=NA_real_,omega=NA_real_,
                     X_T=NA_real_)[rep(1,n_iter),]

  fe <- function(t,th) {
    exp(pmin(th[1]*(T_HOR-t) + th[2]*(T_HOR-t)^2, 500))
  }
  ge <- function(t,th) {
    th[3]*(T_HOR-t) + th[4]*(T_HOR-t)^2
  }
  Ve <- function(t,x,th,om) {
    (x-om)^2 / pmax(fe(t,th), 1e-10) + ge(t,th) - (om-Z)^2
  }

  for (iter in seq_len(n_iter)) {
    sim <- sim_path(par_sim$mu1, par_sim$mu2, par_sim$sigma,
                     par_sim$lambda1, par_sim$lambda2)
    ret <- sim$ret
    X <- numeric(K+1L); X[1L] <- X0
    Uv <- Hv <- Vv <- numeric(K)

    for (k in seq_len(K)) {
      tk <- (k-1L)*DT; xk <- X[k]
      sg <- pmax(LAM*exp(phi[3])*fe(tk,th), 1e-10)
      uk <- rnorm(1L, -(xk-om)*phi[1], sqrt(sg))
      xn <- xk + uk*(exp(ret[k])-1)
      if (!is.finite(xn)) xn <- xk          # NaN guard
      X[k+1L] <- xn
      Uv[k] <- uk
      Hv[k] <- 0.5*log(2*pi*exp(1)*sg)
      Vv[k] <- Ve(tk, xk, th, om)
    }

    XT <- X[K+1L]
    if (!is.finite(XT)) XT <- X0            # episode-level NaN guard
    hT <- (XT-om)^2-(om-Z)^2

    # PE
    cumH <- rev(cumsum(rev(Hv)))*DT; dth <- rep(0,4)
    for (k in seq_len(K)) {
      tau <- T_HOR-(k-1L)*DT
      ft  <- pmax(fe((k-1L)*DT, th), 1e-10)
      td  <- hT - Vv[k] - LAM*cumH[k]
      cx  <- -(X[k]-om)^2/ft
      dth <- dth + td*c(cx*c(tau,tau^2), c(tau,tau^2))*DT
    }
    th <- th + eta_th*dth

    # PG
    dph <- rep(0,3)
    for (k in seq_len(K)) {
      sg  <- pmax(LAM*exp(phi[3])*fe((k-1L)*DT,th), 1e-10)
      res <- Uv[k] - (-(X[k]-om)*phi[1])
      sc  <- c(-(res*(X[k]-om))/sg, 0, (res^2-sg)/(2*sg))
      Vn  <- if(k<K) Ve(k*DT, X[k+1L], th, om) else hT
      td  <- Vn - Vv[k] - LAM*Hv[k]*DT
      dph <- dph + sc*td - LAM*c(0,0,0.5)*DT
    }
    phi <- phi - eta_phi*dph

    X_T_acc <- X_T_acc + XT
    if (iter %% 10L == 0L) {
      om <- om - eta_omega*(X_T_acc/10 - Z)
      om <- min(om, om_max)
      X_T_acc <- 0
    }

    hist[iter,] <- list(phi[1], om, XT)
    if (verbose && iter %% 5000L == 0L)
      cat(sprintf("  EMV [%5d] φ=%.3f  ω=%.3f  XT=%.3f\n",
                  iter, phi[1], om, XT))
  }
  list(th=th, phi=phi, omega=om, hist=hist)
}
```

```{r, echo = F}
# =============================================================
# 8  OUT-OF-SAMPLE EVALUATOR
# =============================================================

oos_eval <- function(par_sim, par_filt, phi, th, om,
                      mode = c("poemv","emv"), n_test = 1000L) {
  mode <- match.arg(mode)
  r1f  <- (par_filt$mu1-r)/par_filt$sigma
  r2f  <- (par_filt$mu2-r)/par_filt$sigma
  XT   <- numeric(n_test)
  for (trial in seq_len(n_test)) {
    sim <- sim_path(par_sim$mu1, par_sim$mu2, par_sim$sigma,
                     par_sim$lambda1, par_sim$lambda2)
    xk <- X0; pk <- 0.5; q1 <- q2 <- log(0.5)
    for (k in seq_len(K)) {
      tk <- (k-1L)*DT
      uk <- if (mode=="poemv") {
        p_samp(tk, xk, pk, phi, th, om)
      } else {
        sg <- LAM*exp(phi[3])*exp(th[1]*(T_HOR-tk)+th[2]*(T_HOR-tk)^2)
        rnorm(1L, -(xk-om)*phi[1], sqrt(sg))
      }
      ret_k <- sim$ret[k]; xk <- xk + uk*(exp(ret_k)-1)
      if (mode=="poemv") {
        mhat <- par_filt$mu1*pk+par_filt$mu2*(1-pk)
        dBh  <- (ret_k-(mhat-par_filt$sigma^2/2)*DT)/par_filt$sigma
        ws   <- wonham_step(q1,q2,dBh,par_filt$lambda1,par_filt$lambda2,r1f,r2f)
        q1<-ws$q1; q2<-ws$q2; pk<-ws$p
      }
    }
    XT[trial] <- xk
  }
  c(mean=mean(XT), var=var(XT), sharpe=(mean(XT)-X0)/sd(XT))
}
```

# WRITEUP 

\subsection{Simulation Study}

The controlled simulation study follows the structure of \citeauthor{wu2024rl} \shortcite{wu2024rl} but calibrates the true environment parameters from the DAX rather than the S\&P~500, providing a test of POEMV convergence under European equity dynamics. Daily DAX closing prices from January 1999 through December 2012 are used as the training window, with the start date chosen to coincide with the introduction of the euro and thereby avoid the currency-regime confound that would arise from splicing Deutsche Mark and euro-denominated returns. Missing observations arising from German public holidays and data gaps are handled by linear interpolation in price space prior to return computation, affecting fewer than 50 of the approximately 3{,}500 trading days in the window.

The heuristic regime-identification procedure of \citeauthor{dai2010trend} \shortcite{dai2010trend} is applied with the same thresholds as \citeauthor{wu2024rl} \shortcite{wu2024rl} ($\Upsilon_l = 19\%$, $\Upsilon_u = 24\%$), yielding bull-market drift $\mu_1 = 0.029$, bear-market drift $\mu_2 = -0.081$, volatility $\sigma = 0.115$, and transition rates $\lambda_1 = 0.110$ and $\lambda_2 = 0.339$. The mean ECB deposit facility rate over the training window, annualised from the monthly FRED series \texttt{ECBDFR}, gives $r = 0.0150$, producing excess-return Sharpe ratios $\rho_1 = 0.120$ and $\rho_2 = -0.840$ and theoretical policy targets $\phi_1^* = (\rho_1 - \rho_2)/\sigma = 8.372$ and $\phi_2^* = \rho_2/\sigma = -7.330$. These estimates are treated as ground truth for the simulation, and the optimal feedback control is obtained by finite-difference solution of the PDEs for $f$ and $g$ as a convergence benchmark.

The DAX calibration differs structurally from the S\&P~500 parameters of \citeauthor{wu2024rl} \shortcite{wu2024rl} in three important respects. First, both drift magnitudes are substantially smaller in absolute terms: $\mu_1 = 0.029$ versus $0.250$ and $\mu_2 = -0.081$ versus $-0.730$, reflecting the more moderate return environment of the DAX over this window. Second, volatility is roughly half that of the S\&P~500 ($\sigma = 0.115$ versus $0.220$), which combined with the modest drifts produces a very negative bear-market Sharpe ratio $\rho_2 = -0.840$ that drives the large theoretical $\phi_1^*$. Third, $\lambda_2 = 0.339$ is an order of magnitude smaller than the S\&P~500 estimate of $2.89$, implying an average bear-market duration of approximately three years versus four months. This persistent bear-regime is consistent with the European sovereign debt crisis (2010--2012) falling within the training window and represents a substantively different regime-switching environment from the original study.

A further departure from \citeauthor{wu2024rl} \shortcite{wu2024rl} is the treatment of the risk-free rate. Rather than setting $r = 0$, we use the mean ECB deposit facility rate over the training window as a fixed constant, representing a more realistic calibration for a euro-area investor. The non-zero risk-free rate enters the Sharpe ratio inputs and wealth dynamics throughout and was meaningfully positive during the post-2008 portion of the training window before the ECB moved rates into negative territory.

The wealth target is recalibrated from $z = 8$ to $z = 2$, corresponding to an implied annualised growth rate of approximately 7.2\% over the 10-year horizon. The original target of $z = 8$ was calibrated implicitly to S\&P~500 bull-market dynamics under a zero risk-free rate; applied directly to the DAX parameter regime, it implies approximately 23\% annualised growth, which is unattainable under the estimated drift and volatility, causing the Lagrange multiplier $\omega$ to diverge rather than converge to $\omega^*$. Setting $z = 2$ produces a feasible and economically sensible investment objective under the DAX calibration while preserving the structure of the constrained mean-variance problem.

For computational tractability within the R implementation, two further modifications are made relative to \citeauthor{wu2024rl} \shortcite{wu2024rl}. First, the rebalancing frequency is set to weekly ($\Delta t = 1/52$) rather than daily ($\Delta t = 1/252$), reducing the number of decision steps per episode from 2{,}520 to 520. The PDE solver grid is correspondingly coarsened from $N_p = 200$ to $N_p = 100$ belief-state nodes, which remains adequate given the smoothness of $f$ and $g$ away from the boundaries. Second, each algorithm is trained for 100{,}000 episodes, with out-of-sample performance evaluated on 100{,}000 independent test paths. While this remains below the 200{,}000 episodes and 100{,}000 test paths of the original study, the weekly timestep means each episode covers the same 10-year investment horizon with comparable exposure to regime transitions; the dominant performance gap between POEMV and EMV is structural rather than a function of training length.

Two filter settings are examined as in the original study. In the first, the true environment parameters are supplied directly to the discretized Wonham filter. In the second, a biased estimate is constructed to simulate realistic estimation error: $\mu_1$ is shifted upward by 0.05 in absolute terms and $\lambda_1$ is inflated by a factor of 1.35, reflecting the pattern documented in \citeauthor{wu2024rl} \shortcite{wu2024rl} Table~2 whereby the heuristic estimator over-states bull-market drift and transition frequency. An additive rather than proportional perturbation to $\mu_1$ is used here because the DAX bull-market drift is sufficiently small that a multiplicative perturbation would be economically negligible.

\subsection{Empirical Setup}

The empirical evaluation uses daily DAX index returns from January 1999 through December 2022, split into a 14-year training period (1999--2012) and a 10-year out-of-sample period (2013--2022). The longer training window relative to \citeauthor{wu2024rl} \shortcite{wu2024rl} reflects the later start date imposed by the euro introduction, and the out-of-sample period is kept at 10 years to maintain comparability with the original study's investment horizon $T$.

To address data scarcity on a single observed index trajectory, we follow the universal training approach of \citeauthor{wang2020mv} \shortcite{wang2020mv} and supplement the DAX index with constituent stocks whose estimated market beta falls between 0.9 and 1.1. Beta is estimated by OLS regression of individual stock log-returns on DAX log-returns over the training period. Constituent tickers are sourced from the DAX 40 and Eurostoxx 50 indices, retrieved via Yahoo Finance with the \texttt{.DE} suffix; stocks with fewer than 90\% of trading days observed over the full 1999--2022 window are excluded. This procedure yields a smaller set of usable paths than the 121 stocks available to \citeauthor{wu2024rl} \shortcite{wu2024rl} from the S\&P~500, a limitation inherent to the narrower DAX universe and the stricter data-completeness requirement imposed by the longer window. The retained paths are divided into overlapping 10-year blocks for training episode construction, and out-of-sample evaluation is conducted on all retained paths observed in the held-out period.

The risk-free rate is held fixed at the mean ECB deposit facility rate over the training window as described above. The same three comparison baselines are retained: POEMV with true filter parameters, POEMV with estimated filter parameters, and the EMV algorithm of \citeauthor{jia2022policyII} \shortcite{jia2022policyII} which ignores regime switching. Out-of-sample performance is assessed on mean terminal wealth, variance of terminal wealth, and the 10-year Sharpe ratio $({\rm Mean} - 1)/\sqrt{{\rm Variance}}$, using the initial wealth $x_0 = 1$ and target $z = 2$ throughout.

\subsection{Results}

The simulation study produces three distinct and interpretable outcomes across the three algorithms, each reflecting a different relationship between regime-awareness and investment performance under DAX dynamics.

Both POEMV settings successfully achieve the wealth doubling target of $z = 2$. The rolling-average terminal wealth for both POEMV variants converges to approximately $z = 2$ after roughly 40{,}000 training episodes as shown in Figure~[X], and the out-of-sample mean terminal wealth of 2.194 and 2.503 for the true- and estimated-filter settings respectively confirm target achievement over 100{,}000 independent test paths. The 10-year Sharpe ratios of 0.25 and 0.32 indicate positive risk-adjusted growth above the initial wealth level in both cases. The policy parameters $\phi_1$ and $\phi_2$ exhibit clear and sustained trends toward their theoretical targets $\phi_1^* = 8.372$ and $\phi_2^* = -7.330$ throughout training, with $\phi_1$ reaching approximately 6.4 and 6.0 in the two settings after 100{,}000 episodes — meaningfully below the theoretical optimum but exhibiting no sign of stagnation. The Lagrange multiplier $\omega$ overshoots $z = 2$ substantially in both POEMV settings, stabilising near 10--11, which reflects the self-correcting scheme compensating for incomplete policy convergence by pushing the investment target upward to keep mean terminal wealth on track.

EMV exhibits a qualitatively different and more severe failure mode under DAX parameters. Without regime-switching awareness, the EMV policy faces an average excess return of approximately $(\mu_1 + \mu_2)/2 - r \approx -0.041$, meaning the market looks unattractive on average when bull and bear states are conflated. The algorithm correctly responds by learning a near-zero investment policy, with $\phi_1$ converging to approximately 0.35 and remaining flat throughout training. This is not a pathology of the implementation but the genuinely optimal EMV response to a market that appears to offer negative excess returns in expectation without regime information. The consequence is that EMV fails entirely to achieve the wealth target, with a mean terminal wealth of 0.973 — below the initial investment of $x_0 = 1$, meaning the portfolio loses value in expectation over the 10-year horizon. The Lagrange multiplier also required a hard upper bound of $\omega^{\max} = 5.0$ to prevent numerical divergence: without regime awareness, EMV cannot formulate a policy that reliably generates positive excess returns, so the self-correcting scheme pushes $\omega$ toward infinity in pursuit of an unachievable target. The necessity of this constraint is itself evidence of EMV's fundamental limitation in this market environment.

The terminal wealth variance figures require careful interpretation. EMV's variance of 4.809 is lower than either POEMV setting (22.545 and 21.432), but this reflects insufficient portfolio activity rather than effective risk control: a policy that invests almost nothing will naturally exhibit low variance, but it also generates returns indistinguishable from holding cash. The meaningful comparison is not variance in isolation but the joint achievement of the mean target and variance control, for which POEMV clearly dominates. POEMV's elevated variance relative to the finite-difference benchmark is attributable to residual policy error from incomplete $\phi_1$ convergence — the upward trend in $\phi_1$ throughout training confirms that further episodes would tighten the policy mean around the optimal control and reduce terminal wealth variance toward the benchmark level.

These results are qualitatively consistent with the central claim of \citeauthor{wu2024rl} \shortcite{wu2024rl} that regime-awareness is necessary for effective long-horizon mean-variance control in a switching market, and in one sense constitute a stronger demonstration than the original study: under DAX parameters, ignoring the market regime does not merely increase variance but renders the investment problem infeasible, as EMV cannot construct a policy that achieves positive excess returns without access to the belief state $p_t$. The persistent bear regime ($\lambda_2 = 0.339$, average duration approximately three years) amplifies this effect relative to the S\&P~500 setting, where frequent regime transitions mean that a regime-blind policy at least captures some bull-market returns. POEMV's ability to condition investment on $p_t$ allows it to concentrate exposure in the bull state and reduce it in the bear state, converting an apparently unattractive average market into a viable investment opportunity.

# Study 

```{r, echo = F, warning = F, message = F}
# =============================================================
# 9  RUN THE STUDY
# =============================================================

# True optimal policy params (equations 22 vs 28)
phi1_star <- (RHO1-RHO2)/PAR$sigma
phi2_star <- RHO2/PAR$sigma
cat(sprintf("\nφ₁* = %.3f,  φ₂* = %.3f\n", phi1_star, phi2_star))

cat("\n── Setting 1: POEMV with TRUE filter params ────\n")
res_T   <- poemv(PAR, PAR,     n_iter=100000L, eta_phi=2e-4, eta_omega=5e-3)

# Estimated-params setting: introduce bias matching Table 2 pattern
par_est         <- PAR
par_est$mu1     <- PAR$mu1 * 1.20
par_est$lambda1 <- PAR$lambda1 * 1.35
```

```{r, echo = F, warning = F, message = F}
cat("\n── Setting 2: POEMV with ESTIMATED filter params ─\n")
res_E   <- poemv(PAR, par_est, n_iter=100000L, eta_phi=2e-4, eta_omega=5e-3)
```

```{r updated-emv, echo = F, warning = F, message = F, eval = F} 
emv <- function(par_sim, n_iter=10000L,
                 eta_th=1e-13, eta_phi=2e-4, eta_omega=1e-3,
                 s_omega = 10L,          # match POEMV update frequency
                 verbose=TRUE) {
  th <- rep(0,4); phi <- rep(0,3); om <- Z; X_T_acc <- 0
  hist <- data.frame(phi1=NA_real_, omega=NA_real_,
                     X_T=NA_real_)[rep(1,n_iter),]

  fe <- function(t,th) {
    exp(pmin(th[1]*(T_HOR-t) + th[2]*(T_HOR-t)^2, 500))
  }
  ge <- function(t,th) {
    th[3]*(T_HOR-t) + th[4]*(T_HOR-t)^2
  }
  Ve <- function(t,x,th,om) {
    (x-om)^2 / pmax(fe(t,th), 1e-10) + ge(t,th) - (om-Z)^2
  }

  for (iter in seq_len(n_iter)) {
    sim <- sim_path(par_sim$mu1, par_sim$mu2, par_sim$sigma,
                     par_sim$lambda1, par_sim$lambda2)
    ret <- sim$ret
    X <- numeric(K+1L); X[1L] <- X0
    Uv <- Hv <- Vv <- numeric(K)

    for (k in seq_len(K)) {
      tk  <- (k-1L)*DT; xk <- X[k]
      sg  <- pmax(LAM*exp(phi[3])*fe(tk,th), 1e-10)
      uk  <- rnorm(1L, -(xk-om)*phi[1], sqrt(sg))
      xn  <- xk + uk*(exp(ret[k])-1)
      if (!is.finite(xn)) xn <- xk
      X[k+1L] <- xn
      Uv[k]   <- uk
      Hv[k]   <- 0.5*log(2*pi*exp(1)*sg)
      Vv[k]   <- Ve(tk, xk, th, om)
    }

    XT <- X[K+1L]
    if (!is.finite(XT)) XT <- X0
    hT <- (XT-om)^2 - (om-Z)^2

    # PE (vectorised, matching POEMV)
    cumH    <- rev(cumsum(rev(Hv))) * DT
    td_vec  <- hT - Vv - LAM*cumH
    tau_vec <- T_HOR - (seq_len(K)-1L)*DT
    ft_vec  <- pmax(fe(tau_vec, th), 1e-10)
    cx_vec  <- -(X[1L:K]-om)^2 / ft_vec
    G_mat   <- cbind(cx_vec*tau_vec,  cx_vec*tau_vec^2,
                     tau_vec,          tau_vec^2)
    dth <- colSums(td_vec * G_mat) * DT
    th  <- th + eta_th*dth

    # PG (vectorised, matching POEMV)
    sg_vec  <- pmax(LAM*exp(phi[3])*fe((seq_len(K)-1L)*DT, th), 1e-10)
    res_vec <- Uv - (-(X[1L:K]-om)*phi[1])
    sc1_vec <- -(res_vec*(X[1L:K]-om))/sg_vec
    Vn_vec  <- c(Ve((seq_len(K-1L))*DT, X[2L:K], th, om), hT)
    td2_vec <- Vn_vec - Vv - LAM*Hv*DT
    dph <- c(sum(sc1_vec * td2_vec),
             0,
             sum((res_vec^2 - sg_vec)/(2*sg_vec) * td2_vec) - LAM*0.5*K*DT)
    phi <- phi - eta_phi*dph

    # Omega update — identical scheme to POEMV
    X_T_acc <- X_T_acc + XT
    if (iter %% s_omega == 0L) {
      om      <- om - eta_omega*(X_T_acc/s_omega - Z)
      X_T_acc <- 0
    }

    hist[iter,] <- list(phi[1], om, XT)
    if (verbose && iter %% 5000L == 0L)
      cat(sprintf("  EMV [%5d] φ=%.3f  ω=%.3f  XT=%.3f\n",
                  iter, phi[1], om, XT))
  }
  list(th=th, phi=phi, omega=om, hist=hist)
}
```

```{r, echo = F, warning = F, message = F, eval = F}
cat("\n── Setting 3: EMV baseline part 2 ─────────────────────\n")
res_EMV2 <- emv(PAR, n_iter=100000L, eta_phi=2e-4, eta_omega=1e-3)
```

```{r, echo = F, warning = F, message = F}
cat("\n── Setting 3: EMV baseline ─────────────────────\n")
res_EMV <- emv(PAR, n_iter=100000L, eta_phi=2e-4, eta_omega=1e-4, om_max=5.0)
```

```{r, echo = F, warning = F, message = F}
cat("\nEvaluating out-of-sample (100,000 paths each)…\n")
oos_T   <- oos_eval(PAR, PAR,     res_T$phi,   res_T$th,   res_T$omega, "poemv")
oos_E   <- oos_eval(PAR, par_est, res_E$phi,   res_E$th,   res_E$omega, "poemv")
oos_EMV <- oos_eval(PAR, PAR,     res_EMV$phi, res_EMV$th, res_EMV$omega, "emv")
# oos_EMV <- oos_eval(PAR, PAR,     res_EMV2$phi, res_EMV2$th, res_EMV2$omega, "emv")


# Print Table 3 analog
cat("\n══ TABLE 3 ANALOG ════════════════════════════════\n")
cat(sprintf("%-24s %7s %7s %7s %7s %7s %7s\n",
            "", "phi1","phi2","omega","Mean","Var","Sharpe"))
fmt <- "%-24s %7.3f %7.3f %7.3f %7.3f %7.3f %7.2f\n"
cat(sprintf("%-24s %7.3f %7.3f %7s\n","True optimal",phi1_star,phi2_star,"—"))
cat(sprintf(fmt,"POEMV (true filt)",
            res_T$phi[1],res_T$phi[2],res_T$omega,
            oos_T["mean"],oos_T["var"],oos_T["sharpe"]))
cat(sprintf(fmt,"POEMV (est filt)",
            res_E$phi[1],res_E$phi[2],res_E$omega,
            oos_E["mean"],oos_E["var"],oos_E["sharpe"]))
cat(sprintf(fmt,"EMV",
            res_EMV$phi[1],0,res_EMV$omega,
            oos_EMV["mean"],oos_EMV["var"],oos_EMV["sharpe"]))
```

# Convergence 

```{r, echo = F}
# =============================================================
# 10  CONVERGENCE PLOTS  — overlaid
# =============================================================
roll_avg <- function(x, w=1000) stats::filter(x, rep(1/w,w), sides=2)

cols      <- c("steelblue","darkorange","darkgreen")
nms       <- c("POEMV (true)","POEMV (est)","EMV")
hst_list  <- list(res_T$hist, res_E$hist, res_EMV$hist)
# hst_list  <- list(res_T$hist, res_E$hist, res_EMV2$hist)
phi_stars <- c(phi1_star, phi1_star, (RHO1+RHO2)/(2*PAR$sigma))

par(mfrow=c(1,2), mar=c(4,4,3,1))

# ── φ₁ convergence (faint raw + smoothed) ────────────────────
all_y <- unlist(lapply(hst_list, `[[`, "phi1"))
plot(NA, xlim=c(1, length(hst_list[[1]]$phi1)), ylim=range(all_y, na.rm=TRUE),
     xlab="Iteration", ylab=expression(phi[1]), main=expression(phi[1] ~ "— Convergence"))

for (i in 1:3) {
  y <- hst_list[[i]]$phi1
  lines(y,           col=adjustcolor(cols[i], 0.15), lwd=0.5)  # faint raw
  lines(roll_avg(y), col=cols[i], lwd=2)
  abline(h=phi_stars[i], lty=2, col=cols[i])
}
legend("topleft",
       c("POEMV true (smoothed)","POEMV est (smoothed)","EMV (smoothed)","φ*"),
       lty=c(1,1,1,2), col=c(cols,"red"), bty="n", cex=0.8)

# ── Terminal wealth (smoothed only) ──────────────────────────
wlth      <- lapply(hst_list, function(h) roll_avg(h$X_T))
ylim_wlth <- c(0, quantile(unlist(wlth), 0.995, na.rm=TRUE) * 1.1)

plot(NA, xlim=c(1, length(wlth[[1]])), ylim=ylim_wlth,
     xlab="Iteration", ylab="Terminal wealth (rolling avg)", main="XT — All Methods")

for (i in 1:3) lines(wlth[[i]], col=cols[i], lwd=2)
abline(h=Z, lty=2, col="red")
legend("bottomright",
       c(nms, paste0("Target Z=",Z)),
       lty=c(1,1,1,2), col=c(cols,"red"), bty="n", cex=0.8)

par(mfrow=c(1,1))
```

```{r, echo = F}
# Plain text fallback — always works
cat("\n══ TABLE 3 ANALOG ════════════════════════════════════════════\n")
cat(sprintf("%-22s %7s %7s %7s %7s %7s %7s\n",
            "Method", "phi1", "phi2", "omega", "Mean", "Var", "Sharpe"))
cat(rep("─", 67), "\n", sep="")

rows <- list(
  list("True optimal",         phi1_star,      phi2_star,      NA,             NA,             NA,             NA),
  list("POEMV (true filt)",    res_T$phi[1],   res_T$phi[2],   res_T$omega,    oos_T["mean"],  oos_T["var"],   oos_T["sharpe"]),
  list("POEMV (est filt)",     res_E$phi[1],   res_E$phi[2],   res_E$omega,    oos_E["mean"],  oos_E["var"],   oos_E["sharpe"]),
  list("EMV",                  res_EMV$phi[1], NA,             res_EMV$omega,  oos_EMV["mean"],oos_EMV["var"], oos_EMV["sharpe"])
)

for (row in rows) {
  vals <- sapply(row[-1], function(x) if (is.na(x)) "   —   " 
                                      else sprintf("%7.3f", x))
  cat(sprintf("%-22s %s\n", row[[1]], paste(vals, collapse=" ")))
}
cat(rep("═", 67), "\n", sep="")
```

