---
output:
  pdf_document: default
  html_document: default
---
[Stat 5000]{.smallcaps}[Homework #3]{.smallcaps}\
[Fall 2024]{.smallcaps} [due Fri, Sep 20th @ 11:59 pm]{.smallcaps}
[Name: Sam Olson]{.smallcaps} \

**Directions:** Type or clearly handwrite your solutions to each of the
following exercises. Partial credit cannot be given unless all work is
shown. You may work in groups provided that each person takes
responsibility for understanding and writing out the solutions.
Additionally, you must give proper credit to your collaborators by
providing their names on the line below (if you worked alone, write "No
Collaborators\"):\

[Collaborators: Ben, Sabrina, Sarah, Vanessa, **The Hat Man**]{.smallcaps} \

# Note to Self 

## Confidence  Intervals

Formula: 
$$\text{Confidence Interval} \space = (\bar{Y_1} - \bar{Y_2}) \pm t_{(df - 2, 1 - \frac{\alpha}{2})} S_p \sqrt{\frac{1}{n_1} + \frac{1}{n_2}}$$
Where: 
$$S_p = \frac{(n_1-1)S_1^2 + (n_2-1)S_2^2}{n_1+n_2-2}$$

Calculation
```{r}
# Load Data 
library(readr)
smsspeed_1 <- read_csv("C:/Users/samue/OneDrive/Desktop/Iowa_State_PS/STAT 5000/Labs/Lab 3/smsspeed-1.csv")

library(dplyr)
data1 <- smsspeed_1 %>% 
  filter(smsspeed_1$AgeGroup == "Over30")

data2 <- smsspeed_1 %>% 
  filter(smsspeed_1$AgeGroup == "Teens")

sampleMean1 <- mean(data1$`Own Phone`)
sampleMean2 <- mean(data2$`Own Phone`)
difference <- sampleMean1 - sampleMean2 
difference

#Step 2: Finding standard deviation 
s1 <- sd(data1$`Own Phone`)
s2 <- sd(data2$`Own Phone`)

#Step 3: Finding sample size 
n1 <- length(data1$`Own Phone`)
n2 <- length(data2$`Own Phone`)

numerator <- (n1-1)*(s1^2) + (n2-1)*(s2^2)
denom <- n1 + n2 - 2
pooled <- sqrt( numerator / denom )  
sqrtFactor <- sqrt(1/n1 + 1/n2)

tStatDf <-  2.0484

rightSide <- tStatDf*pooled*sqrtFactor 

pooled
rightSide 

lb <- difference - rightSide
ub <- difference + rightSide

lb
ub
```

# Assignment 

1.  **\[+21\]:** A major medical center in the Northeastern U.S. conducted a study looking at blood cholesterol levels and incidence of heart attack. Below are summary statistics of blood cholesterol levels from 16 people who had a heart attack and 20 people who did not have a heart attack.

::: center

  Group                  Sample Size ($n$)   Sample Mean ($\bar y$)   Sample Std. Dev. ($s$)
  --------------------- ------------------- ------------------------ ------------------------
  Heart Attack (1)          $n_1 = 16$         $\bar y_1 = 265.4$         $s_1 = 43.645$
  No Heart Attack (0)       $n_2 = 20$         $\bar y_2 = 193.1$         $s_2 = 21.623$
  
:::

> 1.  List the assumptions needed to properly use the $t$-based confidence interval.

  - $Y_{1,1}, Y_{1,2}, ... ,Y_{1,16}$ are i.i.d. $N(\mu_1, \sigma^2)$
  - $Y_{2,1}, Y_{1,2}, ... ,Y_{2,20}$ are i.i.d. $N(\mu_2, \sigma^2)$
  - Population variances between the two groups are equal (both have $\sigma^2$ variance)
  - Independence between observations between groups, i.e. $Y_{1,i}$ and $Y_{2,j}$ are independent for all i and j

\newpage

> 2.  Using the formula from lecture, compute a 95% confidence interval "by hand\" for the population difference in mean cholesterol level between the heart attack and the no heart attack groups.

```{r}
# df <- 16 + 20 - 2 
# tStatistic <- 34, 0.975

tStatistic <- function(n1, n2, mu1, mu2, sd1, sd2) {   
  numerator1 <- (n1-1)*(sd1^2) + (n2-1)*(sd2^2)
  denom1 <- n1 + n2 - 2
  pooled <- sqrt( numerator1 / denom1 ) 
  
  numerator2 <- mu1 - mu2
  denom2 <- pooled * (sqrt ( (1/n1) + (1/n2) ) )
  tStatOutput <- numerator2 / denom2 
  tStatOutput
  
  }

ciEstimate <- function(n1, n2, mu1, mu2, sd1, sd2, tStat) {
  
  numerator <- (n1-1)*(sd1^2) + (n2-1)*(sd2^2)
  denom <- n1 + n2 - 2
  
  pooled <- sqrt( numerator / denom )  
  sqrtFactor <- sqrt(1/n1 + 1/n2)
  rightSide <- tStat*pooled*sqrtFactor 
  
  difference <- mu1 - mu2
  
  lb <- difference - rightSide
  ub <- difference + rightSide
  output <- c(min(lb, ub), max(lb, ub))
  output
  
}
```

```{r}
# tStatistic(n1 = 16, n2 = 20, mu1 = 265.4, mu2 = 193.1, sd1 = 43.645, sd2 = 21.623)
qt(p = 0.975, df = 34)

ciEstimate(n1 = 16, n2 = 20, mu1 = 265.4, mu2 = 193.1, sd1 = 43.645, sd2 = 21.623, tStat = 2.032245)
```

The Interval of the difference between groups (average for Heart attack - average for non-Heart attack) is (49.67554 to 94.92446) or (49.676, 94.924) difference in cholesterol levels. 

\newpage

> 3.  Interpret the confidence interval in the context of the study.

We are 95% confidence the true difference between the two groups (those who had heart attacks compared to those who didn't) is between 49.676, 94.924, meaning we believe those who had a heart attack on average had between 49.676 to 94.924 greater blood pressure compared to those who didn't have a heart attack..

> 4.  Suppose that the researchers want to replicate the study, varying the targeted sample sizes in each group to obtain the *best* (or least variable) estimate of the difference in group means, given the constraint that they can only afford to collect information from 50 total participants. How many subjects should they recruit for each group? Fill in the table below to help you answer the question (as an example, one of the solutions is already provided).

$$Var(\bar Y_1 - \bar Y_2) = \sigma^2 \left( \frac{1}{n_1} + \frac{1}{n_2} \right)$$


   $n_1$   $n_2$   $Var(\bar Y_1 - \bar Y_2) = \sigma^2 \left( \frac{1}{n_1} + \frac{1}{n_2} \right)$ \ 
  ------- ------- ------------------------------------------------------------------------------------
     1      49     1.020$\sigma^2$
     5      45     0.222$\sigma^2$
    10      40     0.125$\sigma^2$
    20      30     0.083$\sigma^2$
    25      25     0.08$\sigma^2$
    30      20     0.083$\sigma^2$
    40      10     0.125$\sigma^2$
    45       5     0.222$\sigma^2$
    49       1     1.020$\sigma^2$
  ------- ------- ------------------------------------------------------------------------------------

They should evenly divide the 50 total participants into two groups, making two groups of 25. 

\newpage

2.  **\[+10\]:** Refer to the data set `cholesterol.csv` (posted in Canvas). This file contains data on 18 randomly sampled individuals diagnosed with high cholesterol who replaced butter in their diets with a brand (A or B) of margarine. The brand of margarine was randomized. Their doctors recorded their blood cholesterol levels at the beginning of the experiment, after four weeks of their diets, and again after eight weeks. The researchers are interested in exploring the question:

> How big is the difference in mean cholesterol reduction after 8 weeks between brand A and B for subjects who replaced butter in their diet with margarine?

> 1.  Compute the 99% confidence interval in R and provide a screenshot of the output.

```{r}
library(readr)
library(dplyr)
cholesterol <- read_csv("cholesterol.csv")

differenceDf <- cholesterol |>
  mutate(difference = Before - After8weeks)

data1 <- differenceDf |>
  filter(Margarine == "A")
  
data2 <- differenceDf |>
  filter(Margarine == "B")
```

```{r}
t.test(differenceDf$difference ~ differenceDf$Margarine, conf.level = 0.99)
```

```{r}
statisticT <- qt(p = 0.995, df = 16)

ciEstimate(n1 = 9, n2 = 9, mu1 = mean(data1$difference), mu2 = mean(data2$difference), sd1 = sd(data1$difference), sd2 = sd(data2$difference), tStat = statisticT)
```

The Confidence Interval is (-0.3876, 0.0587) for the average difference between Group A and Group B where the difference in a particular group is given by "Cholesterol Before" - "Cholesterol After 8 Weeks". 

\newpage

> 2.  Interpret the confidence interval in the context of the study.

We are 99% confident that the true difference in the change in blood cholesterol levels between those who consumed Brand A vs. Brand B during an 8 week period is between (-0.3876, 0.0587) with units being the difference in change in cholesterol levels; we are 99% confident the true difference between the change in cholesterol for consuming Brand A compared to consuming Brand B is between (-0.3876, 0.0587). 

\newpage

3.  **\[+9\]:** Suppose that the researchers want to replicate the cholesterol study. Help them achieve their study design goals by performing the following sample size determinations in R (provide screenshot of output) or by hand-calculation (show work):

> 1.  Given an approximate pooled sample standard deviation of $S_p = 0.16$, what sample size is needed in each of two equally-sized treatment groups in order for the standard error of the difference in average cholesterol reduction to be no more than 0.02?

### Standard Error

Formula

$n = \frac{2S_P^2}{(s.e.)^2}$

Where s.e. is taken as an input

### Calculation

$n = \frac{2S_p^2}{(s.e.)^2}$

Where $S_p = 0.16$ and $s.e. = 0.02$ 

```{r}
n <- 2*(0.16^2)/(0.02^2)
n
```

128 sample size for each treatment group.  

\newpage

> 2.  Given an approximate pooled sample standard deviation of $S_p = 0.16$, what sample size is needed in each of two equally-sized treatment groups in order for the width of a 95% confidence interval for the true difference in mean cholesterol reduction to be no more than 0.04?

### Confidence Intervals 

Formula 

$$n_0 = 8(\frac{z_{1-\frac{\alpha}{2}}S_p}{w})^2$$

$$n = 8(\frac{t_{2(n_0-1),1-\frac{\alpha}{2}}S_p}{w})^2$$

### Calculation 

$$n_0 = 8(\frac{z_{1-\frac{\alpha}{2}}S_p}{w})^2$$

$$n = 8(\frac{t_{2(n_0-1),1-\frac{\alpha}{2}}S_p}{w})^2$$
Where 
$w = 0.04$, $S_p = 0.16$

A 95% Confidence Interval gives us: 

$\alpha = 0.05 \rightarrow z_{0.975} = 1.96$

$n_0 = 491.7248 \rightarrow t_{492, 0.975} \approx  1.984$

```{r}
qt(p = .975, df = 100)
qt(p = .975, df = 492)
```

```{r}
n0 <- 8*((0.16*1.96)/(0.04))^2

n <- 8*((0.16*1.965)/(0.04))^2
n0
n 
```

The sample size needed is 495 participants per group (using $n_0 = 492$). 

\newpage

> 3.  Given an approximate pooled sample standard deviation of $S_p = 0.16$ and an effect size of $\delta=0.03$, what sample size is needed in each of two equally-sized treatment groups in order for a level $\alpha=0.05$ two-sided test to have 80% power?

### Hypothesis Tests

Formula 

$$n_0 = \frac{(z_{1 - \frac{\alpha}{2}} + z_{1-\beta})^2(2S_p^2)}{\delta^2}$$

$$n = \frac{(t_{2(n_0-1), 1-\frac{\alpha}{2}}+ t_{2(n_0-1), 1-\beta })^2(2S_p^2)}{\delta^2}$$

### Calculation

$$n_0 = \frac{(z_{1 - \frac{\alpha}{2}} + z_{1-\beta})^2(2S_p^2)}{\delta^2}$$

$$n = \frac{(t_{2(n_0-1), 1-\frac{\alpha}{2}}+ t_{2(n_0-1), 1-\beta })^2(2S_p^2)}{\delta^2}$$

Where $\delta = 0.03$, $S_p = 0.16$, $\alpha = 0.05$, and $\beta = 0.2 \rightarrow 1-\beta = 0.8$

```{r}
z1 <- qnorm(p = 0.975)
z2 <- qnorm(p = 0.8)
```

```{r}
numerator <- (z1 + z2)^2 * (2 * 0.16^2) 
denom <- 0.03^2 
n0 <- numerator/denom
n0
```

```{r}
t1 <- qt(p = .975, df = 447)
t2 <- qt(p = .80, df = 447)

numerator <- (t1 + t2)^2 * (2 * 0.16^2) 
denom <- 0.03^2 
n <- numerator/denom
n
```

We require a sample size of 449 participants per group. 

\newpage

4.  **\[+10\]:** Refer to the data set `birthweight.csv` (posted in Canvas) containing the weights of a random sample of babies born at a certain hospital and information about whether the birth mother was a smoker.

> 1.  Using R, find a 90% confidence interval for the difference in average birthweight between babies born to mothers who smoke versus non-smoking mothers, and provide a screenshot of the output.

```{r}
birthweight <- read_csv("birthweight.csv")

birthweight1 <- birthweight |> 
  filter(smoker == 0)
birthweight2 <- birthweight |>
  filter(smoker == 1)
```

$90\% \text{ Confidence} \rightarrow \alpha = .10 \rightarrow 1- \alpha/2 = 0.95$

Using the following formula, we have: 

$$\text{Confidence Interval} \space = (\bar{Y_1} - \bar{Y_2}) \pm t_{(df - 2, 1 - \frac{\alpha}{2})} S_p \sqrt{\frac{1}{n_1} + \frac{1}{n_2}}$$

We calculate our t statistic as: 

```{r}
qt(p = .95, df = 40)
qt(p = .05, df = 40)
```

```{r}
ciEstimate(n1 = nrow(birthweight1), 
         n2 = nrow(birthweight2), 
         mu1 = mean(birthweight1$Birthweight), 
         mu2 = mean(birthweight2$Birthweight), 
         sd1 = sd(birthweight1$Birthweight), 
         sd2 = sd(birthweight2$Birthweight), 
         tStat = 1.66342)
```

Giving us a confidence interval (average non-smoking birthweight minus average smoking birthweight) of (0.077, 0.674).

\newpage

> 2.  Interpret the confidence interval in the context of the study.

We are 90% confident the true difference between non-smoking birthweights compared to smoking birthweights is between 0.077 to 0.674. This is evidence in support of the observation that smoking mothers will on average have children with lower birthweight compared to non-smoking mothers. 

**Total:** 50 points **\# correct:** **%:**
