---
title: "AnalyzeData"
output: html_document
date: "2025-05-19"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library(readxl)
library(dplyr)
library(tidyverse)
```

```{r}
voyageDat <- read_excel("C:/Users/samue/OneDrive/Desktop/Iowa_State_PS/SIDI 2025/voyages_20211020.xlsx")
```

```{r}
names(voyageDat)
```

[1] "voyageID"           "voyageRank"         "voyageName"         "port"               "sailingFrom"        "ground"            
[7] "yearOut"            "dayOut"             "yearIn"             "dayIn"              "returnCode"         "agentID"           
[13] "agent"              "bone"               "sperm"              "oil"                "customsDistrict"    "logbookExists"     
[19] "aowlLink"           "crewListLink"       "source"             "masterID"           "master"             "fate"              
[25] "birth"              "birthLocation"      "death"              "deathLocation"      "wife"               "wifeToSea"         
[31] "vitalRecordsSource" "laterOccupation"    "vesselID"           "vessel"             "rig"                "tonnage"           
[37] "builtPlace"         "builtDate"          "end"                "dennisWood"         "logbookScan" 

```{r}
summary(voyageDat)
```

names(voyageDat)[c(1, 2, 5)]

```{r}
voyageDat |> 
  group_by(voyageRank) |> 
  count() |> 
  arrange(desc(n))
```

```{r}
voyageDat |> 
  group_by(sailingFrom) |> 
  count() |> 
  arrange(desc(n))
```

Standardize Dates (Birth Date `birth`, and Death Date `death`)

Check for duplicates `voyageId` and `masterId`, these are possible "grain" of the dataset, i.e., it is a truly unique identifier

```{r}
nrow(voyageDat)
```

```{r}
length(unique(voyageDat$voyageID))
```

```{r}
length(unique(voyageDat$masterID))
```

# Thoughts

- Possible Duplication of records, not tidy 
- Also not tidy due to nesting of information in certain columns
- Lots of missing data
- Lacking information about what fields/columns mean, and their values 

