---
title: "JustQ1"
output: pdf_document
date: "2024-11-13"
---

# Q1 

Suppose that six observations of the yield (Y) of a chemical process were taken at each of four temperature levels (X) for running the process, but you are only given information on the sample means and standard deviations for the observed yields at each temperature. The summary data are

| Temperature (°C) | Sample Mean | Sample Variance | Sample Size |
|------------------|-------------|-----------------|-------------|
| 150              | 66          | 1.15            | 6           |
| 200              | 81          | 1.00            | 6           |
| 250              | 89          | 1.35            | 6           |
| 300              | 92          | 0.90            | 6           |

# Formulas 

General Form
$$
\mu = \frac{1}{n} \sum\limits_{i=1}^{n} x_i 
$$

$$
Var(X) = \sigma^2 = \frac{1}{n-1} \sum\limits_{i=1}^{n} (x_i - \bar{x})^2
$$

Week 10, Slide 10 
$$
b_0 = \bar{Y} - b_1 \bar{x}
$$

$$
b_1 = \frac{\sum\limits_{i=1}^{n} (x_i - \bar{x})(Y_i - \bar{Y})}{\sum\limits_{i=1}^{n}(x_i - \bar{x})^2} = \frac{\sum\limits_{i=1}^{n}(x_i - \bar{x})Y_i}{\sum\limits_{i=1}^{n}(x_i - \bar{x})^2}
$$

```{r}
temperature <- c(150, 200, 250, 300)
sampleMean <- c(66, 81, 89, 92)
sampleVariance <- c(1.15, 1.00, 1.35, 0.90)

temperatureMean <- mean(temperature)
tempVar <- var(temperature)
responseMean <- mean(sampleMean)

num <- sum(6 * (temperature - temperatureMean)*(sampleMean - responseMean))
denom <- sum(6 * (temperature - temperatureMean)^2) 
b1 <- num/denom
b0 <- responseMean - (b1*temperatureMean)

b1
b0 
```

Week 10 Slide 15 

$$
Var(b_0) = \frac{1}{n} + \frac{\bar{x}^2}{\sum\limits_{i=1}^{n}(x_i - \bar{x})^2}
$$

$$
Var(b_1) = \frac{1}{\sum\limits_{i=1}^{n}(x_i - \bar{x})^2}
$$

```{r}
n <- 24 
temperature <- c(150, 200, 250, 300)
sampleMean <- c(66, 81, 89, 92)
sampleVariance <- c(1.15, 1.00, 1.35, 0.90)

temperatureMean <- mean(temperature)
tempVar <- var(temperature)
responseMean <- mean(sampleMean)

varB0 <- 1/n + (temperatureMean^2 / sum(6 * (temperature - temperatureMean)^2) )
SEb0 <- sqrt(varB0)
SEb0

varB0 <- 1 / sum(6 * (temperature - temperatureMean)^2)
SEB1 <- sqrt(varB0)
SEB1
```

Week 10 Slide 23 

$$
SS_{model} = b_1^{2}\sum\limits_{i=1}^{n}(x_i - \bar{x})^2
$$

```{r}
SSModel <- b1 * sum(6 * (temperature - temperatureMean)^2)
SSModel
```

Week 10 Slide 25 

$$
SS_{model} = \sum\limits_{i=1}^{n} (\hat{Y}_i - \bar{Y}_i)^2
$$

$$
SS_{error} = \sum\limits_{i=1}^{n} (Y_i - \hat{Y}_i)^2
$$

Week 11 Slide 40 

$$
SS_{error} = SS_{pure error} + SS_{lack-of-fit}
$$

$$
SS_{lack-of-fit} = \sum\limits_i \sum\limits_j (\bar{Y}_{i.} - \hat{Y}_i)^2 
$$

$$
\hat{Y}_i = b_0 + b_1x_i
$$

```{r}
n <- 24 
temperature <- c(150, 200, 250, 300)
sampleMean <- c(66, 81, 89, 92)
sampleVariance <- c(1.15, 1.00, 1.35, 0.90)

temperatureMean <- mean(temperature)
tempVar <- var(temperature)
responseMean <- mean(sampleMean)

varB0 <- 1/n + (temperatureMean^2 / sum(6 * (temperature - temperatureMean)^2) )
SEb0 <- sqrt(varB0)
varB0 <- 1 / sum(6 * (temperature - temperatureMean)^2)
SEB1 <- sqrt(varB0)

hatY <- b0 + b1*temperature
hatYRep <- rep(hatY, each = 6)
SSLack <- sum((sampleMean - hatYRep)^2)
SSLack
```

# SS_{error} or SS_{pure error} 

$$
SS_{pure error} = \sum\limits_i \sum\limits_j (Y_{ij} - \bar{Y_i}) = \sum_{i=1}^{4} \sum_{j=1}^{6} (Y_{ij} - \bar{Y}_{i.})^2
$$

$$
s_i^2 = \frac{1}{n_i - 1} \sum_{j=1}^{n_i} (Y_{ij} - \bar{Y}_{i.})^2
$$

$$
SS_{pure error} = \sum_{i=1}^{4} \sum_{j=1}^{n_i} (Y_{ij} - \bar{Y}_{i.})^2 = \sum_{i=1}^{4} (n_i - 1) s_i^2
$$
```{r}
n <- 24 
n_i <- 6 

temperature <- c(150, 200, 250, 300)
sampleMean <- c(66, 81, 89, 92)
sampleVariance <- c(1.15, 1.00, 1.35, 0.90)

temperatureMean <- mean(temperature)
tempVar <- var(temperature)
responseMean <- mean(sampleMean)

SSPure <- sum((n_i - 1) * sampleVariance)
SSPure
```

Giving the following table 

| Source of Variation | Degrees of Freedom | Sum of Squares | Mean Square |
|---------------------|--------------------|----------------|-------------|
| Regression on X     | 1                  |  12900         | 12900       |
| Residuals           | 22                 |  4092          | 186         |
| - Lack-of-fit       | 2                  |  4070          | 2035        |
| - Pure error        | 20                 |  22            | 1.1         |
| Total               | 23                 |  16992         |             |


```{r}
temperature <- c(150, 200, 250, 300)  
sample_mean <- c(66, 81, 89, 92)      
sample_variances <- c(1.15, 1.00, 1.35, 0.90) 
n_i <- 6  

SSPE <- SSPure
SSLOF <- SSLack
SSR <- SSModel

temperatureMean <- mean(temperature)
responseMean <- mean(sample_mean)

df_regression <- 1
df_residual <- length(temperature) * n_i - 2
df_lack_of_fit <- length(temperature) - 2
df_pure_error <- df_residual - df_lack_of_fit
df_total <- length(temperature) * n_i - 1

SSE <- SSLOF + SSPE 
SST <- SSR + SSE

MSR <- SSR / df_regression
MSE <- SSE / df_residual
MSLOF <- SSLOF / df_lack_of_fit
MSPE <- SSPE / df_pure_error

anova_table <- data.frame(
  "Source of Variation" = c("Regression on X", "Residuals", "- Lack-of-fit", "- Pure error", "Total"),
  "Degrees of Freedom" = c(df_regression, df_residual, df_lack_of_fit, df_pure_error, df_total),
  "Sum of Squares" = c(SSR, SSE, SSLOF, SSPE, SST),
  "Mean Square" = c(MSR, MSE, MSLOF, MSPE, NA)
)

# Display the table
print(anova_table)
```