---
title: "HW9"
output: pdf_document
date: "Sam Olson"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Q1

Let $X_1, \ldots, X_n$ be iid exponential($\theta$) and let $\hat{\theta}_n \equiv \bar{X}_n \equiv \sum_{i=1}^n X_i/n$ denote the MLE based on $X_1, \ldots, X_n$.

## a)

Determine the limiting distribution of $\sqrt{n}(\hat{\theta}_n - \theta)$ as $n \to \infty$.

### Answer

As given, $X_1, \ldots, X_n$ are iid with $X_i \sim \text{Exponential}(\theta)$. 

This is a known distribution, such that:

$\mathbb{E}[X_i] = \theta$

And: 

$\text{Var}(X_i) = \theta^2$

By the Central Limit Theorem, we also know: 

$$
\sqrt{n}(\bar{X}_n - \theta) \overset{d}{\longrightarrow} N(0, \theta^2)
$$

Substituting values, we get our limiting distribution: 

$$
\sqrt{n}(\hat{\theta}_n - \theta) \overset{d}{\longrightarrow} N(0, \theta^2)
$$

## b)

Find a variance stabilizing transformation (VST) for $\{\hat{\theta}_n\}$ and use this to determine a large sample confidence interval for $\theta$ with approximate confidence coefficient $1 - \alpha$.

### Answer

As given, $X_1, \ldots, X_n \overset{iid}{\sim} \text{Exp}(\theta)$. Given this distribution, we know it's MLE due to meeting the regularity conditions of the CRLB, such that: $\hat{\theta}_n = \bar{X}_n$.

From part a), we know the limiting distribution is given by:

$$
\sqrt{n}(\hat{\theta}_n - \theta) \overset{d}{\longrightarrow} N(0, \theta^2)
$$

We arrive at a VST by using the Delta Method. 

To that end, define a continuous function $g(\cdot)$:

$$
\sqrt{n}(g(\hat{\theta}_n) - g(\theta)) \overset{d}{\longrightarrow} N\left(0, [g'(\theta)]^2 \theta^2\right)
$$

Where:

$$
[g'(\theta)]^2 \theta^2 = 1
$$

Isolating the function g', by taking square root, we have:

$$
g'(\theta) = \frac{1}{\theta}
$$

And integrating to solve for g:

$$
g(\theta) = \log \theta + C
$$

Where $C = 0$ for our purposes.

Thus, a VST via the Delta Method is:

$$
\sqrt{n}(\log \hat{\theta}_n - \log \theta) \overset{d}{\longrightarrow} N(0, 1)
$$

Then, fora large sample confidence interval, we may invert the test to get an approximate $1-\alpha$ confidence interval for $\log(\theta)$:

$$
\left( \log(\hat{\theta}_n) \pm \frac{z_{\alpha/2}}{\sqrt{n}} \right)
$$

Where $z_{\alpha/2}$ is the $1-\alpha/2$ standard normal quantile, and taking advantage of symmetry in the standard normal distribution (and further noting we go in order of ($L = \frac{\alpha}{2}, U = 1 - \frac{\alpha}{2}$) for the confidence interval that follows, as we have an increasing function with respect to $\theta$). 

To then isolate into just an expression of $\theta$ then, we have a confidence interval for $\theta$ with approximate confidence coefficient $1 - \alpha$ given by:

$$
\left( \hat{\theta}_n \exp\left(-\frac{z_{\alpha/2}}{\sqrt{n}}\right), \hat{\theta}_n \exp\left(\frac{z_{1 - \alpha/2}}{\sqrt{n}}\right) \right)
$$

Note: We may simplify further, noting that $\hat{\theta}_n = \bar{X}_n$

Substituting this into the above formula, we then have an equivalent expression of the confidence interval as: 

$$
\left( \bar{X}_n \exp\left(-\frac{z_{\alpha/2}}{\sqrt{n}}\right), \bar{X}_n \exp\left(\frac{z_{1 - \alpha/2}}{\sqrt{n}}\right) \right)
$$

And noting the use of ( instead of [ given the use of "approximate coverage". 

Calculating this explicitly: 

```{r}
x_bar <- 1.835464
n <- 100
z_90 <- qnorm(0.95)

lower_vst <- x_bar * exp(-z_90 / sqrt(n))
upper_vst <- x_bar * exp(z_90 / sqrt(n))

c(lower_vst, upper_vst)
```

A large sample confidence interval for $\theta$ with approximate confidence coefficient $1-\alpha$ is $\theta \in$(1.557079 2.163620). 

\newpage

# Q3

Suppose $X_1, \ldots, X_n$ are a random sample with common cdf given by

$$
P(X_1 \leq x|\theta) = 
\begin{cases}
1 - e^{-(x/\theta)^2} & \text{if } x > 0 \\
0 & \text{otherwise},
\end{cases}
\quad \theta > 0
$$

## a)

Use the Mood-Graybill-Boes Method to derive a CI for $\theta$ with C.C. $1 - \alpha$ based on the statistic $X_{(1)} = \min_{1 \leq i \leq n} X_i$.

### Answer

Since $X_1, \ldots, X_n$ are a random sample with common cdf, they are iid, such that we may write:

$$
P(X_{(1)} \leq x) = 1 - P(X_1 > x, \ldots, X_n > x) = 1 - \left(P(X_1 > x)\right)^n
$$

With the cdf as given this simplifies:

$$
P(X_1 > x) = 1 - P(X_1 \leq x|\theta) = 1 - (1 - e^{-(x/\theta)^2}) = e^{-(x/\theta)^2}
$$

By the definition of $X_{(1)}$ then:

$$
P(X_{(1)} \leq x) = 1 - e^{-n(x/\theta)^2}
$$

Let:

$$
V = n\left( \frac{X_{(1)}}{\theta} \right)^2 \rightarrow P(V \leq v) = 1 - e^{-v}
$$

The above cdf is from an Exponential distribution!

So, $V \sim \text{Exponential}(1)$, and V is a pivotal quantity.

We require quantiles of the $\text{Exponential}(1)$ distribution in order to form confidence intervals. 

Let $q_p$ denote the $p$-th quantile of the $\text{Exponential}(1)$ distribution.

We want coverage coefficient:

$$
P_{\theta}\left( q_{\alpha/2} \leq V \leq q_{1-\alpha/2} \right) = 1-\alpha
$$

In terms of $\theta$, solving:

$$
q_{\alpha/2} \leq n\left( \frac{X_{(1)}}{\theta} \right)^2 \leq q_{1-\alpha/2}
\rightarrow
\sqrt{ \frac{q_{\alpha/2}}{n} } \leq \frac{X_{(1)}}{\theta} \leq \sqrt{ \frac{q_{1-\alpha/2}}{n}}
$$

After some more algebra, we have:

$$
\theta \in \left( \frac{X_{(1)}}{ \sqrt{q_{1-\alpha/2}/n} }, \frac{X_{(1)}}{ \sqrt{q_{\alpha/2}/n} } \right)
$$

I believe the proof may end here, so leaving some space before continuing...

---

We may simplify further with a note: 

By definition, the $p$-th quantile $q_p$ satisfies the expression:

$$
P(X \leq q_p) = p
$$

Substituting the given cdf:

$$
F(q_p) = p \rightarrow 
1 - e^{-q_p} = p
$$

Rearranging:

$$
e^{-q_p} = 1 - p \rightarrow - q_p = \log(1-p) \rightarrow q_p = -\log(1-p)
$$

So, we may write the $p$-th quantile $q_p$ of an $\text{Exponential}(1)$ random variable as:

$$
q_p = -\log(1-p)
$$

So, taking our CI for $\theta$ given above, we may then write: 

For a confidence coefficient of $1 - \alpha$, we have: 

$$
q_{1-\alpha/2} = -\log(\alpha/2)
$$

And: 

$$
q_{\alpha/2} = -\log(1-\alpha/2)
$$

Substituting into the above interval calculation:

$$
\theta \in \left( \frac{X_{(1)}}{ \sqrt{ \frac{-\log(\alpha/2)}{n} } }, \frac{X_{(1)}}{ \sqrt{ \frac{-\log(1-\alpha/2)}{n} } } \right)
$$

Which is equivalent to:

$$
\theta \in \left( X_{(1)} \sqrt{ \frac{n}{-\log(\alpha/2)} }, \, X_{(1)} \sqrt{ \frac{n}{-\log(1-\alpha/2)} } \right)
$$

## b)

Use the Mood-Graybill-Boes Method to derive a CI for $\theta$ with C.C. $1 - \alpha$ based on the statistic $T = \sum_{i=1}^n X_i^2$. Express your confidence interval using chi-squared quantiles.

Note: One can show $X_i^2$ is Exponential($\theta^2$) distributed so that $2T/\theta^2$ is $\chi_{2n}^2$ distributed with $2n$ degrees of freedom.

### Answer

As given, we know:

$$
X_i^2 \sim \text{Exponential}(\theta^2)
$$

Note: 

$$
T = \sum_{i=1}^n X_i^2 \rightarrow T \sim \text{Gamma}(n, \theta^2)
$$

Using the above note, let:

$$
T = \sum_{i=1}^n X_i^2 \rightarrow \frac{2T}{\theta^2} \sim \chi^2_{2n}
$$

Where T is a pivotal quantity. 

Regarding the coverage coefficient, we define:

$$
P_{\theta}\left( \chi^2_{2n,\alpha/2} \leq \frac{2T}{\theta^2} \leq \chi^2_{2n,1-\alpha/2} \right) = 1-\alpha
$$

Solving for $\theta^2$:

$$
\frac{2T}{\chi^2_{2n,1-\alpha/2}} \leq \theta^2 \leq \frac{2T}{\chi^2_{2n,\alpha/2}}
\rightarrow
\sqrt{ \frac{2T}{\chi^2_{2n,1-\alpha/2}} } \leq \theta \leq \sqrt{ \frac{2T}{\chi^2_{2n,\alpha/2}}}
$$

Thus, the confidence interval for $\theta$ is:

$$
\left( \sqrt{ \frac{2 \sum_{i=1}^n X_i^2}{\chi^2_{2n,1-\alpha/2}} }, \, \sqrt{ \frac{2 \sum_{i=1}^n X_i^2}{\chi^2_{2n,\alpha/2}} } \right)
$$

With the desired coverage $1 - \alpha$. 