---
title: "HW9"
output: pdf_document
date: "Sam Olson"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Q1

Let $X_1, \ldots, X_n$ be iid exponential($\theta$) and let $\hat{\theta}_n \equiv \bar{X}_n \equiv \sum_{i=1}^n X_i/n$ denote the MLE based on $X_1, \ldots, X_n$.

## a)

Determine the limiting distribution of $\sqrt{n}(\hat{\theta}_n - \theta)$ as $n \to \infty$.

### Answer

As given, $X_1, \ldots, X_n$ are iid with $X_i \sim \text{Exponential}(\theta)$. 

This is a known distribution, such that:

$\mathbb{E}[X_i] = \theta$

And: 

$\text{Var}(X_i) = \theta^2$

By the Central Limit Theorem, we also know: 

$$
\sqrt{n}(\bar{X}_n - \theta) \overset{d}{\longrightarrow} N(0, \theta^2)
$$

Substituting values, we get our limiting distribution: 

$$
\sqrt{n}(\hat{\theta}_n - \theta) \overset{d}{\longrightarrow} N(0, \theta^2)
$$

## b)

Find a variance stabilizing transformation (VST) for $\{\hat{\theta}_n\}$ and use this to determine a large sample confidence interval for $\theta$ with approximate confidence coefficient $1 - \alpha$.

### Answer

As given, $X_1, \ldots, X_n \overset{iid}{\sim} \text{Exp}(\theta)$. Given this distribution, we know it's MLE due to meeting the regularity conditions of the CRLB, such that: $\hat{\theta}_n = \bar{X}_n$.

From part a), we know the limiting distribution is given by:

$$
\sqrt{n}(\hat{\theta}_n - \theta) \overset{d}{\longrightarrow} N(0, \theta^2)
$$

We arrive at a VST by using the Delta Method. 

To that end, define a continuous function $g(\cdot)$:

$$
\sqrt{n}(g(\hat{\theta}_n) - g(\theta)) \overset{d}{\longrightarrow} N\left(0, [g'(\theta)]^2 \theta^2\right)
$$

Where:

$$
[g'(\theta)]^2 \theta^2 = 1
$$

Isolating the function g', by taking square root, we have:

$$
g'(\theta) = \frac{1}{\theta}
$$

And integrating to solve for g:

$$
g(\theta) = \log \theta + C
$$

Where $C = 0$ for our purposes.

Thus, a VST via the Delta Method is:

$$
\sqrt{n}(\log \hat{\theta}_n - \log \theta) \overset{d}{\longrightarrow} N(0, 1)
$$

Then, fora large sample confidence interval, we may invert the test to get an approximate $1-\alpha$ confidence interval for $\log(\theta)$:

$$
\left( \log(\hat{\theta}_n) \pm \frac{z_{\alpha/2}}{\sqrt{n}} \right)
$$

Where $z_{\alpha/2}$ is the $1-\alpha/2$ standard normal quantile.

To then isolate into just an expression of $\theta$ then, we have the approximate confidence interval for $\theta$ with approximate confidence coefficient $1 - \alpha$ is:

$$
\left( \hat{\theta}_n \exp\left(-\frac{z_{\alpha/2}}{\sqrt{n}}\right), \hat{\theta}_n \exp\left(\frac{z_{\alpha/2}}{\sqrt{n}}\right) \right)
$$

Noting the use of ( instead of [ given the use of "approximate coverage". 

## c)

Suppose a random sample $X_1, \ldots, X_{100}$ of $n = 100$ observations yields $\bar{x}_n = 1.835464$. Use this information to obtain a large sample confidence interval for $\theta$ based on a likelihood ratio statistic, which has approximate confidence coefficient $90\%$. (Use the chi-squared approximation for this; you should be able to then numerically determine the interval.) 

Using this data, compute also a confidence interval with approximate confidence coefficient $90\%$ using the VST approach from part b).

### Answer

For testing $H_0: \theta = \theta_0$, the likelihood ratio statistic satisfies:

$$
-2 \log \Lambda(\theta) \overset{d}{\longrightarrow} \chi^2_1
$$

where:

$$
\Lambda(\theta) = \frac{L(\theta)}{L(\hat{\theta}_n)}
$$

and degrees of freedom 1 (since $\theta$ is a single parameter).

Since the data are iid Exponential$(\theta)$, the joint likelihood function is:

$$
L(\theta) = \prod_{i=1}^n \left( \frac{1}{\theta} e^{-x_i/\theta} \right) = \left( \frac{1}{\theta} \right)^n \exp\left( -\frac{1}{\theta} \sum_{i=1}^n x_i \right)
$$

Taking the log (noting that the logarithm is a monotonic transformation and preserves maxima):

$$
\ell(\theta) = \log L(\theta) = -n \log \theta - \frac{1}{\theta} \sum_{i=1}^n x_i
$$

Since:

$$
\sum_{i=1}^n x_i = n \bar{x}_n
$$

the log-likelihood simplifies to:

$$
\ell(\theta) = -n \log \theta - \frac{n\bar{x}_n}{\theta}
$$

Thus, the likelihood ratio test statistic simplifies to:

$$
-2 \log \Lambda(\theta) = 2n \left[ \log\left( \frac{\theta}{\hat{\theta}_n} \right) + \frac{\hat{\theta}_n}{\theta} - 1 \right],
$$

Where $\hat{\theta}_n = \bar{X}_n$ is the MLE for $\theta$.

We then have:

$$
2n \left[ \log\left( \frac{\theta}{\hat{\theta}_n} \right) + \frac{\hat{\theta}_n}{\theta} - 1 \right] \leq \chi^2_{1,0.90}
$$

Where $\chi^2_{1,0.90} \approx 2.7055$.

We have everything we then need to construct a confidence interval, using $n = 100$ observations with $\bar{x}_n = 1.835464$. 

We have the inequality:

$$
2n\left[\log\left(\frac{\theta}{\bar{x}_n}\right) + \frac{\bar{x}_n}{\theta} - 1\right] \leq 2.7055.
$$

Our goal then is to isolate into $\theta$ terms, when possible. To that end: 

$$
\log\left(\frac{\theta}{1.835464}\right) + \frac{1.835464}{\theta} - 1 \leq 0.0135275
\rightarrow 
\log\left(\frac{\theta}{1.835464}\right) + \frac{1.835464}{\theta} = 1.0135275
$$

This is rather tricky to solve for analytically! I will turn to the computer for help. 

```{r}
n <- 100
x_bar <- 1.835464
chi_sq_crit <- qchisq(0.90, df = 1)

rhs_value <- chi_sq_crit / (2 * n)

lrt_equation <- function(theta) log(theta/x_bar) + (x_bar/theta) - (1 + rhs_value)

lower_theta <- uniroot(lrt_equation, lower = 0.5, upper = x_bar)$root
upper_theta <- uniroot(lrt_equation, lower = x_bar, upper = 5)$root

c(lower_theta, upper_theta)
```

Interestingly, this is similar (but different) to the interval via the VST method. 

#### Aside (VST Method)

Using the VST approach from part b): 

```{r}
x_bar <- 1.835464
n <- 100
z_90 <- qnorm(0.95)

lower_vst <- x_bar * exp(-z_90 / sqrt(n))
upper_vst <- x_bar * exp(z_90 / sqrt(n))

c(lower_vst, upper_vst)
```

\newpage

# Q3

Suppose $X_1, \ldots, X_n$ are a random sample with common cdf given by

$$
P(X_1 \leq x|\theta) = 
\begin{cases}
1 - e^{-(x/\theta)^2} & \text{if } x > 0 \\
0 & \text{otherwise},
\end{cases}
\quad \theta > 0
$$

## a)

Use the Mood-Graybill-Boes Method to derive a CI for $\theta$ with C.C. $1 - \alpha$ based on the statistic $X_{(1)} = \min_{1 \leq i \leq n} X_i$.

### Answer

Since $X_1, \ldots, X_n$ are a random sample with common cdf, they are iid, such that we may write:

$$
P(X_{(1)} \leq x) = 1 - P(X_1 > x, \ldots, X_n > x) = 1 - \left(P(X_1 > x)\right)^n
$$

With the cdf as given this simplifies:

$$
P(X_1 > x) = 1 - P(X_1 \leq x|\theta) = 1 - (1 - e^{-(x/\theta)^2}) = e^{-(x/\theta)^2}
$$

By the definition of $X_{(1)}$ then:

$$
P(X_{(1)} \leq x) = 1 - e^{-n(x/\theta)^2}
$$

Let:

$$
V = n\left( \frac{X_{(1)}}{\theta} \right)^2 \rightarrow P(V \leq v) = 1 - e^{-v}
$$

The above cdf is from an Exponential distribution!

So, $V \sim \text{Exponential}(1)$, and V is a pivotal quantity.

We require quantiles of the $\text{Exponential}(1)$ distribution in order to form confidence intervals. 

Let $q_p$ denote the $p$-th quantile of the $\text{Exponential}(1)$ distribution.

We want coverage coefficient:

$$
P_{\theta}\left( q_{\alpha/2} \leq V \leq q_{1-\alpha/2} \right) = 1-\alpha
$$

In terms of $\theta$, solving:

$$
q_{\alpha/2} \leq n\left( \frac{X_{(1)}}{\theta} \right)^2 \leq q_{1-\alpha/2}
\rightarrow
\sqrt{ \frac{q_{\alpha/2}}{n} } \leq \frac{X_{(1)}}{\theta} \leq \sqrt{ \frac{q_{1-\alpha/2}}{n}}
$$

After some more algebra, we have:

$$
\theta \in \left( \frac{X_{(1)}}{ \sqrt{q_{1-\alpha/2}/n} }, \frac{X_{(1)}}{ \sqrt{q_{\alpha/2}/n} } \right)
$$

I believe the proof may end here, so leaving some space before continuing...

---

We may simplify further with a note: 

By definition, the $p$-th quantile $q_p$ satisfies the expression:

$$
P(X \leq q_p) = p
$$

Substituting the given cdf:

$$
F(q_p) = p
\quad \Rightarrow \quad
1 - e^{-q_p} = p
$$

Rearranging:

$$
e^{-q_p} = 1 - p \rightarrow - q_p = \log(1-p) \rightarrow q_p = -\log(1-p)
$$

So, we may write the $p$-th quantile $q_p$ of an $\text{Exponential}(1)$ random variable as:

$$
q_p = -\log(1-p)
$$

So, taking our CI for $\theta$ given above, we may then write: 

For a confidence coefficient of $1 - \alpha$, we have: 

$$
q_{1-\alpha/2} = -\log(\alpha/2)
$$

And: 

$$
q_{\alpha/2} = -\log(1-\alpha/2)
$$

Substituting into the above interval:

$$
\theta \in \left( \frac{X_{(1)}}{ \sqrt{ \frac{-\log(\alpha/2)}{n} } }, \frac{X_{(1)}}{ \sqrt{ \frac{-\log(1-\alpha/2)}{n} } } \right)
$$

Which is equivalent to:

$$
\theta \in \left( X_{(1)} \sqrt{ \frac{n}{-\log(\alpha/2)} }, \, X_{(1)} \sqrt{ \frac{n}{-\log(1-\alpha/2)} } \right)
$$

## b)

Use the Mood-Graybill-Boes Method to derive a CI for $\theta$ with C.C. $1 - \alpha$ based on the statistic $T = \sum_{i=1}^n X_i^2$. Express your confidence interval using chi-squared quantiles.

Note: One can show $X_i^2$ is Exponential($\theta^2$) distributed so that $2T/\theta^2$ is $\chi_{2n}^2$ distributed with $2n$ degrees of freedom.

### Answer

As given, we know:

$$
X_i^2 \sim \text{Exponential}(\theta^2)
$$

Note: 

$$
T = \sum_{i=1}^n X_i^2 \rightarrow T \sim \text{Gamma}(n, \theta^2)
$$

Using the above note, let:

$$
T = \sum_{i=1}^n X_i^2 \rightarrow \frac{2T}{\theta^2} \sim \chi^2_{2n}
$$

Where T is a pivotal quantity. 

Regarding the coverage coefficient, we define:

$$
P_{\theta}\left( \chi^2_{2n,\alpha/2} \leq \frac{2T}{\theta^2} \leq \chi^2_{2n,1-\alpha/2} \right) = 1-\alpha
$$

Solving for $\theta^2$:

$$
\frac{2T}{\chi^2_{2n,1-\alpha/2}} \leq \theta^2 \leq \frac{2T}{\chi^2_{2n,\alpha/2}}
\rightarrow
\sqrt{ \frac{2T}{\chi^2_{2n,1-\alpha/2}} } \leq \theta \leq \sqrt{ \frac{2T}{\chi^2_{2n,\alpha/2}}}
$$

Thus, the confidence interval for $\theta$ is:

$$
\left( \sqrt{ \frac{2 \sum_{i=1}^n X_i^2}{\chi^2_{2n,1-\alpha/2}} }, \, \sqrt{ \frac{2 \sum_{i=1}^n X_i^2}{\chi^2_{2n,\alpha/2}} } \right)
$$

With the desired coverage $1 - \alpha$. 