---
title: "HW2"
author: "Sam Olson"
output: pdf_document
header-includes:
  - \newcommand{\ve}[1]{\underline{#1}}
  - \newcommand{\bb}{\ve{\beta}}
  - \newcommand{\bhat}{\widehat{\ve{\beta}}}
  - \newcommand{\lam}{\ve{\lambda}}
  - \newcommand{\y}{\ve{y}}
  - \newcommand{\av}{\ve{a}}
  - \newcommand{\bv}{\ve{b}}
  - \newcommand{\dv}{\ve{d}}
  - \newcommand{\hv}{\ve{h}}
  - \newcommand{\rv}{\ve{r}}
  - \newcommand{\tv}{\ve{t}}
  - \newcommand{\wv}{\ve{w}}
  - \newcommand{\zv}{\ve{z}}
  - \newcommand{\rhov}{\ve{\rho}}
  - \newcommand{\eps}{\ve{\epsilon}}
  - \newcommand{\zero}{\ve{0}}
  - \newcommand{\one}{\ve{1}}
  - \newcommand{\tr}{^{\prime}}
  - \newcommand{\C}{\mathcal{C}}
  - \newcommand{\R}{\mathcal{R}}
  - \newcommand{\PX}{P_X}
  - \newcommand{\Real}{\mathbb{R}}
  - \newcommand{\rank}{\operatorname{rank}}
  - \newcommand{\tra}{\operatorname{tr}}
  - \newcommand{\Var}{\operatorname{Var}}
  - \newcommand{\E}{\operatorname{E}}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Q1

Suppose that $\y$ is an $n \times 1$ observable random vector that follows Model II.

## (a)

If $\lam_1\tr\bb$ and $\lam_2\tr\bb$ are both estimable, is $\lam_1\tr\bb + \lam_2\tr\bb$ necessarily estimable? Explain your answer.

**Answer.** Yes.

Note: By definition, $\lam\tr\bb$ is *estimable* if there exists a linear unbiased estimator of it, that is, some $c + \av\tr\y$ with $\E(c + \av\tr\y) = \lam\tr\bb$ for all $\bb$.

With this definition in hand, also note:

- **Lemma 2.1**: Under Models II--VII, a linear estimator $c + \av\tr\y$ of $\lam\tr\bb$ is unbiased if and only if $c = 0$ and $\lam\tr = \av\tr X$.
- **Theorem 2.1**: Under Models II--VII, $\lam\tr\bb$ is estimable if and only if $\lam\tr = \av\tr X$ for some $n \times 1$ vector $\av$; said differently, $\lam\tr\bb$ is estimable if and only if $\lam\tr \in \R(X)$.

Since $\lam_1\tr\bb$ and $\lam_2\tr\bb$ are estimable, Theorem 2.1 gives $n \times 1$ vectors $\av_1$ and $\av_2$ with $\lam_1\tr = \av_1\tr X$ and $\lam_2\tr = \av_2\tr X$.

It then follows that $\lam_1\tr\bb + \lam_2\tr\bb = (\lam_1 + \lam_2)\tr\bb$ is estimable by Theorem 2.1, as

$$
\lam_1\tr + \lam_2\tr = (\av_1 + \av_2)\tr X \in \R(X).
$$
Then, by Lemma 2.1, $(\av_1 + \av_2)\tr\y$ is a linear unbiased estimator of $(\lam_1 + \lam_2)\tr\bb$.

## (b)

If $\lam_1\tr\bb$ and $\lam_2\tr\bb$ are both nonestimable, is $\lam_1\tr\bb + \lam_2\tr\bb$ necessarily nonestimable? Explain your answer.

**Answer.** No. We provide proof via counterexample.

Take $p = 2$ and $X = \one_n \one_2\tr$, the $n \times 2$ matrix whose every row is $(1, 1)$, so that $y_i = \beta_1 + \beta_2 + \epsilon_i$.

For any $n \times 1$ vector $\av$, $\av\tr X = (\av\tr\one_n)\one_2\tr$, so
$$
\R(X) = \{t\,\one_2\tr : t \in \Real\} = \{(t, t) : t \in \Real\}.
$$
Let $\lam_1 = (1, 0)\tr$ and $\lam_2 = (0, 1)\tr$. Then $\lam_1\tr = (1, 0) \notin \R(X)$ and $\lam_2\tr = (0, 1) \notin \R(X)$, so by Theorem 2.1, $\lam_1\tr\bb = \beta_1$ and $\lam_2\tr\bb = \beta_2$ are both nonestimable.

However, $\lam_1\tr\bb + \lam_2\tr\bb = \beta_1 + \beta_2$ is estimable by Theorem 2.1, as

$$
\lam_1\tr + \lam_2\tr = (1, 1) \in \R(X).
$$

The general statement then is: If $\lam\tr\bb$ is nonestimable, then so is $-\lam\tr\bb$ (as $\lam\tr \notin \R(X)$ implies $-\lam\tr \notin \R(X)$), yet $\lam\tr\bb + (-\lam\tr\bb) = 0$ is trivially estimable. So a counterexample exists for any $X$ with $\rank(X) < p$.

## (c)

Can $\lam_1\tr\bb + \lam_2\tr\bb$ be estimable if $\lam_1\tr\bb$ is estimable and $\lam_2\tr\bb$ is nonestimable? Explain your answer.

**Answer.** No.

Note: By **Corollary 2 to Theorem 2.1**, any linear combination of estimable functions is estimable.

For proof, suppose, for contradiction, that $\lam_1\tr\bb + \lam_2\tr\bb$ is estimable. Since $\lam_1\tr\bb$ is estimable by assumption,

$$
\lam_2\tr\bb = (1)\big(\lam_1\tr\bb + \lam_2\tr\bb\big) + (-1)\big(\lam_1\tr\bb\big)
$$
is a linear combination of estimable functions, and so it is estimable by Corollary 2 to Theorem 2.1.

This contradicts the assumption that $\lam_2\tr\bb$ is nonestimable, so we must conclude that $\lam_1\tr\bb + \lam_2\tr\bb$ is nonestimable.

# Q2

Take the model to be Model VII and suppose that

$$
\y = \begin{pmatrix} 2 \\ 3 \\ 4 \end{pmatrix}
\quad \text{and} \quad
X = \begin{pmatrix} 1 & 2 & 3 \\ 2 & 3 & 1 \\ 0 & 0 & 0 \end{pmatrix}.
$$

## (a)

Find the orthogonal projections of $\y$ on $\C(X)$ and on $\C^{\perp}(X)$.

**Answer.** The projections are $(2, 3, 0)\tr$ on $\C(X)$ and $(0, 0, 4)\tr$ on $\C^{\perp}(X)$.

Note the following:

- $\C(X) = \{X\bv : \bv \in \Real^3\}$ is the span of the columns of $X$,
- $\C^{\perp}(X) = \{\av \in \Real^3 : X\tr\av = \zero\}$ is the null space of $X\tr$, and
- **Theorem 2.6**: Any $n \times 1$ vector $\y$ admits a unique decomposition $\y = \y_1 + \y_2$ with $\y_1 \in \C(X)$ and $\y_2 \in \C^{\perp}(X)$. What's more, this decomposition is given by $\y_1 = X\bhat$ and $\y_2 = \y - X\bhat$ for any solution $\bhat$ to the normal equations. And the vectors $\y_1$ and $\y_2$ are the orthogonal projections of $\y$ on $\C(X)$ and $\C^{\perp}(X)$, resp.

Now, to answer the question at hand:

*Finding $\C(X)$.* The columns of $X$ are
$$
\ve{x}_1 = (1, 2, 0)\tr, \qquad \ve{x}_2 = (2, 3, 0)\tr, \qquad \ve{x}_3 = (3, 1, 0)\tr.
$$
Each has third coordinate $0$, so $\C(X) \subseteq \{(u, v, 0)\tr : u, v \in \Real\}$. For the reverse inclusion, note that
$$
-3\ve{x}_1 + 2\ve{x}_2 = (1, 0, 0)\tr \qquad \text{and} \qquad 2\ve{x}_1 - \ve{x}_2 = (0, 1, 0)\tr
$$
both lie in $\C(X)$, and every $(u, v, 0)\tr$ is a linear combination of these two. Hence
$$
\C(X) = \{(u, v, 0)\tr : u, v \in \Real\}.
$$

*Finding $\C^{\perp}(X)$.* For $\av = (a_1, a_2, a_3)\tr$,
$$
X\tr\av = (a_1 + 2a_2,\ 2a_1 + 3a_2,\ 3a_1 + a_2)\tr.
$$
Since $\det\begin{pmatrix} 1 & 2 \\ 2 & 3 \end{pmatrix} = -1 \neq 0$, the first two coordinates vanish only if $a_1 = a_2 = 0$. The third then vanishes as well, and $a_3$ is unrestricted. Hence
$$
\C^{\perp}(X) = \{(0, 0, w)\tr : w \in \Real\}.
$$

*The projections.* Splitting $\y$ accordingly,
$$
\y = \begin{pmatrix} 2 \\ 3 \\ 0 \end{pmatrix} + \begin{pmatrix} 0 \\ 0 \\ 4 \end{pmatrix},
$$
with the first vector in $\C(X)$ and the second in $\C^{\perp}(X)$. By the uniqueness in Theorem 2.6, these are the orthogonal projections of $\y$ on $\C(X)$ and $\C^{\perp}(X)$, resp.

## (b)

Determine $\rank(X\tr X)$.

**Answer.** $\rank(X\tr X) = 2$.

Note: By **Lemma 2.3**, $\rank(A\tr A) = \rank(A)$ for any matrix $A$.

From part (a), $\C(X)$ is spanned by the two linearly independent vectors $(1, 0, 0)\tr$ and $(0, 1, 0)\tr$, so $\rank(X) = \dim \C(X) = 2$. It then follows from Lemma 2.3, with $A = X$, that
$$
\rank(X\tr X) = \rank(X) = 2.
$$

## (c)

Use the following method and find a generalized inverse of $X\tr X$.

Let $A$ be an $n \times m$ matrix of rank $r \geq 1$. Let $B$ be a nonsingular $r \times r$ submatrix of $A$. For matrix $A$, replace the entries of $B$ by those of $(B^{-1})\tr$, and replace all other entries of $A$ by $0$. Then the transpose of the resulting matrix is a generalized inverse of $A$.

**Answer.** We apply the method with $A = X\tr X$ and $B$ its leading $2 \times 2$ submatrix.

Note: By definition, a *generalized inverse* of a matrix $A$ is any matrix $G$ satisfying $AGA = A$.

Here
$$
A = X\tr X
= \begin{pmatrix} 1 & 2 & 0 \\ 2 & 3 & 0 \\ 3 & 1 & 0 \end{pmatrix}
  \begin{pmatrix} 1 & 2 & 3 \\ 2 & 3 & 1 \\ 0 & 0 & 0 \end{pmatrix}
= \begin{pmatrix} 5 & 8 & 5 \\ 8 & 13 & 9 \\ 5 & 9 & 10 \end{pmatrix},
$$
which has rank $r = 2$ by part (b). Then
$$
B = \begin{pmatrix} 5 & 8 \\ 8 & 13 \end{pmatrix}, \qquad
\det(B) = 65 - 64 = 1 \neq 0, \qquad
(B^{-1})\tr = B^{-1} = \begin{pmatrix} 13 & -8 \\ -8 & 5 \end{pmatrix}.
$$
Replacing the entries of $B$ in $A$ by those of $(B^{-1})\tr$, and all other entries by $0$, gives
$$
G = (X\tr X)^{-} = \begin{pmatrix} 13 & -8 & 0 \\ -8 & 5 & 0 \\ 0 & 0 & 0 \end{pmatrix}.
$$
Side Note: Since $B$ is symmetric, $(B^{-1})\tr = B^{-1}$; the resulting matrix is then symmetric too, so the final transpose leaves it unchanged.

*Check.* We have
$$
AG = \begin{pmatrix} 1 & 0 & 0 \\ 0 & 1 & 0 \\ -7 & 5 & 0 \end{pmatrix}
\qquad \text{and} \qquad
AGA = \begin{pmatrix} 5 & 8 & 5 \\ 8 & 13 & 9 \\ 5 & 9 & 10 \end{pmatrix} = A,
$$
so $G$ is a generalized inverse of $X\tr X$.

## (d)

Compute a solution $\bhat$ to the normal equations.

**Answer.** $\bhat = (0, 1, 0)\tr$.

Note the following:

- **Theorem 2.3**: The normal equations $X\tr X\bhat = X\tr\y$ are consistent.
- **Theorem 1.2**: A matrix $G$ is a generalized inverse of a matrix $A$ if and only if $\wv = G\ve{c}$ solves $A\wv = \ve{c}$ for every vector $\ve{c}$ for which these equations are consistent.

With these in hand, since the normal equations are consistent (Theorem 2.3), Theorem 1.2 with $A = X\tr X$ and $\ve{c} = X\tr\y$ gives that $\bhat = (X\tr X)^{-}X\tr\y$ solves them. Since
$$
X\tr\y = \begin{pmatrix} 1 & 2 & 0 \\ 2 & 3 & 0 \\ 3 & 1 & 0 \end{pmatrix}\begin{pmatrix} 2 \\ 3 \\ 4 \end{pmatrix} = \begin{pmatrix} 8 \\ 13 \\ 9 \end{pmatrix},
$$
taking $(X\tr X)^{-} = G$ from part (c) gives
$$
\bhat = \begin{pmatrix} 13 & -8 & 0 \\ -8 & 5 & 0 \\ 0 & 0 & 0 \end{pmatrix}
  \begin{pmatrix} 8 \\ 13 \\ 9 \end{pmatrix}
= \begin{pmatrix} 104 - 104 \\ -64 + 65 \\ 0 \end{pmatrix}
= \begin{pmatrix} 0 \\ 1 \\ 0 \end{pmatrix}.
$$
As a direct check, $X\tr X\bhat$ is the second column of $X\tr X$, namely $(8, 13, 9)\tr = X\tr\y$.

## (e)

Check that the vectors $X\bhat$ and $\y - X\bhat$ agree with your results in part (a).

**Answer.** They agree. From part (d), $X\bhat$ is the second column of $X$, so
$$
X\bhat = \begin{pmatrix} 2 \\ 3 \\ 0 \end{pmatrix}, \qquad
\y - X\bhat = \begin{pmatrix} 0 \\ 0 \\ 4 \end{pmatrix},
$$
which match the projections in part (a), as Theorem 2.6 requires.

A quick numerical check of parts (c)--(e) in R:

```{r q2-check}
X <- matrix(c(1, 2, 3,
              2, 3, 1,
              0, 0, 0), nrow = 3, byrow = TRUE)
y <- c(2, 3, 4)
XtX <- crossprod(X)
G <- matrix(c(13, -8, 0,
              -8,  5, 0,
               0,  0, 0), nrow = 3, byrow = TRUE)

all.equal(XtX %*% G %*% XtX, XtX)  # G is a generalized inverse of X'X
bhat <- drop(G %*% crossprod(X, y))
bhat
drop(X %*% bhat)                    # projection onto C(X)
drop(y - X %*% bhat)                # projection onto C-perp(X)
```

# Q3

For Model VII, partition $\y$, $X$, and $\eps$ correspondingly as follows:
$$
\y = \begin{pmatrix} \y_1 \\ \y_2 \end{pmatrix}, \quad
X = \begin{pmatrix} X_1 \\ X_2 \end{pmatrix}, \quad \text{and} \quad
\eps = \begin{pmatrix} \eps_1 \\ \eps_2 \end{pmatrix}.
$$
Suppose that the row spaces of $X_1$ and $X_2$ are essentially disjoint, that is, $\R(X_1) \cap \R(X_2) = \{\zero\tr\}$. Show that, if $\lam\tr \in \R(X_1)$, then the least squares estimate of $\lam\tr\bb$ based on the model $\y_1 = X_1\bb + \eps_1$ is the same as the least squares estimate of $\lam\tr\bb$ based on the model $\y = X\bb + \eps$.

*Hint: Under the full model, the LSE is $\av\tr X\tr\y = \av\tr X_1\tr\y_1 + \av\tr X_2\tr\y_2$ and, under the partial model, the LSE is $\bv\tr X_1\tr\y_1$ where here $X_1\tr X_1\av + X_2\tr X_2\av = X\tr X\av = \lam = X_1\tr X_1\bv$ (why?); in the last equation, move terms involving $X_2\tr$ and $X_1\tr$ to different sides.*

**Answer.** We provide proof by writing both least squares estimates in terms of solutions to conjugate normal equations, then using the disjointness of $\R(X_1)$ and $\R(X_2)$ to show they coincide.

Note: $X_1$ is $n_1 \times p$ and $X_2$ is $n_2 \times p$, with $n_1 + n_2 = n$. The partial model $\y_1 = X_1\bb + \eps_1$ is itself an instance of Model VII, as $\E(\eps_1) = \zero_{n_1}$ and $\Var(\eps_1)$ is the leading $n_1 \times n_1$ block of $\Var(\eps)$. So every Model II--VII result below applies to both models.

Side Note: The hint's $\av$ and $\bv$ are $p \times 1$; we write them as $\rv$ and $\rv_1$ to keep $\av$ reserved for $n \times 1$ coefficient vectors.

With this in hand, also note the following:

- **Theorem 2.5**: Under Models II--VII, $\lam\tr\bhat = \lam\tr\widetilde{\ve{\beta}}$ for any two solutions $\bhat$ and $\widetilde{\ve{\beta}}$ to the normal equations if and only if $\lam\tr\bb$ is estimable.
- **Theorem 2.7**: Under Models II--VII, $\lam\tr\bb$ is estimable if and only if the conjugate normal equations $X\tr X\rv = \lam$ are consistent.
- **Theorem 2.8**: Under Models II--VII, the least squares estimate of an estimable $\lam\tr\bb$ equals $\rv\tr X\tr\y$, where $\rv$ is any solution to $X\tr X\rv = \lam$.
- **Lemma 1.2**: For any matrix $M$, $\tra(M\tr M) = 0$ implies $M = 0$.
- **Lemma 1.3(ii)**: For any matrices $M$, $P$, and $Q$ of appropriate dimensions, $M\tr MP = M\tr MQ$ implies $MP = MQ$.

*Estimability.* Since $\lam\tr \in \R(X_1)$, we have $\lam\tr = \tv\tr X_1$ for some $n_1 \times 1$ vector $\tv$. Hence
$$
\lam\tr = (\tv\tr, \zero_{n_2}\tr)\begin{pmatrix} X_1 \\ X_2 \end{pmatrix} \in \R(X),
$$
so $\lam\tr\bb$ is estimable under both models by Theorem 2.1 (see Q1(a)), and by Theorem 2.5 its least squares estimate is well defined under *each* model. The two models have different normal equations, though, so it remains to show that the two estimates coincide.

*The two estimates (the hint's "why?").* Since $\lam\tr\bb$ is estimable under both models, Theorem 2.7 gives $p \times 1$ vectors $\rv$ and $\rv_1$ with
$$
X\tr X\rv = \lam \qquad \text{and} \qquad X_1\tr X_1\rv_1 = \lam,
$$
and by Theorem 2.8, the least squares estimates under the full and partial models are $\rv\tr X\tr\y$ and $\rv_1\tr X_1\tr\y_1$, resp. By block multiplication,
$$
X\tr X = X_1\tr X_1 + X_2\tr X_2 \qquad \text{and} \qquad X\tr\y = X_1\tr\y_1 + X_2\tr\y_2,
$$
so
$$
X_1\tr X_1\rv + X_2\tr X_2\rv = \lam = X_1\tr X_1\rv_1,
\qquad \text{that is,} \qquad
X_1\tr X_1(\rv - \rv_1) = -X_2\tr X_2\rv.
$$

*Using disjointness.* Transposing both sides gives an equality of $1 \times p$ row vectors,
$$
\underbrace{[X_1(\rv - \rv_1)]\tr X_1}_{\in\, \R(X_1)} = \underbrace{[-X_2\rv]\tr X_2}_{\in\, \R(X_2)}.
$$
The left side is $\ve{u}\tr X_1$ with $\ve{u} = X_1(\rv - \rv_1) \in \Real^{n_1}$, so it lies in $\R(X_1)$; similarly, the right side lies in $\R(X_2)$. Since $\R(X_1) \cap \R(X_2) = \{\zero\tr\}$, both sides equal $\zero_p\tr$. This gives two facts:

1. $\rv\tr X_2\tr X_2 = \zero_p\tr$, so $(X_2\rv)\tr(X_2\rv) = \rv\tr X_2\tr X_2\rv = 0$. Then, by Lemma 1.2 with $M = X_2\rv$ (an $n_2 \times 1$ matrix, for which $\tra(M\tr M) = M\tr M = 0$), $X_2\rv = \zero_{n_2}$.
2. $X_1\tr X_1\rv = X_1\tr X_1\rv_1$. Then, by Lemma 1.3(ii) with $M = X_1$, $P = \rv$, and $Q = \rv_1$, $X_1\rv = X_1\rv_1$.

*Conclusion.* Combining these,
$$
\rv\tr X\tr\y = \rv\tr X_1\tr\y_1 + \rv\tr X_2\tr\y_2 = (X_1\rv)\tr\y_1 + (X_2\rv)\tr\y_2 = (X_1\rv_1)\tr\y_1 + 0 = \rv_1\tr X_1\tr\y_1,
$$
so the least squares estimates of $\lam\tr\bb$ under the two models coincide. Said differently, since $X_2\rv = \zero_{n_2}$, the observations in $\y_2$ get zero weight in the full-model estimate of $\lam\tr\bb$. \hfill $\square$

# Q4

Show that, under Model II, $\av\tr\y$ is a BLUE of its expectation if and only if $\av \in \C(X)$.

**Answer.** We provide proof by splitting $\av = \PX\av + (I_n - \PX)\av$, where $\PX = X(X\tr X)^{-}X\tr$, and showing that $\av\tr\y$ attains the BLUE's variance exactly when the second piece is zero.

Note: $\av$ is $n \times 1$, and under Model II, $\E(\eps) = \zero_n$ and $\Var(\y) = \Var(\eps) = \sigma^2 I_n$ with $\sigma^2 > 0$.

Also note the following:

- **Lemma 1.4**: For any matrix $X$, $X\tr X(X\tr X)^{-}X\tr = X\tr$.
- **Corollary 1(ii) to Lemma 1.4**: For any matrix $X$, $X(X\tr X)^{-}X\tr X = X$.
- **Corollary 3 to Lemma 1.5**: For any matrix $X$, $X(X\tr X)^{-}X\tr$ is symmetric.
- **Corollary to Lemma 2.4**: For a random vector $\ve{x}$ with $\Var(\ve{x}) = \Sigma$ and a matrix of constants $L$, $\Var(L\tr\ve{x}) = L\tr\Sigma L$.
- **Theorem 2.9 (Gauss--Markov)**: Under Models II--VII, if $\lam\tr\bb$ is estimable and $\bhat$ is any solution to the normal equations, then $\lam\tr\bhat$ is a linear unbiased estimator of $\lam\tr\bb$. What's more, under Model II, the variance of $\lam\tr\bhat$ is uniformly less than that of any other linear unbiased estimator of $\lam\tr\bb$; said differently, $\lam\tr\bhat$ is the BLUE.

*Properties of $\PX$.* The $n \times n$ matrix $\PX$ satisfies

- $\PX X = X$, by Corollary 1(ii) to Lemma 1.4,
- $\PX\tr = \PX$, by Corollary 3 to Lemma 1.5, and
- $\PX^2 = X(X\tr X)^{-}\big[X\tr X(X\tr X)^{-}X\tr\big] = X(X\tr X)^{-}X\tr = \PX$, by Lemma 1.4.

It then follows that $\PX\tr(I_n - \PX) = \PX - \PX^2 = 0_{n \times n}$.

*The BLUE of $\E(\av\tr\y)$.* We have $\E(\av\tr\y) = \av\tr X\bb = \lam\tr\bb$ with $\lam\tr = \av\tr X \in \R(X)$. By Lemma 2.1 (see Q1(a)), $\av\tr\y$ is a linear unbiased estimator of $\lam\tr\bb$, and so $\lam\tr\bb$ is estimable. By Theorems 2.3 and 1.2 (see Q2(d)), $\bhat = (X\tr X)^{-}X\tr\y$ solves the normal equations, so by Theorem 2.9, the BLUE of $\lam\tr\bb$ is
$$
\lam\tr\bhat = \av\tr X(X\tr X)^{-}X\tr\y = \av\tr\PX\y = (\PX\av)\tr\y,
$$
where the last step uses $\PX\tr = \PX$.

*Comparing variances.* Applying the Corollary to Lemma 2.4 with $L = \av$ and $\Sigma = \sigma^2 I_n$, and writing $\av = \PX\av + (I_n - \PX)\av$,
$$
\Var(\av\tr\y) = \sigma^2\av\tr\av
= \sigma^2(\PX\av)\tr(\PX\av) + 2\sigma^2\av\tr\PX\tr(I_n - \PX)\av + \sigma^2[(I_n - \PX)\av]\tr[(I_n - \PX)\av].
$$
The middle term vanishes because $\PX\tr(I_n - \PX) = 0_{n \times n}$, and the first term is $\Var\big((\PX\av)\tr\y\big)$, again by the Corollary to Lemma 2.4. Hence
$$
\Var(\av\tr\y) = \Var\big((\PX\av)\tr\y\big) + \sigma^2[(I_n - \PX)\av]\tr[(I_n - \PX)\av]. \tag{Q4.1}
$$

*When $\av\tr\y$ is a BLUE.* Since $\av\tr\y$ is linear and unbiased for $\lam\tr\bb$, and $(\PX\av)\tr\y$ is the BLUE, $\av\tr\y$ is a BLUE if and only if $\Var(\av\tr\y) = \Var\big((\PX\av)\tr\y\big)$. Because $\sigma^2 > 0$, (Q4.1) shows this holds if and only if $[(I_n - \PX)\av]\tr[(I_n - \PX)\av] = 0$. By Lemma 1.2 (see Q3), applied to the $n \times 1$ matrix $(I_n - \PX)\av$, this holds if and only if $(I_n - \PX)\av = \zero_n$, that is, $\av = \PX\av$.

*When $\av = \PX\av$.* Recall that $\av \in \C(X)$ if and only if $\av = X\bv$ for some $p \times 1$ vector $\bv$.

- ($\Rightarrow$) If $\av = \PX\av$, then $\av = X\bv$ with $\bv = (X\tr X)^{-}X\tr\av$, so $\av \in \C(X)$.
- ($\Leftarrow$) If $\av = X\bv$, then $\PX\av = \PX X\bv = X\bv = \av$, since $\PX X = X$.

Hence $\av\tr\y$ is a BLUE of its expectation if and only if $\av \in \C(X)$. \hfill $\square$

# Q5

## (a)

Show that, under Model II, the problem of determining the BLUE $\av\tr\y$ of an estimable function $\lam\tr\bb$ is equivalent to that of minimizing $\av\tr\av$ subject to the constraints $\av\tr X = \lam\tr$.

**Answer.** The two problems share a feasible set, and their objectives differ only by the constant factor $\sigma^2 > 0$.

Note: $\av$ is $n \times 1$ and $\lam$ is $p \times 1$.

By Lemma 2.1 (see Q1(a)), a linear estimator $c + \av\tr\y$ is unbiased for $\lam\tr\bb$ if and only if $c = 0$ and $\av\tr X = \lam\tr$. So the linear unbiased estimators of $\lam\tr\bb$ are exactly
$$
\{\av\tr\y : \av\tr X = \lam\tr\},
$$
which is nonempty because $\lam\tr\bb$ is estimable. Under Model II, $\Var(\y) = \sigma^2 I_n$, so by the Corollary to Lemma 2.4 (see Q4), each has variance
$$
\Var(\av\tr\y) = \av\tr(\sigma^2 I_n)\av = \sigma^2\av\tr\av.
$$
Since $\sigma^2 > 0$ does not depend on $\av$, it then follows that $\av$ minimizes $\Var(\av\tr\y)$ over this class if and only if it minimizes $\av\tr\av$ subject to $\av\tr X = \lam\tr$ (equivalently, $X\tr\av = \lam$). And since the minimizing $\av$ does not depend on $\sigma^2$, the resulting estimator has minimum variance uniformly in $\sigma^2$. \hfill $\square$

## (b)

The Lagrangian function associated with this minimization problem is
$$
L(\av, \rhov) = \av\tr\av + 2\rhov\tr(X\tr\av - \lam).
$$
Use the vector calculus to differentiate the Lagrangian function with respect to $\av$ and the Lagrangian multipliers, and set these derivatives equal to zero.

**Answer.**

Note: There is one multiplier per constraint in $X\tr\av = \lam$, so $\rhov$ is $p \times 1$.

Also note the following identities from Section 2.3 of the notes, for a vector variable $\bb$:

- For any constant vector $\hv$, $\partial(\hv\tr\bb)/\partial\bb = \hv$.
- For any constant matrix $X$, $\partial(\bb\tr X\tr X\bb)/\partial\bb = 2X\tr X\bb$.

Applying these term by term:

- *With respect to $\av$.* The second identity, with $I_n$ in place of $X$ and $\av$ in place of $\bb$, gives $\partial(\av\tr\av)/\partial\av = 2\av$. Writing $\rhov\tr X\tr\av = (X\rhov)\tr\av$, the first identity with $\hv = X\rhov$ gives $\partial(\rhov\tr X\tr\av)/\partial\av = X\rhov$. The term $-2\rhov\tr\lam$ does not involve $\av$.
- *With respect to $\rhov$.* The term $\av\tr\av$ does not involve $\rhov$. Writing $2\rhov\tr(X\tr\av - \lam) = 2(X\tr\av - \lam)\tr\rhov$, the first identity with $\hv = X\tr\av - \lam$ applies.

Together, the $n \times 1$ and $p \times 1$ gradients are
$$
\frac{\partial L}{\partial \av} = 2\av + 2X\rhov,
\qquad
\frac{\partial L}{\partial \rhov} = 2(X\tr\av - \lam).
$$
Setting these equal to zero gives the system
$$
\av + X\rhov = \zero_n, \qquad X\tr\av = \lam. \tag{Q5.1}
$$

## (c)

Show that the resulting equations have a solution if and only if the conjugate normal equations are consistent.

**Answer.** The link between the two systems is the substitution $\rv = -\rhov$, with $\av = X\rv$.

($\Rightarrow$) Suppose $(\av_0, \rhov_0)$ solves (Q5.1). The first equation gives $\av_0 = -X\rhov_0$, and substituting into the second,
$$
\lam = X\tr\av_0 = X\tr X(-\rhov_0).
$$
So $\rv = -\rhov_0$ solves $X\tr X\rv = \lam$, and the conjugate normal equations are consistent.

($\Leftarrow$) Suppose $X\tr X\rv_0 = \lam$ for some $p \times 1$ vector $\rv_0$. Set $\av_0 = X\rv_0$ and $\rhov_0 = -\rv_0$. Then
$$
\av_0 + X\rhov_0 = X\rv_0 - X\rv_0 = \zero_n
\qquad \text{and} \qquad
X\tr\av_0 = X\tr X\rv_0 = \lam,
$$
so $(\av_0, \rhov_0)$ solves (Q5.1). \hfill $\square$

## (d)

Solve your equations for the minimizing value of $\av$.

**Answer.** $\av_* = X(X\tr X)^{-}\lam$.

Note the following:

- **Theorem 1.3**: All solutions to a consistent matrix equation $AW = Y$ are generated by $W = A^{-}Y + (I - A^{-}A)Z$ as $Z$ ranges over all matrices of appropriate dimensions. (We write $W$ for the notes' unknown to avoid a clash with the design matrix $X$.)
- **Corollary to Lemma 1.1**: For any matrix $X$, $[(X\tr X)^{-}]\tr$ is a generalized inverse of $X\tr X$.

*A solution exists.* Since $\lam\tr\bb$ is estimable, the conjugate normal equations are consistent by Theorem 2.7 (see Q3), so (Q5.1) has a solution by part (c).

*All solutions.* The first equation of (Q5.1) gives $\av = -X\rhov$. Substituting into the second gives $X\tr X(-\rhov) = \lam$, a consistent system in the $p \times 1$ unknown $-\rhov$. Applying Theorem 1.3 with $A = X\tr X$, $W = -\rhov$, and $Y = \lam$, all solutions are
$$
-\rhov = (X\tr X)^{-}\lam + [I_p - (X\tr X)^{-}X\tr X]\zv, \qquad \zv \in \Real^p.
$$
Hence
$$
\av = -X\rhov = X(X\tr X)^{-}\lam + \big[X - X(X\tr X)^{-}X\tr X\big]\zv = X(X\tr X)^{-}\lam,
$$
since $X(X\tr X)^{-}X\tr X = X$ by Corollary 1(ii) to Lemma 1.4 (see Q4). So every solution of (Q5.1) has the same $\av$-component,
$$
\av_* = X(X\tr X)^{-}\lam.
$$

*The stationary point is the minimum.* Equations (Q5.1) are necessary conditions only, so we verify directly that $\av_*$ minimizes $\av\tr\av$ subject to $X\tr\av = \lam$. Let $\av$ be any $n \times 1$ vector with $X\tr\av = \lam$, and set $\dv = \av - \av_*$, so that $X\tr\dv = \lam - \lam = \zero_p$. Since $\av_* = X\rv_*$ with $\rv_* = (X\tr X)^{-}\lam$,
$$
\av_*\tr\dv = \rv_*\tr X\tr\dv = 0,
$$
and therefore
$$
\av\tr\av = (\av_* + \dv)\tr(\av_* + \dv) = \av_*\tr\av_* + \dv\tr\dv \ \ge\ \av_*\tr\av_*.
$$
Equality holds if and only if $\dv\tr\dv = 0$, that is, $\dv = \zero_n$ (by Lemma 1.2, see Q3). So $\av_*$ is the unique minimizer.

Side Note: Uniqueness also means $X(X\tr X)^{-}\lam$ does not depend on the choice of generalized inverse.

*The resulting BLUE.* By part (a), the BLUE is
$$
\av_*\tr\y = \lam\tr[(X\tr X)^{-}]\tr X\tr\y.
$$
By the Corollary to Lemma 1.1, $[(X\tr X)^{-}]\tr$ is itself a generalized inverse of $X\tr X$, so by Theorems 2.3 and 1.2 (see Q2(d)), $\bhat = [(X\tr X)^{-}]\tr X\tr\y$ solves the normal equations, and
$$
\av_*\tr\y = \lam\tr\bhat.
$$
That is, the BLUE of $\lam\tr\bb$ is its least squares estimator, in agreement with Theorem 2.9 (see Q4). \hfill $\square$
