---
title: "HW1"
author: "Sam Olson"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Status 

Q1: Started
Q2: Done (Minor Touch-ups)
Q3: 
Q4: 
Q5: 
Q6: 
Q7: 
Q8: Done 

# Q1: A.3

Let $I = [0,1]$, $\Omega = \mathbb{R}$ and for $\alpha \in \mathbb{R}$, $A_\alpha = (\alpha - 1, \alpha + 1)$, the open interval $\{x : \alpha - 1 < x < \alpha + 1\}$.

## (a) 

Show that $\bigcup_{\alpha \in I} A_\alpha = (-1, 2)$ and $\bigcap_{\alpha \in I} A_\alpha = (0, 1)$.

## Answer 

To start, note that $A_{0} = (-1, 1)$ and $A_{1} = (0, 2)$. 

## (b) 

Suppose $J = \{x : x \in I, x \text{ is rational}\}$. Find $\bigcup_{x \in J} A_x$ and $\bigcap_{x \in J} A_x$.

## Answer 

# Q2: A.5

Show that $X \equiv \{0,1\}^{\mathbb{N}}$, the set of all sequences $\{\omega_i\}_{i \in \mathbb{N}}$ where each $\omega_i \in \{0,1\}$, is **uncountable**. Conclude that $\mathcal{P}(\mathbb{N})$ is uncountable.

## Answer 

## A.5 Solution

Sketch: To prove $X \equiv \{0,1\}^{\mathbb{N}}$ is uncountable, and consequently $\mathcal{P}(\mathbb{N})$ is uncountable, we must first establish (1): $X$ is uncountable and then (2): $\mathcal{P}(\mathbb{N})$ and $X$ are in bijection. 

### (1): $X$ is uncountable

Suppose, for contradiction, that $X$ is countable. Since $X$ is infinite (it contains, e.g., one sequence for every finite subset of $\mathbb{N}$), countability would mean there exists a surjection $h : \mathbb{N} \to X$ enumerating every element of $X$: $h(1), h(2), h(3), \ldots$

Write $h(n) = (h(n)_1, h(n)_2, h(n)_3, \ldots)$ for the $n$-th sequence in the list, with $h(n)_i \in \{0,1\}$ its $i$-th coordinate. 

Define $\delta = (\delta_i)_{i \in \mathbb{N}}$ by flipping the diagonal entries: $\delta_i = 1 - h(i)_i, \qquad i \in \mathbb{N}.$

Then $\delta \in X$, since each $\delta_i \in \{0,1\}$. But for every $n \in \mathbb{N}$, $\delta_n = 1 - h(n)_n \neq h(n)_n,$, so $\delta$ and $h(n)$ disagree in coordinate $n$, i.e. $\delta \neq h(n)$. This holds for **every** $n$, so $\delta \notin \{h(1), h(2), h(3), \ldots\}$, contradicting the assumption that $h$ is surjective (enumerates all of $X$).

Hence no such enumeration exists, and $X$ is uncountable.

### (2): $\mathcal{P}(\mathbb{N})$ and $X$ are in bijection

Define $\Phi : \mathcal{P}(\mathbb{N}) \to X$ by

$$\Phi(A) = \{\omega_i\}_{i \in \mathbb{N}}, \qquad \omega_i = \mathbb{1}\{i \in A\}, \quad \forall A \in \mathcal{P}(\mathbb{N}),$$

i.e. $\Phi(A)$ is the indicator sequence of $A$: coordinate $i$ is $1$ if $i \in A$ and $0$ otherwise. We show $\Phi$ is a bijection.

- **Injective:** if $A \neq B$, then without loss of generality there is some $i \in A \setminus B$, so $\Phi(A)_i = 1 \neq 0 = \Phi(B)_i$, hence $\Phi(A) \neq \Phi(B)$.
- **Surjective:** given any $\omega = (\omega_i)_{i \in \mathbb{N}} \in X$, let $A = \{i \in \mathbb{N} : \omega_i = 1\} \in \mathcal{P}(\mathbb{N})$. Then $\Phi(A) = \omega$ by construction.

Hence $\Phi$ is a bijection, so $|\mathcal{P}(\mathbb{N})| = |X|$.

### Conclusion

Bijections preserve cardinality, so $|\mathcal{P}(\mathbb{N})| = |X|$. By Step 1, $X$ is uncountable. Therefore $\mathcal{P}(\mathbb{N})$ is uncountable.

### Remarks

This recovers the general fact $|\Omega| < |\mathcal{P}(\Omega)|$ (proved for $\Omega = \mathbb{N}$ directly in the first proof) as a special case. That argument builds, for any purported bijection $g : \mathbb{N} \to \mathcal{P}(\mathbb{N})$, the set $A = \{i \in \mathbb{N} : i \notin g(i)\}$, which lies in no $g(i)$: a preimage $x$ with $g(x) = A$ would force $x \in A \iff x \notin g(x) = A$, a contradiction. Translating through $\Phi$, this $A$ is exactly the diagonal sequence $\delta$ constructed above, since $\delta_i = 1 - \Phi(g(i))_i$ says precisely $\delta_i = 1 \iff i \notin g(i)$. The two diagonalizations are the same argument, one on sequences and one on sets.

# Q3: A.9

Apply the principle of induction to establish the following:

## (a) 

For each $n \in \mathbb{N}$, $\sum_{j=1}^n j^2 = \dfrac{n(n+1)(2n+1)}{6}$.

## Answer 

## (b) 

For each $n \in \mathbb{N}$, $x_1, x_2, \ldots, x_k \in \mathbb{R}$,

### (i) 

**(The binomial formula)**. $(x_1 + x_2)^n = \sum_{r=0}^{n} \binom{n}{r} x_1^r x_2^{n-r}$.

### Answer 

## (ii) 

**(The multinomial formula)**.

$$(x_1 + x_2 + \ldots + x_k)^n = \sum \frac{n!}{r_1! r_2! \ldots r_k!} x_1^{r_1} x_2^{r_2} \ldots x_k^{r_k},$$

where the summation extends over all $(r_1, r_2, \ldots, r_k)$ such that $r_i \in \mathbb{N}$, $0 \leq r_i \leq n$, $\sum_{i=1}^{k} r_i = n$.

### Answer 

# Q4: A.11

Show that the set $A = \{r : r \in \mathbb{Q}, r^2 < 2\}$ is bounded above in $\mathbb{Q}$ but has no l.u.b. in $\mathbb{Q}$.

## Answer

# Q5: A.12

Show that for any two sequences $\{x_n\}_{n\geq 1}, \{y_n\}_{n\geq 1} \subset \mathbb{R}$,

$$\varliminf_{n\to\infty} x_n + \varliminf_{n\to\infty} y_n \leq \varliminf_{n\to\infty} (x_n + y_n) \leq \varlimsup_{n\to\infty} (x_n + y_n)$$

$$\leq \varlimsup_{n\to\infty} x_n + \varlimsup_{n\to\infty} y_n.$$

## Answer 

# Q6: A.17 

Find the radius of convergence, $\rho$, for the powers series $A(x) \equiv \sum_{n=0}^{\infty} a_n x^n$ where

## (a) 

$a_n = \dfrac{n}{(n+1)}$, $n \geq 0$.

## Answer 

## (b) 

$a_n = n^p$, $n \geq 0$, $p \in \mathbb{R}$.

## Answer 

## (c) 

$a_n = \dfrac{1}{n!}$, $n \geq 0$ (where $0! = 1$).

## Answer 

# Q7: A.18

## (a) 

Find the Taylor series at $a = 0$ for the function $f(x) = \dfrac{1}{1-x}$ in $I \equiv (-1, +1)$ and show that it converges to $f(x)$ on $I$.

## Answer 

## (b) 

Find the Taylor series of $1 + x + x^2$ in $I = (1,3)$, centered at 2.

## Answer 

## (c) 

Let

$$f(x) = \begin{cases} e^{-\frac{1}{x^2}} & \text{if } |x| < 1, x \neq 0 \\ 0 & \text{if } x = 0. \end{cases}$$

### (i) 

Show that $f$ is infinitely differentiable at 0 and compute $f^{(j)}(0)$ for all $j \geq 1$.

### Answer 

### (ii) 

Show that the Taylor series at $a = 0$ converges but not to $f$ on $(-1,1)$.

### Answer 

# Q8: A.24 (c)

Draw the open unit ball $B_p \equiv \{x : x \in \mathbb{R}^2,\ d_p(x,0) < 1\}$ in $\mathbb{R}^2$ for $p = 1, 2$ and $\infty$.

## Answer 

$p = 1, 2$ and $\infty$ are shown below in black, red, and blue respectively. 

```{r, echo = F, warning = F, message = F}
library(ggplot2)
library(patchwork)

theme_ball <- theme_void() +
  theme(plot.title = element_text(hjust = 0.5, size = 14),
        plot.margin = margin(10, 10, 10, 10))

lims <- coord_fixed(xlim = c(-1.5, 1.5), ylim = c(-1.5, 1.5))

## p = 1 : diamond (black)
diamond <- data.frame(x = c(1, 0, -1, 0, 1), y = c(0, 1, 0, -1, 0))
p1 <- ggplot() +
  geom_path(data = diamond, aes(x, y), color = "black") +
  geom_hline(yintercept = 0) +
  geom_vline(xintercept = 0) +
  lims +
  labs(title = expression(group("||", x, "||")[1])) +
  theme_ball

## p = 2 : circle (red)
theta <- seq(0, 2*pi, length.out = 200)
circle <- data.frame(x = cos(theta), y = sin(theta))
p2 <- ggplot() +
  geom_path(data = circle, aes(x, y), color = "red") +
  geom_hline(yintercept = 0) +
  geom_vline(xintercept = 0) +
  lims +
  labs(title = expression(group("||", x, "||")[2])) +
  theme_ball

## p = infinity : square (blue)
square <- data.frame(x = c(1, 1, -1, -1, 1), y = c(-1, 1, 1, -1, -1))
p3 <- ggplot() +
  geom_path(data = square, aes(x, y), color = "blue") +
  geom_hline(yintercept = 0) +
  geom_vline(xintercept = 0) +
  lims +
  labs(title = expression(group("||", x, "||")[infinity])) +
  theme_ball

# side-by-side (patchwork default with "+")
p1 + p2 + p3
```