---
title: "HW1"
author: "Sam Olson"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Status 

Q1: Done (Minor Touch-ups)

Q2: Done (Minor Touch-ups)

Q3: 

Q4: Done (Minor Touch-ups)

Q5: Done (Minor Touch-ups)

Q6: Done (Minor Touch-ups)

Q7: 

Q8: Done 

# Q1: A.3

Let $I = [0,1]$, $\Omega = \mathbb{R}$ and for $\alpha \in \mathbb{R}$, $A_\alpha = (\alpha - 1, \alpha + 1)$, the open interval $\{x : \alpha - 1 < x < \alpha + 1\}$.

## (a) 

Show that $\bigcup_{\alpha \in I} A_\alpha = (-1, 2)$ and $\bigcap_{\alpha \in I} A_\alpha = (0, 1)$.

## Answer 

Sketch: For each containment we argue both directions. For the union, every $A_\alpha$ with $\alpha\in I=[0,1]$ sits inside $(-1,2)$ since $\alpha-1\geq -1$ and $\alpha+1\leq 2$; conversely, any $x\in(-1,2)$ lies in $A_\alpha$ for $\alpha = \mathrm{clamp}(x,0,1) := \min(\max(x,0),1) \in I$. For the intersection, $A_0\cap A_1=(0,1)$ already gives one containment since $0,1\in I$; the reverse follows because every $\alpha\in I$ satisfies $\alpha - 1 \leq 0$ and $\alpha+1\geq 1$, so $(0,1)\subset A_\alpha$ for every such $\alpha$.

### Union

*Proof.* ($\subset$) For any $\alpha \in [0,1]$, $\alpha - 1 \in [-1,0]$ and $\alpha+1\in[1,2]$, so $A_\alpha = (\alpha-1,\alpha+1) \subset (-1,2)$. Since this holds for every $\alpha\in I$,

$$\bigcup_{\alpha\in I} A_\alpha \subset (-1,2).$$

($\supset$) Let $x \in (-1,2)$ be given, and set $\alpha = \min(\max(x,0),1) \in [0,1] = I$.

- If $x < 0$: $\alpha = 0$, and since $-1<x<0<1$, $x \in A_0 = (-1,1)$.
- If $0 \leq x \leq 1$: $\alpha = x$, and trivially $x - 1 < x < x+1$, so $x \in A_x$.
- If $x > 1$: $\alpha = 1$, and since $0<1<x<2$, $x \in A_1 = (0,2)$.

In every case $x \in A_\alpha$ for some $\alpha \in I$, so $x \in \bigcup_{\alpha\in I} A_\alpha$. Hence

$$(-1,2) \subset \bigcup_{\alpha\in I} A_\alpha.$$

Combining both containments, $\bigcup_{\alpha\in I} A_\alpha = (-1,2)$. $\square$

### Intersection

*Proof.* ($\subset$) Since $0,1 \in I$, $\bigcap_{\alpha\in I} A_\alpha \subset A_0 \cap A_1 = (-1,1)\cap(0,2) = (0,1)$.

($\supset$) Let $x \in (0,1)$ be given, and let $\alpha \in I$ be arbitrary. Since $\alpha \geq 0$, $\alpha - 1 \geq -1$... more precisely, since $0 \le \alpha \le 1$, we have $\alpha - 1 \leq 0 < x$ and $x < 1 \leq \alpha+1$, so $x \in A_\alpha$. As $\alpha\in I$ was arbitrary, $x \in \bigcap_{\alpha\in I} A_\alpha$. Hence

$$(0,1) \subset \bigcap_{\alpha\in I} A_\alpha.$$

Combining both containments, $\bigcap_{\alpha\in I} A_\alpha = (0,1)$. $\square$

## (b) 

Suppose $J = \{x : x \in I, x \text{ is rational}\}$. Find $\bigcup_{x \in J} A_x$ and $\bigcap_{x \in J} A_x$.

## Answer 

Sketch: Since $J \subset I$, both $\bigcup_{x\in J}A_x \subset \bigcup_{\alpha\in I}A_\alpha$ and $\bigcap_{x\in J}A_x \supset \bigcap_{\alpha\in I}A_\alpha$ hold automatically (a union over fewer sets is smaller; an intersection over fewer sets is larger). The key observation is that $0$ and $1$ are themselves rational, so $0,1 \in J$ — meaning $A_0, A_1$ belong to the $J$-indexed family too. That lets us reuse part (a)'s results ($A_0\cup A_1 = (-1,2)$ and $A_0\cap A_1=(0,1)$, both established above) to force the reverse containments, without any density argument.

### Union

*Proof.* Since $J \subset I$,

$$\bigcup_{x\in J} A_x \subset \bigcup_{\alpha\in I} A_\alpha = (-1,2).$$

Conversely, since $0,1 \in J$ (both are rational numbers in $I$),

$$\bigcup_{x\in J} A_x \supset A_0 \cup A_1 = (-1,1)\cup(0,2) = (-1,2),$$

using that the two intervals overlap on $(0,1)$. Combining both containments,

$$\bigcup_{x\in J} A_x = (-1,2).$$

### Intersection

Since $J \subset I$,

$$\bigcap_{x\in J} A_x \supset \bigcap_{\alpha\in I} A_\alpha = (0,1).$$

Conversely, since $0,1 \in J$,

$$\bigcap_{x\in J} A_x \subset A_0 \cap A_1 = (-1,1)\cap(0,2) = (0,1).$$

Combining both containments,

$$\bigcap_{x\in J} A_x = (0,1).$$

Thus $\bigcup_{x\in J}A_x = \bigcup_{\alpha\in I}A_\alpha = (-1,2)$ and $\bigcap_{x\in J}A_x = \bigcap_{\alpha\in I}A_\alpha = (0,1)$: restricting the index set from all of $I$ down to just its rationals changes neither the union nor the intersection. $\square$

# Q2: A.5

Show that $X \equiv \{0,1\}^{\mathbb{N}}$, the set of all sequences $\{\omega_i\}_{i \in \mathbb{N}}$ where each $\omega_i \in \{0,1\}$, is **uncountable**. Conclude that $\mathcal{P}(\mathbb{N})$ is uncountable.

## Answer 

## A.5 Solution

Sketch: To prove $X \equiv \{0,1\}^{\mathbb{N}}$ is uncountable, and consequently $\mathcal{P}(\mathbb{N})$ is uncountable, we must first establish (1): $X$ is uncountable and then (2): $\mathcal{P}(\mathbb{N})$ and $X$ are in bijection. 

### (1): $X$ is uncountable

Suppose, for contradiction, that $X$ is countable. Since $X$ is infinite (it contains, e.g., one sequence for every finite subset of $\mathbb{N}$), countability would mean there exists a surjection $h : \mathbb{N} \to X$ enumerating every element of $X$: $h(1), h(2), h(3), \ldots$

Write $h(n) = (h(n)_1, h(n)_2, h(n)_3, \ldots)$ for the $n$-th sequence in the list, with $h(n)_i \in \{0,1\}$ its $i$-th coordinate. 

Define $\delta = (\delta_i)_{i \in \mathbb{N}}$ by flipping the diagonal entries: $\delta_i = 1 - h(i)_i, \qquad i \in \mathbb{N}.$

Then $\delta \in X$, since each $\delta_i \in \{0,1\}$. But for every $n \in \mathbb{N}$, $\delta_n = 1 - h(n)_n \neq h(n)_n,$, so $\delta$ and $h(n)$ disagree in coordinate $n$, i.e. $\delta \neq h(n)$. This holds for **every** $n$, so $\delta \notin \{h(1), h(2), h(3), \ldots\}$, contradicting the assumption that $h$ is surjective (enumerates all of $X$).

Hence no such enumeration exists, and $X$ is uncountable.

### (2): $\mathcal{P}(\mathbb{N})$ and $X$ are in bijection

Define $\Phi : \mathcal{P}(\mathbb{N}) \to X$ by

$$\Phi(A) = \{\omega_i\}_{i \in \mathbb{N}}, \qquad \omega_i = \mathbb{1}\{i \in A\}, \quad \forall A \in \mathcal{P}(\mathbb{N}),$$

i.e. $\Phi(A)$ is the indicator sequence of $A$: coordinate $i$ is $1$ if $i \in A$ and $0$ otherwise. We show $\Phi$ is a bijection.

- **Injective:** if $A \neq B$, then without loss of generality there is some $i \in A \setminus B$, so $\Phi(A)_i = 1 \neq 0 = \Phi(B)_i$, hence $\Phi(A) \neq \Phi(B)$.
- **Surjective:** given any $\omega = (\omega_i)_{i \in \mathbb{N}} \in X$, let $A = \{i \in \mathbb{N} : \omega_i = 1\} \in \mathcal{P}(\mathbb{N})$. Then $\Phi(A) = \omega$ by construction.

Hence $\Phi$ is a bijection, so $|\mathcal{P}(\mathbb{N})| = |X|$.

### Conclusion

Bijections preserve cardinality, so $|\mathcal{P}(\mathbb{N})| = |X|$. By Step 1, $X$ is uncountable. Therefore $\mathcal{P}(\mathbb{N})$ is uncountable.

### Remarks

This recovers the general fact $|\Omega| < |\mathcal{P}(\Omega)|$ (proved for $\Omega = \mathbb{N}$ directly in the first proof) as a special case. That argument builds, for any purported bijection $g : \mathbb{N} \to \mathcal{P}(\mathbb{N})$, the set $A = \{i \in \mathbb{N} : i \notin g(i)\}$, which lies in no $g(i)$: a preimage $x$ with $g(x) = A$ would force $x \in A \iff x \notin g(x) = A$, a contradiction. Translating through $\Phi$, this $A$ is exactly the diagonal sequence $\delta$ constructed above, since $\delta_i = 1 - \Phi(g(i))_i$ says precisely $\delta_i = 1 \iff i \notin g(i)$. The two diagonalizations are the same argument, one on sequences and one on sets.

# Q3: A.9

Apply the principle of induction to establish the following:

## (a) 

For each $n \in \mathbb{N}$, $\sum_{j=1}^n j^2 = \dfrac{n(n+1)(2n+1)}{6}$.

## Answer 

## (b) 

For each $n \in \mathbb{N}$, $x_1, x_2, \ldots, x_k \in \mathbb{R}$,

### (i) 

**(The binomial formula)**. $(x_1 + x_2)^n = \sum_{r=0}^{n} \binom{n}{r} x_1^r x_2^{n-r}$.

### Answer 

## (ii) 

**(The multinomial formula)**.

$$(x_1 + x_2 + \ldots + x_k)^n = \sum \frac{n!}{r_1! r_2! \ldots r_k!} x_1^{r_1} x_2^{r_2} \ldots x_k^{r_k},$$

where the summation extends over all $(r_1, r_2, \ldots, r_k)$ such that $r_i \in \mathbb{N}$, $0 \leq r_i \leq n$, $\sum_{i=1}^{k} r_i = n$.

### Answer 

# Q4: A.11

Show that the set $A = \{r : r \in \mathbb{Q}, r^2 < 2\}$ is bounded above in $\mathbb{Q}$ but has no l.u.b. in $\mathbb{Q}$.

## Answer

Sketch: To show $A$ is bounded above in $\mathbb{Q}$ but has no l.u.b. in $\mathbb{Q}$, we establish (1): $A$ is bounded above, by exhibiting an explicit upper bound; and (2): no rational $M$ can be the *least* upper bound, by supposing $M = \sup_{\mathbb{Q}}(A)$ exists and deriving a contradiction from a case split on $M^2 < 2$ versus $M^2 > 2$, in each case constructing a rational that violates either "upper bound" or "least."

### (1): $A$ is bounded above

*Proof.* Let $r \in A$ be given. Thus $r^2 < 2$. If $r \geq 2$, then $r^2 \geq 4 > 2$, contradicting $r \in A$. Thus $r < 2$.
By definition of an upper bound, $2$ is an upper bound of $A$. $\square$

### (2): $A$ has no l.u.b. in $\mathbb{Q}$

*Proof.* Suppose to the contrary that $M = \sup_{\mathbb{Q}}(A)$ exists. Since $1 \in A$ (as $1^2 = 1 < 2$), $M \geq 1$, so $M$ is a positive rational. Since $\sqrt{2}$ is irrational, no rational squares to $2$, so $M^2 \neq 2$. Thus either $M^2 < 2$ or $M^2 > 2$; we derive a contradiction in each case.

**Case $M^2 < 2$:** We construct $r \in A$ with $r > M$, contradicting $M$ as an upper bound.
Let $\varepsilon = 2 - M^2 > 0$. Choose a rational $\delta$ with $0 < \delta < \min\left(1, \dfrac{\varepsilon}{2M+1}\right)$, and set $r = M + \delta$. Then
$$r^2 = M^2 + 2M\delta + \delta^2 < M^2 + \delta(2M+1) < M^2 + \varepsilon = 2.$$
So $r \in A$, yet $r = M + \delta > M$ — contradicting that $M$ is an upper bound of $A$.

**Case $M^2 > 2$:** We construct a rational upper bound $b$ of $A$ with $b < M$, contradicting $M$ as the *least* upper bound.
Let $\varepsilon = M^2 - 2 > 0$. Choose a rational $\delta$ with $0 < \delta < \min\left(M, \dfrac{\varepsilon}{2M}\right)$, and set $b = M - \delta$. Then
$$b^2 = M^2 - 2M\delta + \delta^2 > M^2 - 2M\delta > M^2 - \varepsilon = 2.$$
Since $b > 0$ and $b^2 > 2$: for any $a \in A$, $a^2 < 2 < b^2$, and since $a, b > 0$, this gives $a < b$. So $b$ is an upper bound of $A$, yet $b < M$ — contradicting that $M$ is the *least* upper bound.

Both cases are contradictions, so no such $M \in \mathbb{Q}$ exists. Thus $A$ has no l.u.b. in $\mathbb{Q}$. $\square$

# Q5: A.12

Show that for any two sequences $\{x_n\}_{n\geq 1}, \{y_n\}_{n\geq 1} \subset \mathbb{R}$,

$$\varliminf_{n\to\infty} x_n + \varliminf_{n\to\infty} y_n \leq \varliminf_{n\to\infty} (x_n + y_n) \leq \varlimsup_{n\to\infty} (x_n + y_n)$$

$$\leq \varlimsup_{n\to\infty} x_n + \varlimsup_{n\to\infty} y_n.$$

## Answer

Sketch: To establish the chain of four inequalities, we bound $x_n + y_n$ between $\inf_{k\geq n}x_k + \inf_{k\geq n}y_k$ and $\sup_{k\geq n}x_k + \sup_{k\geq n}y_k$ for every $n$, then pass this bound through the definitions of $\varliminf$ and $\varlimsup$ — using that $\inf_{k\geq n}x_k$ is itself monotone increasing in $n$ (so it *converges* to $\varliminf x_n$, rather than merely bounding it) and $\sup_{k\geq n}x_k$ is monotone decreasing (converging to $\varlimsup x_n$). The middle inequality is then just the trivial fact that liminf never exceeds limsup.

### Step 1: The pointwise bound

For every $n \in \mathbb{N}$ and every $k \geq n$,

$$\inf_{k\geq n} x_k \leq x_k \leq \sup_{k\geq n} x_k, \qquad \inf_{k\geq n} y_k \leq y_k \leq \sup_{k\geq n} y_k.$$

Taking $k = n$ and adding the two left- and right-hand inequalities gives, for every $n$, the bound

$$\inf_{k\geq n} x_k + \inf_{k\geq n} y_k \;\leq\; x_n + y_n \;\leq\; \sup_{k\geq n} x_k + \sup_{k\geq n} y_k,$$

which we refer to below as $(\ast)$.

### Step 2: The outer terms converge to $\varliminf$ and $\varlimsup$

By definition, $\varliminf_{n\to\infty} x_n = \lim_{n\to\infty}\inf_{k\geq n} x_k$ and $\varlimsup_{n\to\infty} x_n = \lim_{n\to\infty}\sup_{k\geq n} x_k$, and these limits exist (possibly $\pm\infty$) because $\inf_{k\geq n}x_k$ is non-decreasing in $n$ and $\sup_{k\geq n}x_k$ is non-increasing in $n$. The same holds for $y_n$. Hence the left side of $(\ast)$ converges:

$$\inf_{k\geq n} x_k + \inf_{k\geq n} y_k \;\longrightarrow\; \varliminf_{n\to\infty} x_n + \varliminf_{n\to\infty} y_n,$$

and the right side of $(\ast)$ converges:

$$\sup_{k\geq n} x_k + \sup_{k\geq n} y_k \;\longrightarrow\; \varlimsup_{n\to\infty} x_n + \varlimsup_{n\to\infty} y_n.$$

### Step 3: Pass $(\ast)$ through liminf and limsup

Since $\liminf$ and $\limsup$ are monotone with respect to termwise $\leq$, applying them to $(\ast)$ gives, from the left inequality,

$$\varliminf_{n\to\infty}\Big(\inf_{k\geq n} x_k + \inf_{k\geq n} y_k\Big) \leq \varliminf_{n\to\infty}(x_n+y_n),$$

and by Step 2 the left side is an honest limit, equal to $\varliminf_{n\to\infty} x_n + \varliminf_{n\to\infty} y_n$. This gives

$$\varliminf_{n\to\infty} x_n + \varliminf_{n\to\infty} y_n \;\leq\; \varliminf_{n\to\infty}(x_n+y_n).$$

Symmetrically, applying $\limsup$ to the right inequality of $(\ast)$ and using Step 2 gives

$$\varlimsup_{n\to\infty}(x_n+y_n) \;\leq\; \varlimsup_{n\to\infty} x_n + \varlimsup_{n\to\infty} y_n.$$

### Step 4: The middle inequality

For any real sequence $z_n$, $\inf_{k\geq n} z_k \leq \sup_{k\geq n} z_k$ for every $n$; taking the limit in $n$ gives $\varliminf_{n\to\infty} z_n \leq \varlimsup_{n\to\infty} z_n$. Applying this with $z_n = x_n + y_n$:

$$\varliminf_{n\to\infty}(x_n+y_n) \leq \varlimsup_{n\to\infty}(x_n+y_n).$$

### Conclusion

Step 3 gives the outer two inequalities of the claim, and Step 4 gives the middle one. Chaining all three together yields

$$\varliminf_{n\to\infty} x_n + \varliminf_{n\to\infty} y_n \leq \varliminf_{n\to\infty} (x_n + y_n) \leq \varlimsup_{n\to\infty} (x_n + y_n) \leq \varlimsup_{n\to\infty} x_n + \varlimsup_{n\to\infty} y_n. \qquad \blacksquare$$

# Q6: A.17 

Find the radius of convergence, $\rho$, for the powers series $A(x) \equiv \sum_{n=0}^{\infty} a_n x^n$ where

## (a) 

$a_n = \dfrac{n}{(n+1)}$, $n \geq 0$.

## Answer 

Sketch: Since $\rho = \left(\varlimsup_{n\to\infty}|a_n|^{1/n}\right)^{-1}$ (Cauchy–Hadamard), it suffices to show $a_n^{1/n} \to 1$. Since $a_n \in (0,1)$ for all $n\geq 1$ and $a_n \to 1$, we compute the limit of $\log(a_n^{1/n})$ directly using the standard fact $\log(1+t)/t \to 1$ as $t\to 0$.

For $n \geq 1$,

$$\log\big(a_n^{1/n}\big) = \frac{1}{n}\log\!\left(\frac{n}{n+1}\right) = -\frac{1}{n}\log\!\left(1+\frac{1}{n}\right).$$
Writing $t = 1/n \to 0$ and using $\lim_{t\to 0} \dfrac{\log(1+t)}{t} = 1$,

$$\frac{1}{n}\log\!\left(1+\frac{1}{n}\right) = \frac{1}{n^2}\cdot\frac{\log(1+1/n)}{1/n} \;\longrightarrow\; 0 \cdot 1 = 0.$$

Hence $\log(a_n^{1/n}) \to 0$, so $a_n^{1/n} \to e^0 = 1$. Since this is an honest limit, $\varlimsup_{n\to\infty}|a_n|^{1/n} = 1$.

Thus $\rho = 1^{-1} = 1$.

## (b) 

$a_n = n^p$, $n \geq 0$, $p \in \mathbb{R}$.

## Answer 

Sketch: The key fact is $\log n / n \to 0$, which lets us show $n^{p/n} \to 1$ for *every* fixed $p$ — so the radius of convergence is $1$ regardless of the exponent $p$. We justify $\log n/n \to 0$ by applying L'Hôpital's rule to the continuous function $\log x/x$ as $x \to \infty$; since the resulting limit holds over all reals $x \to \infty$, it holds in particular along the integer sequence $n \to \infty$.

*Proof.* For the continuous function $f(x) = \dfrac{\log x}{x}$, L'Hôpital's rule gives

$$\lim_{x\to\infty} \frac{\log x}{x} = \lim_{x\to\infty} \frac{1/x}{1} = 0,$$

and since this limit holds as $x \to \infty$ through the reals, it holds along the integer sequence: $\dfrac{\log n}{n} \to 0$.

For $n \geq 1$,

$$\log\big(a_n^{1/n}\big) = \log\big(n^{p/n}\big) = p \cdot \frac{\log n}{n} \;\longrightarrow\; p \cdot 0 = 0$$

for every fixed $p \in \mathbb{R}$. Hence $a_n^{1/n} = n^{p/n} \to e^0 = 1$, so $\varlimsup_{n\to\infty}|a_n|^{1/n} = 1$.

(For $p<0$ the term $a_0 = 0^p$ is undefined, but the radius of convergence depends only on the tail behavior as $n\to\infty$, so this does not affect $\rho$.)

Thus $\rho = 1^{-1} = 1$, for every $p \in \mathbb{R}$.

## (c) 

$a_n = \dfrac{1}{n!}$, $n \geq 0$ (where $0! = 1$).

## Answer 

Sketch: We show $(1/n!)^{1/n} \to 0$ using the elementary bound $n! \geq (n/e)^n$, obtained by truncating the power series for $e^n$ to a single term. This avoids relying on Stirling's formula, and avoids the false intermediate claim $(1/n)^{1/n}\to 0$ (in fact $(1/n)^{1/n}\to 1$, which would only give $\varlimsup \leq 1$ — far too weak to conclude $\rho=\infty$).

*Proof.* For $n \geq 1$, the exponential series $e^n = \displaystyle\sum_{k=0}^{\infty} \frac{n^k}{k!}$ has all nonnegative terms, so it is bounded below by its $k=n$ term:

$$e^n \geq \frac{n^n}{n!} \quad\Longrightarrow\quad n! \geq \frac{n^n}{e^n} = \left(\frac{n}{e}\right)^n.$$

Taking $n$-th roots (all quantities positive),

$$(n!)^{1/n} \geq \frac{n}{e}.$$

Hence

$$0 \;<\; \left(\frac{1}{n!}\right)^{1/n} = \frac{1}{(n!)^{1/n}} \;\leq\; \frac{e}{n} \;\longrightarrow\; 0.$$

By the squeeze theorem, $\displaystyle\lim_{n\to\infty}\left(\frac{1}{n!}\right)^{1/n} = 0$, so $\varlimsup_{n\to\infty}|a_n|^{1/n} = 0$.

Thus $\rho = (0)^{-1} = +\infty$: the series converges for every $x \in \mathbb{R}$. $\blacksquare$

# Q7: A.18

## (a) 

Find the Taylor series at $a = 0$ for the function $f(x) = \dfrac{1}{1-x}$ in $I \equiv (-1, +1)$ and show that it converges to $f(x)$ on $I$.

## Answer 

## (b) 

Find the Taylor series of $1 + x + x^2$ in $I = (1,3)$, centered at 2.

## Answer 

## (c) 

Let

$$f(x) = \begin{cases} e^{-\frac{1}{x^2}} & \text{if } |x| < 1, x \neq 0 \\ 0 & \text{if } x = 0. \end{cases}$$

### (i) 

Show that $f$ is infinitely differentiable at 0 and compute $f^{(j)}(0)$ for all $j \geq 1$.

### Answer 

### (ii) 

Show that the Taylor series at $a = 0$ converges but not to $f$ on $(-1,1)$.

### Answer 

# Q8: A.24 (c)

Draw the open unit ball $B_p \equiv \{x : x \in \mathbb{R}^2,\ d_p(x,0) < 1\}$ in $\mathbb{R}^2$ for $p = 1, 2$ and $\infty$.

## Answer 

$p = 1, 2$ and $\infty$ are shown below in black, red, and blue respectively. 

```{r, echo = F, warning = F, message = F}
library(ggplot2)
library(patchwork)

theme_ball <- theme_void() +
  theme(plot.title = element_text(hjust = 0.5, size = 14),
        plot.margin = margin(10, 10, 10, 10))

lims <- coord_fixed(xlim = c(-1.5, 1.5), ylim = c(-1.5, 1.5))

## p = 1 : diamond (black)
diamond <- data.frame(x = c(1, 0, -1, 0, 1), y = c(0, 1, 0, -1, 0))
p1 <- ggplot() +
  geom_path(data = diamond, aes(x, y), color = "black") +
  geom_hline(yintercept = 0) +
  geom_vline(xintercept = 0) +
  lims +
  labs(title = expression(group("||", x, "||")[1])) +
  theme_ball

## p = 2 : circle (red)
theta <- seq(0, 2*pi, length.out = 200)
circle <- data.frame(x = cos(theta), y = sin(theta))
p2 <- ggplot() +
  geom_path(data = circle, aes(x, y), color = "red") +
  geom_hline(yintercept = 0) +
  geom_vline(xintercept = 0) +
  lims +
  labs(title = expression(group("||", x, "||")[2])) +
  theme_ball

## p = infinity : square (blue)
square <- data.frame(x = c(1, 1, -1, -1, 1), y = c(-1, 1, 1, -1, -1))
p3 <- ggplot() +
  geom_path(data = square, aes(x, y), color = "blue") +
  geom_hline(yintercept = 0) +
  geom_vline(xintercept = 0) +
  lims +
  labs(title = expression(group("||", x, "||")[infinity])) +
  theme_ball

# side-by-side (patchwork default with "+")
p1 + p2 + p3
```