---
title: "HW1"
author: "Sam Olson"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Q1: Done 
Q2: Done
Q3: Done
Q4: Done
Q5: Done 
Q6: Done 
Q7: 
Q8: Done 

# Q1: A.3

Let $I = [0,1]$, $\Omega = \mathbb{R}$ and for $\alpha \in \mathbb{R}$, $A_\alpha = (\alpha - 1, \alpha + 1)$, the open interval $\{x : \alpha - 1 < x < \alpha + 1\}$.

## (a) 

Show that $\bigcup_{\alpha \in I} A_\alpha = (-1, 2)$ and $\bigcap_{\alpha \in I} A_\alpha = (0, 1)$.

## Answer 

Sketch: To prove equality we argue both directions for both the Union and Intersection proofs, i.e. $\subset$ and $\supset$. 

Note: we use $\subset \equiv \subseteq$ and $\supset \equiv \supseteq$ for the purposes of set relations, as noted by Dr. Ghosh in lecture.  

### Union

*Proof.* 

($\subset$) For any $\alpha \in [0,1]$, $\alpha - 1 \in [-1,0]$ and $\alpha+1\in[1,2]$, so $A_\alpha = (\alpha-1,\alpha+1) \subset (-1,2)$. Since this holds for every $\alpha\in I$,

$$\bigcup_{\alpha\in I} A_\alpha \subset (-1,2).$$

($\supset$) Let $x \in (-1,2)$ be given, and set $\alpha = \min(\max(x,0),1) \in [0,1] = I$.

- If $x < 0$: $\alpha = 0$, and since $-1<x<0<1$, $x \in A_0 = (-1,1)$.
- If $0 \leq x \leq 1$: $\alpha = x$, and trivially $x - 1 < x < x+1$, so $x \in A_x$.
- If $x > 1$: $\alpha = 1$, and since $0<1<x<2$, $x \in A_1 = (0,2)$.

In every case $x \in A_\alpha$ for some $\alpha \in I$, so $x \in \bigcup_{\alpha\in I} A_\alpha$. Hence

$$(-1,2) \subset \bigcup_{\alpha\in I} A_\alpha.$$

Combining both containments, we conclude $\bigcup_{\alpha\in I} A_\alpha = (-1,2)$.//

### Intersection

*Proof.* 

($\subset$) Since $0,1 \in I$, $\bigcap_{\alpha\in I} A_\alpha \subset A_0 \cap A_1 = (-1,1)\cap(0,2) = (0,1)$.

($\supset$) Let $x \in (0,1)$ be given, and let $\alpha \in I$ be arbitrary. Since $\alpha \geq 0$, $\alpha - 1 \geq -1$. And since $0 \le \alpha \le 1$, it follows that $\alpha - 1 \leq 0 < x$ and $x < 1 \leq \alpha+1$, for $x \in A_\alpha$. As $\alpha\in I$ was arbitrary, $x \in \bigcap_{\alpha\in I} A_\alpha$. Hence

$$(0,1) \subset \bigcap_{\alpha\in I} A_\alpha.$$

Combining both containments, we conclude $\bigcap_{\alpha\in I} A_\alpha = (0,1)$.//

## (b) 

Suppose $J = \{x : x \in I, x \text{ is rational}\}$. Find $\bigcup_{x \in J} A_x$ and $\bigcap_{x \in J} A_x$.

## Answer 

Sketch: The key is that $0$ and $1$ are each rational, so $0,1 \in J$; this lets us use the results from part (a) to force the reverse containments.

### Union

*Proof.* Since $J \subset I$,

$$\bigcup_{x\in J} A_x \subset \bigcup_{\alpha\in I} A_\alpha = (-1,2).$$

Conversely, since $0,1 \in J$ (both are rational numbers in $I$),

$$\bigcup_{x\in J} A_x \supset A_0 \cup A_1 = (-1,1)\cup(0,2) = (-1,2),$$

using that the two intervals overlap on $(0,1)$. Combining both containments,

$$\bigcup_{x\in J} A_x = (-1,2).//$$

### Intersection

Since $J \subset I$,

$$\bigcap_{x\in J} A_x \supset \bigcap_{\alpha\in I} A_\alpha = (0,1).$$

Conversely, since $0,1 \in J$,

$$\bigcap_{x\in J} A_x \subset A_0 \cap A_1 = (-1,1)\cap(0,2) = (0,1).$$

Combining both containments,

$$\bigcap_{x\in J} A_x = (0,1).//$$

# Q2: A.5

Show that $X \equiv \{0,1\}^{\mathbb{N}}$, the set of all sequences $\{\omega_i\}_{i \in \mathbb{N}}$ where each $\omega_i \in \{0,1\}$, is **uncountable**. Conclude that $\mathcal{P}(\mathbb{N})$ is uncountable.

## Answer 

Sketch: To prove $X \equiv \{0,1\}^{\mathbb{N}}$ is uncountable, and that consequently $\mathcal{P}(\mathbb{N})$ is uncountable, we first establish (1): $X$ is uncountable and then (2): $\mathcal{P}(\mathbb{N})$ and $X$ are in bijection. 

### (1): $X$ is uncountable

Suppose, for contradiction, that $X$ is countable. Since $X$ is countable and nonempty (it contains, e.g., the constant sequence $\omega_i = 0$ for all $i$), $\exists$ a surjection $h : \mathbb{N} \to X$ enumerating every element of $X$: $h(1), h(2), h(3), \ldots$

Write $h(n) = (h(n)_1, h(n)_2, h(n)_3, \ldots)$ for the $n$-th sequence in the list, with $h(n)_i \in \{0,1\}$ its $i$-th coordinate.

Define $\delta = (\delta_i)_{i \in \mathbb{N}}$ by flipping the diagonal entries: $\delta_i = 1 - h(i)_i, \qquad i \in \mathbb{N}.$

Then $\delta \in X$, since each $\delta_i \in \{0,1\}$. But for every $n \in \mathbb{N}$, $\delta_n = 1 - h(n)_n \neq h(n)_n,$ so $\delta$ and $h(n)$ disagree in coordinate $n$, i.e. $\delta \neq h(n)$. This holds for **every** $n$, so $\delta \notin \{h(1), h(2), h(3), \ldots\}$, contradicting the assumption that $h$ is surjective (enumerates all of $X$).

Hence no such enumeration exists, and $X$ is uncountable.

### (2): $\mathcal{P}(\mathbb{N})$ and $X$ are in bijection

Define $\Phi : \mathcal{P}(\mathbb{N}) \to X$ by

$$\Phi(A) = \{\omega_i\}_{i \in \mathbb{N}}, \qquad \omega_i = \mathbb{1}\{i \in A\}, \quad \forall A \in \mathcal{P}(\mathbb{N}),$$

i.e. $\Phi(A)$ is the indicator sequence of $A$: coordinate $i$ is $1$ if $i \in A$ and $0$ otherwise. We show $\Phi$ is a bijection.

- **Injective:** if $A \neq B$, then without loss of generality there is some $i \in A \setminus B$, so $\Phi(A)_i = 1 \neq 0 = \Phi(B)_i$, hence $\Phi(A) \neq \Phi(B)$.
- **Surjective:** given any $\omega = (\omega_i)_{i \in \mathbb{N}} \in X$, let $A = \{i \in \mathbb{N} : \omega_i = 1\} \in \mathcal{P}(\mathbb{N})$. Then $\Phi(A) = \omega$ by construction.

Hence $\Phi$ is a bijection, so $|\mathcal{P}(\mathbb{N})| = |X|$.

### Conclusion

By Part (2), $\Phi$ is a bijection, so $|\mathcal{P}(\mathbb{N})| = |X|$. Since bijections preserve cardinality (and $X$ is uncountable by Part (1)), it follows that $\mathcal{P}(\mathbb{N})$ is uncountable as well.//

# Q3: A.9

Apply the principle of induction to establish the following:

## (a) 

For each $n \in \mathbb{N}$, $\sum_{j=1}^n j^2 = \dfrac{n(n+1)(2n+1)}{6}$.

## Answer 

*Proof.* Let $P(n)$ be the statement

$$\sum_{j=1}^n j^2 = \frac{n(n+1)(2n+1)}{6}.$$

*Base case ($n=1$):* The left side is $\sum_{j=1}^1 j^2 = 1$. The right side is $\dfrac{1 \cdot 2 \cdot 3}{6} = 1$. Since equality holds, the base case $P(1)$ holds.

*Inductive step:* Let $n \in \mathbb{N}$ be given, and suppose $P(n)$ holds, i.e.

$$\sum_{j=1}^n j^2 = \frac{n(n+1)(2n+1)}{6}.$$

Then consider $P(n+1)$. By the inductive hypothesis 

$$
\sum_{j=1}^{n+1} j^2 = \sum_{j=1}^n j^2 + (n+1)^2 = \frac{n(n+1)(2n+1)}{6} + (n+1)^2,
$$

Factoring out $(n+1)$,

$$
\sum_{j=1}^{n+1} j^2 = (n+1)\left[\frac{n(2n+1)}{6} + (n+1)\right]
= (n+1) \cdot \frac{n(2n+1) + 6(n+1)}{6}
= (n+1) \cdot \frac{2n^2 + 7n + 6}{6}.
$$

Simplifying, using $2n^2 + 7n + 6 = (n+2)(2n+3)$, gives

$$
\sum_{j=1}^{n+1} j^2 = \frac{(n+1)(n+2)(2n+3)}{6} = \frac{(n+1)\big((n+1)+1\big)\big(2(n+1)+1\big)}{6},
$$

which is exactly the statement $P(n+1)$.

Thus $P(n) \implies P(n+1)$. By induction, $P(n)$ holds for all $n \in \mathbb{N}$.// 

## (b) 

For each $n \in \mathbb{N}$, $x_1, x_2, \ldots, x_k \in \mathbb{R}$,

### (i) 

**(The binomial formula)**. $(x_1 + x_2)^n = \sum_{r=0}^{n} \binom{n}{r} x_1^r x_2^{n-r}$.

### Answer 

*Proof.* Let $P(n)$ be the statement

$$(x_1+x_2)^n = \sum_{r=0}^n \binom{n}{r} x_1^r x_2^{n-r}.$$

*Base case ($n=1$):* The left side is $(x_1+x_2)^1 = x_1 + x_2$. The right side is

$$
\sum_{r=0}^1 \binom{1}{r} x_1^r x_2^{1-r} = \binom{1}{0}x_2 + \binom{1}{1}x_1 = x_1 + x_2.
$$

Since equality holds, the base case $P(1)$ holds.

*Inductive step:* Let $n \in \mathbb{N}$ be given, and suppose $P(n)$ holds. We then consider $P(n+1)$. By the inductive hypothesis 

$$
(x_1+x_2)^{n+1} = (x_1+x_2)(x_1+x_2)^n = (x_1+x_2)\sum_{r=0}^n \binom{n}{r} x_1^r x_2^{n-r}
$$

Distributing,

$$
(x_1+x_2)^{n+1} = \sum_{r=0}^n \binom{n}{r} x_1^{r+1} x_2^{n-r} + \sum_{r=0}^n \binom{n}{r} x_1^r x_2^{n-r+1}.
$$

Re-indexing the first sum with $r \to r-1$,

$$
(x_1+x_2)^{n+1} = \sum_{r=1}^{n+1} \binom{n}{r-1} x_1^{r} x_2^{n+1-r} + \sum_{r=0}^n \binom{n}{r} x_1^r x_2^{n+1-r}.
$$

Separating out the $r=0$ term from the first sum's absence and the $r=n+1$ term from the second sum's absence, and combining,

$$
(x_1+x_2)^{n+1} = x_2^{n+1} + \sum_{r=1}^n \left[\binom{n}{r-1} + \binom{n}{r}\right] x_1^r x_2^{n+1-r} + x_1^{n+1}.
$$

By Pascal's Rule, $\binom{n}{r-1}+\binom{n}{r} = \binom{n+1}{r}$, so

$$
(x_1+x_2)^{n+1} = \binom{n+1}{0}x_2^{n+1} + \sum_{r=1}^n \binom{n+1}{r} x_1^r x_2^{n+1-r} + \binom{n+1}{n+1}x_1^{n+1} = \sum_{r=0}^{n+1} \binom{n+1}{r} x_1^r x_2^{n+1-r},
$$

which is exactly the statement $P(n+1)$.

Thus $P(n) \implies P(n+1)$. By induction, $P(n)$ holds for all $n \in \mathbb{N}$.// 

### (ii) 

**(The multinomial formula)**.

$$(x_1 + x_2 + \ldots + x_k)^n = \sum \frac{n!}{r_1! r_2! \ldots r_k!} x_1^{r_1} x_2^{r_2} \ldots x_k^{r_k},$$

where the summation extends over all $(r_1, r_2, \ldots, r_k)$ such that $r_i \in \mathbb{N}$, $0 \leq r_i \leq n$, $\sum_{i=1}^{k} r_i = n$.

### Answer 

*Proof.* Fix $k \in \mathbb{N}$ and $x_1,\ldots,x_k \in \mathbb{R}$. Let $P(n)$ be the statement

$$
(x_1 + x_2 + \cdots + x_k)^n = \sum \frac{n!}{r_1! r_2! \cdots r_k!} x_1^{r_1} x_2^{r_2} \cdots x_k^{r_k},
$$

where the sum extends over all $(r_1,\ldots,r_k)$ such that $r_i \in \mathbb{N} \cup \{0\}$, $0 \le r_i \le n$, and $\sum_{i=1}^k r_i = n$.

*Base case ($n=1$):* The left side is $x_1+x_2+\cdots+x_k$. On the right side, since $\sum_{i=1}^k r_i = 1$ and each $r_i \ge 0$, exactly one $r_i$ equals $1$ and all others equal $0$. The term corresponding to $r_i = 1$ is

$$
\frac{1!}{0! \cdots 1! \cdots 0!} x_1^0 \cdots x_i^1 \cdots x_k^0 = x_i,
$$

so summing over $i = 1,\ldots,k$ gives $x_1 + x_2 + \cdots + x_k$. Since equality holds, the base case $P(1)$ holds.

*Inductive step:* Let $n \in \mathbb{N}$ be given, and suppose $P(n)$ holds. We then turn to $P(n+1)$. By the inductive hypothesis

$$
(x_1+\cdots+x_k)^{n+1} = (x_1+\cdots+x_k)^n (x_1+\cdots+x_k) = \left(\sum_{r_1+\cdots+r_k=n} \frac{n!}{r_1!\cdots r_k!} x_1^{r_1}\cdots x_k^{r_k}\right)(x_1+\cdots+x_k)
$$

Distributing over the $k$ terms $x_1,\ldots,x_k$,

$$
(x_1+\cdots+x_k)^{n+1} = \sum_{i=1}^k \sum_{r_1+\cdots+r_k=n} \frac{n!}{r_1!\cdots r_k!} x_1^{r_1}\cdots x_i^{r_i+1}\cdots x_k^{r_k}.
$$

For fixed $i$, re-index by setting $s_j = r_j$ for $j \ne i$ and $s_i = r_i + 1$, so that $\sum_{j=1}^k s_j = n+1$ and $s_i \ge 1$. Since $r_i = s_i - 1$ gives $(s_i-1)! = s_i!/s_i$,

$$
\frac{n!}{r_1!\cdots r_i! \cdots r_k!} = \frac{n!}{s_1!\cdots (s_i-1)! \cdots s_k!} = \frac{n! \, s_i}{s_1! \cdots s_i! \cdots s_k!}.
$$

Since the $s_i=0$ terms vanish anyway, we can drop the restriction $s_i \ge 1$ and sum over the full range $s_1+\cdots+s_k=n+1$:

$$
(x_1+\cdots+x_k)^{n+1} = \sum_{i=1}^k \sum_{s_1+\cdots+s_k=n+1} \frac{n!\, s_i}{s_1!\cdots s_k!} x_1^{s_1}\cdots x_k^{s_k}.
$$

Swapping the order of summation and collecting terms with the same exponents $(s_1,\ldots,s_k)$,

$$
(x_1+\cdots+x_k)^{n+1} = \sum_{s_1+\cdots+s_k=n+1} \frac{n!}{s_1!\cdots s_k!}\left(\sum_{i=1}^k s_i\right) x_1^{s_1}\cdots x_k^{s_k}.
$$

Since $\sum_{i=1}^k s_i = n+1$ for every term in the sum, this is

$$
(x_1+\cdots+x_k)^{n+1} = \sum_{s_1+\cdots+s_k=n+1} \frac{(n+1)!}{s_1!\cdots s_k!} x_1^{s_1}\cdots x_k^{s_k},
$$

which is exactly the statement $P(n+1)$.

Thus $P(n) \implies P(n+1)$. By induction, $P(n)$ holds for all $n \in \mathbb{N}$.// 

# Q4: A.11

Show that the set $A = \{r : r \in \mathbb{Q}, r^2 < 2\}$ is bounded above in $\mathbb{Q}$ but has no l.u.b. in $\mathbb{Q}$.

## Answer

Sketch: We establish (1): $A$ is bounded above, and (2): No rational $M$ can be the *least* upper bound, by supposing $M = \sup_{\mathbb{Q}}(A)$ exists and deriving a contradiction (two contradictions, one for each case). 

### (1): $A$ is bounded above

*Proof.* Let $r \in A$ be given. Thus $r^2 < 2$. If $r \geq 2$, then $r^2 \geq 4 > 2$, contradicting $r \in A$. Thus $r < 2$.
By definition of an upper bound, $2$ is an upper bound of $A$.

### (2): $A$ has no l.u.b. in $\mathbb{Q}$

*Proof.* Suppose, for contradiction, that $M = \sup_{\mathbb{Q}}(A)$ exists. Since $1 \in A$ (as $1^2 = 1 < 2$), $M \geq 1$, so $M$ is a positive rational. Since $\sqrt{2}$ is irrational, no rational squares to $2$, so $M^2 \neq 2$. Thus either $M^2 < 2$ or $M^2 > 2$; we then examine each case.

**Case $M^2 < 2$:** Let $\varepsilon = 2 - M^2 > 0$. Choose a rational $\delta$ with $0 < \delta < \min\left(1, \dfrac{\varepsilon}{2M+1}\right)$, and set $r = M + \delta$. Then

$$r^2 = M^2 + 2M\delta + \delta^2 < M^2 + \delta(2M+1) < M^2 + \varepsilon = 2.$$

So $r \in A$, yet $r = M + \delta > M$, contradicting that $M$ is an upper bound of $A$.

**Case $M^2 > 2$:** Let $\varepsilon = M^2 - 2 > 0$. Choose a rational $\delta$ with $0 < \delta < \min\left(M, \dfrac{\varepsilon}{2M}\right)$, and set $b = M - \delta$. Then

$$b^2 = M^2 - 2M\delta + \delta^2 > M^2 - 2M\delta > M^2 - \varepsilon = 2.$$

Since $b > 0$ and $b^2 > 2$: for any $a \in A$, $a^2 < 2 < b^2$, and since $a, b > 0$, this gives $a < b$. So $b$ is an upper bound of $A$, yet $b < M$, contradicting that $M$ is the *least* upper bound.

Both cases are contradictions, leading to the conclusion that no such $M \in \mathbb{Q}$ exists. Thus $A$ has no l.u.b. in $\mathbb{Q}$.//

# Q5: A.12

Show that for any two sequences $\{x_n\}_{n\geq 1}, \{y_n\}_{n\geq 1} \subset \mathbb{R}$,

$$\varliminf_{n\to\infty} x_n + \varliminf_{n\to\infty} y_n \leq \varliminf_{n\to\infty} (x_n + y_n) \leq \varlimsup_{n\to\infty} (x_n + y_n)$$

$$\leq \varlimsup_{n\to\infty} x_n + \varlimsup_{n\to\infty} y_n.$$

## Answer

*Proof*: By definition, for every $n \in \mathbb{N}$ and every $k \geq n$,

$$\inf_{k\geq n} x_k \leq x_k \leq \sup_{k\geq n} x_k, \qquad \inf_{k\geq n} y_k \leq y_k \leq \sup_{k\geq n} y_k.$$

Taking $k=n$ and adding the left- and right-hand inequalities gives, for every $n$,

$$\inf_{k\geq n} x_k + \inf_{k\geq n} y_k \;\leq\; x_n + y_n \;\leq\; \sup_{k\geq n} x_k + \sup_{k\geq n} y_k. \tag{1}$$

Since $\inf_{k\geq n}x_k$ is non-decreasing in $n$ and $\sup_{k\geq n}x_k$ is non-increasing in $n$ (and likewise for $y_n$), both are monotone sequences and hence by definition converge to $\varliminf_{n\to\infty}x_n$ and $\varlimsup_{n\to\infty}x_n$ respectively. So the left side of $(1)$ converges, as $n \to \infty$, to $\varliminf_{n\to\infty} x_n + \varliminf_{n\to\infty} y_n$, and the right side converges to $\varlimsup_{n\to\infty} x_n + \varlimsup_{n\to\infty} y_n$.

Now apply $\liminf$ and $\limsup$ to $(1)$, using that both are monotone with respect to termwise $\leq$. From the left inequality of $(1)$,

$$\varliminf_{n\to\infty}\Big(\inf_{k\geq n} x_k + \inf_{k\geq n} y_k\Big) \leq \varliminf_{n\to\infty}(x_n+y_n),$$

and since the left side is a limit (shown above), it equals $\varliminf_{n\to\infty} x_n + \varliminf_{n\to\infty} y_n$. This gives

$$\varliminf_{n\to\infty} x_n + \varliminf_{n\to\infty} y_n \;\leq\; \varliminf_{n\to\infty}(x_n+y_n).$$
Symmetrically, applying $\limsup$ to the right inequality of $(1)$ gives

$$\varlimsup_{n\to\infty}(x_n+y_n) \;\leq\; \varlimsup_{n\to\infty} x_n + \varlimsup_{n\to\infty} y_n.$$

Finally, for any real sequence $z_n$ we have $\inf_{k\geq n} z_k \leq \sup_{k\geq n} z_k$ for every $n$, and taking $n \to \infty$ gives $\varliminf_{n\to\infty} z_n \leq \varlimsup_{n\to\infty} z_n$. Applying this with $z_n = x_n+y_n$ gives the middle inequality,

$$\varliminf_{n\to\infty}(x_n+y_n) \leq \varlimsup_{n\to\infty}(x_n+y_n).$$

Chaining the three inequalities above gives us:

$$\varliminf_{n\to\infty} x_n + \varliminf_{n\to\infty} y_n \leq \varliminf_{n\to\infty} (x_n + y_n) \leq \varlimsup_{n\to\infty} (x_n + y_n) \leq \varlimsup_{n\to\infty} x_n + \varlimsup_{n\to\infty} y_n.//$$

# Q6: A.17 

Find the radius of convergence, $\rho$, for the powers series $A(x) \equiv \sum_{n=0}^{\infty} a_n x^n$ where

## (a) 
$a_n = \dfrac{n}{(n+1)}$, $n \geq 0$.

## Answer 

Sketch: Since $\rho = \left(\varlimsup_{n\to\infty}|a_n|^{1/n}\right)^{-1}$, it suffices to show $a_n^{1/n} \to 1$.

*Proof.* For $n \geq 1$,

$$\log\big(a_n^{1/n}\big) = \frac{1}{n}\log\!\left(\frac{n}{n+1}\right) = -\frac{1}{n}\log\!\left(1+\frac{1}{n}\right).$$

Writing $t = 1/n \to 0$ and using $\lim_{t\to 0} \dfrac{\log(1+t)}{t} = 1$,

$$\frac{1}{n}\log\!\left(1+\frac{1}{n}\right) = \frac{1}{n^2}\cdot\frac{\log(1+1/n)}{1/n} \;\longrightarrow\; 0 \cdot 1 = 0.$$

Hence $\log(a_n^{1/n}) \to 0$, so $a_n^{1/n} \to e^0 = 1$. Since this is an honest limit, $\varlimsup_{n\to\infty}|a_n|^{1/n} = 1$.

Thus $\rho = 1^{-1} = 1$.//

## (b) 

$a_n = n^p$, $n \geq 0$, $p \in \mathbb{R}$.

## Answer 

Sketch: The key fact is $\log n / n \to 0$, which gives $n^{p/n} \to 1$ for every fixed $p$, so $\rho = 1$ regardless of the exponent.

*Proof.* For the continuous function $f(x) = \dfrac{\log x}{x}$, L'Hopital's rule gives

$$\lim_{x\to\infty} \frac{\log x}{x} = \lim_{x\to\infty} \frac{1/x}{1} = 0,$$

and since this limit holds as $x \to \infty$ through the reals, it holds along the integer sequence: $\dfrac{\log n}{n} \to 0$.

For $n \geq 1$,

$$\log\big(a_n^{1/n}\big) = \log\big(n^{p/n}\big) = p \cdot \frac{\log n}{n} \;\longrightarrow\; p \cdot 0 = 0$$

for every fixed $p \in \mathbb{R}$. Hence $a_n^{1/n} = n^{p/n} \to e^0 = 1$, so $\varlimsup_{n\to\infty}|a_n|^{1/n} = 1$.

(For $p<0$ the term $a_0 = 0^p$ is undefined, but the radius of convergence depends only on the tail behavior as $n\to\infty$, so this does not affect $\rho$.)

Thus $\rho = 1^{-1} = 1$, for every $p \in \mathbb{R}$.//

## (c) 
$a_n = \dfrac{1}{n!}$, $n \geq 0$ (where $0! = 1$).

## Answer 

Sketch: We show $(1/n!)^{1/n} \to 0$ using the elementary bound $n! \geq (n/e)^n$, obtained by truncating the series for $e^n$ to a single term (without invoking Stirling's formula, though that is also an option).

*Proof.* For $n \geq 1$, the exponential series $e^n = \displaystyle\sum_{k=0}^{\infty} \frac{n^k}{k!}$ has all nonnegative terms, so it is bounded below by its $k=n$ term:

$$e^n \geq \frac{n^n}{n!} \quad\Longrightarrow\quad n! \geq \frac{n^n}{e^n} = \left(\frac{n}{e}\right)^n.$$

Taking $n$-th roots (all quantities positive),
$$(n!)^{1/n} \geq \frac{n}{e}.$$

Hence

$$0 \;<\; \left(\frac{1}{n!}\right)^{1/n} = \frac{1}{(n!)^{1/n}} \;\leq\; \frac{e}{n} \;\longrightarrow\; 0.$$

By the squeeze (or sandwich, they are one-and-the-same) theorem, $\displaystyle\lim_{n\to\infty}\left(\frac{1}{n!}\right)^{1/n} = 0$, so $\varlimsup_{n\to\infty}|a_n|^{1/n} = 0$.

Thus $\rho = (0)^{-1} = +\infty$: the series converges for every $x \in \mathbb{R}$.//

# Q7: A.18

## (a) 

Find the Taylor series at $a = 0$ for the function $f(x) = \dfrac{1}{1-x}$ in $I \equiv (-1, +1)$ and show that it converges to $f(x)$ on $I$.

## Answer 

Sketch: We first find a closed form for $f^{(n)}(x)$ by induction on $n$, then evaluate at $x=0$ to read off the Taylor coefficients, which turn out to be exactly those of the geometric series $\sum x^n$. To show the series converges to $f$, we compute the geometric partial sums explicitly and show the error term vanishes for $|x|<1$.

*Claim.* For all $n \in \mathbb{N} \cup \{0\}$, $f^{(n)}(x) = \dfrac{n!}{(1-x)^{n+1}}$.

*Proof.* *Base case ($n=0$):* $f^{(0)}(x) = f(x) = \dfrac{1}{1-x} = \dfrac{0!}{(1-x)^{0+1}}$, so the claim holds.

*Inductive step:* Suppose $f^{(n)}(x) = \dfrac{n!}{(1-x)^{n+1}}$ for some $n \ge 0$. Differentiating,

$$
f^{(n+1)}(x) = \frac{d}{dx}\left[n!(1-x)^{-(n+1)}\right] = n!(n+1)(1-x)^{-(n+2)}(1) = \frac{(n+1)!}{(1-x)^{(n+1)+1}},
$$

which is the claim for $n+1$. By induction, the formula holds for all $n \in \mathbb{N}\cup\{0\}$.

*The Taylor series at $a=0$.* Since $f^{(n)}(0) = n!$ for all $n$, the Taylor series is

$$
\sum_{n=0}^{\infty} \frac{f^{(n)}(0)}{n!}(x-0)^n = \sum_{n=0}^{\infty} \frac{n!}{n!} x^n = \sum_{n=0}^\infty x^n.
$$

*Convergence to $f$ on $I = (-1,1)$.* For $x \in I$, let $S_N(x) = \sum_{n=0}^N x^n$ be the $N$-th partial sum. By the finite geometric sum formula (since $x \ne 1$),

$$
S_N(x) = \frac{1-x^{N+1}}{1-x} = \frac{1}{1-x} - \frac{x^{N+1}}{1-x}.
$$

Since $|x| < 1$, we have $x^{N+1} \to 0$ as $N \to \infty$, so

$$
\lim_{N\to\infty} S_N(x) = \frac{1}{1-x} - 0 = f(x).
$$

Hence $\sum_{n=0}^\infty x^n$ converges to $f(x)$ for every $x \in I = (-1,1)$. 

## (b) 

Find the Taylor series of $1 + x + x^2$ in $I = (1,3)$, centered at 2.

## Answer 

Sketch: Since $g$ is a degree-2 polynomial, all derivatives of order $3$ and higher vanish identically, so the Taylor series centered at any point is just the 3-term Taylor polynomial itself. We compute $g, g', g''$ at $a=2$ and assemble the finite sum, then verify by expanding that it reproduces $g(x)$.

Let $g(x) = 1+x+x^2$, and consider its Taylor expansion centered at $a=2$.

*Derivatives.* We compute

$$
g(x) = 1+x+x^2, \qquad g'(x) = 1+2x, \qquad g''(x) = 2, \qquad g^{(n)}(x) = 0 \text{ for } n \ge 3.
$$

Evaluating at $a=2$:

$$
g(2) = 1+2+4 = 7, \qquad g'(2) = 1+4 = 5, \qquad g''(2) = 2, \qquad g^{(n)}(2)=0 \ (n\ge 3).
$$

*Taylor series.* Since all derivatives of order $\ge 3$ vanish, the Taylor series at $a=2$ is the finite sum

$$
g(x) = \sum_{n=0}^{\infty} \frac{g^{(n)}(2)}{n!}(x-2)^n = g(2) + g'(2)(x-2) + \frac{g''(2)}{2!}(x-2)^2 = 7 + 5(x-2) + (x-2)^2.
$$

*Verification.* Expanding,

$$
7 + 5(x-2) + (x-2)^2 = 7 + 5x - 10 + x^2 - 4x + 4 = x^2 + x + 1,
$$

which recovers $g(x) = 1+x+x^2$ exactly. Since this is a finite sum (a polynomial identity), it holds for all $x \in \mathbb{R}$, and in particular on $I = (1,3)$.

## (c) 

Let

$$f(x) = \begin{cases} e^{-\frac{1}{x^2}} & \text{if } |x| < 1, x \neq 0 \\ 0 & \text{if } x = 0. \end{cases}$$

### (i) 

Show that $f$ is infinitely differentiable at 0 and compute $f^{(j)}(0)$ for all $j \geq 1$.

### Answer 

Sketch: Away from $0$, each derivative of $f$ has the form (rational function of $x$) $\times\, e^{-1/x^2}$, since differentiating this shape via the chain/product rule always produces another term of the same shape. To handle $x=0$, we show $e^{-1/x^2}$ divided by any power of $x$ still $\to 0$ as $x\to 0$ (the exponential decay dominates any polynomial blow-up), then induct on $j$ using the definition of the derivative at a point to show every $f^{(j)}(0) = 0$.

*Claim.* $f$ is infinitely differentiable at $0$, and $f^{(j)}(0) = 0$ for all $j \ge 1$.

*Setup.* For $x \ne 0$ (with $|x|<1$), one can show inductively that each derivative of $f$ has the form

$$
f^{(j)}(x) = \sum_{k=0}^{j-1} a_{k,j}\, x^{2k-3j}\, e^{-1/x^2}, \qquad x \ne 0,
$$

for suitable constants $a_{k,j}$ (this follows by repeated differentiation via the chain and product rules; each differentiation multiplies existing terms by a rational function of $1/x$ and adds new terms of the same general shape). We take this representation as given and focus on differentiability at $x=0$.

*Key Lemma.* For every integer $m \ge 0$,

$$
\lim_{x\to 0} \frac{e^{-1/x^2}}{x^m} = 0.
$$

*Proof:* Substituting $s = 1/|x|$ so that $s\to\infty$ as $x\to 0$, we have $\left|\dfrac{e^{-1/x^2}}{x^m}\right| = s^m e^{-s^2} \to 0$ as $s \to \infty$, since the exponential $e^{-s^2}$ decays faster than any polynomial power $s^m$ grows.

*Proof (by induction on $j$) that $f^{(j)}(0)$ exists and equals $0$ for all $j \ge 0$.*

*Base case ($j=0$):* By definition $f(0) = 0$. Also $\lim_{x\to 0} f(x) = \lim_{x\to 0} e^{-1/x^2} = 0$ by the Lemma (with $m=0$), so $f$ is continuous at $0$ with $f(0)=0$.

*Inductive step:* Suppose $f^{(j)}(0) = 0$ and, for $x\ne 0$, $f^{(j)}(x) = \sum_{k=0}^{j-1} a_{k,j} x^{2k-3j} e^{-1/x^2}$. Then by the definition of the derivative,

$$
f^{(j+1)}(0) = \lim_{h\to 0} \frac{f^{(j)}(h) - f^{(j)}(0)}{h} = \lim_{h\to 0} \frac{f^{(j)}(h)}{h} = \lim_{h\to 0} \sum_{k=0}^{j-1} a_{k,j}\, \frac{e^{-1/h^2}}{h^{3j-2k+1}}.
$$

Each term in this finite sum tends to $0$ by the Key Lemma (with $m = 3j-2k+1 \ge 0$). Hence the whole limit is $0$, so $f^{(j+1)}(0)$ exists and equals $0$.

By induction, $f^{(j)}(0)$ exists and equals $0$ for every $j \ge 0$; in particular, $f$ is infinitely differentiable at $0$ and $f^{(j)}(0) = 0$ for all $j \ge 1$. 

### (ii) 

Show that the Taylor series at $a = 0$ converges but not to $f$ on $(-1,1)$.

### Answer 

Sketch: Since every derivative of $f$ vanishes at $0$ by part (c)(i), the Taylor series at $a=0$ is identically the zero function, which trivially converges everywhere. But $f(x) = e^{-1/x^2} > 0$ for every $x \neq 0$, so the series (constant $0$) disagrees with $f$ at every nonzero point of $(-1,1)$.

*Claim.* The Taylor series of $f$ at $a=0$ converges on $(-1,1)$, but not to $f$.

*Proof.* From part (c)(i), $f^{(n)}(0) = 0$ for all $n \ge 0$ (including $f(0)=0$). Therefore the Taylor series at $a=0$ is

$$
\sum_{n=0}^\infty \frac{f^{(n)}(0)}{n!}(x-0)^n = \sum_{n=0}^\infty \frac{0}{n!}x^n = \sum_{n=0}^\infty 0 = 0,
$$

i.e., the Taylor series is identically the zero function. This series trivially converges (to $0$) for every $x \in \mathbb{R}$, and in particular for every $x \in (-1,1)$.

However, for any $x \in (-1,1)$ with $x \ne 0$, we have $f(x) = e^{-1/x^2} > 0 \ne 0$. Hence the Taylor series converges to $0$, but $f(x) \ne 0$, so the Taylor series does **not** converge to $f(x)$ at any such point. Only at $x=0$ does the series' value ($0$) agree with $f(0) = 0$.

Thus the Taylor series of $f$ at $a=0$ converges on all of $(-1,1)$, but it converges to $f$ only at the single point $x=0$, not on $(-1,1)$ as a whole.

# Q8: A.24 (c)

Draw the open unit ball $B_p \equiv \{x : x \in \mathbb{R}^2,\ d_p(x,0) < 1\}$ in $\mathbb{R}^2$ for $p = 1, 2$ and $\infty$.

## Answer 

$p = 1, 2$ and $\infty$ are shown below in black, red, and blue respectively. 

```{r, echo = F, warning = F, message = F, out.height="40%"}
library(ggplot2)
library(patchwork)

theme_ball <- theme_void() +
  theme(plot.title = element_text(hjust = 0.5, size = 14),
        plot.margin = margin(10, 10, 10, 10))

lims <- coord_fixed(xlim = c(-1.5, 1.5), ylim = c(-1.5, 1.5))

## p = 1 : diamond (black)
diamond <- data.frame(x = c(1, 0, -1, 0, 1), y = c(0, 1, 0, -1, 0))
p1 <- ggplot() +
  geom_path(data = diamond, aes(x, y), color = "black") +
  geom_hline(yintercept = 0) +
  geom_vline(xintercept = 0) +
  lims +
  labs(title = expression(group("||", x, "||")[1])) +
  theme_ball

## p = 2 : circle (red)
theta <- seq(0, 2*pi, length.out = 200)
circle <- data.frame(x = cos(theta), y = sin(theta))
p2 <- ggplot() +
  geom_path(data = circle, aes(x, y), color = "red") +
  geom_hline(yintercept = 0) +
  geom_vline(xintercept = 0) +
  lims +
  labs(title = expression(group("||", x, "||")[2])) +
  theme_ball

## p = infinity : square (blue)
square <- data.frame(x = c(1, 1, -1, -1, 1), y = c(-1, 1, 1, -1, -1))
p3 <- ggplot() +
  geom_path(data = square, aes(x, y), color = "blue") +
  geom_hline(yintercept = 0) +
  geom_vline(xintercept = 0) +
  lims +
  labs(title = expression(group("||", x, "||")[infinity])) +
  theme_ball

# side-by-side (patchwork default with "+")
p1 + p2 + p3
```