---
title: "PS6"
output: pdf_document
author: "Sam Olson"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Q1 

See Nettleton, HW7, Q1 

Consider the dataset `pigs` provided in the R package `emmeans`. The data can be accessed in R with the following commands.

```{r}
library(emmeans)
pigs
```

To learn a more about the data, type `?pigs` at the R prompt. For the purposes of this problem, use the natural logarithm of the variable `conc` as the response. Consider both `source` and `percent` as categorical factors. Assume the cell-means model with one unrestricted treatment mean for each combination of `source` and `percent`.

### a) 

Generate an ANOVA table with Type I (sequential) sums of squares for `source`, `percent`, `source × percent`, `error`, and `corrected total`. In addition to sums of squares, your ANOVA table should include degrees of freedom, mean squares, F statistics, and p-values where appropriate.

### b) 

Generate an ANOVA table with Type II sums of squares for `source`, `percent`, `source × percent`, `error`, and `corrected total`. In addition to sums of squares, your ANOVA table should include degrees of freedom, mean squares, F statistics, and p-values where appropriate.

### c) 

Generate an ANOVA table with Type III sums of squares for `source`, `percent`, `source × percent`, `error`, and `corrected total`. In addition to sums of squares, your ANOVA table should include degrees of freedom, mean squares, F statistics, and p-values where appropriate.

### d) 

Find LSMeans for `source` and `percent`.

### e) 

Consider simplifying the model so that `percent` is treated like a quantitative variable with linear effects on `log(conc)` and linear interactions; i.e.,

```{r, eval = F}
lm(y ~ source + percent + source:percent)
```

where `y=log(conc)` and `percent` is numeric. Does such a model fit adequately relative to the cell-means model? Conduct a lack of fit test and report the results.

### f) 

The reduced model fit in part (e) implies that, for each `source`, there is a linear relationship between the expected log concentration and percentage. Based on the fit of the reduced model in part (e), provide the estimated linear relationship for each `source`.

\newpage

## Q2

See Nettleton, HW6, Q1 

Consider the plant density example discussed in slide set 6.

### a) 

For each of the tests in the ANOVA table on slide 38, provide a vector $c$ so that a test of

$$H_0 : c^T \beta = 0$$

would yield the same statistic and p-value as the ANOVA test. (You can use R to help you with the computations like we did on slides 45 and 46 of slide set 6.) Label these vectors $c_1$, $c_2$, $c_3$, and $c_4$ for the linear, quadratic, cubic, and quartic tests, respectively.

### b) 

Are $c^T_1 \beta$, $c^T_2 \beta$, $c^T_3 \beta$, and $c^T_4 \beta$ contrasts? Explain.

### c) 

Are $c^T_1 \beta$, $c^T_2 \beta$, $c^T_3 \beta$, and $c^T_4 \beta$ orthogonal? Explain.

\newpage 

## Q3 

See Nettleton, HW6, Q2 

Suppose $H$ is a symmetric matrix. Prove that $H$ is nonnegative definite if and only if all its eigenvalues are nonnegative. (If you wish, you may use the Spectral Decomposition Theorem in your proof.)
